### 2012-05-30 paper/chemokine.tex
Initial Commit: tex and figures

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Title}
}
% Insert Author names, affiliations and corresponding author email.
\\
Author1$^{1}$, 
Author2$^{2}$, 
Author3$^{3,\ast}$
\\
\bf{1} Author1 Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{2} Author2 Dept/Program/Center, Institution Name, City, State, Country
\\
\bf{3} Author3 Dept/Program/Center, Institution Name, City, State, Country
\\
$\ast$ E-mail: Corresponding author@institute.edu
\end{flushleft}

% Please keep the abstract between 250 and 300 words
\section*{Abstract}

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLoS ONE authors please skip this step. 
% Author Summary not valid for PLoS ONE submissions.   
\section*{Author Summary}

\section*{Introduction}

The adaptive immune response induced during an acute viral infection is a complex web of defense mechanisms able to control all but the most virulent of infections.   A complete understanding of the evolution of this response will be instrumental in the design of antiviral prophylaxis, therapy and vaccine interventions.   The effector arm of the immune response operates, however, under strict constraints of time and the geography of body structure during the dynamic spread of replicating viruses.  Elegant studies have outlined the role of local dendritic cells in detecting invasion and delivering antigen-specific cargoes to the lymph nodes [1-3].  In situ imaging has provided temporal and spatial details on the communications between dendritic cells and recirculating precursor or memory T and B cells within the lymph node architecture [4-6].  The specificity, activation profile and functional capabilities of resulting effector lymphocytes are well described [7-9].  In contrast to the induction phase of the response, how activated lymphocytes emanating from regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10], has been less well studied.

The immune response to influenza A virus (IAV) is arguably the most well-studied with respect to viral specificity, individual immunocyte components, whole animal models, and mathematical modeling.  Animal models of IAV include the non-human primate, ferret, and mouse [11].   Although not a natural host of influenza viruses, the inbred mouse model has provided a detailed description of viral and immunocyte dynamics.  Mathematical models have used the mouse model data to simulate the evolution of the immune response and its impact on viral kinetics [12-16].  These mathematical models have approached such questions as the relative roles of antibody and cytotoxic T lymphocytes (CTL), number of CTL required, regulation of antigen presentation, and comparing drug intervention strategies.  Such whole-response modeling has the potential advantage of selecting the most promising strategies of vaccination and therapy for subsequent expensive animal and human trials.   Each of these models, however, must make assumptions about the efficiency of each functional component in the model.   Math modeling can also inform us on the behavior of each component.

Here we approach the more focused problems of time and physical constraints on the lymphocyte diaspora.  How do very small foci of infected tissue attract limited numbers of activated CTL released into the systemic vascular system servicing non-infected tissue orders of magnitude larger than the target infection?  A host of chemokines secreted by infected cells, with signal amplification by immigrating inflammatory cells, are critical in directing inflammatory and immune cells to sites of inflammation [17-19].  While each chemokine has been studied in detail for its synthesis, receptor specificity and function in knockout models, the spatial and temporal details of chemokine function and efficiency have not been explicitly modeled in a virus-infected whole animal environment.

Using an agent-based model (ABM) designed to test the temporal and spatial demands during an acute influenza infection in the lung, we examined the homing of CTL to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.   Viral secretion dynamics and detailed chemokine secretion patterns by infected human epithelial cells in vitro [20] provided the data representing local behaviors at the alveolar level.  Data on CTL production and recirculation and the resolution of viral infections was obtained from the literature on mouse infections.   We excluded the role of immigrating inflammatory cells in order to focus on the contribution of two chemokines induced during infection with three influenza viruses, including a seasonal H1N1 virus, the pandemic H1N1 virus, and a virulent avian H5N1 virus.  In our modeling environment we observed the markedly different efficiencies of control for the three influenza viruses, differing effectiveness of the two chemokines, and the importance of the velocity of CTL search within tissue.

% You may title this section "Methods" or "Models". 
% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
% authors may use "Analysis" 
\section*{Models}

\subsection*{Computational Modeling}

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=0.4\textwidth]{SystemChart}
\end{center}
\caption{A region of infected tissue of radius $r$ (shaded region $A$) expressing chemokines and inflammatory signals. This region is surrounded by region $B$ which does not have any infected cells and hence does not express either chemokines or inflammatory signals. The $2^{14}$ capillaries (red circles) are distributed evenly throughout the entire lung (region $A$ and $B$).}
\label{fig:systemchart}
\end{figure}

% FIXME: The main goal of this paper is to compare chemokine signaling effectiveness between infections from different strains of influenza. Our study looks at separate infections from three different strains: seasonal, avian, and swine. Futhermore, we compare the roles of two established signaling molecules: CXCL10 (IP-10) and CCL5 (RANTES).

FIXME: Need to say somewhere that we simulate a generic chemokinetic signal that abstracts X, Y, and Z.

Computational modeling was conducted using Cycells [REFERENCE], a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response [CITATIONS].
%developed for studying the spread of influenza and tuberculosis in tissue [CITATIONS].
%
%It has an interface for defining cell types and molecules. For example, the T cell type can be defined as having a particular size, lifespan, movement rate, change in state in response to contacting a DC or change in motion in response to contacting a chemokine. Cycells will be modified to account for flows between lymph nodes, the circulatory system and tissue compartments.
A simplified model of T-Cell activation and recirculation was implemented in Cycells (Fig 1), and simulations were conducted to measure efficiency of infection clearance under different environmental conditions. To model a growing infection, the lung was represented as a two dimensional sheet of healthy eptithelial cells. Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node. Activated T-cells descend through the vascular tree until chemotactic signal is detected, at which point they exit the vasculature and follow the [FIXME chemokine/inflammatory] gradient to the site of infection. T-cells that do not encounter chemokine recirculate to the lymph node. Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.

The simulation begins when a single cell in the center of the tissue is infected. After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine. Virus diffuses locally, infecting nearby cells, and continuing the cycle. The chemokine diffuses from secreting cells, thus increasing the size of the signaling area. After a delay, to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue. Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations. Figure \ref{fig:modelchart} illustrates the model components 
(epithelial cells, T-cells, and particles) and how they transition between different states.

\subsection*{Model Definition}

\begin{figure}[ht!]
\begin{center}
\includegraphics[width=\textwidth]{ModelChart}
\end{center}
\caption{Computational model: Components and state transitions.}
\label{fig:modelchart}
\end{figure}

Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{infected}, \emph{expressing}, \emph{apoptotic}, and \emph{dead}. Healthy cells remain unchanged unless infected by virus. Once infected, there is a 10h incubation window before the cell transitions to \emph{expressing}. \emph{Expressing} cells secrete virus and chemokine for 17 hours when they die. \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells. Apoptosising cells continue to secrete virus and die one hour after their transition. \emph{Dead} cells remain inert and do not regenerate over the course of an infection.

T-cells have three states: \emph{Searching}, \emph{Recirculating}, and \emph{Activated}. T-Cells begin emerging from the lymph node at five days post infection and continue at the rate of 777 T-cells per hour for the duration of the simulation. T-Cells enter the vascular system in the \emph{Searching} state, arriving at a random location on the lung's surface. \emph{Searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{Circulating} if they fail to encounter chemokine. Circulating cells spend six minutes recirculating to the lymph node where they are converted to \emph{Searching} and reintroduced to a new random location in the lung. If a \emph{Searching} T-Cell encounters chemokine, it immedaiately converts to \emph{Activated} and begins climbing the chemotactic gradient to the source of infection. \emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{Expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}. Activated T-Cells decay exponentially with an average lifespan of two hours. FIXME-DREW[What happens after they decay? Is there a threshold after which they die or what?]

The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{Expressing} epithelial cells.
%As particles, both have their kinetics limited to diffusion and decay.
Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
direct effect beyond activating T-Cells. Both particle types decay over time.
%purpose other than to activate searching T-cells.
FIXME-DREW[Viruses decay over time. Do Chemokines?]

\subsection*{Parameters}

Define CyCells environment.  Talk about the need for spatial model.  Human/mouse scaling.  Parameter definitions.

\begin{table}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Viral Diffusion & .0318 & Stokes-Einstein / Beau \\
  Viral Decay &  1/day & Soumya \\
  Chemokine Diffusion & .318 & Stokes-Einstein \\
  Chemokine Decay &  3.8508e-4/s & 30m half-life \\
  Infection Sensitivity &  2h/virion & [1] \\
  Incubation Time &  10h & Mitchell \\
  Expression Time &  16.7h & Mitchell \\
  Apoptosis Time & 1h & Fred \\
  T-Cell Production Rate & 777/h & Soumya, Miao fits \\ 
  Circulation Time & 6m & Fred, mouse model \\
  Search Time & 10m & Fred, mouse model \\
  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
  T-Cell Sensitivity & $100 ng/ml$ & See Results \\
  T-Cell Expected Kill Rate & 10m & Fred/Drew \\
  Cell Radius & $25 \mu m$ & Fred \\
  T-Cell Age (Active) & 2h & Fred \\
  T-Cell Age (Other) & 3d & Fred \\
  T-Cell Ramp Up Time & 5d & Icaris \\
  IgM Strength & Viral decay of 3/day after day 3 & Soumya [3] \\
  \hline  
\end{tabular}
\caption{caption.}
\label{table:parameters}
\end{table}

\begin{table}
\begin{tabular}{ | c | c | c | }
  \hline                        
  Paramter & Value & Source \\
  \hline
  Chemokine Production Rate &  9.348e-5 (varies by strain) & [2] \\
  Virus Production Rate &  3.783e-4 (varies by strain) & Mitchell \\
  \hline  
\end{tabular}
\caption{Strain specific parameters}
\label{table:strains}
\end{table}

\begin{itemize}
\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/ml.  Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.  This probability scales linearly with the viral concentration.
\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
\end{itemize}

% Results and Discussion can be combined.
\section*{Results}

\subsection*{Chemokine production}

Chemokine data was collected in conjunction with the Mitchell et al (REF) (Fig.~\ref{fig:data}).   Increases in IP-10 concentration are observed starting no later than eight hours post-infection and increases in RANTES concentration are seen no later than 16 hours post-infection.  We extended the ODE model established in Mitchell to allow for chemokine production from infected cells.  A genetic algorithm was used to find per cell production rates for both IP-10 and RANTES for the three different strains of influenza.  The production values for IP-10 are 7.15e-5 ng/cell*s for Avian, 9.35e-5 ng/cell*s for Seasonal, and 6.23e-5 ng/cell*s for Swine.  The values for RANTES are 3.90e-5 ng/cell*s for Avian, 1.27e-6 ng/cell*s for Seasonal, and 1.39e-6 ng/cell*s for Swine.  We observe consistent production values of each chemokine across the different strains with an exception of a larger production rate of RANTES in avian H5N1.  There seems to be no correlation between viral production and induced chemokine production across strains of influenza.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=.9\textwidth]{data}
 \end{center}
\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).  For each strain, the viral load (blue circle) is shown in PFU/ml and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/ml.} 
 \label{fig:data}
\end{figure}

\subsection*{Model variance}

The parameters of the agent-based model are listed in Table~\ref{table:parameters}.  The model was run using calculated values of viral production and chemokine production for the three different strains of influenza.  To measure the variance of the stochastic model, we ran each model ten times using the default parameters (Fig.~\ref{fig:variance}).  The amount of variance between each run was calculated to be X .  The plots show total number of incubating, virus secreting, and apoptotic cells, but do not include dead cells.  Thus the figures approximate the rate of growth of the plaque over time.

Overlaying multiple runs reveals a slight elbow at day three of the infection.  This shift in the growth rate is due to the introduction of IgM antibodies that clear free virus.  Each infection shows a sharp decrease in infected cells at day five due to the T-cell response.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{variance}
 \end{center}
\caption{Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.  Both chemokines were used with the exception of aH5N1 which only produced RANTES.  The ten runs were initialized with the same parameters for each strain.  Plots show the number of infected, expressing, and apoptotic cells.  Dead cells are not included.  Variance arises from the stochastic nature of the agent-based model.  All three simulations are dramatically affected by the T-cell response at day five.  Each simulation uses a T-cell sensitivity to chemokine value of 1e-22.} 
 \label{fig:variance}
\end{figure}

\subsection*{Spatial effects}

Spatial effects of viral and chemokine diffusion play an important role in both the spread and the clearance of the infection.  Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine is also produced by infected cells and serves as chemotactic gradient for T-cells to follow to cells secreting virus.  Although virus is produced at a higher rate, it diffuses much more slowly than the two chemokines because it is a larger particle.  On the other hand, the chemokines are limited in their diffusion by a higher decay rate than virus.  This gives the two particle types similar spatial profiles (Fig.~\ref{fig:cycells}).

At day four, expressing cells make up a significant proportion of the infection and the total number of dead cells is relatively small.  By day seven, most expressing cells have been killed off by T-cells and most of the plaque consists of dead cells.  Furthermore, the pockets of concentrated chemokine lags somewhat behind the cell and virus spatial layout.  It takes time for newly secreting cells to produce chemokine and for old pockets of chemokine to decay away.  Thus, T-cells, which rely on the chemokine gradient to find infected cells, will often lag slightly behind the changes in the plaque.  The delayed response and the low proportion of virus secreting cells make it difficult for T-cells alone to completely clear the infection.

FROM FIGURE~\ref{fig:plaquesize} CAPTION: Early during the infection the plaque is dominated by cells that are active in producing more virus (incubating cells and secreting cells).  As the infection grows, cells in the middle of the plaque die and these active cells make up a decreasing proportion of the total size. At day five T-cells arrive and begin to kill any virus secreting cells they can find.  Initially, secreting cells make up a significant proportion of the total active cell population, but by day six the number of expressing cells has been reduced dramatically.   The H5N1 plaque is concentrated enough that the secreting cells can be found and completely eliminated.  On the other hand, both H1N1 simulations secreting cells are not completely eliminated.  These cells only account for at most 10\% of the active cell population and less than  1\% of the total plaque by day six.  T-cells continue to accumulate at the site of infection, but arrive at a slower rate than the plaque's rate of growth.  Because of this, it becomes more and more difficult for T-cells to find and kill the remaining expressing cells in order to shut down the infection.  Thus, the infection is able to recover and resume growth.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{cycells}
 \end{center}
\caption{Screenshots taken at Day 4, Day 5.5, and Day 7 of a sample simulated sH1N1 infection.  The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).  Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).  Each plot of virus and chemokine corresponds directly with the simultaneous cell figures.} 
 \label{fig:cycells}
\end{figure}

\subsection*{Infection dynamics}

The three strains show markedly different levels of virulence consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).  Furthermore, the pandemic H1N1 grows so fast that the immune response is not able to contain the infection.  By the time the T-Cells arrive at the site of infection the plaque has become so spread out that the T-Cells are unable to cover the entire area.  Because H1N1 is produced at such a high rate, the window of opportunity for the T-Cells to stay ahead of the infection is too small (see discussion).  H5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 eventually recovers and continues to grow.

ADD IP-10 RANTES COMPARISON

\subsection*{T-cell sensitivity to chemokine}

T-Cells exit a capillary when they can sense an inflammatory cytokine signal.  T-Cells then encounter a second signal, the chemokine gradient.  We use the chemokine gradient as an approximation of the inflammatory signal.  As seen in Figure \ref{fig:cycells}, chemokine is concentrated around virus secreting cells.  The threshold sensitivity level determines the minimum concentration of aggregate chemokine signal required for T-cells to detect the gradient.

To test what concentration the cytokine/chemokine signal is enough to recruit T-Cells from the capillaries we examined sensitivity levels that spanned seven orders of magnitude.  There is a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1) after which there is no discernible difference in the infection kinetics.  Because multiple T-cells in the same area do not increase the rate of induction of apoptosis (killing), we hypothesize that there is a critical number of T-cells required to clear an infection after which there is no benefit gained from increasing numbers.  For consistency, all model runs were preformed using a sensitivity level of 1e-22.

NOTE TO SELF: Discuss the threshold in the discussion?  How does this threshold translate into a level of chemokine needed?

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{sensitivity}
 \end{center}
\caption{Plots of H5N1 using only RANTES and sH1N1 and pH1N1 infections using both IP-10 and RANTES with varying values for T-cell sensitivity to chemokine.  The plots show the total number of infected, expressing, and apoptotic cells for each infection.  The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.  We notice a threshold between concentrations of 1e-18 and 1e-19 for H5N1 and sH1N1 and a threshold between 1e-19 and 1e-20 (NOTE TO SELF: 1e-21???) for pH1N1, after which there is no discernible difference in the infection kinetics.} 
 \label{fig:sensitivity}
\end{figure}

\subsection*{Infection resurgence}

As seen in Figure \ref{fig:variance}, the immune response fails to completely clear the seasonal infection and is unable to contain the pandemic infection.  We suggest this is an artifact of the spatial nature of the ABM.  Why can the immune response create an immediate decrease in the number of infected cells of the pandemic infection, only to fail in the same endeavor two days later?  It is important to note that Figure \ref{fig:variance} shows only the number of infected cells and does not include the number of dead cells in the focus of infection (FOI).  Furthermore, T-Cells are only able to induce apoptosis for epithelial cells that are presenting viral peptides.  This means that they do not kill incubating cells that are not yet expressing the virus.  At day 5.5, the proportion of expressing cells to the size of the FOI is high (Fig.~\ref{fig:cycells}).  By day seven however, the number of sH1N1 virus-secreting cells has decreased to a small fraction of the overall FOI size, whereas pH1N1 secreting cells remain a significant proportion of plaque size.  Even though the number of T-Cells at the site of infection continue to increase, the size of the plaque increases at an even faster rate.  Thus, the few remaining expressing cells become dilute, decreasing the efficiency of the T-cell search.

By day seven the ratio of plaque size to the number of virus-secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.  However the greater replication rate of pH1N1 widens the gap between plaque size and fixed T-cell input (Fig.~\ref{plaquesize}C) whereas the lower replication rate of sH1N1 allows a match to T-cell input after day seven and subsequent control (but not elimination) of the sH1N1 infection.

A cell infected with pandemic influenza will produce new virus at the rate of 5.08e-3 particles/s.  This means that it will produce a new viral particle approximately every 200 seconds.  Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the BEST a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.  Thus, with no help from other immune sources, it becomes an impossible task for T-Cells alone to contain the pH1N1 infection.  In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, so the T-cell can limit a single infected cell to produce only 1.3 viral particles in the one hour interval.  The avian virus-secreting cell will produce only 0.2 viral particles in the interval after T-cell detection.

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=0.7\textwidth]{searcharea}
 \end{center}
\caption{Comparison of the number of T-cells at the site of infection and the number of expressing cells left in the plaque.  Although T-cell counts continue to increase, the number of expressing cells the T-cells are targeting become fewer and far between.  By day six, an equilibrium is reached where T-cells are only able to contain (and not clear) the remaining expressing cells. } 
 \label{fig:searcharea}
\end{figure}

\begin{figure}[ht!]
\begin{center}
 \includegraphics[width=\textwidth]{plaquesize}
 \end{center}
\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.  NOTE TO SELF: Text came out tiny!  Legend is outdated.} 
 \label{fig:plaquesize}
\end{figure}

\section*{Discussion}

\section*{Materials}

Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described (ref Mitchell 2010).  Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.  Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Quantitative viral culture was performed by standard plaque assay.  Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.

% Do NOT remove this, even if you are not including acknowledgments
\section*{Acknowledgments}

%\section*{References}
% The bibtex filename
\bibliography{template}

\section*{Figure Legends}
%\begin{figure}[!ht]
%\begin{center}
%%\includegraphics[width=4in]{figure_name.2.eps}
%\end{center}
%\caption{
%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
%should be left justified, as specified by the options to the caption 
%package.
%}
%\label{Figure_label}
%\end{figure}

\section*{Tables}
%\begin{table}[!ht]
%\caption{
%\bf{Table title}}
%\begin{tabular}{|c|c|c|}
%table information
%\end{tabular}
%\begin{flushleft}Table caption
%\end{flushleft}
%\label{tab:label}
% \end{table}

\end{document}

### 2012-05-30 data.m

@@ -1,0 +1,164 @@
+% Creates Data Plot
+
+% --- Avian Data ---
+avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
+av  = [ 63, 125, 825, 2563,  988, 1650, 2838, 17375, 5775];
+
+ait = [120, 480, 600, 960, 1200, 1440, 1680,  1920,  2280,  2880]/60;
+ai  = [ 24,  87, 296, 770, 3048, 3901, 8261, 12935, 26970, 24120];
+
+art = [120, 480, 600, 960, 1200, 1440, 1680, 1920, 2280, 2880]/60;
+ar  = [ 45,  50,  57,  87,  134,  151,  150,  245,  780,  695];
+
+% --- Seasonal Data ---
+svt = [600,  720,  960,  1200,  1440,  1800,  2280,  2880]/60;
+sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
+
+sit = [120,  480,  720,  960, 1200, 1440]/60;
+si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
+
+srt = [120, 480, 720, 960, 1200, 1440, 1800, 2280, 2880]/60;
+sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
+
+% --- Swine Data ---
+wvt = [120,   720,   960,   1200,   1440,   1800,   2880]/60;
+wv  = [ 20, 64620, 84053, 254587, 665559, 814586, 654773];
+
+wit = [480,  720,  960,  1200,  1440,  1920,  2880]/60;
+wi  = [ 75+1, 1835, 1349, 23458, 12073, 30700, 19814] - 75;
+
+wrt = [480, 720,  960, 1200, 1440, 1920, 2880]/60;
+wr  = [ 36,  55,   46,  150,   93,  380, 1224];
+
+% --- Plot ---
+
+% Avian
+subplot(1, 3, 1)
+[haxes, hline1, hline2] = plotyy(avt, av, art, ar, 'semilogy', 'semilogy');
+hold on
+hline3 = plot(ait, ai);
+hold off
+
+set(hline1, 'Marker', 'o')
+set(hline1, 'Color', 'b')
+
+set(hline2, 'Marker', 's')
+set(hline2, 'Color', [0 .5 0])
+
+set(hline3, 'Marker', '^')
[…]

### 2012-06-04 paper/chemokine.tex
polishing pass over results. some fixme's left for Drew

@@
 
-Chemokine data was collected in conjunction with the Mitchell et al (REF) (Fig.~\ref{fig:data}).
-Increases in IP-10 concentration are observed starting no later than eight hours post-infection and increases in RANTES concentration are seen no later than 16 hours post-infection.
-We extended the ODE model established in Mitchell to allow for chemokine production from infected cells.
-A genetic algorithm was used to find per cell production rates for both IP-10 and RANTES for the three different strains of influenza.
-The production values for IP-10 are 7.15e-5 ng/cell*s for Avian, 9.35e-5 ng/cell*s for Seasonal, and 6.23e-5 ng/cell*s for Swine.
+Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{MitchellEtAl2011} (Fig.~\ref{fig:data}).
+IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..
+To estimate per-cell production rates, we extended the ODE model of Ref.
+\cite{MitchellEtAl2011} to represent chemokine production from infected cells.
+Model was fits were generated for three influenza straints (Fig.~\ref{fig:data}) using a genetic algorithm \cite{GA-ref}.
+The model estimates that IP-10 is produced at 7.15e-5 ng/cell*s for Avian, 9.35e-5 ng/cell*s for Seasonal, and 6.23e-5 ng/cell*s for Swine.
 The values for RANTES are 3.90e-5 ng/cell*s for Avian, 1.27e-6 ng/cell*s for Seasonal, and 1.39e-6 ng/cell*s for Swine.
-We observe consistent production values of each chemokine across the different strains with an exception of a larger production rate of RANTES in avian H5N1.
-There seems to be no correlation between viral production and induced chemokine production across strains of influenza.
+These chemokine expression rates are consistent for each chemokine across the different strain, except that the RANTES rate is higher in avian H5N1.
+There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.
 
@@
 
-\subsection*{Model variance}
+\subsection*{Stochastic modeling effects}
 
-The parameters of the agent-based model are listed in Table~\ref{table:parameters}.
-The model was run using calculated values of viral production and chemokine production for the three different strains of influenza.
-To measure the variance of the stochastic model, we ran each model ten times using the default parameters (Fig.~\ref{fig:variance}).
-The amount of variance between each run was calculated to be X .
-The plots show total number of incubating, virus secreting, and apoptotic cells, but do not include dead cells.
-Thus the figures approximate the rate of growth of the plaque over time.
+[FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?]
 
-Overlaying multiple runs reveals a slight elbow at day three of the infection.
-This shift in the growth rate is due to the introduction of IgM antibodies that clear free virus.
-Each infection shows a sharp decrease in infected cells at day five due to the T-cell response.
+Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure \ref{}fig:variance).
+To test the strength of this effect, we ran each model ten times using the default parameters given in Table~\ref{table:parameters}, finding that the variance of infected cells was X.
+Each run took as input the calculated viral production and chemokine production rates for the three different strains of influenza, reporting as output the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
+The figures thus approximate plaque growth over time.
+
+Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the presence of IgM.
+%antibodies that clear free virus. 
+In addition, each infection the number of infected cells declines quickly at day five due to the T-cell response.
 
@@
  \end{center}
-\caption{Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.
-Both chemokines were used with the exception of aH5N1 which only produced RANTES.
-The ten runs were initialized with the same parameters for each strain.
-Plots show the number of infected, expressing, and apoptotic cells.
-Dead cells are not included.
-Variance arises from the stochastic nature of the agent-based model.
-All three simulations are dramatically affected by the T-cell response at day five.
-Each simulation uses a T-cell sensitivity to chemokine value of 1e-22.}
+\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.
+IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
+The ten runs were initialized identically for each strain.
+% Only say this once, either in the text or here.
+%Plots show the number of infected, expressing, and apoptotic cells.  Dead cells are not included.  
+%All three model scenairos respond dramatically to the T-cell response at day five.  
+Each simulation was run with identical T-cell sensitivity to chemokine  (1e-22).} 
  \label{fig:variance}
@@
 
-Spatial effects of viral and chemokine diffusion play an important role in both the spread and the clearance of the infection.
+Spatial effects of viral and chemokine diffusion play an important role in both the spread and clearance of infection.
 Free virus particles diffuse from virus secreting cells and infect healthy cells.
-Chemokine is also produced by infected cells and serves as chemotactic gradient for T-cells to follow to cells secreting virus.
-Although virus is produced at a higher rate, it diffuses much more slowly than the two chemokines because it is a larger particle.
-On the other hand, the chemokines are limited in their diffusion by a higher decay rate than virus.
-This gives the two particle types similar spatial profiles (Fig.~\ref{fig:cycells}).
+Chemokine produced by infected cells attracts T-cells to the infected cells.
+Although virus is produced at a higher rate than chemokine, it diffuses much more slowly because it is a larger particle.
+Chemokine diffusion, however, is limited by higher decay rates than the virus.
+These countervaling effects result in similar
+spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
 
-At day four, expressing cells make up a significant proportion of the infection and the total number of dead cells is relatively small.
-By day seven, most expressing cells have been killed off by T-cells and most of the plaque consists of dead cells.
+%FROM FIGURE~\ref{fig:plaquesize} CAPTION: 
+Early in the infection (until day 4), plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
+Over time, cells in the plaque's interior die, so active cells comprise a decreasing proportion of the total.
+T-cells arrive at day 5 and begin to killing the plentiful virus-expressing cells.
+By day 6 (OR 7?), however,  many of the expressing cells have been eliminated and the plaque is dominated by dead cells.
+In the H5N1
+example, the plaque is dense,  so the secreting cells can be found easily and the infection is eliminated.
+In both H1N1 simulations, however, secreting cells are not eliminated.
+In this case, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
+T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth.
+This leads to a lower probability of T-cells finding and killing the remaining expressing cells, and the infection is able to recover and resume growth.
+The regions of concentrated chemokine trail the cell and virus spatial layout.
+It takes time for newly secreting cells to produce chemokine and for old pockets of chemokine to decay away.
+Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind changes in the plaque.
+The delayed response and the low proportion of virus secreting cells prevent T-cells from clearing the infection completely.
+
+% At day four, expressing cells make up a significant proportion of the infection and the total number of dead cells is relatively small.  
+% By day seven, most expressing cells have been killed off by T-cells and most of the plaque consists of dead cells.
 Furthermore, the pockets of concentrated chemokine lags somewhat behind the cell and virus spatial layout.
@@
 
-FROM FIGURE~\ref{fig:plaquesize} CAPTION: Early during the infection the plaque is dominated by cells that are active in producing more virus (incubating cells and secreting cells).
-As the infection grows, cells in the middle of the plaque die and these active cells make up a decreasing proportion of the total size.
-At day five T-cells arrive and begin to kill any virus secreting cells they can find.
-Initially, secreting cells make up a significant proportion of the total active cell population, but by day six the number of expressing cells has been reduced dramatically.
-The H5N1 plaque is concentrated enough that the secreting cells can be found and completely eliminated.
-On the other hand, both H1N1 simulations secreting cells are not completely eliminated.
-These cells only account for at most 10\% of the active cell population and less than  1\% of the total plaque by day six.
-T-cells continue to accumulate at the site of infection, but arrive at a slower rate than the plaque's rate of growth.
-Because of this, it becomes more and more difficult for T-cells to find and kill the remaining expressing cells in order to shut down the infection.
-Thus, the infection is able to recover and resume growth.
-
 \begin{figure}[ht!]
@@
  \end{center}
-\caption{Screenshots taken at Day 4, Day 5.5, and Day 7 of a sample simulated sH1N1 infection.
+\caption{Example simulated sH1N1 infection.
+Screenshots taken at Day 4, Day 5.5, and Day 7.
 The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).
 Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).
-Each plot of virus and chemokine corresponds directly with the simultaneous cell figures.}
+%Each plot of virus and chemokine corresponds directly to the cell population
+%the simultaneous cell figures.
+FIXME_DREW (mols/ml) ??} 
  \label{fig:cycells}
@@
 
-The three strains show markedly different levels of virulence consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
-Furthermore, the pandemic H1N1 grows so fast that the immune response is not able to contain the infection.
-By the time the T-Cells arrive at the site of infection the plaque has become so spread out that the T-Cells are unable to cover the entire area.
-Because H1N1 is produced at such a high rate, the window of opportunity for the T-Cells to stay ahead of the infection is too small (see discussion).
-H5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 eventually recovers and continues to grow.
+The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
+The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
+By the time T-Cells arrive, the plaque is too large for them  to cover effectively.
+Because H1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
+H5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
 
@@
 
-T-Cells exit a capillary when they can sense an inflammatory cytokine signal.
-T-Cells then encounter a second signal, the chemokine gradient.
-We use the chemokine gradient as an approximation of the inflammatory signal.
-As seen in Figure \ref{fig:cycells}, chemokine is concentrated around virus secreting cells.
-The threshold sensitivity level determines the minimum concentration of aggregate chemokine signal required for T-cells to detect the gradient.
+T-Cells exit a capillary when they sense inflammatory cytokine and then follow a second signal, the chemokine gradient.
+We use the chemokine gradient as a proxy for the inflammatory signal.
+Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
+%threshold sensitivity level 
+T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
 
-To test what concentration the cytokine/chemokine signal is enough to recruit T-Cells from the capillaries we examined sensitivity levels that spanned seven orders of magnitude.
-There is a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1) after which there is no discernible difference in the infection kinetics.
-Because multiple T-cells in the same area do not increase the rate of induction of apoptosis (killing), we hypothesize that there is a critical number of T-cells required to clear an infection after which there is no benefit gained from increasing numbers.
-For consistency, all model runs were preformed using a sensitivity level of 1e-22.
+To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
+We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1), above which there is no discernible effect on infection kinetics.
+Because multiple T-cells in the same area do not increase the rate at which  apoptosis is induced, we hypothesize that a critical number of T-cells is required to clear an infection, above which there is no additional benefit from increasing numbers.
+For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.
+
+% We notice a threshold between concentrations of 1e-18 and 1e-19 for H5N1 and sH1N1 and a threshold between 1e-19 and 1e-20 (NOTE TO SELF: 1e-21???) for pH1N1, after which there is no discernible difference in the infection kinetics.
 
@@
  \end{center}
-\caption{Plots of H5N1 using only RANTES and sH1N1 and pH1N1 infections using both IP-10 and RANTES with varying values for T-cell sensitivity to chemokine.
-The plots show the total number of infected, expressing, and apoptotic cells for each infection.
+\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+Total number of infected, expressing and apoptotic cells are plotted for each infection.
 The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.
-We notice a threshold between concentrations of 1e-18 and 1e-19 for H5N1 and sH1N1 and a threshold between 1e-19 and 1e-20 (NOTE TO SELF: 1e-21???) for pH1N1, after which there is no discernible difference in the infection kinetics.}
+}
  \label{fig:sensitivity}
@@
 
-\subsection*{Infection resurgence}
+\subsection*{Infection rebound}
 
-As seen in Figure \ref{fig:variance}, the immune response fails to completely clear the seasonal infection and is unable to contain the pandemic infection.
-We suggest this is an artifact of the spatial nature of the ABM.
-Why can the immune response create an immediate decrease in the number of infected cells of the pandemic infection, only to fail in the same endeavor two days later?
-It is important to note that Figure \ref{fig:variance} shows only the number of infected cells and does not include the number of dead cells in the focus of infection (FOI).
-Furthermore, T-Cells are only able to induce apoptosis for epithelial cells that are presenting viral peptides.
-This means that they do not kill incubating cells that are not yet expressing the virus.
-At day 5.5, the proportion of expressing cells to the size of the FOI is high (Fig.~\ref{fig:cycells}).
-By day seven however, the number of sH1N1 virus-secreting cells has decreased to a small fraction of the overall FOI size, whereas pH1N1 secreting cells remain a significant proportion of plaque size.
-Even though the number of T-Cells at the site of infection continue to increase, the size of the plaque increases at an even faster rate.
-Thus, the few remaining expressing cells become dilute, decreasing the efficiency of the T-cell search.
+FIXME-DREW-FRED: expressing, actively presenting viral peptides, secreting, viral-secreting, etc.  Pls. choose one of these terms, introduce its synomys early, and then
+be consistent for the rest of the paper.
 
-By day seven the ratio of plaque size to the number of virus-secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
-However the greater replication rate of pH1N1 widens the gap between plaque size and fixed T-cell input (Fig.~\ref{plaquesize}C) whereas the lower replication rate of sH1N1 allows a match to T-cell input after day seven and subsequent control (but not elimination) of the sH1N1 infection.
+The immune response fails to clear sH1N1 completely and is unable to contain the pH1N1 (Fig.
+\ref{fig:variance}).
+This appears to be an artifact of the spatial nature of the ABM.
+Why can the immune response create an immediate decrease in the number of pH1N1 infected cells at day 5, only to fail in the same endeavor two days later?
+Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).
+Further, T-Cells induce apoptosis
+%for epithelial cells 
+in cells that are actively presenting viral peptides, but not in incubating cells.  At day 5.5, the ratio of expressing cells to the total FOI is high (Fig.~\ref{fig:cycells}).  
+By day 7, the ratio of plaque size to the number of virus-secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
+However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.
+%By day 7,  the number of sH1N1-secreting cells is only a small fraction of the overall FOI, whereas pH1N1-secreting cells remain a significant proportion.
+Even though the number of T-Cells on the plaque continues to increase, the plaque increases more quickly.
+The few remaining expressing cells become dilute, decreasing the efficiency of the T-cell search.
 
-A cell infected with pandemic influenza will produce new virus at the rate of 5.08e-3 particles/s.
-This means that it will produce a new viral particle approximately every 200 seconds.
-Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the BEST a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.
-Thus, with no help from other immune sources, it becomes an impossible task for T-Cells alone to contain the pH1N1 infection.
-In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, so the T-cell can limit a single infected cell to produce only 1.3 viral particles in the one hour interval.
-The avian virus-secreting cell will produce only 0.2 viral particles in the interval after T-cell detection.
+A cell infected with pH1N1  produces new virus at the rate of 5.08e-3 particles/s [FIXME: Where does this come from?
+IS this our data, model results, or what?].
+That is,   a new viral particle is produced approximately every 200 seconds.
+Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do on average is to limit a single infected cell to producing 18 new viral particles.
+Without additional immune responses,  T-Cells alone cannot  contain the pH1N1 infection.
+In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, so the T-cell can limit a single infected cell to producing 1.3 viral particles in the one-hour window.
+Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
 
@@
  \end{center}
-\caption{Comparison of the number of T-cells at the site of infection and the number of expressing cells left in the plaque.
-Although T-cell counts continue to increase, the number of expressing cells the T-cells are targeting become fewer and far between.
-By day six, an equilibrium is reached where T-cells are only able to contain (and not clear) the remaining expressing cells.
+\caption{T-cell frequency in plaque: T-cells compared to expressing cells at the FOI.
+Total  T-cell counts increase, but the number of expressing cells are fewer and harder to find.
+Equilibrium is reached at day 6, when T-cells contain (but cannot clear) expressing cells.
 }

### 2012-06-05 paper/chemokine.tex
Intermediate changes

--- 2012-05-30 paper/chemokine.tex
@@
 
-The adaptive immune response induced during an acute viral infection is a complex web of defense mechanisms able to control all but the most virulent of infections.
-A complete understanding of the evolution of this response will be instrumental in the design of antiviral prophylaxis, therapy and vaccine interventions.
-The effector arm of the immune response operates, however, under strict constraints of time and the geography of body structure during the dynamic spread of replicating viruses.
-Elegant studies have outlined the role of local dendritic cells in detecting invasion and delivering antigen-specific cargoes to the lymph nodes [1-3].
-In situ imaging has provided temporal and spatial details on the communications between dendritic cells and recirculating precursor or memory T and B cells within the lymph node architecture [4-6].
-The specificity, activation profile and functional capabilities of resulting effector lymphocytes are well described [7-9].
-In contrast to the induction phase of the response, how activated lymphocytes emanating from regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10], has been less well studied.
+The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
+Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes [1-3], temporal and spatial details of precursor T and B cell activation within the lymph node architecture [4-6], and functional capabilities of resulting effector lymphocytes [7-9].
+In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10].
 
-The immune response to influenza A virus (IAV) is arguably the most well-studied with respect to viral specificity, individual immunocyte components, whole animal models, and mathematical modeling.
-Animal models of IAV include the non-human primate, ferret, and mouse [11].
-Although not a natural host of influenza viruses, the inbred mouse model has provided a detailed description of viral and immunocyte dynamics.
-Mathematical models have used the mouse model data to simulate the evolution of the immune response and its impact on viral kinetics [12-16].
-These mathematical models have approached such questions as the relative roles of antibody and cytotoxic T lymphocytes (CTL), number of CTL required, regulation of antigen presentation, and comparing drug intervention strategies.
-Such whole-response modeling has the potential advantage of selecting the most promising strategies of vaccination and therapy for subsequent expensive animal and human trials.
-Each of these models, however, must make assumptions about the efficiency of each functional component in the model.
-Math modeling can also inform us on the behavior of each component.
+Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics [12-16].
+Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
+Models using differential equations to predict temporal events do not account for spatial characteristics of cell behavior, and must make assumptions about the efficiency of each functional component in the model.
 
-Here we approach the more focused problems of time and physical constraints on the lymphocyte diaspora.
-How do very small foci of infected tissue attract limited numbers of activated CTL released into the systemic vascular system servicing non-infected tissue orders of magnitude larger than the target infection?
-A host of chemokines secreted by infected cells, with signal amplification by immigrating inflammatory cells, are critical in directing inflammatory and immune cells to sites of inflammation [17-19].
-While each chemokine has been studied in detail for its synthesis, receptor specificity and function in knockout models, the spatial and temporal details of chemokine function and efficiency have not been explicitly modeled in a virus-infected whole animal environment.
+Here we focus on the time and physical constraints of the lymphocyte diaspora.
+How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control (17a), after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
+A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection [17-19].
+While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.
 
-Using an agent-based model (ABM) designed to test the temporal and spatial demands during an acute influenza infection in the lung, we examined the homing of CTL to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.
-Viral secretion dynamics and detailed chemokine secretion patterns by infected human epithelial cells in vitro [20] provided the data representing local behaviors at the alveolar level.
-Data on CTL production and recirculation and the resolution of viral infections was obtained from the literature on mouse infections.
-We excluded the role of immigrating inflammatory cells in order to focus on the contribution of two chemokines induced during infection with three influenza viruses, including a seasonal H1N1 virus, the pandemic H1N1 virus, and a virulent avian H5N1 virus.
-In our modeling environment we observed the markedly different efficiencies of control for the three influenza viruses, differing effectiveness of the two chemokines, and the importance of the velocity of CTL search within tissue.
+Using an agent-based model (ABM) designed to test the spatial demands on the effector phase, we examined the homing of activated cytotoxic T lymphocytes (CTL) to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.
+Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains [20], combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
+In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.
 
@@
 
+\subsection*{Model Validation}
+
+The ABM produces infection dynamics consistent with \textit{in vivo} observations.
+We observe FOI diameters on the order of 1mm in seasonal H1N1 simulation.
+The T-cell response at day five post-infection rapidly decreases the number of infected cells over the first 24 hours.
+
+Scaling to humans?
+
+\subsection*{ABM Advantages}
+
+The use of an ABM has advantages over a spatially homogeneous ODE model.
+An ODE model assumes that any virus particle is capable of infecting any healthy cell.
+Figure \ref{fig:cycells} shows that this is clearly not the case.
+In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
+ODE models account for this discrepancy by lowering rates of infection, but this simply results in improper parameter values.
+
+The spatial nature of the simulation also reveals challenges to T-cell detection rates that are not apparent in ODE models.
+T-cells find infected cells by climbing the chemokine gradient.
+This results in T-cell clustering at local maxima of chemokine concentration.
+Thus, T-cells do not increase in effectiveness as the number of T-cells increases.
+Furthermore, infected cells grow more disperse as the size of the infection grows.
+Because T-cells cluster together they become incapable of covering the increasing plaque size.
+This allows for the infection to recover and accelerate once its size crosses a certain threshold.
+This is another effect that is not seen in an ODE model.
+
+\subsection*{Caveats and Limitations}
+
+In both H1N1 strains the T-cell resposne fails to clear the infection by day 10.
+In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.
+We suggest that this is due to the absence of several innate immune agents that were not included in the model.
+We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.
+Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.
+We see that the viral production rate is a major factor in the strength of the infection.
+
+\subsection*{Future Work}
+
 \section*{Materials}

### 2012-06-05 paper/chemokine.tex
Fixed up Steph's changes

@@
 
-Computational modeling was conducted using Cycells [REFERENCE], a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response [CITATIONS].
-%developed for studying the spread of influenza and tuberculosis in tissue [CITATIONS].
-%
-%It has an interface for defining cell types and molecules.
-For example, the T cell type can be defined as having a particular size, lifespan, movement rate, change in state in response to contacting a DC or change in motion in response to contacting a chemokine.
-Cycells will be modified to account for flows between lymph nodes, the circulatory system and tissue compartments.
-A simplified model of T-Cell activation and recirculation was implemented in Cycells (Fig 1), and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
-To model a growing infection, the lung was represented as a two dimensional sheet of healthy eptithelial cells.
+Computational modeling was conducted using Cycells [REF], a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response [REF].
+A simplified model of T-Cell activation and recirculation was implemented in Cycells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
+To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells.
 Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
-Activated T-cells descend through the vascular tree until chemotactic signal is detected, at which point they exit the vasculature and follow the [FIXME chemokine/inflammatory] gradient to the site of infection.
+Activated T-cells descend through the vascular tree until chemotactic signal is detected, at which point they exit the vasculature and follow the [FIXME chemotactic/inflammatory] gradient to the site of infection.
 T-cells that do not encounter chemokine recirculate to the lymph node.
@@
 
-Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{infected}, \emph{expressing}, \emph{apoptotic}, and \emph{dead}.
-Healthy cells remain unchanged unless infected by virus.
-Once infected, there is a 10h incubation window before the cell transitions to \emph{expressing}.
-\emph{Expressing} cells secrete virus and chemokine for 17 hours when they die.
+Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
+\emph{healthy} cells remain unchanged unless infected by virus.
+Once infected, there is a 10h incubation window before the cell transitions from {incubating} to \emph{expressing}.
+\emph{Expressing} cells secrete virus and chemokine for 17 hours and then die.
 \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells.
-Apoptosising cells continue to secrete virus and die one hour after their transition.
+Apoptoic cells continue to secrete virus and die one hour after their transition.
 \emph{Dead} cells remain inert and do not regenerate over the course of an infection.
 
-T-cells have three states: \emph{Searching}, \emph{Recirculating}, and \emph{Activated}.
-T-Cells begin emerging from the lymph node at five days post infection and continue at the rate of 777 T-cells per hour for the duration of the simulation.
-T-Cells enter the vascular system in the \emph{Searching} state, arriving at a random location on the lung's surface.
-\emph{Searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{Circulating} if they fail to encounter chemokine.
-Circulating cells spend six minutes recirculating to the lymph node where they are converted to \emph{Searching} and reintroduced to a new random location in the lung.
-If a \emph{Searching} T-Cell encounters chemokine, it immedaiately converts to \emph{Activated} and begins climbing the chemotactic gradient to the source of infection.
-\emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{Expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}.
-Activated T-Cells decay exponentially with an average lifespan of two hours.
-FIXME-DREW[What happens after they decay?
-Is there a threshold after which they die or what?]
+T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{activated}.
+T-Cells begin emerging from the lymph node at five days post infection at the rate of 777 T-cells per hour for the duration of the simulation.
+T-Cells enter the vascular system in the \emph{searching} state, arriving at a random location on the lung's surface.
+\emph{searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{circulating} if they fail to encounter chemokine.
+\emph{Circulating} cells spend six minutes recirculating to the lymph node where they are converted to \emph{searching} and reintroduced to a new random location in the lung.
+If a \emph{searching} T-Cell encounters chemokine, it immedaiately converts to \emph{activated} and begins climbing the chemotactic gradient to the source of infection.
+\emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}.
+Activated T-cells are removed from the simulation using an exponential decay rate with an average lifespan of two hours.
 
-The model contains two kinds of particles: virus and chemokine. Both are produced at constant rates by \emph{Expressing} epithelial cells.
-%As particles, both have their kinetics limited to diffusion and decay.
-Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration. Chemokine diffuses across the tissue but has no
+The model contains two kinds of particles: virus and chemokine.
+Both are produced at constant rates by \emph{expressing} epithelial cells.
+Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration.
+Chemokine diffuses across the tissue but has no
 direct effect beyond activating T-Cells. Both particle types decay over time.
-%purpose other than to activate searching T-cells.
-FIXME-DREW[Viruses decay over time. Do Chemokines?]
 
@@
 \end{tabular}
-\caption{caption.}
+\caption{Generic model parameters}
 \label{table:parameters}
@@
 \begin{table}
-\begin{tabular}{ | c | c | c | }
+\begin{center}
+\begin{tabular}{ | r | c | c | }
   \hline                        
@@
   \hline
-  Chemokine Production Rate &  9.348e-5 (varies by strain) & [2] \\
-  Virus Production Rate &  3.783e-4 (varies by strain) & Mitchell \\
+  IP-10 Production Rate &  & [2]\\
+  Avian H5N1 & 7.2e-5 ng/s$\cdot$cell & \\
+  Seasonal H1N1 & 9.3e-5 ng/s$\cdot$cell& \\
+  Pandemic H1N1 & 6.3e-5 ng/s$\cdot$cell& \\
+  \hline
+  RANTES Production Rate & & [2] \\
+  Avian H5N1 & 3.9e-5 ng/s$\cdot$cell& \\
+  Seasonal H1N1 & 1.3e-6 ng/s$\cdot$cell& \\
+  Pandemic H1N1 & 1.4e-6 ng/s$\cdot$cell& \\
+  \hline
+  Virus Production Rate &  & Mitchell \\
+  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
+  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell& \\
+  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell& \\  
   \hline  
 \end{tabular}
-\caption{Strain specific parameters}
+\caption{Strain specific model parameters}
 \label{table:strains}
+\end{center}
 \end{table}
@@
 \cite{MitchellEtAl2011} to represent chemokine production from infected cells.
-Model was fits were generated for three influenza straints (Fig.~\ref{fig:data}) using a genetic algorithm \cite{GA-ref}.
-The model estimates that IP-10 is produced at 7.15e-5 ng/cell*s for Avian, 9.35e-5 ng/cell*s for Seasonal, and 6.23e-5 ng/cell*s for Swine.
-The values for RANTES are 3.90e-5 ng/cell*s for Avian, 1.27e-6 ng/cell*s for Seasonal, and 1.39e-6 ng/cell*s for Swine.
-These chemokine expression rates are consistent for each chemokine across the different strain, except that the RANTES rate is higher in avian H5N1.
+Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{GA-ref}.
+The model estimates are shown in Table~\ref{table:strains}.
+These chemokine expression rates are consistent for each chemokine across the different strains except that the RANTES rate is higher in avian H5N1.
 There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.
@@
 
-[FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?]
+[FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?   Drew: Sure, I can always do more runs.]
 
@@
 Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the presence of IgM.
-%antibodies that clear free virus. 
-In addition, each infection the number of infected cells declines quickly at day five due to the T-cell response.
+In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.
 
@@
 The ten runs were initialized identically for each strain.
-% Only say this once, either in the text or here.
-%Plots show the number of infected, expressing, and apoptotic cells.  Dead cells are not included.  
-%All three model scenairos respond dramatically to the T-cell response at day five.  
-Each simulation was run with identical T-cell sensitivity to chemokine  (1e-22).} 
+Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/ml).}
  \label{fig:variance}
@@
 Chemokine diffusion, however, is limited by higher decay rates than the virus.
-These countervaling effects result in similar
-spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
+These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
 
-%FROM FIGURE~\ref{fig:plaquesize} CAPTION: 
-Early in the infection (until day 4), plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
+Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
 Over time, cells in the plaque's interior die, so active cells comprise a decreasing proportion of the total.
 T-cells arrive at day 5 and begin to killing the plentiful virus-expressing cells.
-By day 6 (OR 7?), however,  many of the expressing cells have been eliminated and the plaque is dominated by dead cells.
-In the H5N1
-example, the plaque is dense,  so the secreting cells can be found easily and the infection is eliminated.
+By day 6 many of the expressing cells have been eliminated and the plaque is dominated by dead cells.
+In the H5N1 example, the plaque is dense, so the secreting cells can be found easily and the infection is eliminated.
 In both H1N1 simulations, however, secreting cells are not eliminated.
-In this case, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
+In these cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
 T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth.
@@
 The delayed response and the low proportion of virus secreting cells prevent T-cells from clearing the infection completely.
-
-% At day four, expressing cells make up a significant proportion of the infection and the total number of dead cells is relatively small.  
-% By day seven, most expressing cells have been killed off by T-cells and most of the plaque consists of dead cells.
-Furthermore, the pockets of concentrated chemokine lags somewhat behind the cell and virus spatial layout.
-It takes time for newly secreting cells to produce chemokine and for old pockets of chemokine to decay away.
-Thus, T-cells, which rely on the chemokine gradient to find infected cells, will often lag slightly behind the changes in the plaque.
-The delayed response and the low proportion of virus secreting cells make it difficult for T-cells alone to completely clear the infection.
 
@@
 The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).
-Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).
-%Each plot of virus and chemokine corresponds directly to the cell population
-%the simultaneous cell figures.
-FIXME_DREW (mols/ml) ??} 
+Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).}
  \label{fig:cycells}
@@
 Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-%threshold sensitivity level 
 T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
@@
 For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.
-
-% We notice a threshold between concentrations of 1e-18 and 1e-19 for H5N1 and sH1N1 and a threshold between 1e-19 and 1e-20 (NOTE TO SELF: 1e-21???) for pH1N1, after which there is no discernible difference in the infection kinetics.
-
-NOTE TO SELF: Discuss the threshold in the discussion?  How does this threshold translate into a level of chemokine needed?
 
@@
 Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).
-Further, T-Cells induce apoptosis
-%for epithelial cells 
-in cells that are actively presenting viral peptides, but not in incubating cells.  At day 5.5, the ratio of expressing cells to the total FOI is high (Fig.~\ref{fig:cycells}).  
-By day 7, the ratio of plaque size to the number of virus-secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
+Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.
+At day 5.5, the ratio of secreting cells to the total FOI is high (Fig.~\ref{fig:cycells}).
+By day 7, the ratio of plaque size to the number of secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
 However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.
-%By day 7,  the number of sH1N1-secreting cells is only a small fraction of the overall FOI, whereas pH1N1-secreting cells remain a significant proportion.
-Even though the number of T-Cells on the plaque continues to increase, the plaque increases more quickly.
-The few remaining expressing cells become dilute, decreasing the efficiency of the T-cell search.
 
-A cell infected with pH1N1  produces new virus at the rate of 5.08e-3 particles/s [FIXME: Where does this come from?
-IS this our data, model results, or what?].
+A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s [FIXME: Where does this come from?
+IS this our data, model results, or what?
+Drew: from Mitchell et al].
 That is, a new viral particle is produced approximately every 200 seconds.
-Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do on average is to limit a single infected cell to producing 18 new viral particles.
+Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.
 Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.

### 2012-06-08 paper/chemokine.tex
Added Fred's discussion and started bibtex

@@
 The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
-Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes [1-3], temporal and spatial details of precursor T and B cell activation within the lymph node architecture [4-6], and functional capabilities of resulting effector lymphocytes [7-9].
-In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [10].
+Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture [Legge and Braciale 2006], and functional capabilities of resulting effector lymphocytes [La Gruta and Doherty 2006].
+In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [Marshall 2001].
 
-Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics [12-16].
+Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics [Handel 2008; Lee 2009; Miao 2010; Handel 2010; Wu 2011].
 Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using differential equations to predict temporal events do not account for spatial characteristics of cell behavior, and must make assumptions about the efficiency of each functional component in the model.
+Models using differential equations to predict temporal events, however, do not account for spatial characteristics of cell behavior, and must make assumptions about the efficiency of each functional component in the model.
 
 Here we focus on the time and physical constraints of the lymphocyte diaspora.
-How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control (17a), after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
-A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection [17-19].
-While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.
+How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control (Allan 1990; La Gruta and Doherty 2006), after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
+A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection [Kawai 1999; Sallustro 2000; Zhao 2000; Dufour 2002; Thelen 2008; Sigmundsdottir 2008; Sallustro 2008; Bromley 2008; Kohlmeier 2008].
+While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine local concentration, function and efficiency are difficult to examine in a virus-infected whole animal model.
 
 Using an agent-based model (ABM) designed to test the spatial demands on the effector phase, we examined the homing of activated cytotoxic T lymphocytes (CTL) to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.
-Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains [20], combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
+Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains [Mitchell 2011], combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
 In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.
@@
 
-[FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?   Drew: Sure, I can always do more runs.]
+% [FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?   Drew: Sure, I can always do more runs, I did 10 just to have a figure to look at.]
 
@@
 
+FIXME: Needs editing.
+
 The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
@@
 By the time T-Cells arrive, the plaque is too large for them  to cover effectively.
-Because H1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
-H5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
+Because pH1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
+aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
 
-ADD IP-10 RANTES COMPARISON
+Because aH5N1 has been shown to surpress the production of interferon (\ref{ref:mitchell}) we hypothesize that it renders IP-10 is ineffectual (FIXME: need to clean this up).
+This may explain why aH5N1 has a much higher secretion rate for RANTES than the other two strains (Table~\ref{table:strains}).
+Because of this behavior, IP-10 was not included in the aH5N1 models.
+
+Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the efficacy of the simulated immune response (Fig.~\ref{fig:chemokine}).
+The lack of both chemokines leads to runaway infections in all three strains.
+The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
+IP-10 alone proves to be just as effective and both IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{chemokine}
+	\caption{Effects of different chemokine combinations.
+A) aH5N1 does not stimulate an IP-10 response.
+B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
+	\label{fig:chemokine}
+\end{center}
+\end{figure}
 
@@
 
-FIXME-DREW-FRED: expressing, actively presenting viral peptides, secreting, viral-secreting, etc.  Pls. choose one of these terms, introduce its synomys early, and then
-be consistent for the rest of the paper.
+%FIXME-DREW-FRED: expressing, actively presenting viral peptides, secreting, viral-secreting, etc.
+Pls.
+choose one of these terms, introduce its synomys early, and then be consistent for the rest of the paper.
+Drew: Healthy, Virus-Incubating or Incubating, Virus-Secreting or Secreting, Apoptotic, Dead.
 
@@
 \begin{center}
- \includegraphics[width=0.7\textwidth]{searcharea}
+ \includegraphics[width=0.7\textwidth]{tcellratio}
  \end{center}
-\caption{T-cell frequency in plaque: T-cells compared to expressing cells at the FOI.
-Total  T-cell counts increase, but the number of expressing cells are fewer and harder to find.
-Equilibrium is reached at day 6, when T-cells contain (but cannot clear) expressing cells.
-}
- \label{fig:searcharea}
+\caption{A logarithmic plot of the ratio of T-Cells at the focus of infection vs the number of cells that make up the focus of infection.
+aH5N1 exhibits unstable behavior once the infection is and T-cell counts are reduced to negligible levels.}
+ \label{fig:tcellratio}
 \end{figure}
@@
 \section*{Discussion}
+
+Hypercytokinemia is a characteristic of virulent influenza strains [de Jong 2006] but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.
+We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported [Mitchell 2011].
+In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the initial infection naturally.
+The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h post-infection.
+When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.
+Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains [Chan 2005; Chan 2010; Zeng 2011] and pandemic strains [Zeng 2011].
+MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion [Chan 2010].
+However, in spite of the attenuated type 1 interferon response induced by H5N1 strains [Zeng 2007], the chemokine secretion is not attenuated.
+
+Chemokines with a clear role directing T cell traffic include CXCL10 [Dufour 2002], CCL5 [Kawai 1999] and CCL3 [Awai 1997].
+The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial in the accelerated recruitment of CD8+ T cells to lung airways [Kohlmeier 2008].
+Other chemokines are likely to play a role in T cell traffic.
+Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures [Matsukura 1996; Arndt 2002].
+In the intact animal immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals [Julkunen 2001].
+Finally, antigen-specific CD8+ T cell recognition of its target will induced lung injury mediated by TNFa and trigger addition alveolar epithelial cell chemokine secretion [Zhao 2000].
+Thus our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
+
+A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal to induce directed motion of the cell up the chemical gradient [Thelen 2008].
+The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 100 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
+This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell [Francis and Palsson 1997].
 
@@
 
-%\section*{References}
+\section*{References}
 % The bibtex filename
-\bibliography{template}
+\bibliography{references}
 

### 2012-06-13 paper/chemokine.tex
Finished refs, switched some results sections around

@@
 Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture [Legge and Braciale 2006], and functional capabilities of resulting effector lymphocytes [La Gruta and Doherty 2006].
-In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora [Marshall 2001].
+In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora \cite{Marshall2001}.
 
-Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics [Handel 2008; Lee 2009; Miao 2010; Handel 2010; Wu 2011].
+Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics \cite{Handel2008, Lee2009, Miao2010, Handel2010, Wu2011}.
 Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
@@
 Here we focus on the time and physical constraints of the lymphocyte diaspora.
-How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control (Allan 1990; La Gruta and Doherty 2006), after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
-A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection [Kawai 1999; Sallustro 2000; Zhao 2000; Dufour 2002; Thelen 2008; Sigmundsdottir 2008; Sallustro 2008; Bromley 2008; Kohlmeier 2008].
+How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite{Allan1990} [La Gruta and Doherty 2006], after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
+A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Kawai1999, Sallusto2000, Zhao2000, Dufour2002, Thelen2008, Sigmundsdottir2008, Sallusto2008, Bromley2008, Kohlmeier2008}.
 While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine local concentration, function and efficiency are difficult to examine in a virus-infected whole animal model.
@@
 Using an agent-based model (ABM) designed to test the spatial demands on the effector phase, we examined the homing of activated cytotoxic T lymphocytes (CTL) to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.
-Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains [Mitchell 2011], combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
+Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains \cite{Mitchell2011}, combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
 In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.
@@
 
-Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{MitchellEtAl2011} (Fig.~\ref{fig:data}).
+Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
 IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..
 To estimate per-cell production rates, we extended the ODE model of Ref.
-\cite{MitchellEtAl2011} to represent chemokine production from infected cells.
-Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{GA-ref}.
+\cite{Mitchell2011} to represent chemokine production from infected cells.
+Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm [REF: GA-ref].
 The model estimates are shown in Table~\ref{table:strains}.
@@
 \begin{center}
- \includegraphics[width=.9\textwidth]{data}
+ \includegraphics[width=\textwidth]{data}
  \end{center}
@@
 
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure \ref{}fig:variance).
-To test the strength of this effect, we ran each model ten times using the default parameters given in Table~\ref{table:parameters}, finding that the variance of infected cells was X.
+Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).
+To test the strength of this effect, we ran each model ten (MORE?) times using the default parameters given in Table~\ref{table:parameters}, finding that the variance of infected cells was X.
 Each run took as input the calculated viral production and chemokine production rates for the three different strains of influenza, reporting as output the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
@@
 In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.
+
+The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
+The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
+By the time T-Cells arrive, the plaque is too large for them  to cover effectively.
+Because pH1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
+aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
 
@@
  \label{fig:variance}
+\end{figure}
+
+\subsection*{T-cell sensitivity to chemokine}
+
+T-Cells exit a capillary when they sense inflammatory cytokine and then follow a second signal, the chemokine gradient.
+We use the chemokine gradient as a proxy for the inflammatory signal.
+Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
+T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
+
+To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
+We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1), above which there is no discernible effect on infection kinetics.
+Because multiple T-cells in the same area do not increase the rate at which  apoptosis is induced, we hypothesize that a critical number of T-cells is required to clear an infection, above which there is no additional benefit from increasing numbers.
+For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.
+
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{sensitivity}
+ \end{center}
+\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+Total number of infected, expressing and apoptotic cells are plotted for each infection.
+The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.
+}
+ \label{fig:sensitivity}
+\end{figure}
+
+\subsection*{Chemokine combinations}
+
+Because aH5N1 has been shown to surpress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 is ineffectual (FIXME: need to clean this up).
+This may explain why aH5N1 has a much higher secretion rate for RANTES than the other two strains (Table~\ref{table:strains}).
+Because of this behavior, IP-10 was not included in the aH5N1 models.
+
+Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the efficacy of the simulated immune response (Fig.~\ref{fig:chemokine}).
+The lack of both chemokines leads to runaway infections in all three strains.
+The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
+IP-10 alone proves to be just as effective and both IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{chemokine}
+	\caption{Effects of different chemokine combinations.
+A) aH5N1 does not stimulate an IP-10 response.
+B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
+	\label{fig:chemokine}
+\end{center}
 \end{figure}
@@
 
-\subsection*{Infection dynamics}
-
-FIXME: Needs editing.
-
-The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
-The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
-By the time T-Cells arrive, the plaque is too large for them  to cover effectively.
-Because pH1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
-aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
-
-Because aH5N1 has been shown to surpress the production of interferon (\ref{ref:mitchell}) we hypothesize that it renders IP-10 is ineffectual (FIXME: need to clean this up).
-This may explain why aH5N1 has a much higher secretion rate for RANTES than the other two strains (Table~\ref{table:strains}).
-Because of this behavior, IP-10 was not included in the aH5N1 models.
-
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the efficacy of the simulated immune response (Fig.~\ref{fig:chemokine}).
-The lack of both chemokines leads to runaway infections in all three strains.
-The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
-IP-10 alone proves to be just as effective and both IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
-
-\begin{figure}[ht!]
-\begin{center}
-	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
-B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
-\end{figure}
-
-\subsection*{T-cell sensitivity to chemokine}
-
-T-Cells exit a capillary when they sense inflammatory cytokine and then follow a second signal, the chemokine gradient.
-We use the chemokine gradient as a proxy for the inflammatory signal.
-Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
-
-To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1), above which there is no discernible effect on infection kinetics.
-Because multiple T-cells in the same area do not increase the rate at which  apoptosis is induced, we hypothesize that a critical number of T-cells is required to clear an infection, above which there is no additional benefit from increasing numbers.
-For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{sensitivity}
- \end{center}
-\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
-Total number of infected, expressing and apoptotic cells are plotted for each infection.
-The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.
-}
- \label{fig:sensitivity}
-\end{figure}
-
 \subsection*{Infection rebound}
@@
 By day 7, the ratio of plaque size to the number of secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
-However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.
+However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.
 
@@
 
-Hypercytokinemia is a characteristic of virulent influenza strains [de Jong 2006] but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.
-We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported [Mitchell 2011].
+Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006} but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.
+We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.
 In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the initial infection naturally.
@@
 When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.
-Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains [Chan 2005; Chan 2010; Zeng 2011] and pandemic strains [Zeng 2011].
-MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion [Chan 2010].
-However, in spite of the attenuated type 1 interferon response induced by H5N1 strains [Zeng 2007], the chemokine secretion is not attenuated.
+Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains \cite{Chan2005, Chan2010, Zeng2011} and pandemic strains \cite{Zeng2011}.
+MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion \cite{Chan2010}.
+However, in spite of the attenuated type 1 interferon response induced by H5N1 strains \cite{Zeng2007}, the chemokine secretion is not attenuated.
 
-Chemokines with a clear role directing T cell traffic include CXCL10 [Dufour 2002], CCL5 [Kawai 1999] and CCL3 [Awai 1997].
-The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial in the accelerated recruitment of CD8+ T cells to lung airways [Kohlmeier 2008].
+Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999} and CCL3 \cite{Kawai1999}.
+The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial in the accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
 Other chemokines are likely to play a role in T cell traffic.
-Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures [Matsukura 1996; Arndt 2002].
-In the intact animal immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals [Julkunen 2001].
-Finally, antigen-specific CD8+ T cell recognition of its target will induced lung injury mediated by TNFa and trigger addition alveolar epithelial cell chemokine secretion [Zhao 2000].
+Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
+In the intact animal immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
+Finally, antigen-specific CD8+ T cell recognition of its target will induced lung injury mediated by TNFa and trigger addition alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
 Thus our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
 
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal to induce directed motion of the cell up the chemical gradient [Thelen 2008].
+A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
 The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
 For the threshold of 100 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
-This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell [Francis and Palsson 1997].
+This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
 
@@
 
-Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described (ref Mitchell 2010).
+Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
 Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.

### 2012-06-20 paper/chemokine.tex
Fixed up the ABM discussion.  ABM allows us to visuall observe spatial effects.

@@
 \caption{A logarithmic plot of the ratio of T-Cells at the focus of infection vs the number of cells that make up the focus of infection.
-aH5N1 exhibits unstable behavior once the infection is and T-cell counts are reduced to negligible levels.}
+aH5N1 exhibits unstable behavior once the infection is cleared and T-cell counts are reduced to negligible levels.}
  \label{fig:tcellratio}
@@
 \section*{Discussion}
+
+\subsection*{ABM Advantages}
+
+The use of an ABM has advantages over a spatially homogeneous ODE model.
+An ODE model assumes that any virus particle is capable of infecting any healthy cell.
+Figure \ref{fig:cycells} shows that this is clearly not the case.
+In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
+ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
+This is clearly false as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
+
+Another advantage of our ABM is the ability to render the model using OpenGL (Fig.~\ref{fig:cycells}).
+The ability to see the model in action reveals spatial dynamics that are absent in ODE models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.
+For example, observing the simulation reveals challenges to T-cell detection of infected cells that are not apparent using standard population plots.
+By watching the model progress, we observed three key phenomena.
+First, because T-cells find infected cells by climbing the chemokine gradient, we see T-cell clustering at local maxima of chemokine concentration.
+Thus, we have an explanation for why T-cells do not increase in effectiveness as their numbers increase.
+Second, we notice that clumps of T-cells remain in place even after all virus-secreting cells have been killed off.
+This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.
+Finall, we can see that infected cells grow more disperse as the size of the infection grows.
+Because T-cells cluster together they become incapable of covering the increasing plaque size.
+These spatial observations provide explanations for the pH1N1 resurgence that would be hidden without the visualization tools provided by the ABM.
+
+\subsection*{Model Comparison}
 
@@
 This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-
-\subsection*{Model Validation}
-
-The ABM produces infection dynamics consistent with \textit{in vivo} observations.
-We observe FOI diameters on the order of 1mm in seasonal H1N1 simulation.
-The T-cell response at day five post-infection rapidly decreases the number of infected cells over the first 24 hours.
-
-Scaling to humans?
-
-\subsection*{ABM Advantages}
-
-The use of an ABM has advantages over a spatially homogeneous ODE model.
-An ODE model assumes that any virus particle is capable of infecting any healthy cell.
-Figure \ref{fig:cycells} shows that this is clearly not the case.
-In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
-ODE models account for this discrepancy by lowering rates of infection, but this simply results in improper parameter values.
-
-The spatial nature of the simulation also reveals challenges to T-cell detection rates that are not apparent in ODE models.
-T-cells find infected cells by climbing the chemokine gradient.
-This results in T-cell clustering at local maxima of chemokine concentration.
-Thus, T-cells do not increase in effectiveness as the number of T-cells increases.
-Furthermore, infected cells grow more disperse as the size of the infection grows.
-Because T-cells cluster together they become incapable of covering the increasing plaque size.
-This allows for the infection to recover and accelerate once its size crosses a certain threshold.
-This is another effect that is not seen in an ODE model.
 

### 2012-07-05 paper/chemokine.tex
Lots of minor edits to figures and text

@@
 The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
-Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture [Legge and Braciale 2006], and functional capabilities of resulting effector lymphocytes [La Gruta and Doherty 2006].
+Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture \cite[Chapter 5]{Wormser2006a}, and functional capabilities of resulting effector lymphocytes \cite[Chapter 6]{Wormser2006a}.
 In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora \cite{Marshall2001}.
@@
 Here we focus on the time and physical constraints of the lymphocyte diaspora.
-How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite{Allan1990} [La Gruta and Doherty 2006], after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
+How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite[Chapter 6]{Wormser2006a}\cite{Allan1990}, after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
 A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Kawai1999, Sallusto2000, Zhao2000, Dufour2002, Thelen2008, Sigmundsdottir2008, Sallusto2008, Bromley2008, Kohlmeier2008}.
@@
 In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.
+
+We hypothesize that the variation of virulence between the three strains of influenza arise from deviations in viral production and chemokine production from infected epithelial cells as well as differences in the local interactions of T-cells with infected cells and chemokines.
+We find that while viral production does vary greatly between strains, chemokine production does not.
+Two chemokines show increased production during infections: CXCL10 (IP-10) and CCL5 (RANTES).
+Of the two, IP-10 consistently shows higher per cell production across all strains while RANTES is only produced at a comparable rate by aH5N1 influenza, which we hypothesize is in compensation for the inability of IP-10 to function in the absence of interferon (which is suppressed by aH5N1 \cite{Mitchell2011}).
+Furthermore, the spatial nature of our model reveals difficulties in T-cell search not exposed in spatially homogeneous ODE models.
+We see that an increase in the number of T-cells at the focus of infection past a certain threshold does not improve the rate of clearance of the infection and that T-cells face an increasing challenge as plaque sizes increase in diameter over the course of an infection.
 
@@
 
-% FIXME: The main goal of this paper is to compare chemokine signaling effectiveness between infections from different strains of influenza.
-Our study looks at separate infections from three different strains: seasonal, avian, and swine.
-Futhermore, we compare the roles of two established signaling molecules: CXCL10 (IP-10) and CCL5 (RANTES).
-
-FIXME: Need to say somewhere that we simulate a generic chemokinetic signal that abstracts X, Y, and Z.
-
-Computational modeling was conducted using Cycells [REF], a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response [REF].
-A simplified model of T-Cell activation and recirculation was implemented in Cycells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
+Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response.
+A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
 To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells.
 Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
-Activated T-cells descend through the vascular tree until chemotactic signal is detected, at which point they exit the vasculature and follow the [FIXME chemotactic/inflammatory] gradient to the site of infection.
-T-cells that do not encounter chemokine recirculate to the lymph node.
+Activated T-cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
+T-cells that do not encounter cytokine recirculate to the lymph node.
 Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.
@@
 The chemokine diffuses from secreting cells, thus increasing the size of the signaling area.
-After a delay, to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue.
+After a delay to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue.
 Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
@@
   \hline
-  Viral Diffusion & .0318 & Stokes-Einstein / Beau \\
-  Viral Decay &  1/day & Soumya \\
-  Chemokine Diffusion & .318 & Stokes-Einstein \\
-  Chemokine Decay &  3.8508e-4/s & 30m half-life \\
-  Infection Sensitivity &  2h/virion & [1] \\
-  Incubation Time &  10h & Mitchell \\
-  Expression Time &  16.7h & Mitchell \\
-  Apoptosis Time & 1h & Fred \\
-  T-Cell Production Rate & 777/h & Soumya, Miao fits \\ 
-  Circulation Time & 6m & Fred, mouse model \\
-  Search Time & 10m & Fred, mouse model \\
+  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
+  Viral Decay &  $1/day$ & Soumya \\
+  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
+  Infection Sensitivity &  $2 h/virion$ & [1*] \\
+  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
+  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
+  Apoptosis Time & $1 h$ & Fred \\
+  T-Cell Production Rate & $777/h$ & Soumya, Miao fits \\ 
+  Circulation Time & $6 m$ & Fred, mouse model \\
+  Search Time & $10 m$ & Fred, mouse model \\
   T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
   T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
-  T-Cell Sensitivity & $100 ng/ml$ & See Results \\
-  T-Cell Expected Kill Rate & 10m & Fred/Drew \\
-  Cell Radius & $25 \mu m$ & Fred \\
-  T-Cell Age (Active) & 2h & Fred \\
-  T-Cell Age (Other) & 3d & Fred \\
-  T-Cell Ramp Up Time & 5d & Icaris \\
-  IgM Strength & Viral decay of 3/day after day 3 & Soumya [3] \\
+  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
+  T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
+  Cell Radius & $25 \mu m$ & Drew? \\
+  T-Cell Age (Active) & $2 h$ & Fred \\
+  T-Cell Age (Other) & $3 d$ & Fred \\
+  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
+  IgM Strength & Viral decay of $3/day$ after day 3 & Soumya [3*] \\
   \hline  
@@
   \hline
-  IP-10 Production Rate &  & [2]\\
+  IP-10 Production Rate &  & [2*]\\
   Avian H5N1 & 7.2e-5 ng/s$\cdot$cell & \\
@@
   \hline
-  RANTES Production Rate & & [2] \\
+  RANTES Production Rate & & [2*] \\
   Avian H5N1 & 3.9e-5 ng/s$\cdot$cell& \\
@@
   \hline
-  Virus Production Rate &  & Mitchell \\
+  Virus Production Rate &  & \cite{Mitchell2011} \\
   Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
@@
 \begin{itemize}
-\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/ml.
+\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/mL.
 Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.
@@
 \cite{Mitchell2011} to represent chemokine production from infected cells.
-Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm [REF: GA-ref].
+Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.
 The model estimates are shown in Table~\ref{table:strains}.
-These chemokine expression rates are consistent for each chemokine across the different strains except that the RANTES rate is higher in avian H5N1.
+These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.
 There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.
@@
 \caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).
-For each strain, the viral load (blue circle) is shown in PFU/ml and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/ml.}
+For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
+sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $ng/mL$ (dotted line).
+These values were not included in model fitting.}
  \label{fig:data}
@@
 
-% [FIXME-DREW: Shouldn't we do 30 runs instead of 10 to get statistical significance?   Drew: Sure, I can always do more runs, I did 10 just to have a figure to look at.]
+Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).
+To test the strength of this effect, we ran each model fifty times using the default parameters given in Table~\ref{table:parameters}.
+$R^2$ values were calculated for individual runs versus the average of all runs.
+Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9931 and standard deviation of 0.0027.
+sH1N1 had a mean $R^2$ 0.9848 with a standard deviation of 0.0120.
+aH5N1 had a mean $R^2$ of 0.9075 with a standard deviation of 0.1087.
 
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).
-To test the strength of this effect, we ran each model ten (MORE?) times using the default parameters given in Table~\ref{table:parameters}, finding that the variance of infected cells was X.
-Each run took as input the calculated viral production and chemokine production rates for the three different strains of influenza, reporting as output the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
-The figures thus approximate plaque growth over time.
+Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
+Therefore the figures approximate plaque growth over time.
 
@@
 By the time T-Cells arrive, the plaque is too large for them to cover effectively.
-Because pH1N1 is produced so quickly, the window of opportunity for T-Cells to gain ground is too small (Discussion).
+Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.
 aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
@@
 The ten runs were initialized identically for each strain.
-Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/ml).}
+Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/mL).}
  \label{fig:variance}
@@
 
-T-Cells exit a capillary when they sense inflammatory cytokine and then follow a second signal, the chemokine gradient.
-We use the chemokine gradient as a proxy for the inflammatory signal.
+T-Cells exit a capillary when they sense inflammatory cytokine and then follow the chemokine gradient in tissue.
+We use the chemokine gradient as a proxy for the inflammatory response.
 Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
@@
 To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e-18 and 1e-19 (1e-19 and 1e-20 for pH1N1), above which there is no discernible effect on infection kinetics.
-Because multiple T-cells in the same area do not increase the rate at which  apoptosis is induced, we hypothesize that a critical number of T-cells is required to clear an infection, above which there is no additional benefit from increasing numbers.
-For consistency, we standardized all subsequent model runs to  sensitivity of 1e-22.
+We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e6 and 1e5 $ng/ml$, above which there is no discernible effect on infection kinetics.
+Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.
+Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $ng/ml$, therefore we use that sensitivity for each model run.
 
@@
 
-Because aH5N1 has been shown to surpress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 is ineffectual (FIXME: need to clean this up).
-This may explain why aH5N1 has a much higher secretion rate for RANTES than the other two strains (Table~\ref{table:strains}).
-Because of this behavior, IP-10 was not included in the aH5N1 models.
+Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.
+We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).
+Because of this behavior IP-10 was not included in the aH5N1 model runs.
 
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the efficacy of the simulated immune response (Fig.~\ref{fig:chemokine}).
+Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
 The lack of both chemokines leads to runaway infections in all three strains.
 The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
-IP-10 alone proves to be just as effective and both IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
+IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
 
@@
 Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
-Over time, cells in the plaque's interior die, so active cells comprise a decreasing proportion of the total.
-T-cells arrive at day 5 and begin to killing the plentiful virus-expressing cells.
-By day 6 many of the expressing cells have been eliminated and the plaque is dominated by dead cells.
-In the H5N1 example, the plaque is dense, so the secreting cells can be found easily and the infection is eliminated.
-In both H1N1 simulations, however, secreting cells are not eliminated.
-In these cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
+Over time, cells in the plaque's interior die causing active cells to comprise a decreasing proportion of the plaque.
+T-cells arrive at day 5 and begin to killing the virus-expressing cells.
+By day 6 a majority of the expressing cells have been eliminated as the plaque becomes dominated by dead cells.
+In the H5N1 example the plaque is dense, allowing T-cells to find secreting easily.
+Thus the infection can be eliminated.
+However, in both H1N1 simulations secreting cells are not eliminated.
+In these two cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
 T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth.
-This leads to a lower probability of T-cells finding and killing the remaining expressing cells, and the infection is able to recover and resume growth.
-The regions of concentrated chemokine trail the cell and virus spatial layout.
-It takes time for newly secreting cells to produce chemokine and for old pockets of chemokine to decay away.
-Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind changes in the plaque.
-The delayed response and the low proportion of virus secreting cells prevent T-cells from clearing the infection completely.
+This leads to a lower average T-cell killing rate.
+Furthermore, the regions of concentrated chemokine lag behind the cell and virus spatial layout.
+It takes time for newly secreting cells to produce chemokine and for pockets of high chemokine density to decay.
+Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind cellular changes in the plaque.
+The delayed response and the low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
 
@@
 The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).
-Free virus and chemokine particles are represented by compartmentalized concentrations of mols/ml and ng/ml (see the legend in the bottom right).}
+Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
  \label{fig:cycells}
@@
 
-The immune response fails to clear sH1N1 completely and is unable to contain the pH1N1 (Fig.
+The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
 \ref{fig:variance}).
@@
 Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.
-At day 5.5, the ratio of secreting cells to the total FOI is high (Fig.~\ref{fig:cycells}).
-By day 7, the ratio of plaque size to the number of secreting cells is approximately 100:1 for both the pH1N1 and sH1N1 strains.
-However the greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to keep up, even after day seven, leading to control (but not elimination) of the sH1N1 infection.
+At day 5.5, the ratio of secreting cells to the total number of cells in the FOI is high (Fig.~\ref{fig:cycells}).
+By day 7, the ratio of secreting cells to total plaque size is approximately 1:100 for both the pH1N1 and sH1N1 strains.
+The greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to control (but not eliminate) the infection.
 
-A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s [FIXME: Where does this come from?
-IS this our data, model results, or what?
-Drew: from Mitchell et al].
+A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
 That is, a new viral particle is produced approximately every 200 seconds.
@@
 Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.
-In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, so the T-cell can limit a single infected cell to producing 1.3 viral particles in the one-hour window.
+In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing a T-cell to limit a single infected cell to 1.3 viral particles in the one-hour window.
 Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=0.7\textwidth]{tcellratio}
- \end{center}
-\caption{A logarithmic plot of the ratio of T-Cells at the focus of infection vs the number of cells that make up the focus of infection.
-aH5N1 exhibits unstable behavior once the infection is cleared and T-cell counts are reduced to negligible levels.}
- \label{fig:tcellratio}
-\end{figure}
 
@@
  \end{center}
-\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.
-NOTE TO SELF: Text came out tiny!
-Legend is outdated.}
+\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.}
  \label{fig:plaquesize}
@@
 This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.
-Finall, we can see that infected cells grow more disperse as the size of the infection grows.
+Finally, we can see that infected cells grow more disperse as the size of the infection grows.
 Because T-cells cluster together they become incapable of covering the increasing plaque size.
@@
 
-In both H1N1 strains the T-cell resposne fails to clear the infection by day 10.
+In both H1N1 strains the T-cell response fails to clear the infection by day 10.
 In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.

### 2012-07-05 chemokine.m

@@ -1,0 +1,72 @@
+% Chemokine Effects% Sensitivity Plot
+
+figure('Position',[100 500 1800 700])
+
+% Avian
+
+subplot(1, 3, 1)
+
+colors = [.75  0  0 ;
+          .75  .5  0 ;
+          0  .5 0 ;
+          0 0 .75];
+      
+plot(time, infect(an22), 'color', colors(1,:), 'LineWidth', 2)
+hold on
+plot(time, infect(ar22), 'color', colors(2,:), 'LineWidth', 2)
+hold off
+
+set(gca, 'FontSize', 16)
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [1 1200])
+set(gca, 'YTick', [0 200 400 600 800 1000 1200])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Time (hours)')
+ylabel('Infected Cell Count')
+title('aH5N1 Influenza')
+legend('None', 'RANTES', 2)
+
+% Seasonal
+
+subplot(1, 3, 2)
+
+plot(time, infect(sn22), 'color', colors(1,:), 'LineWidth', 2)
+hold on
+plot(time, infect(sr22), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(si22), 'color', colors(3,:), 'LineWidth', 2)
+plot(time, infect(sb22), 'color', colors(4,:), 'LineWidth', 2) 
+hold off
+
+set(gca, 'FontSize', 16)
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [1 1200])
+set(gca, 'YTick', [0 200 400 600 800 1000 1200])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Time (hours)')
+%ylabel('Infected Cell Count')
+title('sH1N1 Influenza')
+legend('None', 'RANTES', 'IP-10', 'Both', 2)
+
+% Swine
+
+subplot(1, 3, 3)
+
+plot(time, infect(wn22), 'color', colors(1,:), 'LineWidth', 2)
+hold on
+plot(time, infect(wr22), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(wi22), 'color', colors(3,:), 'LineWidth', 2)
+plot(time, infect(wb22), 'color', colors(4,:), 'LineWidth', 2) 
[…]

### 2012-07-05 data.m

@@ -1,5 +1,7 @@
 % Creates Data Plot
 
+figure('Position',[100 500 1800 700])
+
 % --- Avian Data ---
 avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
 av  = [ 63, 125, 825, 2563,  988, 1650, 2838, 17375, 5775];
@@ -16,6 +18,10 @@ sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
 
 sit = [120,  480,  720,  960, 1200, 1440]/60;
 si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
+
+siit  = [1440, 1800, 2280, 2880]/60;
+%sii   = [8164, 10200, 12913, 16306] - 300;
+sii   = [8164, 8500, 8500, 8500] - 300;
 
 srt = [120, 480, 720, 960, 1200, 1440, 1800, 2280, 2880]/60;
 sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
@@ -39,13 +45,19 @@ hold on
 hline3 = plot(ait, ai);
 hold off
 
+set(hline1, 'LineWidth', 2)
 set(hline1, 'Marker', 'o')
+set(hline1, 'MarkerSize', 10)
 set(hline1, 'Color', 'b')
 
+set(hline2, 'LineWidth', 2)
 set(hline2, 'Marker', 's')
+set(hline2, 'MarkerSize', 10)
 set(hline2, 'Color', [0 .5 0])
 
+set(hline3, 'LineWidth', 2)
 set(hline3, 'Marker', '^')
+set(hline3, 'MarkerSize', 10)
 set(hline3, 'Color', 'r')
 
 axes(haxes(1))
@@ -67,30 +79,41 @@ set(haxes(2), 'YTick', [1 10 100 1000 10000 100000 1000000])
 set(haxes(2), 'XLim', [0 48])
 set(haxes(2), 'XTick', 0:4:48)
 set(haxes(2), 'XTickLabel', {'0', '', '8', '', '16', '', '24', '', '32', '', '40', '', '48'})
-ylabel('Chemokine (ng/ml)')
+%ylabel('Chemokine (ng/ml)')
 
 lhandle = legend('Virus', 'IP-10', 'RANTES', 4);
 set(lhandle, 'Color', 'w')
 
-title('Avian H5N1 Influenza')
[…]

### 2012-07-05 ftest.m

@@ -1,0 +1,122 @@
+% Calculate F Ratio for Avian, Seasonal, and Swine runs
+
+% Avian
+aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infect(vara04) + ...
+        infect(vara05) + infect(vara06) + infect(vara07) + infect(vara08) + infect(vara09) + ...
+        infect(vara10) + infect(vara11) + infect(vara12) + infect(vara13) + infect(vara14) + ...
+        infect(vara15) + infect(vara16) + infect(vara17) + infect(vara18) + infect(vara19) + ...
+        infect(vara20) + infect(vara21) + infect(vara22) + infect(vara23) + infect(vara24) + ...
+        infect(vara25) + infect(vara26) + infect(vara27) + infect(vara28) + infect(vara29) + ...
+        infect(vara30) + infect(vara31) + infect(vara32) + infect(vara33) + infect(vara34) + ...
+        infect(vara35) + infect(vara36) + infect(vara37) + infect(vara38) + infect(vara39) + ...
+        infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
+        infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
+
+vara = [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
+        infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
+        infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
+        infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
+        infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
+        infect(vara25) infect(vara26) infect(vara27) infect(vara28) infect(vara29) ...
+        infect(vara30) infect(vara31) infect(vara32) infect(vara33) infect(vara34) ...
+        infect(vara35) infect(vara36) infect(vara37) infect(vara38) infect(vara39) ...
+        infect(vara40) infect(vara41) infect(vara42) infect(vara43) infect(vara44) ...
+        infect(vara45) infect(vara46) infect(vara47) infect(vara48) infect(vara49)];
+    
+
[…]

### 2012-07-05 infect.m

@@ -1,0 +1,3 @@
+function [infect] = inf(data)
+  infect = data(:,3) + data(:,4) + data(:,6);
+end

### 2012-07-05 plaquesize.m

@@ -1,0 +1,95 @@
+% Plaque Size Plots
+
+figure('Position',[100 500 1800 1400])
+
+colors = [.75  0  0 ;
+          .75  .5  0 ;
+          0  .75 0 ;
+          0 0 .75];
+
+% Avian
+
+subplot(2, 3, 1)
+semilogy(time, infect(ar26)+ar26(:,5), 'color', colors(1,:), 'LineWidth', 2)
+hold on
+semilogy(time, ar26(:,3), 'color', colors(2,:), 'LineWidth', 2)
+semilogy(time, ar26(:,4), 'color', colors(3,:), 'LineWidth', 2)
+semilogy(time, ar26(:,7), 'color', colors(4,:), 'LineWidth', 2)
+hold off
+
+set(gca, 'FontSize', 16)
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [0 6000])
+set(gca, 'YTick', [1 10 100 1000])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Days Post-Infection')
+ylabel('Infected Cell Count (log10)')
+title('aH5N1')
+%set(gca, 'FontSize', 12)
+legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2)
+
+% Seasonal
+
+subplot(2, 3, 2)
+semilogy(time, infect(sb26)+sb26(:,5), 'color', colors(1,:), 'LineWidth', 2)
+hold on
+semilogy(time, sb26(:,3), 'color', colors(3,:), 'LineWidth', 2)
+semilogy(time, sb26(:,4), 'color', colors(2,:), 'LineWidth', 2)
+semilogy(time, sb26(:,7), 'color', colors(4,:), 'LineWidth', 2)
+hold off
+
+set(gca, 'FontSize', 16)
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [0 6000])
+set(gca, 'YTick', [1 10 100 1000])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Days Post-Infection')
+ylabel('Infected Cell Count (log10)')
+title('sH1N1')
+set(gca, 'FontSize', 12)
[…]

### 2012-07-05 rsquared.m

@@ -1,0 +1,146 @@
+% Calculate R^2 for Avian, Seasonal, and Swine runs
+
+% Avian
+aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infect(vara04) + ...
+        infect(vara05) + infect(vara06) + infect(vara07) + infect(vara08) + infect(vara09) + ...
+        infect(vara10) + infect(vara11) + infect(vara12) + infect(vara13) + infect(vara14) + ...
+        infect(vara15) + infect(vara16) + infect(vara17) + infect(vara18) + infect(vara19) + ...
+        infect(vara20) + infect(vara21) + infect(vara22) + infect(vara23) + infect(vara24) + ...
+        infect(vara25) + infect(vara26) + infect(vara27) + infect(vara28) + infect(vara29) + ...
+        infect(vara30) + infect(vara31) + infect(vara32) + infect(vara33) + infect(vara34) + ...
+        infect(vara35) + infect(vara36) + infect(vara37) + infect(vara38) + infect(vara39) + ...
+        infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
+        infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
+
+vara = [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
+        infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
+        infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
+        infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
+        infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
+        infect(vara25) infect(vara26) infect(vara27) infect(vara28) infect(vara29) ...
+        infect(vara30) infect(vara31) infect(vara32) infect(vara33) infect(vara34) ...
+        infect(vara35) infect(vara36) infect(vara37) infect(vara38) infect(vara39) ...
+        infect(vara40) infect(vara41) infect(vara42) infect(vara43) infect(vara44) ...
[…]

### 2012-07-05 sensitivity.m

@@ -1,0 +1,100 @@
+% Sensitivity Plot
+
+figure('Position',[100 500 1800 700])
+
+% Avian
+
+subplot(1, 3, 1)
+
+colors = winter(10);
+colors = [.75  0  0 ;
+          .8  .4  0 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .1  .1 .9 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ];
+
+hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
+hold on
+hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);
+hline3 = plot(time, infect(ar19), 'color', colors(3,:), 'LineWidth', 2);
+plot(time, infect(ar20), 'color', colors(4,:), 'LineWidth', 2) 
+plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2) 
+plot(time, infect(ar22), 'color', colors(6,:), 'LineWidth', 2) 
+plot(time, infect(ar23), 'color', colors(7,:), 'LineWidth', 2)
+plot(time, infect(ar25), 'color', colors(9,:), 'LineWidth', 2) 
+plot(time, infect(ar26), 'color', colors(10,:), 'LineWidth', 2)
+hline4 = plot(time, infect(ar24), 'color', colors(8,:), 'LineWidth', 2);
+hold off
+
+set(gca, 'FontSize', 16)
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [1 1200])
+set(gca, 'YTick', [0 200 400 600 800 1000 1200])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Time (hours)')
+ylabel('Infected Cell Count')
+title('aH5N1 Influenza')
+legend([hline1 hline2 hline4 hline3], '1e7 ng/mL', '1e6 ng/mL', '1 ng/mL', '1e5 to .01 ng/mL', 2)
+
+% Seasonal
+
+subplot(1, 3, 2)
+
+hline1 = plot(time, infect(sb17), 'color', colors(1,:), 'LineWidth', 2);
+hold on
[…]

### 2012-07-05 variance.m

@@ -1,0 +1,142 @@
+figure('Position',[100 500 1800 700])
+
+% Avian
+
+subplot(1, 3, 1)
+
+aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infect(vara04) + ...
+        infect(vara05) + infect(vara06) + infect(vara07) + infect(vara08) + infect(vara09) + ...
+        infect(vara10) + infect(vara11) + infect(vara12) + infect(vara13) + infect(vara14) + ...
+        infect(vara15) + infect(vara16) + infect(vara17) + infect(vara18) + infect(vara19) + ...
+        infect(vara20) + infect(vara21) + infect(vara22) + infect(vara23) + infect(vara24) + ...
+        infect(vara25) + infect(vara26) + infect(vara27) + infect(vara28) + infect(vara29) + ...
+        infect(vara30) + infect(vara31) + infect(vara32) + infect(vara33) + infect(vara34) + ...
+        infect(vara35) + infect(vara36) + infect(vara37) + infect(vara38) + infect(vara39) + ...        
+        infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
+        infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
+    
+plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
+            infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
+            infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
+            infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
+            infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
+            infect(vara25) infect(vara26) infect(vara27) infect(vara28) infect(vara29) ...
+            infect(vara30) infect(vara31) infect(vara32) infect(vara33) infect(vara34) ...
+            infect(vara35) infect(vara36) infect(vara37) infect(vara38) infect(vara39) ...
+            infect(vara40) infect(vara41) infect(vara42) infect(vara43) infect(vara44) ...
[…]

### 2012-07-13 paper/chemokine.tex
Redid fits (matlab files added).  Added Fred's intro and new discussion section.  Wrote abstract.

@@
 {\Large
-\textbf{Title}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration}
 }
@@
 \\
-Author1$^{1}$, 
-Author2$^{2}$, 
-Author3$^{3,\ast}$
+Drew Levin$^{1,\ast}$, 
+Stephanie Forrest$^{1}$, 
+Soumya Banerjee$^{1}$
+Candice Clay$^{2}$, 
+Melanie Moses$^{1}$, 
+Frederick Koster$^{2}$, 
 \\
-\bf{1} Author1 Dept/Program/Center, Institution Name, City, State, Country
+\bf{1} Department of Computer Science, University of New Mexico, Albuquerque, NM, USA
 \\
-\bf{2} Author2 Dept/Program/Center, Institution Name, City, State, Country
+\bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
 \\
-\bf{3} Author3 Dept/Program/Center, Institution Name, City, State, Country
-\\
-$\ast$ E-mail: Corresponding author@institute.edu
+$\ast$ E-mail: Corresponding drew@cs.unm.edu
 \end{flushleft}
@@
 \section*{Abstract}
+
+During the primary immune response clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.
+T-cell navigation through the complex branching bronchial vascular network is made possible by local cytokine and chemokine production from infected epithelial cells.
+Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
+Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (lab results only) with IP-10 always dominating the migration direction.
+A delayed differential equation model was fit to the empirical data revealing per cell chemokine production rates for each strain.
+These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects inter-strain variation of chemokine production on T-cell recruitment.
+The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.
+Expanding plaque sizes further isolate infected cells, making T-cell discovery and subsequent clearance of the infection less effective over time.
+The immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.
 
@@
 The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
-Elegant studies in the mouse model have described in detail the induction phase of the adaptive response, including dendritic cell- delivery of antigen-specific cargos to the lymph node immunocytes \cite{Maines2008, Saenz2010, Hatta2010}, temporal and spatial details of precursor T and B cell activation within the lymph node architecture \cite[Chapter 5]{Wormser2006a}, and functional capabilities of resulting effector lymphocytes \cite[Chapter 6]{Wormser2006a}.
-In contrast to the induction phase of the response, less is known about the temporal and spatial constraints on the effector phase of the response and how activated lymphocytes leaving regional lymph nodes efficiently ‘search’ for the infected tissue, often called the lymphocyte diaspora \cite{Marshall2001}.
+Understanding all components of defense mechanisms employed will lead to improved vaccine design and strategies to control immunopathology.
+Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including the functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.
+CD8+ T cells are necessary for the resolution of pneumonia and the complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.
 
-Mathematical models using mouse model data have simulated the evolution of the immune response and its impact on viral kinetics \cite{Handel2008, Lee2009, Miao2010, Handel2010, Wu2011}.
+The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve T and B cell precursors are encountered and activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.
+T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.
+In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
+Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
+
+A number of mathematical models have used detailed mouse model data to abstract the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.
 Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using differential equations to predict temporal events, however, do not account for spatial characteristics of cell behavior, and must make assumptions about the efficiency of each functional component in the model.
+Models using ordinary differential equations to predict temporal events, however, can not account for the spatial constraints on cell behavior, and therefore make assumptions about the efficiency of each functional component in the model.
 
-Here we focus on the time and physical constraints of the lymphocyte diaspora.
-How do very small foci of infected tissue attract limited numbers of activated CD8 T cells, necessary for infection control \cite[Chapter 6]{Wormser2006a}\cite{Allan1990}, after release into the systemic vascular system which services non-infected tissue orders of magnitude larger than the target infected tissue?
-A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Kawai1999, Sallusto2000, Zhao2000, Dufour2002, Thelen2008, Sigmundsdottir2008, Sallusto2008, Bromley2008, Kohlmeier2008}.
-While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine local concentration, function and efficiency are difficult to examine in a virus-infected whole animal model.
+Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
+To develop a model with tractable simulations the model does not incorporate some key components of host defense and therefore does not aim to predict the true outcome for each strain examined.
+We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
+The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.
+A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.
+While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.
 
-Using an agent-based model (ABM) designed to test the spatial demands on the effector phase, we examined the homing of activated cytotoxic T lymphocytes (CTL) to infected foci in the lung and how chemokines secreted by infected epithelial cells contributed to infection control.
-Our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains \cite{Mitchell2011}, combined with mouse model data on CTL production and the resolution of influenza infections reported in the literature, were used to construct the model.
-In our modeling environment we observed different efficiencies of control for the three influenza viruses dependent on chemokine kinetics and CTL search constraints.
+To provide data on the local chemokine-secretion environment in which T cells migrate, we used our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
+We selected in vitro chemokine data since this tissue environment surrogate may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and also permits the creation of sustained chemokine gradients within the model lung.
+Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells.
 
-We hypothesize that the variation of virulence between the three strains of influenza arise from deviations in viral production and chemokine production from infected epithelial cells as well as differences in the local interactions of T-cells with infected cells and chemokines.
-We find that while viral production does vary greatly between strains, chemokine production does not.
-Two chemokines show increased production during infections: CXCL10 (IP-10) and CCL5 (RANTES).
-Of the two, IP-10 consistently shows higher per cell production across all strains while RANTES is only produced at a comparable rate by aH5N1 influenza, which we hypothesize is in compensation for the inability of IP-10 to function in the absence of interferon (which is suppressed by aH5N1 \cite{Mitchell2011}).
-Furthermore, the spatial nature of our model reveals difficulties in T-cell search not exposed in spatially homogeneous ODE models.
-We see that an increase in the number of T-cells at the focus of infection past a certain threshold does not improve the rate of clearance of the infection and that T-cells face an increasing challenge as plaque sizes increase in diameter over the course of an infection.
+The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
+While we found this to be true, we observed in this spatially explicit model three interesting characteristics of chemokine-directed migration not revealed from experimental data.
+First, of the only two chemokines stimulated by all 3 strains, the influence of IP-10 consistently dominated over the effect of RANTES, regardless of strain.
+Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at higher concentrations above the threshold.
+Finally, when the search space (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient and may be reflected in the loss of control of the pandemic strain.
 
@@
   IP-10 Production Rate &  & [2*]\\
-  Avian H5N1 & 7.2e-5 ng/s$\cdot$cell & \\
-  Seasonal H1N1 & 9.3e-5 ng/s$\cdot$cell& \\
-  Pandemic H1N1 & 6.3e-5 ng/s$\cdot$cell& \\
+  Avian H5N1 & 2.0e-4 pg/s$\cdot$cell & \\
+  Seasonal H1N1 & 1.8e-4 pg/s$\cdot$cell& \\
+  Pandemic H1N1 & 8.7e-5 pg/s$\cdot$cell& \\
   \hline
   RANTES Production Rate & & [2*] \\
-  Avian H5N1 & 3.9e-5 ng/s$\cdot$cell& \\
-  Seasonal H1N1 & 1.3e-6 ng/s$\cdot$cell& \\
-  Pandemic H1N1 & 1.4e-6 ng/s$\cdot$cell& \\
+  Avian H5N1 & 1.3e-5 pg/s$\cdot$cell& \\
+  Seasonal H1N1 & 8.9e-7pg/s$\cdot$cell& \\
+  Pandemic H1N1 & 4.4e-6 pg/s$\cdot$cell& \\
   \hline
@@
 \end{itemize}
+
+\section*{Materials}
+
+Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
+Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
+Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
+Quantitative viral culture was performed by standard plaque assay.
+Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
 
@@
 For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $ng/mL$ (dotted line).
+sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $pg/mL$ (dotted line).
 These values were not included in model fitting.}
@@
 $R^2$ values were calculated for individual runs versus the average of all runs.
-Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9931 and standard deviation of 0.0027.
-sH1N1 had a mean $R^2$ 0.9848 with a standard deviation of 0.0120.
-aH5N1 had a mean $R^2$ of 0.9075 with a standard deviation of 0.1087.
+Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.
+sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.
+aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.
 
@@
 To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 1e6 and 1e5 $ng/ml$, above which there is no discernible effect on infection kinetics.
+We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 100 and 10 $ng/ml$, below which there is no discernible effect on infection kinetics.
 Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.
-Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $ng/ml$, therefore we use that sensitivity for each model run.
+Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $nM$ ($10 ng/mL$ assuming a chemokine weight of $10 kDa$), therefore we use that sensitivity for each model run.
 
@@
 
-%FIXME-DREW-FRED: expressing, actively presenting viral peptides, secreting, viral-secreting, etc.
-Pls.
-choose one of these terms, introduce its synomys early, and then be consistent for the rest of the paper.
-Drew: Healthy, Virus-Incubating or Incubating, Virus-Secreting or Secreting, Apoptotic, Dead.
-
 The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
@@
 
-\subsection*{Model Comparison}
+\subsection*{General Immune Response Modeling}
+
+NEW
+
+The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
+Recent whole-animal models using a data-fitting approach and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
+We have focused exclusively on the problem of the search of activated CTL for infected foci in the lung, and as in other models have made a number of assumptions.
+
+The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve precursors are encountered and activated.
+T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood [Marshall 2001].
+In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
+Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
+One of the model simplifications we utilized was the equating cytokine and chemokine signals.
+
+A large fraction of the T cells entering lung tissue have come from the spleen where they proliferated under stimulation from antigen-loaded DC.
+This has been demonstrated by the significant loss of T cells entering lung following splenectomy differential [Tripp 1997] and by math modeling results that make sense only if there is a significant proliferation of activated cells outside the lymph node prior to arrival in the lung [Wu 2011].
+Our model did not include the proliferation in the temporary splenic reservoir as this uncertain degree of time delay in arrival between LN and lung would have complicated our ABM model.
+Using numbers of cells egressing from LN likely underestimated the rate of CTL population entering the lung, but our derivations of parameters impacting the T cell search and subsequent function are probably not affected.
+
+The fundamental mechanisms regulating the expansion of viral load and subsequent clearance of virus remain unclear, with current hypotheses focusing on antigen-driven, target cell susceptibility-driven, innate immune response-driven, antigen presentation limited, and adaptive immunity responses.
+We used the data from strain-specific viral replication in vitro in human bronchial epithelial cells [Mitchell 2011] to structure the regulation of viral abundance and viral clearance as essentially a model of antigen-driven, innate immunity-driven regulation.
+In a differential equation model, regulation of viral load was primarily driven by the availability of susceptible target cells [Baccam 2006].
+The question of target-driven versus antigen-driven regulation was addressed in a separate modeling study of infected equine lungs [Saenz 2010], finding that regulation was primarily innate immunity-driven since no more than 27\% of bronchial epithelial cells were ever infected during the course of the lung infection.
+In a comprehensive model of murine influenza rate-limiting antigen presentation was the primary influence on viral clearance [Lee 2009].
+In a subsequent model based on a large murine database, CD8+ T cells were critical components of the adaptive response in limiting the number of infected cells, with neutralizing antibodies responsible for clearing free virus [Miao 2010].
+Most modeling studies used a mouse-adapted virus to enable enumeration of epitope-specific T cells, making direct comparisons of replication efficiency with our human virus isolates more difficult.
+The mechanism of regulation of animal viral load may be specific to viral replication efficiencies, strain-specific suppression of innate immunity, and the phase of the immune response.
+We compared the three strains for their demands placed on the T cell search mechanism to find and kill infected cells.
+The multiplicity of potential regulation mechanisms does not invalidate our approach to the T cell search mechanisms.
+
+Our focus on the T cell search necessitated the simplification or removal of selected features of the mammalian immune response, in order to permit tractable simulations.
+Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
+Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
+The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
+All of these rates may in fact be time-variable rates as indicated by superior data-fitting models [Wu 2011].
+Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection [Miao 2010].
+Since known critical components were omitted or simplified, our model was not designed to demonstrate clearance of virus from the lung.
+Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
+
+\subsection*{Chemokine Directed T-cell Search}
 
@@
 
+NEW
+
+In our model directed migration and viral infection control was not limited over a wide range of T cell sensitivity to the chemotactic signal.
+The density of CCR5 receptors on the surface of CD8+ T cells was a determinant of the strength of chemotactic attraction for CCR5 ligands such as CCL5 in an in vitro assay [Desmetz 2006].
+The density of CCR5 receptors on human CD8+ T cells vary over an order of magnitude, suggesting that sensitivity to the chemotactic signal could play a role in susceptibility to infection.
+CCR5-knockout mice are susceptible to fatal Cryptococcal encephalitis/meningitis [Huffnagle 1999] and West Nile virus encephalitis [Glass 2005].
+In the Cryptococcus-infected mice T cell recruitment to the brain, but not to the lung, was impaired.
+In CCR5-KO mice influenza A virus pneumonia was more severe associated with decreased T cell recruitment to the lung, while in CCR2-KO mice decreased T cell recruitment was due to decreased macrophage influx into the lung [Dawson 2000].
+On the other hand, CCL5-KO mice and CXCR3-KO mice (the receptor for CXCL10/IP-10) had similar viral kinetics and leukocyte recruitment compared to the wild-type mice [Wareing 2004].
+Finally, humans with the CCR5-D32 deletion allele may have increased susceptibility to the pandemic 2009 H1N1 influenza in a small sample [Keynan 2010].
+The question of limiting sensitivity has practical implications, since immunomodulation of cytokine/chemokine signaling to decrease influenza mortality is currently under investigation [Oslund 2011; Alleva 2011; Bassaganya-Riera 2010; Budd 2007; Moseley 2010].
+While blocking chemokine function may decrease inflammation and death from viral pneumonia, it may have an unintended consequence of impairing adaptive immunity.
+Thus, specific T cell recruitment to the influenza virus-infected lung should be quantitatively assessed in chemokine receptor-knockout mice.
+
 \subsection*{Caveats and Limitations}
@@
 We see that the viral production rate is a major factor in the strength of the infection.
-
-\subsection*{Future Work}
-
-\section*{Materials}
-
-Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
-Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
-Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
-Quantitative viral culture was performed by standard plaque assay.
-Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
 

### 2012-07-13 chemokine.m

@@ -11,9 +11,9 @@ colors = [.75  0  0 ;
           0  .5 0 ;
           0 0 .75];
       
-plot(time, infect(an22), 'color', colors(1,:), 'LineWidth', 2)
+plot(time, infect(an20), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-plot(time, infect(ar22), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(ar20), 'color', colors(2,:), 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 16)
@@ -22,7 +22,7 @@ set(gca, 'YLim', [1 1200])
 set(gca, 'YTick', [0 200 400 600 800 1000 1200])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
-xlabel('Time (hours)')
+xlabel('Days Post-Infection')
 ylabel('Infected Cell Count')
 title('aH5N1 Influenza')
 legend('None', 'RANTES', 2)
@@ -31,11 +31,11 @@ legend('None', 'RANTES', 2)
 
 subplot(1, 3, 2)
 
-plot(time, infect(sn22), 'color', colors(1,:), 'LineWidth', 2)
+plot(time, infect(sn20), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-plot(time, infect(sr22), 'color', colors(2,:), 'LineWidth', 2)
-plot(time, infect(si22), 'color', colors(3,:), 'LineWidth', 2)
-plot(time, infect(sb22), 'color', colors(4,:), 'LineWidth', 2) 
+plot(time, infect(sr20), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(si20), 'color', colors(3,:), 'LineWidth', 2)
+plot(time, infect(sb20), 'color', colors(4,:), 'LineWidth', 2) 
 hold off
 
 set(gca, 'FontSize', 16)
@@ -44,7 +44,7 @@ set(gca, 'YLim', [1 1200])
 set(gca, 'YTick', [0 200 400 600 800 1000 1200])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
-xlabel('Time (hours)')
+xlabel('Days Post-Infection')
 %ylabel('Infected Cell Count')
 title('sH1N1 Influenza')
 legend('None', 'RANTES', 'IP-10', 'Both', 2)
[…]

### 2012-07-13 data.m

@@ -69,7 +69,7 @@ set(haxes(1), 'XLim', [0 48])
 set(haxes(1), 'XTick', 0:4:48)
 set(haxes(1), 'XTickLabel', {'0', '', '8', '', '16', '', '24', '', '32', '', '40', '', '48'})
 xlabel('Time (hours)')
-ylabel('Virus (PFU/ml)')
+ylabel('Virus (PFU/mL)')
 
 axes(haxes(2))
 set(haxes(2), 'FontSize', 16)
@@ -79,7 +79,7 @@ set(haxes(2), 'YTick', [1 10 100 1000 10000 100000 1000000])
 set(haxes(2), 'XLim', [0 48])
 set(haxes(2), 'XTick', 0:4:48)
 set(haxes(2), 'XTickLabel', {'0', '', '8', '', '16', '', '24', '', '32', '', '40', '', '48'})
-%ylabel('Chemokine (ng/ml)')
+%ylabel('Chemokine (pg/mL)')
 
 lhandle = legend('Virus', 'IP-10', 'RANTES', 4);
 set(lhandle, 'Color', 'w')
@@ -124,7 +124,7 @@ set(haxes(1), 'XLim', [0 48])
 set(haxes(1), 'XTick', 0:4:48)
 set(haxes(1), 'XTickLabel', {'0', '', '8', '', '16', '', '24', '', '32', '', '40', '', '48'})
 xlabel('Time (hours)')
-%ylabel('Virus (PFU/ml)')
+%ylabel('Virus (PFU/mL)')
 
 axes(haxes(2))
 set(haxes(2), 'FontSize', 16)
@@ -134,7 +134,7 @@ set(haxes(2), 'YTick', [1 10 100 1000 10000 100000 1000000])
 set(haxes(2), 'XLim', [0 48])
 set(haxes(2), 'XTick', 0:4:48)
 set(haxes(2), 'XTickLabel', {'0', '', '8', '', '16', '', '24', '', '32', '', '40', '', '48'})
-%ylabel('Chemokine (ng/ml)')
+%ylabel('Chemokine (pg/mL)')
 
 lhandle = legend([hline1 hline3 hline2], 'Virus', 'IP-10', 'RANTES', 4);
 set(lhandle, 'Color', 'w')
@@ -175,7 +175,7 @@ set(haxes(1), 'XLim', [0 48])
 set(haxes(1), 'XTick', 0:4:48)
[…]

### 2012-07-13 infection.m

@@ -1,0 +1,35 @@
+% infection.m
+%   Returns expected infection values given a set of time points and
+%     coefficients
+% Inputs:
+%   a - List of free parameter values
+%   x - List of time points of interest
+% Output:
+%   y - A list of expected number of infected people for each time
+%       point in 'x'.
+function y = infection(a, timepoints)
+
+% Define beta and delta as globals so we can pass the values into the
+% 'sir' function.
+global IP10 RANTES;
+IP10 = a(1);
+%RANTES = a(1);
+
+% Solve our model.  The initial population was 50,000 healthy
+% individuals and one infected person. Run the model until the
+% last time point we are concerned about.
+% (we are assuming x is ordered).
+
+options = ddeset('InitialY', [1e6 0 0 1e4 0 0 0]);
+
+%sol = dde23('ode', [600 600 1292.28 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Avian
+%sol = dde23('ode', [600 600 1416.73 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Seasonal
+sol = dde23('ode', [600 600 1262.99 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Swine
+
+% Recall that we are interested only in the infected population.
+% We will use the built-in interpolation function interp1 to get
+% values of infected people at that correspond to the proper time
+% points.
+
+y = interp1(sol.x, log10(sol.y(6,:)), timepoints); % IP10
+%y = interp1(sol.x, log10(sol.y(7,:)), timepoints);  % RANTES

### 2012-07-13 log_squared_error.m

@@ -1,0 +1,20 @@
+% log_squared_error(data_time, data_points, fn_time, fn_points)
+%   Calculates the log squared error between a set of data points and a
+% series of points representing a function output.
+%
+% Inputs:
+%   data_time   - A 1D array containing the time points of the data
+%   data_points - A 1D array of the data values
+%   fit_time    - A 1D array of the time points for the function output
+%   fit_points  - A 1D array of the function values
+% Output:
+%   log_error = the sum of the squared residuals of the log data vs. 
+%     the log of the function
+
+function log_error = log_squared_error(data_time, data_points, fn_time, fn_points)
+
+    % Intepolate the fit to match the time points of the data
+    fn_values = interp1(fn_time, fn_points, data_time);
+    
+    % Square the error values and sum them up
+    log_error = sum((log10(fn_values) - log10(data_points)).^2);

### 2012-07-13 ode.m

@@ -1,0 +1,42 @@
+function dx = ode(t, x, lag)
+    
+    global IP10 RANTES;
+
+    ylag1 = lag(:,1);
+    ylag2 = lag(:,2);
+    ylag3 = lag(:,3);
+    ylag4 = lag(:,4);
+    ylag5 = lag(:,5);
+
+     % Avian
+%    B = 8.82931e-9;
+%    X = .001001;
+%    P = .0032662;
+%    E = 1e-8;
+%    Y = .0231;
+
+
+    % Seasonal
+%    B = 1.013e-8;
+%    X = .001;
+%    P = .0227;
+%    E = 1.6174e-6;
+%    Y = .0231;    
+
+    % Swine
+    B = 4.555e-8;
+    X = .001;
+    P = .304882;
+    E = .00344675;
+    Y = .0231;    
+
+    dx = zeros(7, 1);
+
+  	dx(1) = -B * x(1) * x(4);
+	dx(2) = B * x(1) * x(4) - B * ylag1(1) * ylag2(4);
+	dx(3) = B * ylag1(1) * ylag2(4) - X * x(3);
+	dx(4) = P * x(3) / (1 + E * x(5)) - B * x(1) * x(4);
+    dx(5) = ylag3(2);
+    dx(6) = IP10 * ylag4(2) - Y * x(6);
+	dx(7) = RANTES * ylag5(2) - Y * x(7);
+end

### 2012-07-13 plaquesize.m

@@ -10,11 +10,11 @@ colors = [.75  0  0 ;
 % Avian
 
 subplot(2, 3, 1)
-semilogy(time, infect(ar26)+ar26(:,5), 'color', colors(1,:), 'LineWidth', 2)
+semilogy(time, infect(ar20)+ar20(:,5), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-semilogy(time, ar26(:,3), 'color', colors(2,:), 'LineWidth', 2)
-semilogy(time, ar26(:,4), 'color', colors(3,:), 'LineWidth', 2)
-semilogy(time, ar26(:,7), 'color', colors(4,:), 'LineWidth', 2)
+semilogy(time, ar20(:,3), 'color', colors(2,:), 'LineWidth', 2)
+semilogy(time, ar20(:,4), 'color', colors(3,:), 'LineWidth', 2)
+semilogy(time, ar20(:,7), 'color', colors(4,:), 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 16)
@@ -25,18 +25,18 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Infected Cell Count (log10)')
-title('aH5N1')
+title('aH5N1 Influenza')
 %set(gca, 'FontSize', 12)
 legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2)
 
 % Seasonal
 
 subplot(2, 3, 2)
-semilogy(time, infect(sb26)+sb26(:,5), 'color', colors(1,:), 'LineWidth', 2)
+semilogy(time, infect(sb20)+sb20(:,5), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-semilogy(time, sb26(:,3), 'color', colors(3,:), 'LineWidth', 2)
-semilogy(time, sb26(:,4), 'color', colors(2,:), 'LineWidth', 2)
-semilogy(time, sb26(:,7), 'color', colors(4,:), 'LineWidth', 2)
+semilogy(time, sb20(:,3), 'color', colors(3,:), 'LineWidth', 2)
+semilogy(time, sb20(:,4), 'color', colors(2,:), 'LineWidth', 2)
+semilogy(time, sb20(:,7), 'color', colors(4,:), 'LineWidth', 2)
 hold off
 
[…]

### 2012-07-13 sensitivity.m

@@ -8,13 +8,13 @@ subplot(1, 3, 1)
 
 colors = winter(10);
 colors = [.75  0  0 ;
-          .8  .4  0 ;
-          .5  .5 .5 ;
-          .5  .5 .5 ;
-          .5  .5 .5 ;
-          .5  .5 .5 ;
-          .5  .5 .5 ;
+          .8  .2  0 ;
+          .85 .4  0 ;
           .1  .1 .9 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
+          .5  .5 .5 ;
           .5  .5 .5 ;
           .5  .5 .5 ;
           .5  .5 .5 ];
@@ -23,13 +23,13 @@ hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
 hold on
 hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);
 hline3 = plot(time, infect(ar19), 'color', colors(3,:), 'LineWidth', 2);
-plot(time, infect(ar20), 'color', colors(4,:), 'LineWidth', 2) 
-plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2) 
-plot(time, infect(ar22), 'color', colors(6,:), 'LineWidth', 2) 
-plot(time, infect(ar23), 'color', colors(7,:), 'LineWidth', 2)
-plot(time, infect(ar25), 'color', colors(9,:), 'LineWidth', 2) 
-plot(time, infect(ar26), 'color', colors(10,:), 'LineWidth', 2)
-hline4 = plot(time, infect(ar24), 'color', colors(8,:), 'LineWidth', 2);
+hline4 = plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2); 
+plot(time, infect(ar22), 'color', colors(6,:), 'LineWidth', 2); 
+plot(time, infect(ar23), 'color', colors(7,:), 'LineWidth', 2);
+plot(time, infect(ar24), 'color', colors(8,:), 'LineWidth', 2);
+plot(time, infect(ar25), 'color', colors(9,:), 'LineWidth', 2); 
+plot(time, infect(ar26), 'color', colors(10,:), 'LineWidth', 2);
[…]

### 2012-07-13 squared_error.m

@@ -1,0 +1,19 @@
+% squared_error(data_time, data_points, fn_time, fn_points)
+%   Calculates the squared error between a set of data points and a
+% series of points representing a function output.
+%
+% Inputs:
+%   data_time   - A 1D array containing the time points of the data
+%   data_points - A 1D array of the data values
+%   fit_time    - A 1D array of the time points for the function output
+%   fit_points  - A 1D array of the function values
+% Output:
+%   error = the sum of the squared residuals of the data vs. the function
+
+function error = squared_error(data_time, data_points, fn_time, fn_points)
+
+    % Intepolate the fit to match the time points of the data
+    fn_values = interp1(fn_time, fn_points, data_time);
+    
+    % Square the error values and sum them up
+    error = sum((fn_values - data_points).^2);

### 2012-07-13 stderror.m

@@ -1,0 +1,102 @@
+% Calculate F Ratio for Avian, Seasonal, and Swine runs
+
+% Avian
+aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infect(vara04) + ...
+        infect(vara05) + infect(vara06) + infect(vara07) + infect(vara08) + infect(vara09) + ...
+        infect(vara10) + infect(vara11) + infect(vara12) + infect(vara13) + infect(vara14) + ...
+        infect(vara15) + infect(vara16) + infect(vara17) + infect(vara18) + infect(vara19) + ...
+        infect(vara20) + infect(vara21) + infect(vara22) + infect(vara23) + infect(vara24) + ...
+        infect(vara25) + infect(vara26) + infect(vara27) + infect(vara28) + infect(vara29) + ...
+        infect(vara30) + infect(vara31) + infect(vara32) + infect(vara33) + infect(vara34) + ...
+        infect(vara35) + infect(vara36) + infect(vara37) + infect(vara38) + infect(vara39) + ...
+        infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
+        infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
+
+vara = [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
+        infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
+        infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
+        infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
+        infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
+        infect(vara25) infect(vara26) infect(vara27) infect(vara28) infect(vara29) ...
+        infect(vara30) infect(vara31) infect(vara32) infect(vara33) infect(vara34) ...
+        infect(vara35) infect(vara36) infect(vara37) infect(vara38) infect(vara39) ...
+        infect(vara40) infect(vara41) infect(vara42) infect(vara43) infect(vara44) ...
+        infect(vara45) infect(vara46) infect(vara47) infect(vara48) infect(vara49)];
+    
+
+[rows, cols] = size(vara);
+astderr = zeros(rows, 1);
+
[…]

### 2012-07-13 variance.m

@@ -15,7 +15,7 @@ aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infe
         infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
         infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
     
-plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
+hline1 = plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
             infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
             infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
             infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
@@ -27,25 +27,13 @@ plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(v
             infect(vara45) infect(vara46) infect(vara47) infect(vara48) infect(vara49)], 'color', [.75 .75 .75]);
         
 hold on
-%plot(time, infect(vara00), 'color', [.75 0 0]) % .8992
-%plot(time, infect(vara01), 'color', [.75 0 0]) % .8142
-%plot(time, infect(vara05), 'color', [.75 0 0]) % .8712
-plot(time, infect(vara08), 'color', [.75 0 0]) % .7343
-plot(time, infect(vara12), 'color', [.75 0 0]) % .7980
-%plot(time, infect(vara13), 'color', [.75 0 0]) % .8703
-%plot(time, infect(vara14), 'color', [.75 0 0]) % .8464
-%plot(time, infect(vara17), 'color', [.75 0 0]) % .8752
-%plot(time, infect(vara18), 'color', [.75 0 0]) % .8871
-%plot(time, infect(vara20), 'color', [.75 0 0]) % .8392
-%plot(time, infect(vara28), 'color', [.75 0 0]) % .8347
-%plot(time, infect(vara30), 'color', [.75 0 0]) % .8856
-%plot(time, infect(vara34), 'color', [.75 0 0]) % .8626
[…]

### 2012-07-17 paper/chemokine.tex
Getting ready for distribution to outside readers

@@
   Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
-  Viral Decay &  $1/day$ & Soumya \\
+  Viral Decay &  $1/day$ & Phenomenological \\
   Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
   Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
-  Infection Sensitivity &  $2 h/virion$ & [1*] \\
+  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
   Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
   Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
-  Apoptosis Time & $1 h$ & Fred \\
-  T-Cell Production Rate & $777/h$ & Soumya, Miao fits \\ 
-  Circulation Time & $6 m$ & Fred, mouse model \\
-  Search Time & $10 m$ & Fred, mouse model \\
-  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
-  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
+  Apoptosis Time & $1 h$ & Phenomenological \\
+  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
+  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
+  Search Time & $10 m$ & \cite{Banerjee2010b} \\
+  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
+  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
   T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
   T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
-  Cell Radius & $25 \mu m$ & Drew? \\
-  T-Cell Age (Active) & $2 h$ & Fred \\
-  T-Cell Age (Other) & $3 d$ & Fred \\
+  Cell Radius & $25 \mu m$ & Phenomenological \\
+  T-Cell Age (at FOI) & $2 h$ & [REF] \\
+  T-Cell Age (Other) & $3 d$ & [REF] \\
   T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3/day$ after day 3 & Soumya [3*] \\
+  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
   \hline  
@@
   \hline
-  IP-10 Production Rate &  & [2*]\\
+  IP-10 Production Rate &  & ODE Fits \\
   Avian H5N1 & 2.0e-4 pg/s$\cdot$cell & \\
@@
   \hline
-  RANTES Production Rate & & [2*] \\
+  RANTES Production Rate & & ODE Fits \\
   Avian H5N1 & 1.3e-5 pg/s$\cdot$cell& \\
@@
 \end{table}
-
-\begin{itemize}
-\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/mL.
-Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.
-This probability scales linearly with the viral concentration.
-\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
-\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
-\end{itemize}
 
@@
  \end{center}
-\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.
+\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
 IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-The ten runs were initialized identically for each strain.
-Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/mL).}
+Each run was initialized identically for each strain.
+The middle line shows the average while the outer lines show the 95\% confidence interval.}
  \label{fig:variance}
@@
 
+Our model combines the dynamics of human epithelial tissue with the scaling of a mouse's immune system.
+We choose to use mouse scale models because of the computational limitations of our ABM.
+By limiting our model to mice, we are able to keep the active agent population below 100,000, which allows a single model to be run in less than a day.
+A human-sized model would be computationally intractable.
+We find this practice acceptable for two reasons.
+First, the size differences and scaling principles between mice and human has been well documented [REF?] and can be accounted for.
+Second, this paper focuses on relative differences between strains so an exact recreation of the immune response is not necessary.
+
+In our model, the ability of T-cells to detect a chemokine signal and their subsequent motility upon doing so rests entirely on the T-cell's sensitivity to chemokine parameter.
+Once a T-cell encounters a concentration of chemokine above the sensitivity threshold it will immediately climb the chemotactic gradient to the local maxima.
+Studies of T-cell chemotaxis show that T-cell movement depends more on the magnitude of the chemokine gradient rather than the concentration \cite{Gao2003}.
+Furthermore, T-cells will perform a biased random walk rather than moving in a straight line directly up the concentration gradient [REF?].
+We experimented with biased random walks in our model and noticed no change in any results.
+Thus, we chose to keep the model simple and limited T-cell movement to directly climbing the chemokine gradient.
+
 % Do NOT remove this, even if you are not including acknowledgments

### 2012-08-02 paper/chemokine.tex
Paper split into Main and Supplement

@@
 \begin{center}
-\includegraphics[width=0.4\textwidth]{SystemChart}
+\includegraphics[width=0.6\textwidth]{SystemChart}
 \end{center}
@@
 \end{center}
-\caption{Computational model: Components and state transitions.}
+\caption{A visual representation of the model.
+Healthy epithelial cells are infected by virus.
+These infected cells secrete more virus after an incubation delay.
+T-cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.
+Once inside the lung tissue, T-cells climb the chemokine gradient and induce apoptosis in infected cells.}
 \label{fig:modelchart}

### 2012-08-02 paper/chemokine_main.tex
Paper split into Main and Supplement

--- 2012-07-13 paper/chemokine.tex
@@
 
-During the primary immune response clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.
-T-cell navigation through the complex branching bronchial vascular network is made possible by local cytokine and chemokine production from infected epithelial cells.
+During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.
+T-cell navigation through the complex branching bronchial vascular network is enabled by local cytokine and chemokine production from infected epithelial cells.
 Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (lab results only) with IP-10 always dominating the migration direction.
-A delayed differential equation model was fit to the empirical data revealing per cell chemokine production rates for each strain.
-These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects inter-strain variation of chemokine production on T-cell recruitment.
+Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.
+A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.
+These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment.
+In the model immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.
 The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.
-Expanding plaque sizes further isolate infected cells, making T-cell discovery and subsequent clearance of the infection less effective over time.
-The immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.
+Expanding plaque sizes isolate infected cells, causing T-cell discovery and subsequent clearance of the infection to become less effective over time.
 
@@
 
-The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
-Understanding all components of defense mechanisms employed will lead to improved vaccine design and strategies to control immunopathology.
-Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including the functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.
-CD8+ T cells are necessary for the resolution of pneumonia and the complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.
+The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that controls all but the most virulent strains.
+Understanding its components and their interactions are relevant to improving vaccines and strategies to control immunopathology.
+Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.
+CD8+ T cells are necessary for the resolution of pneumonia and complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.
 
-The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve T and B cell precursors are encountered and activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.
-T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.
+The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naive T and B cell precursors are activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.
+T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneously in the lung, spleen and other organs 4-5 days post-infection, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.
 In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
-Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
+Once in the tissue, T cells follow a chemokine gradient to the infected epithelial cells secreting the chemokine.
 
-A number of mathematical models have used detailed mouse model data to abstract the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.
-Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using ordinary differential equations to predict temporal events, however, can not account for the spatial constraints on cell behavior, and therefore make assumptions about the efficiency of each functional component in the model.
+Mathematical models have used detailed mouse data to characterize the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.
+Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
+Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the spread and interaction of populations.
 
 Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
-To develop a model with tractable simulations the model does not incorporate some key components of host defense and therefore does not aim to predict the true outcome for each strain examined.
+To develop a model with tractable simulations, some key components of host defense are not incorporated explicitly and the model therefore does not aim to predict the true outcome for each strain examined.
 We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
@@
 
-To provide data on the local chemokine-secretion environment in which T cells migrate, we used our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
-We selected in vitro chemokine data since this tissue environment surrogate may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and also permits the creation of sustained chemokine gradients within the model lung.
+To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
+We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.
 Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells.
+[REF]
 
 The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
-While we found this to be true, we observed in this spatially explicit model three interesting characteristics of chemokine-directed migration not revealed from experimental data.
-First, of the only two chemokines stimulated by all 3 strains, the influence of IP-10 consistently dominated over the effect of RANTES, regardless of strain.
-Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at higher concentrations above the threshold.
-Finally, when the search space (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient and may be reflected in the loss of control of the pandemic strain.
+While we found this to be true, we observed three interesting characteristics of chemokine-directed migration not revealed from experimental data.
+First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.
+Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.
+Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.
 
@@
 \begin{center}
-\includegraphics[width=0.4\textwidth]{SystemChart}
+\includegraphics[width=0.5\textwidth]{SystemChart}
 \end{center}
-\caption{A region of infected tissue of radius $r$ (shaded region $A$) expressing chemokines and inflammatory signals.
-This region is surrounded by region $B$ which does not have any infected cells and hence does not express either chemokines or inflammatory signals.
-The $2^{14}$ capillaries (red circles) are distributed evenly throughout the entire lung (region $A$ and $B$).}
+\caption{A simplified model of T cell search.
+Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.
+Upon reaching a capillary, T cells will exit into the tissue in the presence of a cytokine signal.
+T cells that do not sense a signal will continue to recirculate either through the lymph network or through the pulmonary vein back to the top of the network.}
 \label{fig:systemchart}
@@
 
-Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response.
-A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
+Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
+A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
 To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells.
 Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
-Activated T-cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
-T-cells that do not encounter cytokine recirculate to the lymph node.
-Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.
+Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
+T cells that do not encounter cytokine recirculate to the lymph node.
+At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.
 
@@
 Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-The chemokine diffuses from secreting cells, thus increasing the size of the signaling area.
-After a delay to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue.
-Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
-Figure \ref{fig:modelchart} illustrates the model components
-(epithelial cells, T-cells, and particles) and how they transition between different states.
+Chemokine also diffuses from secreting cells, creating a ball of stimulation around the infected region.
+After a delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin exiting the lymph node and circulating through the vasculature to the tissue.
+Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
 
@@
 \end{center}
-\caption{Computational model: Components and state transitions.}
+\caption{A visual representation of the model.
+Healthy epithelial cells infected by virus begin secreting virus after an incubation delay.
+T cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.
+Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.
+Solid arrows represent the transition of a cell from one behavioral state to another.
+Dashed arrows display the mechanism used to induce a transition.
+Dotted arrows show the production of new virus.}
 \label{fig:modelchart}
@@
 
-Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
-\emph{healthy} cells remain unchanged unless infected by virus.
-Once infected, there is a 10h incubation window before the cell transitions from {incubating} to \emph{expressing}.
+Epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
+\emph{Healthy} cells remain unchanged unless infected by virus.
+Once infected, the cell transitions from {incubating} to \emph{expressing}.
 \emph{Expressing} cells secrete virus and chemokine for 17 hours and then die.
 \emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells.
-Apoptoic cells continue to secrete virus and die one hour after their transition.
+Apoptoic cells continue to secrete virus until they die one hour after their transition.
 \emph{Dead} cells remain inert and do not regenerate over the course of an infection.
 
-T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{activated}.
-T-Cells begin emerging from the lymph node at five days post infection at the rate of 777 T-cells per hour for the duration of the simulation.
-T-Cells enter the vascular system in the \emph{searching} state, arriving at a random location on the lung's surface.
-\emph{searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{circulating} if they fail to encounter chemokine.
-\emph{Circulating} cells spend six minutes recirculating to the lymph node where they are converted to \emph{searching} and reintroduced to a new random location in the lung.
-If a \emph{searching} T-Cell encounters chemokine, it immedaiately converts to \emph{activated} and begins climbing the chemotactic gradient to the source of infection.
-\emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}.
-Activated T-cells are removed from the simulation using an exponential decay rate with an average lifespan of two hours.
+T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{chemotaxing}.
+\emph{Searching} T-Cells begin emerging from the lymph node at five days post infection.
+These T-Cells arrive at a random location on the lung's surface, perform a random walk in the tissue for 10 minutes and transition to \emph{circulating} in the absence of chemokine.
+\emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung.
+When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins climbing the chemotactic gradient to the source of infection.
+\emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells.
+\emph{Chemotaxing} decay exponentially with an average lifespan of two hours.
 
@@
 Chemokine diffuses across the tissue but has no
-direct effect beyond activating T-Cells. Both particle types decay over time.
+direct effect beyond activating T-Cells. Both particle types decay exponentially.
 
-\subsection*{Parameters}
-
-Define CyCells environment.  Talk about the need for spatial model.  Human/mouse scaling.  Parameter definitions.
+Parameters that are consistent between every model are shown in Table \ref{table:parameters}.  Strain-specific values are shown in Table \ref{table:strains}.
 
 \begin{table}
+\begin{center}
 \begin{tabular}{ | c | c | c | }
@@
   Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
-  Viral Decay &  $1/day$ & Soumya \\
+  Viral Decay &  $1/day$ & Phenomenological \\
   Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
-  Infection Sensitivity &  $2 h/virion$ & [1*] \\
+  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
+  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
   Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
   Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
-  Apoptosis Time & $1 h$ & Fred \\
-  T-Cell Production Rate & $777/h$ & Soumya, Miao fits \\ 
-  Circulation Time & $6 m$ & Fred, mouse model \\
-  Search Time & $10 m$ & Fred, mouse model \\
-  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
-  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
+  Apoptosis Time & $1 h$ & Phenomenological \\
+  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
+  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
+  Search Time & $10 m$ & \cite{Banerjee2010b} \\
+  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
+  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
   T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
-  T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
-  Cell Radius & $25 \mu m$ & Drew? \\
-  T-Cell Age (Active) & $2 h$ & Fred \\
-  T-Cell Age (Other) & $3 d$ & Fred \\
+  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
+  Cell Radius & $25 \mu m$ & Phenomenological \\
+  T-Cell Age (at FOI) & $2 h$ & [REF] \\
+  T-Cell Age (Other) & $3 d$ & [REF] \\
   T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3/day$ after day 3 & Soumya [3*] \\
+  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
   \hline  
@@
 \label{table:parameters}
+\end{center}
 \end{table}
@@
 
-\begin{itemize}
-\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/mL.
-Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.
-This probability scales linearly with the viral concentration.
-\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
-\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
-\end{itemize}
-
 \section*{Materials}
@@
 Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
-Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
+Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
 Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
@@
 \cite{Mitchell2011} to represent chemokine production from infected cells.
-Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.
-The model estimates are shown in Table~\ref{table:strains}.
+Model fits, shown in Table \ref{table:strains}, were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.
 These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.
-There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.
+There is no discernible correlation between viral production and induced chemokine production across the three strains.
 
@@
 For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $pg/mL$ (dotted line).
-These values were not included in model fitting.}
+sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (dotted line).}
  \label{fig:data}
-\end{figure}
-
-\subsection*{Stochastic modeling effects}
-
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).
-To test the strength of this effect, we ran each model fifty times using the default parameters given in Table~\ref{table:parameters}.
-$R^2$ values were calculated for individual runs versus the average of all runs.
-Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.
-sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.
-aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.
-
-Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
-Therefore the figures approximate plaque growth over time.
-
-Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the presence of IgM.
-In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.
-
-The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
-The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
-By the time T-Cells arrive, the plaque is too large for them to cover effectively.
-Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.
-aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{variance}
- \end{center}
-\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.
-IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-The ten runs were initialized identically for each strain.
-Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/mL).}
- \label{fig:variance}
 \end{figure}
@@
 
-T-Cells exit a capillary when they sense inflammatory cytokine and then follow the chemokine gradient in tissue.
-We use the chemokine gradient as a proxy for the inflammatory response.
+T-Cells exit a capillary when they sense inflammatory cytokine and follow the chemokine gradient.
+The model uses chemokine gradient as a surrogate for the inflammatory response.
 Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
@@
 To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 100 and 10 $ng/ml$, below which there is no discernible effect on infection kinetics.
-Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.
-Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $nM$ ($10 ng/mL$ assuming a chemokine weight of $10 kDa$), therefore we use that sensitivity for each model run.
+We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on infection kinetics.
+Because multiple T-cells in the same area do not increase apoptosis, we hypothesize that above a critical threshold no additional benefit is conferred by increasing T cell presence.
+We use the sensitivity of 10 $ng/mL$ for each run, which corresponds to a $1 nM$ concentration assuming a chemokine molecular weight of 10 $kDa$ \cite{Gao2003}.
 
@@
 
-\subsection*{Chemokine combinations}
-
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.
-We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).
-Because of this behavior IP-10 was not included in the aH5N1 model runs.
-
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
-The lack of both chemokines leads to runaway infections in all three strains.
-The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
-IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
-
-\begin{figure}[ht!]
-\begin{center}
-	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
-B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
-\end{figure}
-
 \subsection*{Spatial effects}
 
-Spatial effects of viral and chemokine diffusion play an important role in both the spread and clearance of infection.
+Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing infections.
 Free virus particles diffuse from virus secreting cells and infect healthy cells.
 Chemokine produced by infected cells attracts T-cells to the infected cells.
-Although virus is produced at a higher rate than chemokine, it diffuses much more slowly because it is a larger particle.
-Chemokine diffusion, however, is limited by higher decay rates than the virus.
+Although virus is produced at a higher rate than chemokine, the larger particle diffuses much more slowly.
+Chemokine diffusion decays more quickly.
 These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
 
-Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
-Over time, cells in the plaque's interior die causing active cells to comprise a decreasing proportion of the plaque.
-T-cells arrive at day 5 and begin to killing the virus-expressing cells.
-By day 6 a majority of the expressing cells have been eliminated as the plaque becomes dominated by dead cells.
-In the H5N1 example the plaque is dense, allowing T-cells to find secreting easily.
-Thus the infection can be eliminated.
-However, in both H1N1 simulations secreting cells are not eliminated.
-In these two cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
-T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth.
-This leads to a lower average T-cell killing rate.
-Furthermore, the regions of concentrated chemokine lag behind the cell and virus spatial layout.
-It takes time for newly secreting cells to produce chemokine and for pockets of high chemokine density to decay.
-Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind cellular changes in the plaque.
-The delayed response and the low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
+Up to day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
+Over time, cells in the plaque's interior die, and active cells are a decreasing proportion of the plaque.
+T-cells arrive at day 5 and begin killing the virus-secreting cells.
+By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
+In H5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.
+However, in both H1N1 simulations secreting cells were not eliminated.
+Expressing cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
+T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.
+Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.
+It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.
+Thus, T-cells can lag behind cellular changes in the plaque.
+Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
+
+The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
+\ref{fig:sensitivity}).
+T-cells are unable to kill cells that have not yet presented antigen.
+At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}).
+By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.
+The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.
+
+A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
+That is, a new viral particle is produced approximately every 200 seconds.
+Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit production to 18 new viral particles.
+T-Cells alone cannot contain the pH1N1 infection.
+In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
+Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
 
@@
  \end{center}
-\caption{Example simulated sH1N1 infection.
+\caption{A sH1N1 infection in CyCells.
 Screenshots taken at Day 4, Day 5.5, and Day 7.
@@
 \end{figure}
-
-\subsection*{Infection rebound}
-
-The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
-\ref{fig:variance}).
-This appears to be an artifact of the spatial nature of the ABM.
-Why can the immune response create an immediate decrease in the number of pH1N1 infected cells at day 5, only to fail in the same endeavor two days later?
-Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).
-Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.
-At day 5.5, the ratio of secreting cells to the total number of cells in the FOI is high (Fig.~\ref{fig:cycells}).
-By day 7, the ratio of secreting cells to total plaque size is approximately 1:100 for both the pH1N1 and sH1N1 strains.
-The greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to control (but not eliminate) the infection.
-
-A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
-That is, a new viral particle is produced approximately every 200 seconds.
-Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.
-Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.
-In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing a T-cell to limit a single infected cell to 1.3 viral particles in the one-hour window.
-Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
 
@@
 \section*{Discussion}
-
-\subsection*{ABM Advantages}
-
-The use of an ABM has advantages over a spatially homogeneous ODE model.
-An ODE model assumes that any virus particle is capable of infecting any healthy cell.
-Figure \ref{fig:cycells} shows that this is clearly not the case.
-In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
-ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
-This is clearly false as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
-
-Another advantage of our ABM is the ability to render the model using OpenGL (Fig.~\ref{fig:cycells}).
-The ability to see the model in action reveals spatial dynamics that are absent in ODE models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.
-For example, observing the simulation reveals challenges to T-cell detection of infected cells that are not apparent using standard population plots.
-By watching the model progress, we observed three key phenomena.
-First, because T-cells find infected cells by climbing the chemokine gradient, we see T-cell clustering at local maxima of chemokine concentration.
-Thus, we have an explanation for why T-cells do not increase in effectiveness as their numbers increase.
-Second, we notice that clumps of T-cells remain in place even after all virus-secreting cells have been killed off.
-This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.
-Finally, we can see that infected cells grow more disperse as the size of the infection grows.
-Because T-cells cluster together they become incapable of covering the increasing plaque size.
-These spatial observations provide explanations for the pH1N1 resurgence that would be hidden without the visualization tools provided by the ABM.
-
-\subsection*{General Immune Response Modeling}
-
-NEW
-
-The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
-Recent whole-animal models using a data-fitting approach and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
-We have focused exclusively on the problem of the search of activated CTL for infected foci in the lung, and as in other models have made a number of assumptions.
-
-The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve precursors are encountered and activated.
-T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood [Marshall 2001].
-In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
-Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
-One of the model simplifications we utilized was the equating cytokine and chemokine signals.
-
-A large fraction of the T cells entering lung tissue have come from the spleen where they proliferated under stimulation from antigen-loaded DC.
-This has been demonstrated by the significant loss of T cells entering lung following splenectomy differential [Tripp 1997] and by math modeling results that make sense only if there is a significant proliferation of activated cells outside the lymph node prior to arrival in the lung [Wu 2011].
-Our model did not include the proliferation in the temporary splenic reservoir as this uncertain degree of time delay in arrival between LN and lung would have complicated our ABM model.
-Using numbers of cells egressing from LN likely underestimated the rate of CTL population entering the lung, but our derivations of parameters impacting the T cell search and subsequent function are probably not affected.
-
-The fundamental mechanisms regulating the expansion of viral load and subsequent clearance of virus remain unclear, with current hypotheses focusing on antigen-driven, target cell susceptibility-driven, innate immune response-driven, antigen presentation limited, and adaptive immunity responses.
-We used the data from strain-specific viral replication in vitro in human bronchial epithelial cells [Mitchell 2011] to structure the regulation of viral abundance and viral clearance as essentially a model of antigen-driven, innate immunity-driven regulation.
-In a differential equation model, regulation of viral load was primarily driven by the availability of susceptible target cells [Baccam 2006].
-The question of target-driven versus antigen-driven regulation was addressed in a separate modeling study of infected equine lungs [Saenz 2010], finding that regulation was primarily innate immunity-driven since no more than 27\% of bronchial epithelial cells were ever infected during the course of the lung infection.
-In a comprehensive model of murine influenza rate-limiting antigen presentation was the primary influence on viral clearance [Lee 2009].
-In a subsequent model based on a large murine database, CD8+ T cells were critical components of the adaptive response in limiting the number of infected cells, with neutralizing antibodies responsible for clearing free virus [Miao 2010].
-Most modeling studies used a mouse-adapted virus to enable enumeration of epitope-specific T cells, making direct comparisons of replication efficiency with our human virus isolates more difficult.
-The mechanism of regulation of animal viral load may be specific to viral replication efficiencies, strain-specific suppression of innate immunity, and the phase of the immune response.
-We compared the three strains for their demands placed on the T cell search mechanism to find and kill infected cells.
-The multiplicity of potential regulation mechanisms does not invalidate our approach to the T cell search mechanisms.
-
-Our focus on the T cell search necessitated the simplification or removal of selected features of the mammalian immune response, in order to permit tractable simulations.
-Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
-Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
-The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable rates as indicated by superior data-fitting models [Wu 2011].
-Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection [Miao 2010].
-Since known critical components were omitted or simplified, our model was not designed to demonstrate clearance of virus from the lung.
-Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
 
@@
 
-NEW
+\subsection*{General Immune Response Modeling}
 
-In our model directed migration and viral infection control was not limited over a wide range of T cell sensitivity to the chemotactic signal.
-The density of CCR5 receptors on the surface of CD8+ T cells was a determinant of the strength of chemotactic attraction for CCR5 ligands such as CCL5 in an in vitro assay [Desmetz 2006].
-The density of CCR5 receptors on human CD8+ T cells vary over an order of magnitude, suggesting that sensitivity to the chemotactic signal could play a role in susceptibility to infection.
-CCR5-knockout mice are susceptible to fatal Cryptococcal encephalitis/meningitis [Huffnagle 1999] and West Nile virus encephalitis [Glass 2005].
-In the Cryptococcus-infected mice T cell recruitment to the brain, but not to the lung, was impaired.
-In CCR5-KO mice influenza A virus pneumonia was more severe associated with decreased T cell recruitment to the lung, while in CCR2-KO mice decreased T cell recruitment was due to decreased macrophage influx into the lung [Dawson 2000].
-On the other hand, CCL5-KO mice and CXCR3-KO mice (the receptor for CXCL10/IP-10) had similar viral kinetics and leukocyte recruitment compared to the wild-type mice [Wareing 2004].
-Finally, humans with the CCR5-D32 deletion allele may have increased susceptibility to the pandemic 2009 H1N1 influenza in a small sample [Keynan 2010].
-The question of limiting sensitivity has practical implications, since immunomodulation of cytokine/chemokine signaling to decrease influenza mortality is currently under investigation [Oslund 2011; Alleva 2011; Bassaganya-Riera 2010; Budd 2007; Moseley 2010].
-While blocking chemokine function may decrease inflammation and death from viral pneumonia, it may have an unintended consequence of impairing adaptive immunity.
-Thus, specific T cell recruitment to the influenza virus-infected lung should be quantitatively assessed in chemokine receptor-knockout mice.
+The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
+Recent whole-animal models using a data-fitting approach and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
+We have focused exclusively on the problem of the search of activated CTL for infected foci in the lung, and as in other models have made a number of assumptions.
 
-\subsection*{Caveats and Limitations}
+Our focus on the T cell search necessitated the simplification or removal of selected features of the mammalian immune response, in order to permit tractable simulations.
+Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
+Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
+The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
+All of these rates may in fact be time-variable rates as indicated by superior data-fitting models [Wu 2011].
+Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection [Miao 2010].
+Since known critical components were omitted or simplified, our model was not designed to demonstrate clearance of virus from the lung.
+Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
 
-In both H1N1 strains the T-cell response fails to clear the infection by day 10.
-In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.
-We suggest that this is due to the absence of several innate immune agents that were not included in the model.
-We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.
-Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.
-We see that the viral production rate is a major factor in the strength of the infection.
+\subsection*{ABM Advantages}
+
+The use of an ABM has certain advantages over a spatially homogeneous ODE model.
+An ODE model assumes that any virus particle is capable of infecting any healthy cell.
+Figure \ref{fig:cycells} shows that this is clearly not the case.
+In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
+ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
+This is problematic as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
+
+Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).
+Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.
+For example, observing the simulation reveals the spatial dynamics discussed in the Results.
+We observed three key phenomena.
+First, because T-cells find infected cells by climbing a gradient, we see T-cell clustering at local maxima of chemokine concentration, a posssible explanation for why T-cells do not increase in effectiveness as their numbers increase.
+Second, T-cells bunching persists after all virus-secreting cells have been eliminated.
+The local chemokine maxima takes time to diffuse and decay so that T-cells can climb the gradient to a new maximum.
+Finally, we can see that infected cells are more disperse as infection size grows.
+Because T-cells are clustered they cannot cover the increasing plaque effectively.
+These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
 

### 2012-08-02 paper/chemokine_supplement.tex
Paper split into Main and Supplement

--- 2012-07-13 paper/chemokine.tex
@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration (SUPPLEMENT)}
 }
@@
 
-% Please keep the abstract between 250 and 300 words
-\section*{Abstract}
-
-During the primary immune response clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.
-T-cell navigation through the complex branching bronchial vascular network is made possible by local cytokine and chemokine production from infected epithelial cells.
-Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (lab results only) with IP-10 always dominating the migration direction.
-A delayed differential equation model was fit to the empirical data revealing per cell chemokine production rates for each strain.
-These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects inter-strain variation of chemokine production on T-cell recruitment.
-The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.
-Expanding plaque sizes further isolate infected cells, making T-cell discovery and subsequent clearance of the infection less effective over time.
-The immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.
-
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step. 
-% Author Summary not valid for PLoS ONE submissions.   
-\section*{Author Summary}
-
-\section*{Introduction}
-
-The adaptive immune response induced during an acute influenza A virus infection is a complex web of defense mechanisms able to control all but the most virulent strains.
-Understanding all components of defense mechanisms employed will lead to improved vaccine design and strategies to control immunopathology.
-Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including the functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.
-CD8+ T cells are necessary for the resolution of pneumonia and the complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.
-
-The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve T and B cell precursors are encountered and activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.
-T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.
-In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
-Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
-
-A number of mathematical models have used detailed mouse model data to abstract the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.
-Such whole-response modeling has considerable potential in selecting the most promising strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using ordinary differential equations to predict temporal events, however, can not account for the spatial constraints on cell behavior, and therefore make assumptions about the efficiency of each functional component in the model.
-
-Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
-To develop a model with tractable simulations the model does not incorporate some key components of host defense and therefore does not aim to predict the true outcome for each strain examined.
-We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
-The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.
-A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.
-While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.
-
-To provide data on the local chemokine-secretion environment in which T cells migrate, we used our data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
-We selected in vitro chemokine data since this tissue environment surrogate may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and also permits the creation of sustained chemokine gradients within the model lung.
-Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells.
-
-The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
-While we found this to be true, we observed in this spatially explicit model three interesting characteristics of chemokine-directed migration not revealed from experimental data.
-First, of the only two chemokines stimulated by all 3 strains, the influence of IP-10 consistently dominated over the effect of RANTES, regardless of strain.
-Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at higher concentrations above the threshold.
-Finally, when the search space (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient and may be reflected in the loss of control of the pandemic strain.
-
 % You may title this section "Methods" or "Models". 
@@
 
-\subsection*{Computational Modeling}
+\subsection*{Ordinary Differential Equation Models}
+Here we use a differential equation model:
 
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics[width=0.4\textwidth]{SystemChart}
-\end{center}
-\caption{A region of infected tissue of radius $r$ (shaded region $A$) expressing chemokines and inflammatory signals.
-This region is surrounded by region $B$ which does not have any infected cells and hence does not express either chemokines or inflammatory signals.
-The $2^{14}$ capillaries (red circles) are distributed evenly throughout the entire lung (region $A$ and $B$).}
-\label{fig:systemchart}
-\end{figure}
+\begin{equation}
+\frac{dN_{c}}{dt} = \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} ,  \pi r'^{2} = a \sqrt{I} + b
+\end{equation}
+\begin{equation}
+\frac{dN'_{f}}{dt} = \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} ,  \pi r^{2} = \pi r_{cell}^{2} \cdot I
+\end{equation}
+\begin{equation}
+\frac{dN_{f}}{dt} = \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2}
+\end{equation}
+\begin{equation}
+\label{tcl_eqn1}
+\frac{dT}{dt} = \rho T -\beta TV
+\end{equation}
+\begin{equation}
+\label{tcl_eqn2}
+\frac{dI}{dt} = \beta TV - \delta I - k_{e} N_{f} I
+\end{equation}
+\begin{equation}
+\label{tcl_eqn4}
+\frac{dV}{dt} = pI - \beta TV - \gamma (t) V
+\end{equation}
+\begin{equation}
+\label{is_eqn5}
+\gamma (t) = \left\{ \begin{array}{rcl}
+	1/\mbox{day} & \mbox{,}  & t < 5  \\
+	3/\mbox{day} & \mbox{,} & t \geq 5  
+	\end{array}\right.
+\end{equation}
 
-Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for implementing two- or three-dimensional agent-based simulations of immune response.
-A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig 1) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
-To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells.
-Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
-Activated T-cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
-T-cells that do not encounter cytokine recirculate to the lymph node.
-Once at the site of infection, a T-Cell may encounter an infected epithelial cell and induce apoptosis.
+where $N_{c}$ is the number of circulating activated antigen-specific T-cells, $N'_{f}$ is the number of circulating T-cells that have found and exit into a region of lung tissue expressing chemokines and $N_{f}$ is the number of circulating T-cells that have found an infected region.
+$T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
+The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
+The lung is assumed to be a circular region with radius $R$.
+The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
+The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
 
-The simulation begins when a single cell in the center of the tissue is infected.
-After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine.
-Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-The chemokine diffuses from secreting cells, thus increasing the size of the signaling area.
-After a delay to simulate lymph node stimulation and T-Cell proliferation, activated T-cells begin exiting the lymph node and circulating through the vasculature to the tissue.
-Because T-Cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
-Figure \ref{fig:modelchart} illustrates the model components
-(epithelial cells, T-cells, and particles) and how they transition between different states.
-
-\subsection*{Model Definition}
-
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics[width=\textwidth]{ModelChart}
-\end{center}
-\caption{Computational model: Components and state transitions.}
-\label{fig:modelchart}
-\end{figure}
-
-Epithelial cells are stationary and can be in of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
-\emph{healthy} cells remain unchanged unless infected by virus.
-Once infected, there is a 10h incubation window before the cell transitions from {incubating} to \emph{expressing}.
-\emph{Expressing} cells secrete virus and chemokine for 17 hours and then die.
-\emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells.
-Apoptoic cells continue to secrete virus and die one hour after their transition.
-\emph{Dead} cells remain inert and do not regenerate over the course of an infection.
-
-T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{activated}.
-T-Cells begin emerging from the lymph node at five days post infection at the rate of 777 T-cells per hour for the duration of the simulation.
-T-Cells enter the vascular system in the \emph{searching} state, arriving at a random location on the lung's surface.
-\emph{searching} cells perform a random walk in the tissue for 10 minutes, after which they transition to \emph{circulating} if they fail to encounter chemokine.
-\emph{Circulating} cells spend six minutes recirculating to the lymph node where they are converted to \emph{searching} and reintroduced to a new random location in the lung.
-If a \emph{searching} T-Cell encounters chemokine, it immedaiately converts to \emph{activated} and begins climbing the chemotactic gradient to the source of infection.
-\emph{Activated} T-Cells move continuously up the gradient and if they encounter \emph{expressing} epithelial cells, the epithelial cell transitions \emph{apotosis}.
-Activated T-cells are removed from the simulation using an exponential decay rate with an average lifespan of two hours.
-
-The model contains two kinds of particles: virus and chemokine.
-Both are produced at constant rates by \emph{expressing} epithelial cells.
-Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration.
-Chemokine diffuses across the tissue but has no
-direct effect beyond activating T-Cells. Both particle types decay over time.
-
-\subsection*{Parameters}
-
-Define CyCells environment.  Talk about the need for spatial model.  Human/mouse scaling.  Parameter definitions.
-
-\begin{table}
-\begin{tabular}{ | c | c | c | }
-  \hline                        
-  Paramter & Value & Source \\
-  \hline
-  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
-  Viral Decay &  $1/day$ & Soumya \\
-  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Chemokine Decay &  $3.8508e-4/s$ & 30m half-life \\
-  Infection Sensitivity &  $2 h/virion$ & [1*] \\
-  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
-  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
-  Apoptosis Time & $1 h$ & Fred \\
-  T-Cell Production Rate & $777/h$ & Soumya, Miao fits \\ 
-  Circulation Time & $6 m$ & Fred, mouse model \\
-  Search Time & $10 m$ & Fred, mouse model \\
-  T-Cell Speed (Search) & $30 \mu m/s$ & Fred ? \\
-  T-Cell Speed (Activated) & $3 \mu m/s$ & Fred ?\\
-  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
-  T-Cell Expected Kill Rate & $10 m$ & Fred/Drew \\
-  Cell Radius & $25 \mu m$ & Drew? \\
-  T-Cell Age (Active) & $2 h$ & Fred \\
-  T-Cell Age (Other) & $3 d$ & Fred \\
-  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3/day$ after day 3 & Soumya [3*] \\
-  \hline  
-\end{tabular}
-\caption{Generic model parameters}
-\label{table:parameters}
-\end{table}
-
-\begin{table}
-\begin{center}
-\begin{tabular}{ | r | c | c | }
-  \hline                        
-  Paramter & Value & Source \\
-  \hline
-  IP-10 Production Rate &  & [2*]\\
-  Avian H5N1 & 2.0e-4 pg/s$\cdot$cell & \\
-  Seasonal H1N1 & 1.8e-4 pg/s$\cdot$cell& \\
-  Pandemic H1N1 & 8.7e-5 pg/s$\cdot$cell& \\
-  \hline
-  RANTES Production Rate & & [2*] \\
-  Avian H5N1 & 1.3e-5 pg/s$\cdot$cell& \\
-  Seasonal H1N1 & 8.9e-7pg/s$\cdot$cell& \\
-  Pandemic H1N1 & 4.4e-6 pg/s$\cdot$cell& \\
-  \hline
-  Virus Production Rate &  & \cite{Mitchell2011} \\
-  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
-  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell& \\
-  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell& \\  
-  \hline  
-\end{tabular}
-\caption{Strain specific model parameters}
-\label{table:strains}
-\end{center}
-\end{table}
-
-\begin{itemize}
-\item[1] The molecular concentration where there is one molecule per eptithelial cell volume is 1.53e-17 mol/mL.
-Thus, if you multiply the local viral concentration by 9.05e12 you get an infection event after two hours given that concentration.
-This probability scales linearly with the viral concentration.
-\item[2] These are calculated off of Fred's new data sets using the Mitchell ODE model and parameters as a basis.
-\item[3] IgM is represented by increasing the viral decay rate to 3/day after the third day of the simulation.
-\end{itemize}
-
-\section*{Materials}
-
-Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
-Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either a seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or an avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
-Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
-Quantitative viral culture was performed by standard plaque assay.
-Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
+Circulating activated antigen-specific T-cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$.
+These circulating cells then transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$).
+Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
+Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T-cells ($N_{f}$) at a rate $k_{e}$.
+Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
+Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
+The initial viral titer and the initial number of target cells are denoted $V_{0}$ and $T_{0}$, respectively.
+The initial number of infected cells is assumed to be zero.
 
@@
 \subsection*{Chemokine production}
-
-Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
-IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..
-To estimate per-cell production rates, we extended the ODE model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells.
-Model fits were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.
-The model estimates are shown in Table~\ref{table:strains}.
-These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.
-There is no discernible correlation between viral production and induced chemokine production across the three strains of influenza.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{data}
- \end{center}
-\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).
-For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values above the measurement threshold of 8,500 $pg/mL$ (dotted line).
-These values were not included in model fitting.}
- \label{fig:data}
-\end{figure}
 
@@
  \end{center}
-\caption{Model results: Time series plots of ten runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections.
+\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
 IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-The ten runs were initialized identically for each strain.
-Each simulation was run with identical T-cell sensitivity to chemokine (100 ng/mL).}
+Each run was initialized identically for each strain.
+The middle line shows the average while the outer lines show the 95\% confidence interval.}
  \label{fig:variance}
-\end{figure}
-
-\subsection*{T-cell sensitivity to chemokine}
-
-T-Cells exit a capillary when they sense inflammatory cytokine and then follow the chemokine gradient in tissue.
-We use the chemokine gradient as a proxy for the inflammatory response.
-Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
-
-To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of (FIXME: new values) 100 and 10 $ng/ml$, below which there is no discernible effect on infection kinetics.
-Because multiple T-cells in the same area do not increase the rate at which apoptosis is induced, we hypothesize that a critical number of T-cells are required to clear an infection above which there is no additional benefit gained from increasing T-cell presence.
-Reference \cite{Gao2003} reports T-cell chemotaxis at concentrations as low as 1 $nM$ ($10 ng/mL$ assuming a chemokine weight of $10 kDa$), therefore we use that sensitivity for each model run.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{sensitivity}
- \end{center}
-\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
-Total number of infected, expressing and apoptotic cells are plotted for each infection.
-The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.
-}
- \label{fig:sensitivity}
 \end{figure}
@@
 
-\subsection*{Spatial effects}
-
-Spatial effects of viral and chemokine diffusion play an important role in both the spread and clearance of infection.
-Free virus particles diffuse from virus secreting cells and infect healthy cells.
-Chemokine produced by infected cells attracts T-cells to the infected cells.
-Although virus is produced at a higher rate than chemokine, it diffuses much more slowly because it is a larger particle.
-Chemokine diffusion, however, is limited by higher decay rates than the virus.
-These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
-
-Early in the infection (up to day 4) the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
-Over time, cells in the plaque's interior die causing active cells to comprise a decreasing proportion of the plaque.
-T-cells arrive at day 5 and begin to killing the virus-expressing cells.
-By day 6 a majority of the expressing cells have been eliminated as the plaque becomes dominated by dead cells.
-In the H5N1 example the plaque is dense, allowing T-cells to find secreting easily.
-Thus the infection can be eliminated.
-However, in both H1N1 simulations secreting cells are not eliminated.
-In these two cases, expressing cells account for at most 10\% of the active cell population and less than  1\% of the total plaque at 6 days p.i.
-T-cells still accumulate, but they arrive at a slower rate than the plaque's rate of growth.
-This leads to a lower average T-cell killing rate.
-Furthermore, the regions of concentrated chemokine lag behind the cell and virus spatial layout.
-It takes time for newly secreting cells to produce chemokine and for pockets of high chemokine density to decay.
-Thus, T-cells, which rely on the chemokine gradient to find infected cells, often lag behind cellular changes in the plaque.
-The delayed response and the low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{cycells}
- \end{center}
-\caption{Example simulated sH1N1 infection.
-Screenshots taken at Day 4, Day 5.5, and Day 7.
-The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).
-Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
- \label{fig:cycells}
-\end{figure}
-
-\subsection*{Infection rebound}
-
-The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
-\ref{fig:variance}).
-This appears to be an artifact of the spatial nature of the ABM.
-Why can the immune response create an immediate decrease in the number of pH1N1 infected cells at day 5, only to fail in the same endeavor two days later?
-Figure \ref{fig:variance} reports only the number of infected cells, not counting dead cells in the focus of infection (FOI).
-Further, T-Cells induce apoptosis in virus-secreting cells, but not in virus-incubating cells.
-At day 5.5, the ratio of secreting cells to the total number of cells in the FOI is high (Fig.~\ref{fig:cycells}).
-By day 7, the ratio of secreting cells to total plaque size is approximately 1:100 for both the pH1N1 and sH1N1 strains.
-The greater replication rate of pH1N1 widens the gap (Fig.~\ref{fig:plaquesize}C) whereas the lower replication rate of sH1N1 allows the T-cells to control (but not eliminate) the infection.
-
-A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
-That is, a new viral particle is produced approximately every 200 seconds.
-Assuming that virus particle secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit a single infected cell to producing 18 new viral particles.
-Without additional immune responses, T-Cells alone cannot contain the pH1N1 infection.
-In contrast, the sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing a T-cell to limit a single infected cell to 1.3 viral particles in the one-hour window.
-Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{plaquesize}
- \end{center}
-\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.}
- \label{fig:plaquesize}
-\end{figure}
-
 \section*{Discussion}
 
-\subsection*{ABM Advantages}
-
-The use of an ABM has advantages over a spatially homogeneous ODE model.
-An ODE model assumes that any virus particle is capable of infecting any healthy cell.
-Figure \ref{fig:cycells} shows that this is clearly not the case.
-In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
-ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
-This is clearly false as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
-
-Another advantage of our ABM is the ability to render the model using OpenGL (Fig.~\ref{fig:cycells}).
-The ability to see the model in action reveals spatial dynamics that are absent in ODE models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.
-For example, observing the simulation reveals challenges to T-cell detection of infected cells that are not apparent using standard population plots.
-By watching the model progress, we observed three key phenomena.
-First, because T-cells find infected cells by climbing the chemokine gradient, we see T-cell clustering at local maxima of chemokine concentration.
-Thus, we have an explanation for why T-cells do not increase in effectiveness as their numbers increase.
-Second, we notice that clumps of T-cells remain in place even after all virus-secreting cells have been killed off.
-This is because it takes time for the local chemokine maxima to diffuse and decay to a point where the T-cells can once again climb the gradient to a new maxima.
-Finally, we can see that infected cells grow more disperse as the size of the infection grows.
-Because T-cells cluster together they become incapable of covering the increasing plaque size.
-These spatial observations provide explanations for the pH1N1 resurgence that would be hidden without the visualization tools provided by the ABM.
-
 \subsection*{General Immune Response Modeling}
-
-NEW
 
@@
 
-\subsection*{Caveats and Limitations}
-
-In both H1N1 strains the T-cell response fails to clear the infection by day 10.
-In the case of pandemic H1N1, the infection continues to accelerate after a short period of clearance.
-We suggest that this is due to the absence of several innate immune agents that were not included in the model.
-We chose to remove most of the innate response in order to better focus on the relative effects of the cytokine signaling and T-cell response.
-Although the infections do not clear by day 10, the models show clear differences in the ability of the immune system to deal with influenza strains of varying replication rates.
-We see that the viral production rate is a major factor in the strength of the infection.
-
-% Do NOT remove this, even if you are not including acknowledgments
-\section*{Acknowledgments}
-
 \section*{References}

### 2012-08-02 paper/chemokine_main.tex
small edits

@@
 The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 100 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
+For the threshold of 10 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
 This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.

### 2012-08-02 paper/chemokine_supplement.tex
small edits

@@
 
+\subsection*{Caveats and Limitations}
+
+In both H1N1 strains, the simulated T-cell response fails to clear the infection by day 10.
+In the case of pandemic H1N1, the infection rebounds after initial clearance.
+We suggest that this is due to the absence of several innate immune agents that were not included in the model.
+By removing most of the innate response we were able to focus on the relative effects of the cytokine signaling and T-cell response.
+Although H1N1 infections do not clear by day 10, the models show clear differences in capability of managing influenza strains of varying replication rates.
+In this model, viral production rate is a major factor in the strength of the infection.
+
+Our model combines the dynamics of human epithelial tissue with the scaling of a mouse's immune system.
+We choose to use mouse scale models because ofcomputational limitations.
+In our murine-based model the number of cells remained below 100,000, allowing a single model to be run in less than a day.
+A human-sized model would be computationally intractable at our level of specificity with current technology.
+We use known scaling principals between mice and humans to account for the differences in size \cite{Banerjee2010b}.
+
+In our model T-cell motion depends on the T-cells' sensitivity to chemokine.
+Once a T-cell encounters a chemokine concentration above the sensitivity threshold it climbs the chemotactic gradient to the local maxima.
+Studies of T-cell chemotaxis show that T-cell movement depends more on the magnitude of the chemokine gradient rather than the concentration \cite{Gao2003}.
+Furthermore, T-cells will perform a biased random walk rather than moving in a straight line directly up the concentration gradient [REF].
+Biased random walks showed no change in model results, thus we limit T-cell movement to direct motion along the gradient for simplicity.
+
 \section*{References}

### 2012-08-09 paper/chemokine_main.tex
Formatting changes, working on table 2

@@
 \begin{center}
-\begin{tabular}{ | r | c | c | }
+\begin{tabular}{ | r | c | c | c | }
   \hline                        
-  Paramter & Value & Source \\
+  Parameter & Value & 95\% Confidence & Source \\
+   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \\
   \hline
-  IP-10 Production Rate &  & ODE Fits \\
-  Avian H5N1 & 2.0e-4 pg/s$\cdot$cell & \\
-  Seasonal H1N1 & 1.8e-4 pg/s$\cdot$cell& \\
-  Pandemic H1N1 & 8.7e-5 pg/s$\cdot$cell& \\
+  \textbf{IP-10 Production Rate} & & & ODE Fits \\
+  Avian H5N1 & 2.0e-4 & \small{8.4e-5 --- 4.2e-4} & \\
+  Seasonal H1N1 & 1.8e-4 & \small{1.2e-4 --- 3.0e-4} & \\
+  Pandemic H1N1 & 8.7e-5 & \small{1.7e-5 --- 7.1e-4}  & \\
   \hline
-  RANTES Production Rate & & ODE Fits \\
-  Avian H5N1 & 1.3e-5 pg/s$\cdot$cell& \\
-  Seasonal H1N1 & 8.9e-7pg/s$\cdot$cell& \\
-  Pandemic H1N1 & 4.4e-6 pg/s$\cdot$cell& \\
-  \hline
-  Virus Production Rate &  & \cite{Mitchell2011} \\
-  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell& \\
-  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell& \\
-  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell& \\  
+  \textbf{RANTES Production Rate} & & & ODE Fits \\
+  Avian H5N1 & 1.3e-5 & \small{7.9e-6 --- 1.9e-5} & \\
+  Seasonal H1N1 & 8.9e-7 & \small{4.8e-7 --- 1.6e-6} & \\
+  Pandemic H1N1 & 4.4e-6 & \small{5.0e-7 --- 3.6e-5} & \\
+%  \hline
+  \textbf{Virus Production Rate} & & & \cite{Mitchell2011} \\
+  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell & & \\
+  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell & & \\
+  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell & & \\  
   \hline  
@@
 For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (dotted line).}
+sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
+Solid lines show model fits to IP-10 and RANTES data.}
  \label{fig:data}
@@
 
-Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing infections.
+Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.
 Free virus particles diffuse from virus secreting cells and infect healthy cells.
@@
 Although virus is produced at a higher rate than chemokine, the larger particle diffuses much more slowly.
-Chemokine diffusion decays more quickly.
+Chemokine decays more quickly.
 These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
@@
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-In H5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.
+In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.
 However, in both H1N1 simulations secreting cells were not eliminated.
-Expressing cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
+Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
 T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.
@@
 T-cells are unable to kill cells that have not yet presented antigen.
-At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}).
+At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
 By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.
@@
 In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
-Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after T-cell detection.
+Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.
 
@@
  \end{center}
-\caption{Comparison of total plaque size (green), number of virus incubating cells (blue), number of virus secreting cells (red), and T-cells (indigo) for the simulated infections of H5N1, sH1N1, and pH1N1.}
+\caption{Comparison of total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red) for the simulated infections of aH5N1, sH1N1, and pH1N1.
+T cells are able to clear both secreting and incubating cells in aH5N1, are unable to contain incubating cells in sH1N1, and are unable to contain either type of infected cell in pH1N1.}
  \label{fig:plaquesize}

### 2012-08-09 bootstrap.m

@@ -1,0 +1,29 @@
+% Bootstrapping
+
+results = zeros(1000, 1);
+
+% REMINDER: Set appropriate parameters in ode.m and infection.m
+
+t = wit;
+r = wir;
+[row, col] = size(t);
+
+statopts = statset('Display', 'off');
+
+for i=1:1000
+   x = log10(wi);
+   for j=1:col
+       x(j) = x(j) + r(ceil(rand*col));
+   end
+%   x = max(0, x);
+   
+   results(i) = nlinfit(t*60, x, 'infection', .00025, statopts);
+   disp(i)
+end
+
+results = sort(results);
+low = results(25);
+high = results(976);
+disp(low/60)
+disp(high/60)
+

### 2012-08-09 data.m

@@ -1,6 +1,6 @@
 % Creates Data Plot
 
-figure('Position',[100 500 1800 700])
+figure('Position',[100 500 1800 600])
 
 % --- Avian Data ---
 avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
@@ -36,6 +36,15 @@ wi  = [ 75+1, 1835, 1349, 23458, 12073, 30700, 19814] - 75;
 wrt = [480, 720,  960, 1200, 1440, 1920, 2880]/60;
 wr  = [ 36,  55,   46,  150,   93,  380, 1224];
 
+colors = [.75   0   .2 ;
+          1    .75   0 ;
+          .28  .625  0 ;
+          .1   .1   .9 ;
+          .5    0   .1 ;
+          .75  .625  0 ;
+          .2   .375  0 ;
+          .075 .075 .7 ];
+
 % --- Plot ---
 
 % Avian
@@ -43,25 +52,45 @@ subplot(1, 3, 1)
 [haxes, hline1, hline2] = plotyy(avt, av, art, ar, 'semilogy', 'semilogy');
 hold on
 hline3 = plot(ait, ai);
+hline4 = plot(aft/60, aif);
+hline5 = plot(aft/60, arf);
 hold off
 
 set(hline1, 'LineWidth', 2)
+set(hline1, 'LineStyle', 'none')
 set(hline1, 'Marker', 'o')
 set(hline1, 'MarkerSize', 10)
-set(hline1, 'Color', 'b')
+set(hline1, 'Color', colors(4,:))
+set(hline1, 'MarkerEdgeColor', colors(8,:))
+set(hline1, 'MarkerFaceColor', colors(4,:))
 
 set(hline2, 'LineWidth', 2)
+set(hline2, 'LineStyle', 'none')
 set(hline2, 'Marker', 's')
 set(hline2, 'MarkerSize', 10)
-set(hline2, 'Color', [0 .5 0])
+set(hline2, 'Color', colors(2,:))
+set(hline2, 'MarkerEdgeColor', colors(6,:))
+set(hline2, 'MarkerFaceColor', colors(2,:))
 
 set(hline3, 'LineWidth', 2)
+set(hline3, 'LineStyle', 'none')
 set(hline3, 'Marker', '^')
 set(hline3, 'MarkerSize', 10)
[…]

### 2012-08-09 infection.m

@@ -22,9 +22,9 @@ IP10 = a(1);
 
 options = ddeset('InitialY', [1e6 0 0 1e4 0 0 0]);
 
-%sol = dde23('ode', [600 600 1292.28 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Avian
+sol = dde23('ode', [600 600 1292.28 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Avian
 %sol = dde23('ode', [600 600 1416.73 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Seasonal
-sol = dde23('ode', [600 600 1262.99 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Swine
+%sol = dde23('ode', [600 600 1262.99 480 960], [0 0 0 0 0 0 0], [0 2880], options); % Swine
 
 % Recall that we are interested only in the infected population.
 % We will use the built-in interpolation function interp1 to get

### 2012-08-09 odea.m

@@ -1,6 +1,7 @@
-function dx = ode(t, x, lag)
+function dx = odea(t, x, lag)
     
-    global IP10 RANTES;
+    IP10 = 0.0118;
+    RANTES = 7.6064e-4;
 
     ylag1 = lag(:,1);
     ylag2 = lag(:,2);
@@ -14,21 +15,6 @@ function dx = ode(t, x, lag)
     P = .0032662;
     E = 1e-8;
     Y = .0231;
-
-
-    % Seasonal
-%    B = 1.013e-8;
-%    X = .001;
-%    P = .0227;
-%    E = 1.6174e-6;
-%    Y = .0231;    
-
-    % Swine
-%    B = 4.555e-8;
-%    X = .001;
-%    P = .304882;
-%    E = .00344675;
-%    Y = .0231;    
 
     dx = zeros(7, 1);
 

### 2012-08-09 odes.m

@@ -1,6 +1,7 @@
-function dx = ode(t, x, lag)
+function dx = odes(t, x, lag)
     
-    global IP10 RANTES;
+    IP10 = 0.01062;
+    RANTES = 5.34e-5;
 
     ylag1 = lag(:,1);
     ylag2 = lag(:,2);
@@ -8,27 +9,12 @@ function dx = ode(t, x, lag)
     ylag4 = lag(:,4);
     ylag5 = lag(:,5);
 
-     % Avian
-    B = 8.82931e-9;
-    X = .001001;
-    P = .0032662;
-    E = 1e-8;
-    Y = .0231;
-
-
     % Seasonal
-%    B = 1.013e-8;
-%    X = .001;
-%    P = .0227;
-%    E = 1.6174e-6;
-%    Y = .0231;    
-
-    % Swine
-%    B = 4.555e-8;
-%    X = .001;
-%    P = .304882;
-%    E = .00344675;
-%    Y = .0231;    
+    B = 1.013e-8;
+    X = .001;
+    P = .0227;
+    E = 1.6174e-6;
+    Y = .0231;    
 
     dx = zeros(7, 1);
 

### 2012-08-09 odew.m

@@ -1,6 +1,7 @@
-function dx = ode(t, x, lag)
+function dx = odew(t, x, lag)
     
-    global IP10 RANTES;
+    IP10 = 0.005202;
+    RANTES = 2.61e-4;
 
     ylag1 = lag(:,1);
     ylag2 = lag(:,2);
@@ -8,27 +9,12 @@ function dx = ode(t, x, lag)
     ylag4 = lag(:,4);
     ylag5 = lag(:,5);
 
-     % Avian
-    B = 8.82931e-9;
-    X = .001001;
-    P = .0032662;
-    E = 1e-8;
-    Y = .0231;
-
-
-    % Seasonal
-%    B = 1.013e-8;
-%    X = .001;
-%    P = .0227;
-%    E = 1.6174e-6;
-%    Y = .0231;    
-
     % Swine
-%    B = 4.555e-8;
-%    X = .001;
-%    P = .304882;
-%    E = .00344675;
-%    Y = .0231;    
+    B = 4.555e-8;
+    X = .001;
+    P = .304882;
+    E = .00344675;
+    Y = .0231;    
 
     dx = zeros(7, 1);
 

### 2012-08-09 plaquesize.m

@@ -1,23 +1,24 @@
 % Plaque Size Plots
 
-figure('Position',[100 500 1800 1400])
+figure('Position',[100 500 1800 500])
 
-colors = [.75  0  0 ;
-          .75  .5  0 ;
-          0  .75 0 ;
-          0 0 .75];
+colors = [.625  0  .2 ;
+          1 .75  0 ;
+          .28  .625 0 ;
+          .1 .1 .9];
 
 % Avian
 
-subplot(2, 3, 1)
-semilogy(time, infect(ar20)+ar20(:,5), 'color', colors(1,:), 'LineWidth', 2)
+subplot(1, 3, 1)
+semilogy(time, infect(ar20)+ar20(:,5), 'color', colors(4,:), 'LineWidth', 2)
 hold on
 semilogy(time, ar20(:,3), 'color', colors(2,:), 'LineWidth', 2)
 semilogy(time, ar20(:,4), 'color', colors(3,:), 'LineWidth', 2)
-semilogy(time, ar20(:,7), 'color', colors(4,:), 'LineWidth', 2)
+semilogy(time, ar20(:,7), 'color', colors(1,:), 'LineWidth', 2)
 hold off
 
-set(gca, 'FontSize', 16)
+set(gca, 'FontSize', 20)
+set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [0 6000])
 set(gca, 'YTick', [1 10 100 1000])
@@ -26,20 +27,21 @@ set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Infected Cell Count (log10)')
 title('aH5N1 Influenza')
-%set(gca, 'FontSize', 12)
-legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2)
-
+lh = legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2);
+set(lh, 'FontSize', 16)
+set(lh, 'FontWeight', 'normal')
 % Seasonal
 
-subplot(2, 3, 2)
-semilogy(time, infect(sb20)+sb20(:,5), 'color', colors(1,:), 'LineWidth', 2)
+subplot(1, 3, 2)
[…]

### 2012-08-09 sensitivity.m

@@ -1,15 +1,15 @@
 % Sensitivity Plot
 
-figure('Position',[100 500 1800 700])
+figure('Position',[100 500 1800 1200])
 
 % Avian
 
-subplot(1, 3, 1)
+subplot(2, 3, 1)
 
-colors = winter(10);
-colors = [.75  0  0 ;
-          .8  .2  0 ;
-          .85 .4  0 ;
+%colors = winter(10);
+colors = [.625  0  .2 ;
+          1  .75  0 ;
+          .28 .625  0 ;
           .1  .1 .9 ;
           .5  .5 .5 ;
           .5  .5 .5 ;
@@ -19,9 +19,9 @@ colors = [.75  0  0 ;
           .5  .5 .5 ;
           .5  .5 .5 ];
 
-hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
+hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);      
 hold on
-hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);
+hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
 hline3 = plot(time, infect(ar19), 'color', colors(3,:), 'LineWidth', 2);
 hline4 = plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2); 
 plot(time, infect(ar22), 'color', colors(6,:), 'LineWidth', 2); 
@@ -32,7 +32,8 @@ plot(time, infect(ar26), 'color', colors(10,:), 'LineWidth', 2);
 hline5 = plot(time, infect(ar20), 'color', colors(4,:), 'LineWidth', 2);
 hold off
 
-set(gca, 'FontSize', 16)
+set(gca, 'FontSize', 18)
+set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [1 1200])
 set(gca, 'YTick', [0 200 400 600 800 1000 1200])
@@ -41,26 +42,28 @@ set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Infected Cell Count')
 title('aH5N1 Influenza')
[…]

### 2012-08-15 paper/chemokine_main.tex
Current Draft after Steph Edits

@@
 
+\usepackage{multirow}
+\renewcommand{\arraystretch}{1.1}
+
 % figure files reside in the figures/ directory
@@
 
-During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci representing a tiny fraction of the total lung.
-T-cell navigation through the complex branching bronchial vascular network is enabled by local cytokine and chemokine production from infected epithelial cells.
+During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.
+T-cell navigation to the infected region is enabled by local cytokine and chemokine production from infected epithelial cells.
 Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the eight chemokines tested only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.
+Of the eight chemokines tested, only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.
 A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.
-These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment.
-In the model immune response is unable to clear the infection of pandemic H1N1 influenza due to the high rate of viral production and plaque size expansion.
+These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment and infection control.
+T cells appear to be sufficiently sensitive to chemokine to function as well as they possibly can.
+The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.
 The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.
-Expanding plaque sizes isolate infected cells, causing T-cell discovery and subsequent clearance of the infection to become less effective over time.
+Expanding plaque sizes isolate infected cells, impeding T-cell discovery and subsequent clearance of the infection over time.
 
@@
 
-Here we focus on the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
-To develop a model with tractable simulations, some key components of host defense are not incorporated explicitly and the model therefore does not aim to predict the true outcome for each strain examined.
-We ask how do very small foci of infected tissue attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
+Here we study the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
+To develop a model with tractable simulations, some components of host defense are abstracted and the model therefore does not aim to predict the true outcome for each strain examined.
+We ask how very small foci of infected tissue can attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
 The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.
@@
 We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.
-Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to populate the ABM model with activated CD8+ T cells.
+Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model.
 [REF]
 
-The standard hypothesis would posit that in this model the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
-While we found this to be true, we observed three interesting characteristics of chemokine-directed migration not revealed from experimental data.
+The standard hypothesis posits that the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
+Our model confirms this hypothesis, but we also observed three interesting characteristics of chemokine-directed migration not revealed by earlier data.
 First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.
@@
 \end{center}
-\caption{A simplified model of T cell search.
+\caption{Model of T cell search.
 Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.
-Upon reaching a capillary, T cells will exit into the tissue in the presence of a cytokine signal.
-T cells that do not sense a signal will continue to recirculate either through the lymph network or through the pulmonary vein back to the top of the network.}
+Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.
+In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
 \label{fig:systemchart}
@@
 
-Computational modeling was conducted using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations were conducted to measure efficiency of infection clearance under different environmental conditions.
-To model a growing infection, the lung was represented as a two dimensional sheet of healthy epithelial cells.
+Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
+A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
+The lung was represented as a two-dimensional sheet of healthy epithelial cells.
 Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
@@
 Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-Chemokine also diffuses from secreting cells, creating a ball of stimulation around the infected region.
-After a delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin exiting the lymph node and circulating through the vasculature to the tissue.
+Chemokine diffuses from secreting cells, creating a ball of stimulation around the infected region.
+After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue.
 Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
@@
 \end{center}
-\caption{A visual representation of the model.
-Healthy epithelial cells infected by virus begin secreting virus after an incubation delay.
-T cells are activated in the lymph node and traverse the bronchial vascular network to the site of infection where they are recruited by inflammatory cytokine.
+\caption{Visual representation of the model.
+Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.
+Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.
 Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.
-Solid arrows represent the transition of a cell from one behavioral state to another.
+Solid arrows represent a cell state transition from one behavior to another.
 Dashed arrows display the mechanism used to induce a transition.
-Dotted arrows show the production of new virus.}
+Dotted arrows indicate the production of new virus.}
 \label{fig:modelchart}
@@
 
-Epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
+In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
 \emph{Healthy} cells remain unchanged unless infected by virus.
@@
 
-T-cells have three states: \emph{searching}, \emph{recirculating}, and \emph{chemotaxing}.
+T-cells have three states.
 \emph{Searching} T-Cells begin emerging from the lymph node at five days post infection.
-These T-Cells arrive at a random location on the lung's surface, perform a random walk in the tissue for 10 minutes and transition to \emph{circulating} in the absence of chemokine.
+\textit{Searching} T-Cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine.
 \emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung.
-When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins climbing the chemotactic gradient to the source of infection.
+When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection.
 \emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells.
-\emph{Chemotaxing} decay exponentially with an average lifespan of two hours.
+\emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.
 
@@
 \end{tabular}
-\caption{Generic model parameters}
+\caption{Model parameters}
 \label{table:parameters}
@@
 \begin{table}
-\begin{center}
+\centering
 \begin{tabular}{ | r | c | c | c | }
   \hline                        
-  Parameter & Value & 95\% Confidence & Source \\
-   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \\
+  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
+   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \footnotesize{(PFU/s$\cdot$cell)} \\
   \hline
-  \textbf{IP-10 Production Rate} & & & ODE Fits \\
-  Avian H5N1 & 2.0e-4 & \small{8.4e-5 --- 4.2e-4} & \\
-  Seasonal H1N1 & 1.8e-4 & \small{1.2e-4 --- 3.0e-4} & \\
-  Pandemic H1N1 & 8.7e-5 & \small{1.7e-5 --- 7.1e-4}  & \\
+  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & \multirow{2}{*}{5.4e-5} \\
+   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \\ 
    \hline
-  \textbf{RANTES Production Rate} & & & ODE Fits \\
-  Avian H5N1 & 1.3e-5 & \small{7.9e-6 --- 1.9e-5} & \\
-  Seasonal H1N1 & 8.9e-7 & \small{4.8e-7 --- 1.6e-6} & \\
-  Pandemic H1N1 & 4.4e-6 & \small{5.0e-7 --- 3.6e-5} & \\
-%  \hline
-  \textbf{Virus Production Rate} & & & \cite{Mitchell2011} \\
-  Avian H5N1 & 5.4e-5 PFU/s$\cdot$cell & & \\
-  Seasonal H1N1 & 3.8e-4 PFU/s$\cdot$cell & & \\
-  Pandemic H1N1 & 5.1e-3 PFU/s$\cdot$cell & & \\  
+  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & \multirow{2}{*}{3.8e-4} \\
+   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} &  \\
+   \hline
+  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & \multirow{2}{*}{5.1e-3} \\
+   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} &  \\
   \hline
 \end{tabular}
-\caption{Strain specific model parameters}
+\caption{Strain-specific model parameters.
+Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
+Viral production values are taken from \cite{Mitchell2011}.}
 \label{table:strains}
-\end{center}
 \end{table}
@@
 Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
-IP-10 concentration increases were observed by 8h p.i., and RANTES by 16h p.i..
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
 To estimate per-cell production rates, we extended the ODE model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells.
-Model fits, shown in Table \ref{table:strains}, were generated for three influenza strains (Fig.~\ref{fig:data}) using a genetic algorithm \cite{Mitchell2011}.
-These expression rates are similar across different strains with the exception of significantly higher RANTES production in aH5N1.
-There is no discernible correlation between viral production and induced chemokine production across the three strains.
+\cite{Mitchell2011} to represent chemokine production from infected cells (Supplemental Methods).
+Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
+There is no positive correlation between viral production and induced chemokine production across the three strains.
 
@@
  \end{center}
-\caption{Empirical data of viral and cytokine titer for three strains of influenza: Avian H5N1 (A), Seasonal sH1N1 (B), and Pandemic pH1N1 (C).
-For each strain, the viral load (blue circle) is shown in PFU/mL and the chemokines IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values not included in model fitting above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
-Solid lines show model fits to IP-10 and RANTES data.}
+\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.
+Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
+sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
+Dashed lines show model fits to IP-10 and RANTES data.}
  \label{fig:data}
@@
 
-T-Cells exit a capillary when they sense inflammatory cytokine and follow the chemokine gradient.
-The model uses chemokine gradient as a surrogate for the inflammatory response.
+The model uses chemokine gradient as a surrogate for the inflammatory cytokines required for extravasation.
 Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-T-cell sensitivity determines how much aggregate chemokine signal is required to detect the gradient.
+However, T-cell sensitivity determines how strong the aggregate chemokine signal must be.
 
-To test for the minimum concentration of cytokine/chemokine signal required to recruit T-Cells from capillary, we simulated sensitivity levels ranging over seven orders of magnitude.
-We find a threshold (Fig.~\ref{fig:sensitivity}) between the concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on infection kinetics.
-Because multiple T-cells in the same area do not increase apoptosis, we hypothesize that above a critical threshold no additional benefit is conferred by increasing T cell presence.
-We use the sensitivity of 10 $ng/mL$ for each run, which corresponds to a $1 nM$ concentration assuming a chemokine molecular weight of 10 $kDa$ \cite{Gao2003}.
+We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement.
+Because multiple T-cells in the same area do not enhance control of infection, we hypothesize more T cells are not better above a critical threshold.
+We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
+Single runs were used due to computational limitations.
+Model variance is discussed in the Supplemental Results.
 
@@
  \end{center}
-\caption{Varying T-cell sensitivity to chemokine in the model: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+\caption{Varying T-cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
 Total number of infected, expressing and apoptotic cells are plotted for each infection.
@@
 Chemokine produced by infected cells attracts T-cells to the infected cells.
-Although virus is produced at a higher rate than chemokine, the larger particle diffuses much more slowly.
-Chemokine decays more quickly.
+Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.
 These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
-
-Up to day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
-Over time, cells in the plaque's interior die, and active cells are a decreasing proportion of the plaque.
+Until day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
+Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque.
 T-cells arrive at day 5 and begin killing the virus-secreting cells.
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily and infection is eliminated.
+In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily, and infection is eliminated.
 However, in both H1N1 simulations secreting cells were not eliminated.
@@
 It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.
-Thus, T-cells can lag behind cellular changes in the plaque.
+Thus, T-cells can fail to detect cellular changes in the plaque.
 Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
 
-The immune response fails to clear sH1N1 completely and is unable to contain pH1N1 (Fig.
-\ref{fig:sensitivity}).
 T-cells are unable to kill cells that have not yet presented antigen.
@@
 A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
-That is, a new viral particle is produced approximately every 200 seconds.
-Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell can do is to limit production to 18 new viral particles.
-T-Cells alone cannot contain the pH1N1 infection.
+That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.
+Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell could do is limit production to 18 new viral particles.
+Thus, T-Cells alone cannot contain the pH1N1 infection.
 In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
@@
  \end{center}
-\caption{A sH1N1 infection in CyCells.
-Screenshots taken at Day 4, Day 5.5, and Day 7.
-The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells introduced at day 5 (green).
+\caption{Simulated sH1N1 infection.
+Screenshots from day 4, day 5.5, and day 7.
+The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells arriving at day 5 (green).
 Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
@@
 
+\subsection*{Caveats}
+
+Parameters are estimated to the best of our knowledge when values could not be found in literature.
+We choose to ignore most aspects of the innate immune response to avoid over-parameterizing the model.
+The unbounded growth of pH1N1 can partially be attributed to the lack of many innate mechanisms.
+T cells are assumed to have the same level of sensitivity to both IP-10 and RANTES.
+Furthermore, T cells climb chemokine gradient in straight lines as our experiments with biased random walks showed no effect on model behavior.
+
 \subsection*{Chemokine Directed T-cell Search}
 
-Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006} but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.
+Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006}, but bloodstream levels of cytokines and chemokines may not reflect local concentrations or dynamics critical in guiding T cells to infected epithelial cells.
 We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.
-In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the initial infection naturally.
-The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h post-infection.
+In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the natural initial infection.
+The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h p.i..
 When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.
@@
 Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999} and CCL3 \cite{Kawai1999}.
-The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial in the accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
-Other chemokines are likely to play a role in T cell traffic.
+The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
+Other chemokines likely play a role in T cell trafficking.
 Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
-In the intact animal immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
-Finally, antigen-specific CD8+ T cell recognition of its target will induced lung injury mediated by TNFa and trigger addition alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
+In the intact animal, immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
+Finally, antigen-specific CD8+ T cell recognition of its target induces lung injury mediated by TNFa and triggers additional alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
 Thus our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
 
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
 The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
 For the threshold of 10 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
-This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+This calculation agrees with the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
 
-\subsection*{General Immune Response Modeling}
+\subsection*{Immune Response Modeling}
 
 The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
-Recent whole-animal models using a data-fitting approach and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
-We have focused exclusively on the problem of the search of activated CTL for infected foci in the lung, and as in other models have made a number of assumptions.
-
-Our focus on the T cell search necessitated the simplification or removal of selected features of the mammalian immune response, in order to permit tractable simulations.
+Recent whole-animal models using a data-fitting approach, and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
+We have focused exclusively on the problem of search by activated CTL for infected foci in the lung, and as in other models have made a number of assumptions and simplifications to permit tractable simulations.
 Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
@@
 The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable rates as indicated by superior data-fitting models [Wu 2011].
+All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.
 Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection [Miao 2010].
-Since known critical components were omitted or simplified, our model was not designed to demonstrate clearance of virus from the lung.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.
+Thus, our model is not intended to predict clearance of virus from the lung.
 Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
@@
 ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
-This is problematic as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
+This is limiting as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
 
 Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).
-Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems.
-For example, observing the simulation reveals the spatial dynamics discussed in the Results.
-We observed three key phenomena.
+Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlie.
 First, because T-cells find infected cells by climbing a gradient, we see T-cell clustering at local maxima of chemokine concentration, a posssible explanation for why T-cells do not increase in effectiveness as their numbers increase.
-Second, T-cells bunching persists after all virus-secreting cells have been eliminated.
+Second, T-cell clustering persists after all virus-secreting cells have been eliminated.
 The local chemokine maxima takes time to diffuse and decay so that T-cells can climb the gradient to a new maximum.

### 2012-08-15 paper/chemokine_supplement.tex
Formatting changes, working on table 2
Current Draft after Steph Edits

@@
 
+\usepackage{multirow}
+
 %% ** EDIT HERE **
@@
 \subsection*{Ordinary Differential Equation Models}
-Here we use a differential equation model:
 
-\begin{equation}
-\frac{dN_{c}}{dt} = \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} ,  \pi r'^{2} = a \sqrt{I} + b
-\end{equation}
-\begin{equation}
-\frac{dN'_{f}}{dt} = \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} ,  \pi r^{2} = \pi r_{cell}^{2} \cdot I
-\end{equation}
-\begin{equation}
-\frac{dN_{f}}{dt} = \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2}
-\end{equation}
-\begin{equation}
-\label{tcl_eqn1}
-\frac{dT}{dt} = \rho T -\beta TV
-\end{equation}
-\begin{equation}
-\label{tcl_eqn2}
-\frac{dI}{dt} = \beta TV - \delta I - k_{e} N_{f} I
-\end{equation}
-\begin{equation}
-\label{tcl_eqn4}
-\frac{dV}{dt} = pI - \beta TV - \gamma (t) V
-\end{equation}
-\begin{equation}
-\label{is_eqn5}
-\gamma (t) = \left\{ \begin{array}{rcl}
+To find a value for $\sigma$ we use a differential equation model from \cite{Miao20101}:
+
+\begin{equation*}
+\begin{aligned}
+\frac{dN_{c}}{dt} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}}    & \pi r'^{2} &= a \sqrt{I} + b \\
+\frac{dN'_{f}}{dt} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
+\frac{dN_{f}}{dt} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
+\frac{dT}{dt} &= \rho T -\beta TV \\
+\frac{dI}{dt} &= \beta TV - \delta I - k_{e} N_{f} I \\
+\frac{dV}{dt} &= pI - \beta TV - \gamma (t) V \\
+\gamma (t) &= \left\{ \begin{array}{rcl}
 	1/\mbox{day} & \mbox{,}  & t < 5  \\
 	3/\mbox{day} & \mbox{,} & t \geq 5  
-	\end{array}\right.
-\end{equation}
+	\end{array}\right. \\
+\end{aligned}
+\end{equation*}
 
@@
 The initial number of infected cells is assumed to be zero.
+This ode was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
 
@@
 \section*{Results}
-
-\subsection*{Chemokine production}
 
@@
 
-Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the presence of IgM.
+Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the addition of IgM.
 In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.
 
-The three strains show different levels of virulence, consistent with the results found in Mitchell (Fig.~\ref{fig:variance}).
+The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~\ref{fig:variance}).
 The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
-By the time T-Cells arrive, the plaque is too large for them to cover effectively.
 Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.
@@
 IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-Each run was initialized identically for each strain.
+Each run was initialized identically for each strain save for the random seed.
 The middle line shows the average while the outer lines show the 95\% confidence interval.}
@@
 
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.
+Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.
 We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).
-Because of this behavior IP-10 was not included in the aH5N1 model runs.
-
+Because of this behavior IP-10 was not included in the aH5N1 model runs.-
 Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
@@
 
-\section*{Discussion}
+\section*{Miscellaneous Text}
 
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{ | c | c | c | }
+  \hline                        
+  Paramter & Value & Source \\
+  \hline
+  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
+  Viral Decay &  $1/day$ & Phenomenological \\
+  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
+  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
+  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
+  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
+  Apoptosis Time & $1 h$ & Phenomenological \\
+  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
+  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
+  Search Time & $10 m$ & \cite{Banerjee2010b} \\
+  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
+  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
+  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
+  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
+  Cell Radius & $25 \mu m$ & Phenomenological \\
+  T-Cell Age (at FOI) & $2 h$ & [REF] \\
+  T-Cell Age (Other) & $3 d$ & [REF] \\
+  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
+  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
+  \hline  
+\end{tabular}
+\caption{Model parameters}
+\label{table:parameters}
+\end{center}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{ | r | c | c | c | }
+  \hline                        
+  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
+   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \footnotesize{(PFU/s$\cdot$cell)} \\
+  \hline
+  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & \multirow{2}{*}{5.4e-5} \\
+   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \\ 
+   \hline
+  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & \multirow{2}{*}{3.8e-4} \\
+   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} &  \\
+   \hline
+  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & \multirow{2}{*}{5.1e-3} \\
+   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} &  \\
+  \hline
+\end{tabular}
+\caption{Strain-specific model parameters.
+Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
+Viral production values are taken from \cite{Mitchell2011}.}
+\label{table:strains}
+\end{table}
+
 \end{document}

### 2012-08-15 plaquesize.m

@@ -73,7 +73,7 @@ set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 %ylabel('Infected Cell Count (log10)')
 title('pH1N1 Influenza')
-set(gca, 'FontSize', 12)
+%set(gca, 'FontSize', 12)
 % legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2)
 
 % T-Cell Ratios

### 2012-08-15 variance.m

@@ -27,10 +27,10 @@ hline1 = plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03)
             infect(vara45) infect(vara46) infect(vara47) infect(vara48) infect(vara49)], 'color', [.75 .75 .75]);
         
 hold on
-hline2 = plot(time, infect(vara12), 'color', [.75 0 0]); % .7363
-plot(time, infect(vara13), 'color', [.75 0 0]) % .7504
-plot(time, infect(vara33), 'color', [.75 0 0]) % .7988
-plot(time, infect(vara35), 'color', [.75 0 0]) % .5881
+%hline2 = plot(time, infect(vara12), 'color', [.75 0 0]); % .7363
+%plot(time, infect(vara13), 'color', [.75 0 0]) % .7504
+%plot(time, infect(vara33), 'color', [.75 0 0]) % .7988
+%plot(time, infect(vara35), 'color', [.75 0 0]) % .5881
 plot(time, aavg, '-k', 'LineWidth', 2)
 hline3 = plot(time, alow, 'Color', 'k', 'LineStyle', '--');
 plot(time, ahigh, 'Color', 'k', 'LineStyle', '--')
@@ -74,7 +74,7 @@ plot(time, [infect(vars00) infect(vars01) infect(vars02) infect(vars03) infect(v
             infect(vars45) infect(vars46) infect(vars47) infect(vars48) infect(vars49)], 'color', [.75 .75 .75]);
         
 hold on
-plot(time, infect(vars16), 'color', [.75 0 0])
+%plot(time, infect(vars16), 'color', [.75 0 0])
 plot(time, savg, '-k', 'LineWidth', 2)
 plot(time, slow, 'Color', 'k', 'LineStyle', '--')
 plot(time, shigh, 'Color', 'k', 'LineStyle', '--')

### 2012-08-23 paper/chemokine_main.tex
Version to be sent out for internal review

@@
 Melanie Moses$^{1}$, 
-Frederick Koster$^{2}$, 
+Frederick Koster$^{1,2}$, 
 \\
@@
 During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.
-T-cell navigation to the infected region is enabled by local cytokine and chemokine production from infected epithelial cells.
+T-cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.
+We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).
+
 Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the eight chemokines tested, only CXCL10 (IP-10) and CCL5 (RANTES) were secreted in tissue  (results only) with IP-10 always dominating the affected T-cell migration direction.
-A delayed differential equation model was fit to the empirical data, revealing per cell chemokine production rates for each strain.
-These values were coupled with known biological constants to calibrate a spatially explicit agent-based model to explore the effects of inter-strain variation of chemokine production on T-cell recruitment and infection control.
+Of the eight chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.
+A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment.
 T cells appear to be sufficiently sensitive to chemokine to function as well as they possibly can.
@@
   \hline
-  Viral Diffusion & .0318 & \cite{Beauchemin2006} \\
-  Viral Decay &  $1/day$ & Phenomenological \\
-  Chemokine Diffusion & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Chemokine Decay &  $3.8508$e-4$/s$ & 30m half-life \\
-  Infection Sensitivity &  $2 h/virion$ & Phenomenological \\
-  Incubation Time &  $10 h$ & \cite{Mitchell2011} \\
-  Expression Time &  $16.7 h$ & \cite{Mitchell2011} \\
-  Apoptosis Time & $1 h$ & Phenomenological \\
-  T-Cell Production Rate & $777/h$ & \cite{Miao2010} \\ 
-  Circulation Time & $6 m$ & \cite{Banerjee2010b} \\
-  Search Time & $10 m$ & \cite{Banerjee2010b} \\
+  Viral Diffusion in Airway & $.0318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Viral Decay in Airway &  $1/day$ & \cite{Lee2009} \\
+  Chemokine Diffusion Rate & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & Selected$^1$\\
+  Infection Sensitivity Rate &  $2$ \textit{hour/virion} &  Selected$^2$ \\
+  Incubation Time &  $10$ \textit{hours} & \cite{Mitchell2011} \\
+  Expression Time &  $16.7$ \textit{hours} & \cite{Mitchell2011} \\
+  Apoptosis Time & $1$ \textit{hour} & \cite{Ganusov2008} \\
+  T-Cell Production Rate & $777$\textit{/hour} & \cite{Miao2010} \\ 
+  Blood Circulation Time & $6$ \textit{min} & \cite{Banerjee2010b} \\
+  Search Time In Chemokine-Free Tissue & $10$ \textit{min} & \cite{Banerjee2010b} \\
   T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
-  T-Cell Speed (Activated) & $3 \mu m/s$ & \cite{Miller2003} \\
-  T-Cell Sensitivity & $1 ng/mL$ & \cite{Gao2003} \\
-  T-Cell Expected Kill Rate & $10 m$ & [REF] \\
-  Cell Radius & $25 \mu m$ & Phenomenological \\
-  T-Cell Age (at FOI) & $2 h$ & [REF] \\
-  T-Cell Age (Other) & $3 d$ & [REF] \\
-  T-Cell Ramp Up Time & $5 d$ & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3/day$ & Phenomenological \\
+  T-Cell Speed (Chemotaxis) & $3 \mu m/s$ & \cite{Miller2003} \\
+  T-Cell Sensitivity to Chemokine & $10 ng/mL$ & \cite{Gao2003} \\
+  T-Cell Expected Kill Time & $10$ \textit{min} & Selected$^3$ \\
+  Epithelial Cell Radius & $25 \mu m$ & Selected$^4$ \\
+  T-Cell Age (at FOI) & $2$ \textit{hours} & Selected$^5$ \\
+  T-Cell Age (in Blood) & $3$ \textit{days} & Selected$^5$ \\
+  Onset of ATC Lymph Node Exit & $5$ \textit{days} & \cite{Banerjee2011} \\
+  IgM Strength & Viral decay of $3$\textit{/day} & Selected$^6$ \\
   \hline  
 \end{tabular}
-\caption{Model parameters}
+\caption{Static model parameters.
+1) Chosen to correspond to a 30 minute half-life.
+2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.
+This scales linearly with the number of virions in the cell's vicinity.
+3) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.
+This rate does not scale with T cell numbers.
+4)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
+5) Parameters chosen after discussions with David Woodland, Trudeau Institute.
+6) IgM presence is abstracted by increasing viral decay by a factor of three.
+}
 \label{table:parameters}
@@
   \hline
-  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & \multirow{2}{*}{5.4e-5} \\
-   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \\ 
+  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
+   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
    \hline
-  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & \multirow{2}{*}{3.8e-4} \\
-   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} &  \\
+  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
+   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
    \hline
-  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & \multirow{2}{*}{5.1e-3} \\
-   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} &  \\
+  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
+   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
   \hline
@@
 
-We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement.
+We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement (model variance discussed in \textbf{S}2.1.).
+Seasonal and pandemic H1N1 models simulated both IP-10 and RANTES; avian H5N1 included only RANTES (\textbf{S2.2}).
 Because multiple T-cells in the same area do not enhance control of infection, we hypothesize more T cells are not better above a critical threshold.
@@
 Single runs were used due to computational limitations.
-Model variance is discussed in the Supplemental Results.
 
@@
 
-\subsection*{Caveats}
+%\subsection*{Caveats}
 
-Parameters are estimated to the best of our knowledge when values could not be found in literature.
+%Parameters are estimated to the best of our knowledge when values could not be found in literature.
 We choose to ignore most aspects of the innate immune response to avoid over-parameterizing the model.
@@
 
-\subsection*{Chemokine Directed T-cell Search}
+\subsection*{Modeling Methodology}
 
-Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006}, but bloodstream levels of cytokines and chemokines may not reflect local concentrations or dynamics critical in guiding T cells to infected epithelial cells.
-We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.
-In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the natural initial infection.
-The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h p.i..
-When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.
-Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains \cite{Chan2005, Chan2010, Zeng2011} and pandemic strains \cite{Zeng2011}.
-MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion \cite{Chan2010}.
-However, in spite of the attenuated type 1 interferon response induced by H5N1 strains \cite{Zeng2007}, the chemokine secretion is not attenuated.
-
-Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999} and CCL3 \cite{Kawai1999}.
-The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
-Other chemokines likely play a role in T cell trafficking.
-Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
-In the intact animal, immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
-Finally, antigen-specific CD8+ T cell recognition of its target induces lung injury mediated by TNFa and triggers additional alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
-Thus our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
-
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
-The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 10 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
-This calculation agrees with the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-
-\subsection*{Immune Response Modeling}
-
-The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
-Recent whole-animal models using a data-fitting approach, and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
-We have focused exclusively on the problem of search by activated CTL for infected foci in the lung, and as in other models have made a number of assumptions and simplifications to permit tractable simulations.
+Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.
 Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
@@
 Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
-
-\subsection*{ABM Advantages}
 
@@
 These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
+The behavior of searching T cells described in this paper
+can enhance future global immune response models.
+
+\subsection*{Chemokine Directed T-cell Search}
+
+Hypercytokinemia, or 'cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.
+The use of in vitro human bronchial epithelial cells \cite{Mitchell2011} infected at low cell:virus infection ratio (MOI) was intended to mimic initial natural infection and subsequent local infection spread.
+MOI and degree of differentiation of NHBE cells are critical determinants of cytokine/chemokine secretion \cite{Chan2010}.
+Yet, as we noted at low MOI, greater CCL5 secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.
+Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controling the infection.
+
+Our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
+Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999}, CCL3 \cite{Kawai1999}, and likely others.
+The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
+Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
+In the intact animal, immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
+Finally, antigen-specific CD8+ T cell recognition of its target induces lung injury mediated by TNFa and triggers additional alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
+
+A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 10 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 150 microns in our model.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
 
@@
 
-\section*{References}
+%\section*{References}
 % The bibtex filename

### 2012-08-23 paper/chemokine_supplement.tex
Version to be sent out for internal review

--- 2012-08-02 paper/chemokine_supplement.tex
@@
 % authors may use "Analysis" 
-\section*{Models}
+\section{Models}
 
-\subsection*{Ordinary Differential Equation Models}
-Here we use a differential equation model:
+\subsection{Delay Differential Equation Model}
 
-\begin{equation}
-\frac{dN_{c}}{dt} = \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} ,  \pi r'^{2} = a \sqrt{I} + b
-\end{equation}
-\begin{equation}
-\frac{dN'_{f}}{dt} = \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} ,  \pi r^{2} = \pi r_{cell}^{2} \cdot I
-\end{equation}
-\begin{equation}
-\frac{dN_{f}}{dt} = \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2}
-\end{equation}
-\begin{equation}
-\label{tcl_eqn1}
-\frac{dT}{dt} = \rho T -\beta TV
-\end{equation}
-\begin{equation}
-\label{tcl_eqn2}
-\frac{dI}{dt} = \beta TV - \delta I - k_{e} N_{f} I
-\end{equation}
-\begin{equation}
-\label{tcl_eqn4}
-\frac{dV}{dt} = pI - \beta TV - \gamma (t) V
-\end{equation}
-\begin{equation}
-\label{is_eqn5}
-\gamma (t) = \left\{ \begin{array}{rcl}
+We adapt the delay differential equations first presented in \cite{Mitchell2011} to find values for chemokine production rates.
+We use initial populations and parameter values from the previous study while adding an extra equation to model chemokine production (final equation):
+
+\begin{equation*}
+\label{EqS1}
+\begin{aligned}
+\dot{T} &= - \beta T V \\
+\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
+\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
+\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
+\dot{F} &=  I_1[t-\tau_2] \\
+\dot{C} &= r I[t-\tau_3] - d C \\
+\end{aligned}
+\tag{\textbf{S}1}
+\end{equation*}
+\vspace{.05in}
+
+$C$ represents the chemokine population.
+Two parameters, $r$ and $d$, control the rate of chemokine production and the rate of chemokine decay.
+Chemokine decay, $d$, was set to $3.8508\cdot10^{-4}$/s, which equates to a 30 minute half-life.
+$r$, being the only free parameter, was fit to the empirical data shown in Fig.
+1 of the main paper using Matlab's \texttt{nlinfit} function.
+This process was repeated six times, once for each set of data.
+$\tau_3$, the chemokine production delay, was set to 8 hours for IP-10 fits and 16 hours for RANTES fits.
+
+\subsection{T Cell Production Rate}
+
+To find a value for $\sigma$ we use a differential equation model from \cite{Miao20101}:
+
+\begin{equation*}
+\label{EqS2}
+\begin{aligned}
+\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}}    & \pi r'^{2} &= a \sqrt{I} + b \\
+\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
+\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
+\dot{T} &= \rho T -\beta TV \\
+\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
+\dot{V} &= pI - \beta TV - \gamma (t) V \\
+\gamma (t) &= \left\{ \begin{array}{rcl}
 	1/\mbox{day} & \mbox{,}  & t < 5  \\
 	3/\mbox{day} & \mbox{,} & t \geq 5  
-	\end{array}\right.
-\end{equation}
+	\end{array}\right. \\
+\end{aligned}
+\tag{\textbf{S}2}
+\end{equation*}
+\vspace{.05in}
 
@@
 The initial number of infected cells is assumed to be zero.
+This ode was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
 
 % Results and Discussion can be combined.
-\section*{Results}
+\section{Results}
 
-\subsection*{Chemokine production}
-
-\subsection*{Stochastic modeling effects}
+\subsection{Stochastic modeling effects}
 
@@
 
-\subsection*{Chemokine combinations}
+\subsection{Chemokine combinations}
 
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011} we hypothesize that it renders IP-10 ineffectual.
+Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.
 We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).
 Because of this behavior IP-10 was not included in the aH5N1 model runs.
-
 Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
@@
 
-\section*{Discussion}
-
-\subsection*{General Immune Response Modeling}
-
-The adaptive immune response to influenza in the mouse is a complex network of molecular and cellular elements, and incorporation of all of these into a comprehensive predictive model is not yet achieved.
-Recent whole-animal models using a data-fitting approach and extensive databases have enhanced our global understanding of response metrics that are difficult or impossible to measure.
-We have focused exclusively on the problem of the search of activated CTL for infected foci in the lung, and as in other models have made a number of assumptions.
-
-The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naïve precursors are encountered and activated.
-T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneous in the lung, spleen and other organs 4-5 days post-infection after to extravasation from the blood [Marshall 2001].
-In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
-Once in the tissue, T cells are guided up a chemokine gradient to the infected epithelial cell secreting the chemokine.
-One of the model simplifications we utilized was the equating cytokine and chemokine signals.
-
-A large fraction of the T cells entering lung tissue have come from the spleen where they proliferated under stimulation from antigen-loaded DC.
-This has been demonstrated by the significant loss of T cells entering lung following splenectomy differential [Tripp 1997] and by math modeling results that make sense only if there is a significant proliferation of activated cells outside the lymph node prior to arrival in the lung [Wu 2011].
-Our model did not include the proliferation in the temporary splenic reservoir as this uncertain degree of time delay in arrival between LN and lung would have complicated our ABM model.
-Using numbers of cells egressing from LN likely underestimated the rate of CTL population entering the lung, but our derivations of parameters impacting the T cell search and subsequent function are probably not affected.
-
-The fundamental mechanisms regulating the expansion of viral load and subsequent clearance of virus remain unclear, with current hypotheses focusing on antigen-driven, target cell susceptibility-driven, innate immune response-driven, antigen presentation limited, and adaptive immunity responses.
-We used the data from strain-specific viral replication in vitro in human bronchial epithelial cells [Mitchell 2011] to structure the regulation of viral abundance and viral clearance as essentially a model of antigen-driven, innate immunity-driven regulation.
-In a differential equation model, regulation of viral load was primarily driven by the availability of susceptible target cells [Baccam 2006].
-The question of target-driven versus antigen-driven regulation was addressed in a separate modeling study of infected equine lungs [Saenz 2010], finding that regulation was primarily innate immunity-driven since no more than 27\% of bronchial epithelial cells were ever infected during the course of the lung infection.
-In a comprehensive model of murine influenza rate-limiting antigen presentation was the primary influence on viral clearance [Lee 2009].
-In a subsequent model based on a large murine database, CD8+ T cells were critical components of the adaptive response in limiting the number of infected cells, with neutralizing antibodies responsible for clearing free virus [Miao 2010].
-Most modeling studies used a mouse-adapted virus to enable enumeration of epitope-specific T cells, making direct comparisons of replication efficiency with our human virus isolates more difficult.
-The mechanism of regulation of animal viral load may be specific to viral replication efficiencies, strain-specific suppression of innate immunity, and the phase of the immune response.
-We compared the three strains for their demands placed on the T cell search mechanism to find and kill infected cells.
-The multiplicity of potential regulation mechanisms does not invalidate our approach to the T cell search mechanisms.
-
-Our focus on the T cell search necessitated the simplification or removal of selected features of the mammalian immune response, in order to permit tractable simulations.
-Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
-Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
-The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable rates as indicated by superior data-fitting models [Wu 2011].
-Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection [Miao 2010].
-Since known critical components were omitted or simplified, our model was not designed to demonstrate clearance of virus from the lung.
-Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
-
-\subsection*{Chemokine Directed T-cell Search}
-
-Hypercytokinemia is a characteristic of virulent influenza strains \cite{DeJong2006} but bloodstream levels of cytokines and chemokines may not reflect local concentrations and dynamics critical in guiding T cells to infected epithelial cells.
-We elected to use our data on in vitro secretion of chemokines by human bronchial epithelial (NHBE) cells obtained from the same cultures of the three virus strains previously reported \cite{Mitchell2011}.
-In this study the replication kinetics and productivity of the three virus strains were markedly different, but monolayer infection was initiated with a low MOI of 0.01 or 0.001, mimicking the initial infection naturally.
-The levels of the chemokines CCL5 and CXCL10 in the basal media induced by the 3 strains were approximately equivalent at 24h post-infection.
-When the productivity per cell was calculated, the avian H5N1 strain induced higher secretion of CCL5 (RANTES) than the seasonal or pandemic strains.
-Greater CCL5 secretion by H5N1 strains has been noted in in vitro NHBE cultures when a relatively high MOI is used compared to seasonal strains \cite{Chan2005, Chan2010, Zeng2011} and pandemic strains \cite{Zeng2011}.
-MOI and degree of differentiation of NHBE are critical determinants of cyto/chemokine secretion \cite{Chan2010}.
-However, in spite of the attenuated type 1 interferon response induced by H5N1 strains \cite{Zeng2007}, the chemokine secretion is not attenuated.
-
-Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999} and CCL3 \cite{Kawai1999}.
-The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial in the accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
-Other chemokines are likely to play a role in T cell traffic.
-Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
-In the intact animal immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
-Finally, antigen-specific CD8+ T cell recognition of its target will induced lung injury mediated by TNFa and trigger addition alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
-Thus our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
-
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
-The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 100 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 250 microns in our model.
-This calculation matches the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-
-NEW
-
-In our model directed migration and viral infection control was not limited over a wide range of T cell sensitivity to the chemotactic signal.
-The density of CCR5 receptors on the surface of CD8+ T cells was a determinant of the strength of chemotactic attraction for CCR5 ligands such as CCL5 in an in vitro assay [Desmetz 2006].
-The density of CCR5 receptors on human CD8+ T cells vary over an order of magnitude, suggesting that sensitivity to the chemotactic signal could play a role in susceptibility to infection.
-CCR5-knockout mice are susceptible to fatal Cryptococcal encephalitis/meningitis [Huffnagle 1999] and West Nile virus encephalitis [Glass 2005].
-In the Cryptococcus-infected mice T cell recruitment to the brain, but not to the lung, was impaired.
-In CCR5-KO mice influenza A virus pneumonia was more severe associated with decreased T cell recruitment to the lung, while in CCR2-KO mice decreased T cell recruitment was due to decreased macrophage influx into the lung [Dawson 2000].
-On the other hand, CCL5-KO mice and CXCR3-KO mice (the receptor for CXCL10/IP-10) had similar viral kinetics and leukocyte recruitment compared to the wild-type mice [Wareing 2004].
-Finally, humans with the CCR5-D32 deletion allele may have increased susceptibility to the pandemic 2009 H1N1 influenza in a small sample [Keynan 2010].
-The question of limiting sensitivity has practical implications, since immunomodulation of cytokine/chemokine signaling to decrease influenza mortality is currently under investigation [Oslund 2011; Alleva 2011; Bassaganya-Riera 2010; Budd 2007; Moseley 2010].
-While blocking chemokine function may decrease inflammation and death from viral pneumonia, it may have an unintended consequence of impairing adaptive immunity.
-Thus, specific T cell recruitment to the influenza virus-infected lung should be quantitatively assessed in chemokine receptor-knockout mice.
-
-\section*{References}
 % The bibtex filename
@@
 
-\section*{Figure Legends}
 %\begin{figure}[!ht]
@@
 
-\section*{Tables}
+%\section*{Tables}
 %\begin{table}[!ht]

### 2012-09-14 paper/chemokine_main.tex
Near final draft, waiting for author summary and acknowledgements edits

@@
 During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.
-T-cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.
+T cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.
 We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).
@@
 Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the eight chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.
-A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment.
-T cells appear to be sufficiently sensitive to chemokine to function as well as they possibly can.
+Of the chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.
+A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment in the lungs.
+We observe that T cell sensitivity to chemokine is just high enough to maximize their efficiency.
 The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.
-The spatial nature of the model reveals unique challenges to T-cell recruitment and motility not visible in spatially homogeneous ODE models.
-Expanding plaque sizes isolate infected cells, impeding T-cell discovery and subsequent clearance of the infection over time.
+
+The spatial nature of the model reveals unique challenges to T cell recruitment and motility not visible in spatially homogeneous ODE models.
+Expanding plaque sizes isolate infected cells, impeding T cell discovery and infection control.
+In the model, T cells executed an efficient search with high sensitivity to either chemokine detected.
+A key limitation imposed on T cells was illustrated in their failure to clear the most rapidly replicating pandemic H1N1 virus after day 6, when T cells became ‘lost’ and inefficient in large infected foci as revealed in cinematic displays.
+This spatially explicit model is consistent with an efficient T cell recruitment to infected lung, without apparent means to induce increased efficiency, leaving vaccination as the key to shortening the interval between initial infection and T cell activation.
 
@@
 \section*{Author Summary}
+
+Seasonal influenza infects up to 10,000,000 people and kills between 250,000 and 500,00 people every year.
+Recent mutant strains of flu with high morbidity such as Avian H5N1 and 2009 Pandemic H1N1 raise the possibility of a global pandemic in the near future.
+In order to clear an infection in the lung, T cells produced in the lymph nodes must navigate the branching bronchial vascular system in order to reach the site of infection.
+Using an agent-based model, we look at how molecular signaling by cells infected with three differing strains of influenza aid this search process.
+Our experimental data shows that two chemokines, IP-10 and RANTES, are help improve T cell search.
+The spatial nature of our model reveals challenges to this search process not seen in conventional ordinary differential equation (ODE) models.
+The simulated immune response is not able to contain the highly virulent 2009 pandemic H1N1 strain, suggesting that early intervention, i.e.
+vaccination, is key to containing the infection.
 
@@
 Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the spread and interaction of populations.
+Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the viral spread and interaction of different populations.
 
@@
 A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.
-While each chemokine has been studied in detail for receptor specificity and function in knockout models, the details of chemokine function and efficiency have not been examined in a virus-infected whole animal model.
+While each chemokine has been studied in detail for receptor specificity and function in knockout models, the local advantage brought by chemokine secretion in an inflammatory context have not been examined in detail in a virus-infected whole animal model.
 
@@
 First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.
-Second, each chemokine exhibited a clear threshold to direct migration, with lesser impact at increasing concentrations above the threshold.
+Second, each chemokine exhibited a clear threshold to direct migration, with diminishing improvements at increasing concentrations above the threshold.
 Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.
@@
 
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics[width=0.5\textwidth]{SystemChart}
-\end{center}
-\caption{Model of T cell search.
-Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.
-Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.
-In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
-\label{fig:systemchart}
-\end{figure}
-
 Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of T-cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
+A simplified model of T cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
 The lung was represented as a two-dimensional sheet of healthy epithelial cells.
 Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
-Activated T cells descend through the vascular tree until cytokine signal is detected, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
+Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
 T cells that do not encounter cytokine recirculate to the lymph node.
@@
 
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics[width=\textwidth]{ModelChart}
-\end{center}
-\caption{Visual representation of the model.
-Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.
-Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.
-Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.
-Solid arrows represent a cell state transition from one behavior to another.
-Dashed arrows display the mechanism used to induce a transition.
-Dotted arrows indicate the production of new virus.}
-\label{fig:modelchart}
-\end{figure}
-
 In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
@@
 Once infected, the cell transitions from {incubating} to \emph{expressing}.
-\emph{Expressing} cells secrete virus and chemokine for 17 hours and then die.
-\emph{Expressing} cells become \emph{apoptotic} if they encounter activated T-cells.
+\emph{Expressing} cells secrete virus and chemokine for approximately 17 hours and then die.
+\emph{Expressing} cells become \emph{apoptotic} if they encounter activated T cells.
 Apoptoic cells continue to secrete virus until they die one hour after their transition.
@@
 
-T-cells have three states.
-\emph{Searching} T-Cells begin emerging from the lymph node at five days post infection.
-\textit{Searching} T-Cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine.
-\emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{searching} and are reintroduced to a new random location in the lung.
-When a \emph{searching} T-Cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection.
-\emph{Chemotaxing} T-Cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells.
+T cells have three states.
+\emph{Circulating} T cells begin to emerge from the lymph node at five days post infection.
+\textit{Emigrating} T cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine.
+\emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{emigrating} and are reintroduced to a new random location in the lung.
+When a \emph{emigrating} T cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection.
+\emph{Chemotaxing} T cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells.
 \emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.
@@
 Chemokine diffuses across the tissue but has no
-direct effect beyond activating T-Cells. Both particle types decay exponentially.
+direct effect beyond activating T cells. Both particle types decay exponentially.
 
 Parameters that are consistent between every model are shown in Table \ref{table:parameters}.  Strain-specific values are shown in Table \ref{table:strains}.
-
-\begin{table}
-\begin{center}
-\begin{tabular}{ | c | c | c | }
-  \hline                        
-  Paramter & Value & Source \\
-  \hline
-  Viral Diffusion in Airway & $.0318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Viral Decay in Airway &  $1/day$ & \cite{Lee2009} \\
-  Chemokine Diffusion Rate & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & Selected$^1$\\
-  Infection Sensitivity Rate &  $2$ \textit{hour/virion} &  Selected$^2$ \\
-  Incubation Time &  $10$ \textit{hours} & \cite{Mitchell2011} \\
-  Expression Time &  $16.7$ \textit{hours} & \cite{Mitchell2011} \\
-  Apoptosis Time & $1$ \textit{hour} & \cite{Ganusov2008} \\
-  T-Cell Production Rate & $777$\textit{/hour} & \cite{Miao2010} \\ 
-  Blood Circulation Time & $6$ \textit{min} & \cite{Banerjee2010b} \\
-  Search Time In Chemokine-Free Tissue & $10$ \textit{min} & \cite{Banerjee2010b} \\
-  T-Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
-  T-Cell Speed (Chemotaxis) & $3 \mu m/s$ & \cite{Miller2003} \\
-  T-Cell Sensitivity to Chemokine & $10 ng/mL$ & \cite{Gao2003} \\
-  T-Cell Expected Kill Time & $10$ \textit{min} & Selected$^3$ \\
-  Epithelial Cell Radius & $25 \mu m$ & Selected$^4$ \\
-  T-Cell Age (at FOI) & $2$ \textit{hours} & Selected$^5$ \\
-  T-Cell Age (in Blood) & $3$ \textit{days} & Selected$^5$ \\
-  Onset of ATC Lymph Node Exit & $5$ \textit{days} & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3$\textit{/day} & Selected$^6$ \\
-  \hline  
-\end{tabular}
-\caption{Static model parameters.
-1) Chosen to correspond to a 30 minute half-life.
-2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.
-This scales linearly with the number of virions in the cell's vicinity.
-3) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.
-This rate does not scale with T cell numbers.
-4)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
-5) Parameters chosen after discussions with David Woodland, Trudeau Institute.
-6) IgM presence is abstracted by increasing viral decay by a factor of three.
-}
-\label{table:parameters}
-\end{center}
-\end{table}
-
-\begin{table}
-\centering
-\begin{tabular}{ | r | c | c | c | }
-  \hline                        
-  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
-   & \footnotesize{(pg/s$\cdot$cell)}  & \footnotesize{(pg/s$\cdot$cell)} &  \footnotesize{(PFU/s$\cdot$cell)} \\
-  \hline
-  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
-   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
-   \hline
-  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
-   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
-   \hline
-  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
-   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
-  \hline
-\end{tabular}
-\caption{Strain-specific model parameters.
-Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
-Viral production values are taken from \cite{Mitchell2011}.}
-\label{table:strains}
-\end{table}
 
@@
 Quantitative viral culture was performed by standard plaque assay.
-Quantitative chemokine levels were performed in 30 µl aliquots for a panel of 17 chemokines and cytokines (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
+Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
+Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.
+TNF was ignored as its effects are outside the scope of this paper.
 
@@
 
-Chemokine data were collected from epithelial monolayers infected with influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
+Chemokine data were collected from epithelial monolayers infected with three different influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
 IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
 To estimate per-cell production rates, we extended the ODE model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells (Supplemental Methods).
+\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.
+\textbf{S}1).
 Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+The resulting chemokine production values were used in the CyCells ABM.
 Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
@@
 
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{data}
- \end{center}
-\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.
-Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
-Dashed lines show model fits to IP-10 and RANTES data.}
- \label{fig:data}
-\end{figure}
-
-\subsection*{T-cell sensitivity to chemokine}
+\subsection*{T cell sensitivity to chemokine}
 
@@
 Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-However, T-cell sensitivity determines how strong the aggregate chemokine signal must be.
+However, T cell sensitivity determines how strong the aggregate chemokine signal must be.
 
@@
 Seasonal and pandemic H1N1 models simulated both IP-10 and RANTES; avian H5N1 included only RANTES (\textbf{S2.2}).
-Because multiple T-cells in the same area do not enhance control of infection, we hypothesize more T cells are not better above a critical threshold.
+Because multiple T cells in the same area do not enhance control of infection, we hypothesize more T cells are not more efficient above a critical threshold.
 We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
 Single runs were used due to computational limitations.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{sensitivity}
- \end{center}
-\caption{Varying T-cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
-Total number of infected, expressing and apoptotic cells are plotted for each infection.
-The sensitivity value specifies the minimum level of chemokine concentration required for T-cells to detect it.
-}
- \label{fig:sensitivity}
-\end{figure}
 
@@
 Free virus particles diffuse from virus secreting cells and infect healthy cells.
-Chemokine produced by infected cells attracts T-cells to the infected cells.
+Chemokine produced by infected cells attracts T cells to the infected cells.
 Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.
 These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
-Until day 4 the plaque is dominated by active (incubating and secreting) cells, and dead cells are rare.
+Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare.
 Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque.
-T-cells arrive at day 5 and begin killing the virus-secreting cells.
+T cells arrive at day 5 and begin killing the virus-secreting cells.
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-In aH5N1, the plaque is dense, allowing T-cells to find secreting cells easily, and infection is eliminated.
+In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
 However, in both H1N1 simulations secreting cells were not eliminated.
@@
 It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.
-Thus, T-cells can fail to detect cellular changes in the plaque.
-Taken together, the delayed response and low proportion of virus secreting cells prevent T-cells from completely clearing the infection.
+Thus, T cells can fail to detect cellular changes in the plaque.
+Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.
 
-T-cells are unable to kill cells that have not yet presented antigen.
+T cells are unable to kill cells that have not yet presented antigen.
 At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
 By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.
-The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T-cells can control (but not eliminate) the sH1N1 infection.
+The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.
 
@@
 That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.
-Assuming that secretion continues for one hour after apoptosis is initiated, the best a T-Cell could do is limit production to 18 new viral particles.
-Thus, T-Cells alone cannot contain the pH1N1 infection.
-In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T-cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
+Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.
+Thus, T cells alone cannot contain the pH1N1 infection.
+In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
 Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.
 
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{cycells}
- \end{center}
-\caption{Simulated sH1N1 infection.
-Screenshots from day 4, day 5.5, and day 7.
-The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T-cells arriving at day 5 (green).
-Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
- \label{fig:cycells}
-\end{figure}
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{plaquesize}
- \end{center}
-\caption{Comparison of total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red) for the simulated infections of aH5N1, sH1N1, and pH1N1.
-T cells are able to clear both secreting and incubating cells in aH5N1, are unable to contain incubating cells in sH1N1, and are unable to contain either type of infected cell in pH1N1.}
- \label{fig:plaquesize}
-\end{figure}
-
 \section*{Discussion}
-
-%\subsection*{Caveats}
-
-%Parameters are estimated to the best of our knowledge when values could not be found in literature.
-We choose to ignore most aspects of the innate immune response to avoid over-parameterizing the model.
-The unbounded growth of pH1N1 can partially be attributed to the lack of many innate mechanisms.
-T cells are assumed to have the same level of sensitivity to both IP-10 and RANTES.
-Furthermore, T cells climb chemokine gradient in straight lines as our experiments with biased random walks showed no effect on model behavior.
 
@@
 Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.
-Antigen presentation and peripheral lymphoid tissue amplification of the T cell clonal response is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
+Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
 Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
@@
 All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.
-Immigration and contributions of virus-nonspecific immune cells such as macrophages are not represented in our model.
+Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.
 Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.
@@
 ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
-This is limiting as can been seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
+This is limiting as can be seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
 
 Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).
-Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlie.
-First, because T-cells find infected cells by climbing a gradient, we see T-cell clustering at local maxima of chemokine concentration, a posssible explanation for why T-cells do not increase in effectiveness as their numbers increase.
-Second, T-cell clustering persists after all virus-secreting cells have been eliminated.
-The local chemokine maxima takes time to diffuse and decay so that T-cells can climb the gradient to a new maximum.
+Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlier.
+First, because T cells find infected cells by climbing a gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
+Second, T cell clustering persists after all virus-secreting cells have been eliminated.
+The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.
 Finally, we can see that infected cells are more disperse as infection size grows.
-Because T-cells are clustered they cannot cover the increasing plaque effectively.
+Because T cells are clustered they cannot cover the increasing plaque effectively.
 These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
@@
 
-\subsection*{Chemokine Directed T-cell Search}
+\subsection*{Chemokine Directed T Cell Search}
 
@@
 MOI and degree of differentiation of NHBE cells are critical determinants of cytokine/chemokine secretion \cite{Chan2010}.
-Yet, as we noted at low MOI, greater CCL5 secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.
-Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controling the infection.
+Yet, as we noted at low MOI, greater RANTES secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.
+Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controlling the infection.
 
 Our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
-Chemokines with a clear role directing T cell traffic include CXCL10 \cite{Dufour2002}, CCL5 \cite{Kawai1999}, CCL3 \cite{Kawai1999}, and likely others.
-The chemokine receptor CCR5, binding CCL5 and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
+Chemokines with a clear role directing T cell traffic include IP-10 \cite{Dufour2002}, RANTES \cite{Kawai1999}, CCL3 \cite{Kawai1999}, and likely others.
+The chemokine receptor CCR5, binding CCL5 (RANTES) and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
 Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
@@
 The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 10 pg/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 150 microns in our model.
+For the threshold of 10 ng/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 150 microns in our model.
 This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+
+Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of infants prior vaccination?
+The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation.
+Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity.
+T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance.
+We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency.
+Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.
 
@@
 \section*{Acknowledgments}
+
+We thank Judy Cannon and Francois Asperti-Boursin (University of New Mexico, Department of Pathology) and Sheldon Jordan (University of New Mexico, Center for Biomedical Engineering) for discussions and suggestions concerning cellular and cytokine/chemokine behavior.
+We thank George Bezerra, Ben Edwards, and George Stelle (University of New Mexico, Department of Computer Science) for discussions relating to the design and parameterization of the agent-based model.
+
+This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.) and by NSF grant XXX (to S.F and M.M.).
 
@@
 
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics[width=0.5\textwidth]{SystemChart}
+\end{center}
+\caption{Model of T cell search.
+Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.
+Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.
+In the absence of signal, the T cell recirculates either through the lymph network or through the pulmonary vein back to the top of the network.}
+\label{fig:systemchart}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics[width=\textwidth]{ModelChart}
+\end{center}
+\caption{Visual representation of the model.
+Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.
+Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.
+Chemotaxing T cells climb the chemokine gradient and induce apoptosis in infected cells.
+Solid arrows represent a cell state transition from one behavior to another.
+Dashed arrows display the mechanism used to induce a transition.
+Dotted arrows indicate the production of new virus.}
+\label{fig:modelchart}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{data}
+ \end{center}
+\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.
+Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
+sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
+Dashed lines show model fits to IP-10 and RANTES data.
+Human bronchial epithelial cells were infected at an MOI of 0.01 with one of the three strains of influenza.
+Apical fluid for viral secretion and basal media for chemokine secretion was collected at the given time intervals post infection.
+Viral culture was performed by a standard plaque assay and chemokine levels were measured using 30 µl aliots for a panel of 17 chemokines and cytokines (not shown).}
+ \label{fig:data}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{sensitivity}
+ \end{center}
+\caption{Varying T cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+Total number of incubating, secreting and apoptotic cells are plotted for each infection.
+The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it.
+}
+ \label{fig:sensitivity}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{cycells}
+ \end{center}
+\caption{Simulated sH1N1 infection.
+Screenshots from day 4, day 5.5, and day 7.
+The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T cells arriving at day 5 (green).
+Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
+ \label{fig:cycells}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{plaquesize}
+ \end{center}
+\caption{Comparison of total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red) for the simulated infections of aH5N1, sH1N1, and pH1N1.
+T cells are able to clear both secreting and incubating cells in aH5N1, are unable to contain incubating cells in sH1N1, and are unable to contain either type of infected cell in pH1N1.
+We notice a threshold at day 5: pH1N1 has more expressing cells (green) by a factor of only two to three, yet unlike aH5N1 and sH1N1, the T cell response is unable to clear the expressing population.}
+ \label{fig:plaquesize}
+\end{figure}
+
 \section*{Tables}
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{ | c | c | c | }
+  \hline                        
+  Paramter & Value & Source \\
+  \hline
+  Viral Diffusion in Airway & $.0318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Viral Decay in Airway &  $1/day$ & \cite{Lee2009} \\
+  Chemokine Diffusion Rate & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
+  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & Selected$^1$\\
+  Infection Sensitivity Rate &  $2 hour/virion$ &  Selected$^2$ \\
+  Incubation Time &  $10 hours$ & \cite{Mitchell2011} \\
+  Expression Time &  $16.7 hours$ & \cite{Mitchell2011}$^3$ \\
+  Apoptosis Time & $1 hour$ & \cite{Ganusov2008} \\
+  T cell Production Rate & $777/hour$ & \cite{Miao2010} \\ 
+  Blood Circulation Time & $6 min$ & \cite{Banerjee2010b} \\
+  Search Time In Chemokine-Free Tissue & $10 min$ & \cite{Banerjee2010b} \\
+  T Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
+  T Cell Speed (Chemotaxis) & $3 \mu m/s$ & \cite{Miller2003} \\
+  T Cell Sensitivity to Chemokine & $10 ng/mL$ & \cite{Gao2003} \\
+  T Cell Expected Kill Time & $10 min$ & Selected$^4$ \\
+  Epithelial Cell Radius & $25 \mu m$ & Selected$^5$ \\
+  T Cell Age (at FOI) & $2 hours$ & Selected$^6$ \\
+  T Cell Age (in Blood) & $3 days$ & Selected$^6$ \\
+  Onset of ATC Lymph Node Exit & $5 days$ & \cite{Banerjee2011} \\
+  IgM Strength & Viral decay of $3/day$ & Selected$^7$ \\
+  \hline  
+\end{tabular}
+\caption{Static model parameters.
+1) Chosen to correspond to a 30 minute half-life.
+2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.
+This scales linearly with the number of virions in the cell's vicinity.
+3) Chosen to be a plausible median time (1,000 minutes) between 6 hours and 24 hours.
+4) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.
+This rate does not scale with T cell numbers.
+5)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
+6) Parameters chosen after discussions with David Woodland, Trudeau Institute.
+7) IgM presence is abstracted by increasing viral decay by a factor of three.
+}
+\label{table:parameters}
+\end{center}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{ | r | c | c | c | }
+  \hline                        
+  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
+   & \footnotesize{$(pg/s\cdot cell)$}  & \footnotesize{$(pg/s\cdot cell)$} &  \footnotesize{$(PFU/s\cdot cell)$} \\
+  \hline
+  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
+   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
+   \hline
+  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
+   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
+   \hline
+  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
+   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
+  \hline
+\end{tabular}
+\caption{Strain-specific model parameters.
+Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
+Viral production values are taken from \cite{Mitchell2011}.}
+\label{table:strains}
+\end{table}
+
 \end{document}

### 2012-09-14 paper/chemokine_supplement.tex
Near final draft, waiting for author summary and acknowledgements edits

@@
 
-where $N_{c}$ is the number of circulating activated antigen-specific T-cells, $N'_{f}$ is the number of circulating T-cells that have found and exit into a region of lung tissue expressing chemokines and $N_{f}$ is the number of circulating T-cells that have found an infected region.
+where $N_{c}$ is the number of circulating activated antigen-specific T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines and $N_{f}$ is the number of circulating T cells that have found an infected region.
 $T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
@@
 
-Circulating activated antigen-specific T-cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$.
+Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$.
 These circulating cells then transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
@@
 Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
-Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T-cells ($N_{f}$) at a rate $k_{e}$.
+Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
 Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
 Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
-The initial viral titer and the initial number of target cells are denoted $V_{0}$ and $T_{0}$, respectively.
+The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
 The initial number of infected cells is assumed to be zero.
-This ode was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
+This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
 
@@
 Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the addition of IgM.
-In addition, for each infection the number of infected cells declines quickly at day five due to the T-cell response.
+In addition, for each infection the number of infected cells declines quickly at day five due to the T cell response.
 
 The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~\ref{fig:variance}).
-The rapid growth of the pandemic H1N1 prevents the immune response from containing the infection.
-Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T-Cells to gain ground is too small.
+The rapid production of the pandemic H1N1 virus prevents the immune response from containing the infection.
+Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T Cells to gain ground is too small.
 aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
-
-\begin{figure}[ht!]
-\begin{center}
- \includegraphics[width=\textwidth]{variance}
- \end{center}
-\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
-IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-Each run was initialized identically for each strain save for the random seed.
-The middle line shows the average while the outer lines show the 95\% confidence interval.}
- \label{fig:variance}
-\end{figure}
 
@@
 IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
-
-\begin{figure}[ht!]
-\begin{center}
-	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
-B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
-\end{figure}
+We assume receptor sensitivity is the same for both chemokines and that the chemokines in combination work additively.
 
@@
 \bibliography{references}
+
+\section*{Figure Legends}
 
@@
 
-%\section*{Tables}
+\begin{figure}[ht!]
+\begin{center}
+ \includegraphics[width=\textwidth]{variance}
+ \end{center}
+\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
+IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
+Each run was initialized identically for each strain save for the random seed.
+The middle line shows the average while the outer lines show the 95\% confidence interval.}
+ \label{fig:variance}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{chemokine}
+	\caption{Effects of different chemokine combinations.
+A) aH5N1 does not stimulate an IP-10 response.
+B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
+	\label{fig:chemokine}
+\end{center}
+\end{figure}
+
+\section*{Tables}
+
 %\begin{table}[!ht]
@@
 
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{ | c | c | c | }
+  \hline                        
+  Paramter & Value & Source \\
+  \hline
+  T(0) & $1e6$ & \cite{Mitchell2011} \\
+  V(0) &  $1e4$ & \cite{Mitchell2011} \\
+  $r_{cell}$ &  $10 \mu m$ & \cite{Miao2010} \\
+  $t_{rc}$ & $6s$ & \cite{Peters1983} \\
+  $v_{tcell}$ & $0.3 \mu m/s$ & \cite{Miller2003} \\
+  $k_e$ & $6.4e-5/cell/day$ & \cite{Miao2010} \\
+  $\beta$ & $4.8e-7/cell/PFU$ & \cite{Mitchell2011} \\
+  $p$ & $0.18 PFU/h$ & \cite{Mitchell2011} \\
+  $\delta$ & $16.7 hours$ & \cite{Mitchell2011} \\
+  $a$ & $20.9$ & ABM fits \\
+  $b$ & $70.8$ & ABM fits \\
+  \hline  
+\end{tabular}
+\caption{Supplemental model parameters.
+Most values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
+Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
+\label{table:supplement}
+\end{center}
+\end{table}
+
 \end{document}

### 2012-09-19 paper/chemokine_main.tex
Paper Version submitted to PLoS CB

@@
 Melanie Moses$^{1}$, 
-Frederick Koster$^{1,2}$, 
+Frederick Koster$^{1,2}$
 \\
@@
 
-During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells during the lymphocyte diaspora from regional lymph nodes to small infected foci comprising a tiny fraction of the total lung.
-T cell navigation to the infected region is undirected but focused migration is made possible by local cytokine and chemokine production from infected epithelial cells.
-We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using simulations in a purpose-built agent-based mathematical model (CyCells).
+During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.
+T cell navigation to this region is undirected but migration is made possible by local cytokine and chemokine production from infected epithelial cells.
+We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using an agent-based model (CyCells).
 
-Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used separately to induce chemokine production \textit{in vitro}.
-Of the chemokines tested, only secretion of CXCL10 (IP-10) and CCL5 (RANTES) was stimulated by infection with these three influenza strains, with IP-10 dominating the directed migration due to higher concentrations.
-A delayed differential equation model was fit to the empirical per cell chemokine production rates, coupled with published T cell migration parameters, to calibrate the spatially explicit model in order to test inter-strain variation on T cell recruitment in the lungs.
-We observe that T cell sensitivity to chemokine is just high enough to maximize their efficiency.
-The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production and large plaque size.
+Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used to induce chemokine production \textit{in vitro}.
+Of the chemokines tested, secretion of CXCL10 (IP-10) and CCL5 (RANTES) were stimulated by infection, with IP-10 dominating the migration due to higher concentrations.
+A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.
+We observe that T cell sensitivity to chemokine is high enough to maximize their efficiency.
+The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.
 
-The spatial nature of the model reveals unique challenges to T cell recruitment and motility not visible in spatially homogeneous ODE models.
-Expanding plaque sizes isolate infected cells, impeding T cell discovery and infection control.
-In the model, T cells executed an efficient search with high sensitivity to either chemokine detected.
-A key limitation imposed on T cells was illustrated in their failure to clear the most rapidly replicating pandemic H1N1 virus after day 6, when T cells became ‘lost’ and inefficient in large infected foci as revealed in cinematic displays.
-This spatially explicit model is consistent with an efficient T cell recruitment to infected lung, without apparent means to induce increased efficiency, leaving vaccination as the key to shortening the interval between initial infection and T cell activation.
+The spatial nature of the model reveals unique challenges to T cell recruitment not visible in spatially homogeneous ODE models.
+Infected cells can become isolated in expanding plaques, impeding infection control.
+In the model, T cells executed an efficient search with sensitivity to either chemokine.
+A key limitation imposed on T cells was illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci as revealed in cinematics.
+This spatial model is consistent with efficient T cell recruitment to infected lung.
+Without an apparent means to induce increased efficiency, vaccination remains the key to shortening the interval between initial infection and T cell activation.
 
@@
 
-Seasonal influenza infects up to 10,000,000 people and kills between 250,000 and 500,00 people every year.
-Recent mutant strains of flu with high morbidity such as Avian H5N1 and 2009 Pandemic H1N1 raise the possibility of a global pandemic in the near future.
-In order to clear an infection in the lung, T cells produced in the lymph nodes must navigate the branching bronchial vascular system in order to reach the site of infection.
-Using an agent-based model, we look at how molecular signaling by cells infected with three differing strains of influenza aid this search process.
-Our experimental data shows that two chemokines, IP-10 and RANTES, are help improve T cell search.
-The spatial nature of our model reveals challenges to this search process not seen in conventional ordinary differential equation (ODE) models.
-The simulated immune response is not able to contain the highly virulent 2009 pandemic H1N1 strain, suggesting that early intervention, i.e.
-vaccination, is key to containing the infection.
+Clearance of influenza from the lung depends strongly on the efficiency with which T cells travel from the lymph node to the site of infection.
+However, T cells in the lung must navigate the large branching bronchial network.
+This maze-like structure is a challenge for T cell search, and the vast majority of T cells exit the lung without encountering infection.
+Infected cells produce cytokines and chemokines, which signal T cells to exit the vasculature and enter tissue.
+We developed an agent-based model to quantify how these molecular signals affect search.
+Our experimental data show that two chemokines, IP-10 and RANTES, are important for T cell search.
+We incorporated the data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.
+We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.
+The simulated immune response contains the first two infections but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.
+Direct interventions to improve this T cell search process are unlikely to be effective, and therefore vaccination is required to contain more virulent strains.
 
@@
 To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
-We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients within the model lung.
-Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model.
-[REF]
+We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients in our model.
+Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model \cite{Miao2010, Ganusov2011}.
 
@@
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
+Refer to the videos in Figures S3-S5 to see a representative simulation of seasonal influenza.
+
 In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
-However, in both H1N1 simulations secreting cells were not eliminated.
+However, secreting cells were not eliminated in either H1N1 simulations.
 Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
@@
 The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.
+The video in Figure S6 illustrates this event.
 
@@
 
-Hypercytokinemia, or 'cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.
+Hypercytokinemia, or `cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.
 The use of in vitro human bronchial epithelial cells \cite{Mitchell2011} infected at low cell:virus infection ratio (MOI) was intended to mimic initial natural infection and subsequent local infection spread.
@@
 
-Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of infants prior vaccination?
+Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?
 The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation.
@@
 
-We thank Judy Cannon and Francois Asperti-Boursin (University of New Mexico, Department of Pathology) and Sheldon Jordan (University of New Mexico, Center for Biomedical Engineering) for discussions and suggestions concerning cellular and cytokine/chemokine behavior.
-We thank George Bezerra, Ben Edwards, and George Stelle (University of New Mexico, Department of Computer Science) for discussions relating to the design and parameterization of the agent-based model.
+We thank J.
+Cannon (UNM Pathology and Molecular Genetics \& Microbiology), F.
+Asperti-Boursin (UNM Pathology and Computer Science), S.
+Jordan (UNM Computer Science and Biology), and G.
+Bezerra, B.
+Edwards, and G.
+Stelle (UNM Computer Science) for suggestions concerning the manuscript, cellular and chemokine behavior, and agent-based modeling.
 
-This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.) and by NSF grant XXX (to S.F and M.M.).
+This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F and M.M), and NSF grant EF-1038682 (to S.F and M.M.).
 

### 2012-09-19 paper/chemokine_supplement.tex
Paper Version submitted to PLoS CB

@@
 
+\section*{Video}
+
+\begin{figure}
+\caption{First of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
+T cells show half way through the simulation.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
+ \label{video:cell_view}
+\end{figure}
+
+\begin{figure}
+\caption{Second of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the virus concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:virus_view}
+\end{figure}
+
+\begin{figure}
+\caption{Third of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the chemokine concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:chemokine_view}
+\end{figure}
+
+\begin{figure}
+\caption{A closer look at the 2009 pandemic simulation.
+This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
+Note the high proportion of virus-secreting cells (orange) early on.
+As time passes, secreting cells are gradually contained to the point where they become very sparse.
+T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
+\end{figure}
+
 \end{document}

### 2012-09-19 paper/cover_letter.tex
Paper Version submitted to PLoS CB

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

\usepackage{multirow}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

\begin{flushleft}Dear Editors, \end{flushleft}

We submit this manuscript entitled "Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration" for consideration of publication in PLoS Computational Biology.  We feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.

Many insights have derived from modeling the immune response using differential equations (ODEs) and delay differential equations (DDEs) to simulate linear and non-linear components.  We have taken a different approach, using agent-based spatially-explicit modeling to examine the recruitment of antigen-specific T cells to influenza-infected foci in the lung during a primary immune response.   ODE and DDE models do not take spatial effects into account, but given that the search occurs within a very large uninfected tissue space, we felt the efficiency of this search, depending largely on chemokine signaling of infected regions, needed to be carefully examined.  We found that, similar to the highly efficient recruitment of precursor T cells to the lymph nodes, activated T cell search is indeed efficient, with one caveat applicable to large infected foci.  We wish to stimulate other investigators to use agent-based modeling to carefully examine other aspects of the immune response in a spatially explicit manner, thus improving the comprehensive models of the immune response. \\

This work has not been published elsewhere, and is complementary to a recent publication from our group (H Mitchell, D Levin, et al. J Virol 2011;85(2):1125-1135). \\

All authors have read and approved the manuscript.

\begin{flushleft}Respectfully yours,\end{flushleft}
Drew Levin \\
Stephanie Forrest \\
Soumya Banerjee \\
Candice Clay \\
Melanie Moses \\
Frederick Koster

\end{document}

### 2012-12-16 paper/chemokine_main.tex
Updated with Fred's new Intro and Discussion, plus various minor touchups

@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals constraints on chemokine directed migration}
 }
@@
 
-The spatial nature of the model reveals unique challenges to T cell recruitment not visible in spatially homogeneous ODE models.
+The spatial nature of the model reveals unique challenges to T cell recruitment not visible in standard differential equation models.
 Infected cells can become isolated in expanding plaques, impeding infection control.
@@
 
-The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that controls all but the most virulent strains.
-Understanding its components and their interactions are relevant to improving vaccines and strategies to control immunopathology.
-Elegant studies in the mouse model have described the induction and effector phases of the adaptive immune response, including functional capabilities of the innate interferon-directed response, macrophages, specific antibody, and cytotoxic effector lymphocytes \cite{Sallusto2000, Joo2008, Mackay2008}.
-CD8+ T cells are necessary for the resolution of pneumonia and complete clearance of mouse-adapted influenza strains \cite{Sallusto2000, Joo2008, Mackay2008, Miao2010}, but gaps remain in our understanding of the behavior of activated T cells in infected tissue.
+The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that control viral replication of all but the most virulent strains.
+Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.
+The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.
+Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.
+The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.
+On approximately day 5 after infection activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.
 
-The generally held view of the sequence of CD8+ T cell events during a primary immune response to influenza in the lung involves the following steps: dendritic cells bearing specific antigen migrate from the lung to the draining lymph node where naive T and B cell precursors are activated in specialized architecture \cite{Saenz2010, Beltman2007, Handel2008, Zheng2008, Ingulli2009, Allan1990}.
-T cells proliferate in the lymph node and are released into the bloodstream, appearing simultaneously in the lung, spleen and other organs 4-5 days post-infection, a whole-body distribution known as the lymphocyte diaspora \cite{Thelen2008}.
-In the lung, extravasation from capillaries is signaled by inflammatory cytokine-mediated integrin-activation on endothelial cells.
-Once in the tissue, T cells follow a chemokine gradient to the infected epithelial cells secreting the chemokine.
+The process of recruitment of activated CD8 effector cells to infected sites in the lung is not completely understood.
+Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into tissue \cite{Topham2001}.
+Effector cells must localize to sites of viral replication [4], but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
+Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005} where recruited T cells mediate either protection or immunopathology.
+Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
+Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.
 
-Mathematical models have used detailed mouse data to characterize the evolution of the immune response and predict its impact on viral kinetics \cite{Thomas-Vaslin2008, Beauchemin2008, Smith2010, Thakar2010, Burrowes2004}.
-Such whole-response modeling has considerable potential to inform strategies of vaccination and therapy, thus avoiding subsequent expensive animal and human trials.
-Models using ordinary differential equations to predict temporal events, however, do not account for spatial constraints, and must make assumptions about the efficiency of the viral spread and interaction of different populations.
+The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.
+For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediated T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediated the opposite effect \cite{DeLemos2005}.
+In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of CXCL10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.
+In the herpesvirus model CXCL10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.
+In the parainfluenza virus model CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.
+In the influenza A model initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.
+The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency but viral clearance was not altered \cite{Fadel2008}.
+The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.
 
-Here we study the time and physical constraints of the lymphocyte diaspora using an agent-based model (ABM).
-To develop a model with tractable simulations, some components of host defense are abstracted and the model therefore does not aim to predict the true outcome for each strain examined.
-We ask how very small foci of infected tissue can attract limited numbers of activated CD8 T cells after release into the systemic vascular system.
-The circulating blood supply services non-infected tissue orders of magnitude larger than the target infected tissue, necessitating an efficient mechanism to focus cell migration.
-A host of cytokines and chemokines secreted by infected cells are critical in directing immune cells to sites of infection \cite{Miao2010, Zhao2000, LiJeon2002}.
-While each chemokine has been studied in detail for receptor specificity and function in knockout models, the local advantage brought by chemokine secretion in an inflammatory context have not been examined in detail in a virus-infected whole animal model.
+Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, T cells have not be shown to climb chemokine gradients.
+Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.
+Finally the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.
 
-To characterize the local chemokine-secretion environment in which T cells migrate, we used data on viral secretion and chemokine secretion patterns from human epithelial cells infected in vitro by three different influenza virus strains of avian, seasonal and pandemic origins \cite{Mitchell2011}.
-We selected in vitro chemokine data as a tissue environment surrogate, because it may be more accurate than published chemokine levels in mouse blood and bronchial alveolar fluid, and it allowed us to represent sustained chemokine gradients in our model.
-Extensive literature on mouse model data of infections with mouse-adapted strains PR8 and HK-31 were used to calibrate the ABM model \cite{Miao2010, Ganusov2011}.
+A number of mathematical models have constructed the entire adaptive response and key components \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
+Focused math models can examine each event in detail and narrow the range of possible solutions.
+We sought to examine the CD8 T cell diaspora and recruitment to infected lung in detail.
+Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail.
+Activated T cells searching for and homing into infected tissue do so in a spatially complex environment.
+We therefore utilized the strategy of agent-based modeling (ABM) to represent the physical environment of the searching T cell.
+Thus T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
 
-The standard hypothesis posits that the marked strain-specific differences in replication efficiency and tissue spread previously reported \cite{Mitchell2011} would largely determine the outcome of infection as eradication or lack of control.
-Our model confirms this hypothesis, but we also observed three interesting characteristics of chemokine-directed migration not revealed by earlier data.
-First, of the only two chemokines stimulated by all three strains, the influence of IP-10 consistently dominated the effect of RANTES, regardless of strain.
-Second, each chemokine exhibited a clear threshold to direct migration, with diminishing improvements at increasing concentrations above the threshold.
-Finally, when the target (infection focus) was large, T cell search became inefficient due to the disappearance of the chemokine gradient, possibly contributing to the loss of control of the pandemic strain.
+We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.
+Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
+The model does not aim to predict control of viral replication but only compare simulated outcome among three influenza strains.
+To achieve this focused goal on the efficacy of the chemotactic signals alone, we use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.
+Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.
 
@@
 
+\subsection*{Delay Differential Equation Model}
+
+We estimate chemokine production rates by adapting the delay differential equation model described in \cite{Mitchell2011} by adding one new equation to model chemokine production.
+Initial population sizes and parameter values are taken from the previous study.
+{\footnotesize
+\begin{equation}
+\begin{aligned}
+\dot{T} &= - \beta T V \\
+\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
+\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
+\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
+\dot{F} &=  I_1[t-\tau_2] \\
+\dot{C} &= r I[t-\tau_3] - d C \\
+\end{aligned}
+\label{eq:dde}
+\end{equation}
+\vspace{.05in}
+}
+
+Table \ref{tab:dde} summarizes population and parameter values and descriptions.
+Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table \ref{tab:strains}).
+
 \subsection*{Computational Modeling}
@@
 Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of T cell activation and recirculation was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
+A simplified model of T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
 The lung was represented as a two-dimensional sheet of healthy epithelial cells.
@@
 Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration.
-Chemokine diffuses across the tissue but has no
-direct effect beyond activating T cells. Both particle types decay exponentially.
+Chemokine diffuses across the tissue but has no direct effect beyond activating T cells.
+Both particle types decay exponentially.
+IgM is modeled by increasing the viral decay rate by a factor of three after the third day.
 
-Parameters that are consistent between every model are shown in Table \ref{table:parameters}.  Strain-specific values are shown in Table \ref{table:strains}.
+Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.
 
@@
 
-Chemokine data were collected from epithelial monolayers infected with three different influenza strains, as described in \cite{Mitchell2011} (Fig.~\ref{fig:data}).
+To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
+The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.
 IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
-To estimate per-cell production rates, we extended the ODE model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.
-\textbf{S}1).
-Model fits, (Table \ref{table:strains}), were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+To estimate per-cell production rates, we extended the ordinary differential equation (ODE) model of Ref.
+\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~\ref{eq:dde}).
+Model fits (Table \ref{tab:strains}) were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
 The resulting chemokine production values were used in the CyCells ABM.
@@
 
-The model uses chemokine gradient as a surrogate for the inflammatory cytokines required for extravasation.
-Figure \ref{fig:cycells} shows chemokine concentrated around virus secreting cells.
-However, T cell sensitivity determines how strong the aggregate chemokine signal must be.
+The model simulates a chemokine gradient surrounding an infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).
+T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.
 
-We simulated T cell sensitivity levels ranging over seven orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) between concentrations of 1 $\mu g/ml$ and 100 $ng/ml$ (100 $ng/ml$ and 10 $ng/ml$ in aH5N1), below which there is no discernible effect on T cell movement (model variance discussed in \textbf{S}2.1.).
-Seasonal and pandemic H1N1 models simulated both IP-10 and RANTES; avian H5N1 included only RANTES (\textbf{S2.2}).
-Because multiple T cells in the same area do not enhance control of infection, we hypothesize more T cells are not more efficient above a critical threshold.
-We set the sensitivity to 10 $ng/mL$ (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
-Single runs were used due to computational limitations.
+Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (single runs were used due to computational limitations, model variance is discussed in S2.1.).
+We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
+
+% Because more than one T cell at each infected epithelial cell would not increasing effective target cell death, there may be a threshold of immigrant T cells above which no additional control of viral replication is possible.
 
@@
 In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
-However, secreting cells were not eliminated in either H1N1 simulations.
+However, secreting cells were not eliminated in either H1N1 simulations (Fig.
+S1).
 Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
@@
 Thus, T cells alone cannot contain the pH1N1 infection.
-In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,643 seconds, allowing T cells to limit a single infected cell to 1.3 viral particles in the one-hour window.
+In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,632 seconds, allowing T cells to limit a single infected cell to 1.4 viral particles in the one-hour window.
 Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.
@@
 
-Hypercytokinemia, or `cytokine storm' is described in virulent influenza infections \cite{DeJong2006} but bloodstream cytokine and chemokine levels may not reflect local tissue concentrations.
-The use of in vitro human bronchial epithelial cells \cite{Mitchell2011} infected at low cell:virus infection ratio (MOI) was intended to mimic initial natural infection and subsequent local infection spread.
-MOI and degree of differentiation of NHBE cells are critical determinants of cytokine/chemokine secretion \cite{Chan2010}.
-Yet, as we noted at low MOI, greater RANTES secretion by H5N1 strains has been found in NHBE cultures infected at high MOI with seasonal \cite{Chan2005, Chan2010, Zeng2011} and pandemic \cite{Zeng2011} strains.
-Thus chemokine secretion stimulated by H5N1 strains does not appear to be attenuated in parallel with the type 1 interferon response \cite{Zeng2007}, placing greater importance on the role of T cells in controlling the infection.
-
-Our model represents only a portion of the complete chemokine signal complex operating in the infected animal.
-Chemokines with a clear role directing T cell traffic include IP-10 \cite{Dufour2002}, RANTES \cite{Kawai1999}, CCL3 \cite{Kawai1999}, and likely others.
-The chemokine receptor CCR5, binding CCL5 (RANTES) and other chemokines, is crucial to accelerated recruitment of CD8+ T cells to lung airways \cite{Kohlmeier2008}.
-Although we did not detect significant levels of other chemokines in our cultures, the neutrophil- and T cell-attractant CXCL8/IL-8 is secreted into seasonal influenza-infected NHBE cultures \cite{Matsukura1996, Arndt2002}.
-In the intact animal, immigrant inflammatory cells particularly macrophages augment the multiple chemokine signals \cite{Julkunen2000}.
-Finally, antigen-specific CD8+ T cell recognition of its target induces lung injury mediated by TNFa and triggers additional alveolar epithelial cell chemokine secretion \cite{Zhao2000}.
+ The actual quantitative chemokine environment in the infected lung is difficult to estimate for a math model.
+The hypercytokinemia documented in virulent influenza infections \cite{DeJong2006} likely does not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells in vitro depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate actual levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 strains is not associated with attenuation of chemokine secretion \cite{Zeng2007}.
+The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
 
 A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
-The diameter of this cloud generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 10 ng/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance as approximately 150 microns in our model.
+The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 10 ng/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance to be approximately 150 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
 This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+Spatially explicit modeling can provide support for not only communication distance but also the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
-Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?
-The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation.
+% - removed - Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?
+The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (S2.2).
 Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity.
@@
 
-This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F and M.M), and NSF grant EF-1038682 (to S.F and M.M.).
+% This publication was funded by NIH grants R21-AI-73607 (to F.K.) and U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F and M.M), and NSF grant EF-1038682 (to S.F and M.M.).
 
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=0.5\textwidth]{SystemChart}
+%\includegraphics[width=4in]{SystemChart}
 \end{center}
-\caption{Model of T cell search.
-Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through the 14 layers of the branching bronchial network.
+\caption{{\bf Model of T cell search.}  Activated T cells originate in the lymph node and enter the bloodstream after which they randomly navigate through 14 vascular bifurcations of the bronchial network.
 Upon reaching a capillary, T cells exit into tissue if cytokine signal is present.
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
-\includegraphics[width=\textwidth]{ModelChart}
+%\includegraphics[width=4in]{ModelChart}
 \end{center}
-\caption{Visual representation of the model.
-Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.
+\caption{{\bf Visual representation of the model.}  Healthy epithelial cells infected by virus begin secreting virus after the incubation delay.
 Activated T cells traverse the bronchial vascular network and may be recruited by inflammatory cytokine.
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
- \includegraphics[width=\textwidth]{data}
+% \includegraphics[width=4in]{data}
  \end{center}
-\caption{Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.
-Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
+\caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
 sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
- \includegraphics[width=\textwidth]{sensitivity}
+% \includegraphics[width=4in]{sensitivity}
  \end{center}
-\caption{Varying T cell sensitivity to chemokine: H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
 Total number of incubating, secreting and apoptotic cells are plotted for each infection.
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
- \includegraphics[width=\textwidth]{cycells}
+% \includegraphics[width=4in]{cycells}
  \end{center}
-\caption{Simulated sH1N1 infection.
-Screenshots from day 4, day 5.5, and day 7.
+\caption{{\bf Simulated sH1N1 infection.} Screenshots from day 4, day 5.5, and day 7.
 The top row shows the spreading focus of infection  through the color coding of individual cells:  healthy cells in uninfected tissue (gray),  virus-incubating cells (yellow), virus-secreting cells (orange), apoptotic cells (red), dead cells (brown), and T cells arriving at day 5 (green).
-Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL (see the legend in the bottom right).}
+Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL.
+The individual cell images for days 5.5 and 7 have been enhanced by hand for improved T cell visibility.
+Individual groups of T cells appear larger than they are and only cover an area of one or two epithelial cells in the model.
+Original images available upon request.
+}
  \label{fig:cycells}
@@
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
- \includegraphics[width=\textwidth]{plaquesize}
+% \includegraphics[width=\textwidth]{plaquesize}
  \end{center}
-\caption{Comparison of total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red) for the simulated infections of aH5N1, sH1N1, and pH1N1.
-T cells are able to clear both secreting and incubating cells in aH5N1, are unable to contain incubating cells in sH1N1, and are unable to contain either type of infected cell in pH1N1.
-We notice a threshold at day 5: pH1N1 has more expressing cells (green) by a factor of only two to three, yet unlike aH5N1 and sH1N1, the T cell response is unable to clear the expressing population.}
+\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1.} Plotted values: total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red).
+T cells clear secreting and incubating cells in aH5N1, fail to clear incubating cells in sH1N1, and fail to clear either type of infected cell in pH1N1.
+The number of expressing cells (green) after day 5 differs markedly among the three strains indicating that the T cells have differing success at controlling the infection.}
  \label{fig:plaquesize}
 \end{figure}
+
+% In the simulation, increasing chemokine concentration fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.
+
+\pagebreak
 
@@
 \begin{center}
-\begin{tabular}{ | c | c | c | }
+\begin{tabular}{|  l  l  l  |}
+  \hline
+  Population & Description & Initial Value\\
+  \hline
+  $T$ & Healthy target eptithelial cells & 1,000,000 \\
+  $I_1$ & Virus-incubating cells & 0 \\
+  $I_2$ & Virus-secreting cells & 0 \\
+  $V$ & Free virus particles & 10,000 \\
+  $F$ & Interferon quantity & 0 \\
+  $C$ & Chemokine quantity (ng/mL) & 0 \\
+  \hline
+  Parameter & Description & Value \\
+  \hline
+  $\beta$ & Viral infection rates &  \\
+    & \hspace{2em} Avian & 5.3e-7 (PFU$\cdot$h)$^{-1}$ \\
+    & \hspace{2em} Seasonal & 6.1e-7 (PFU$\cdot$h)$^{-1}$ \\
+    & \hspace{2em} Pandemic &  2.7e-6 (PFU$\cdot$h)$^{-1}$\\
+  $p$ & Viral production rates & \\
+    & \hspace{2em} Avian & 0.20 (PFU/h) \\
+    & \hspace{2em} Seasonal & 1.4 (PFU/h) \\
+    & \hspace{2em} Pandemic & 18.3 (PFU/h) \\
+  $e$ & Interferon strengths & \\
+    & \hspace{2em} Avian & 1.0e-8 \\
+    & \hspace{2em} Seasonal & 1.6e-6 \\
+    & \hspace{2em} Pandemic & 3.4e-3 \\
+  $\tau_2$ & Interferon production delays &  \\
+    & \hspace{2em} Avian & 21.5 (h) \\
+    & \hspace{2em} Seasonal & 23.6 (h) \\
+    & \hspace{2em} Pandemic & 21.0 (h) \\
+  $\delta$ & Virus-secreting cell decay rate & 0.6 (h$^{-1}$) \\
+  $d$ & Chemokine decay rate & 1.386 (h$^{-1}$) \\
+  $\tau_1$ & Viral incubation time & 10 (h) \\
+  $\tau_3$ & Chemokine production delays & \\
+    & \hspace{2em} IP-10 & 8 (h)\\
+    & \hspace{2em} RANTES & 16 (h)\\
+  \hline
+\end{tabular}
+\caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were take from the observed chemokine data (Fig.~\ref{fig:data}).
+Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
+\label{tab:dde}
+\end{center}
+\end{table}
+
+\begin{table}
+\begin{center}
+\begin{tabular}{| l l l |}
   \hline                        
@@
   Epithelial Cell Radius & $25 \mu m$ & Selected$^5$ \\
-  T Cell Age (at FOI) & $2 hours$ & Selected$^6$ \\
+  T Cell Age (at FOI$^\ast$) & $2 hours$ & Selected$^6$ \\
   T Cell Age (in Blood) & $3 days$ & Selected$^6$ \\
-  Onset of ATC Lymph Node Exit & $5 days$ & \cite{Banerjee2011} \\
+  Onset of ATC$^\ddagger$ Lymph Node Exit & $5 days$ & \cite{Banerjee2011} \\
   IgM Strength & Viral decay of $3/day$ & Selected$^7$ \\
@@
 \end{tabular}
-\caption{Static model parameters.
-1) Chosen to correspond to a 30 minute half-life.
+\caption{1) Corresponds to a 30 minute half-life.
 2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.
 This scales linearly with the number of virions in the cell's vicinity.
-3) Chosen to be a plausible median time (1,000 minutes) between 6 hours and 24 hours.
+3) Chosen as a plausible median time (1,000 minutes) between 6 hours and 24 hours.
 4) T cells induce apoptosis in nearby virus-secreting epithelial cells at a probabilistic rate such that the expected time to induce apoptosis is 10 minutes.
@@
 5)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
-6) Parameters chosen after discussions with David Woodland, Trudeau Institute.
+6) Parameters used after discussions with David Woodland, Trudeau Institute.
 7) IgM presence is abstracted by increasing viral decay by a factor of three.
-}
-\label{table:parameters}
+$^\ast$Focus of Infection.
+$^\ddagger$Activated T Cell.}
+\label{tab:parameters}
 \end{center}
@@
 Viral production values are taken from \cite{Mitchell2011}.}
-\label{table:strains}
+\label{tab:strains}
 \end{table}

### 2012-12-16 paper/chemokine_supplement.tex
Updated with Fred's new Intro and Discussion, plus various minor touchups

@@
 
-\subsection{Delay Differential Equation Model}
-
-We adapt the delay differential equations first presented in \cite{Mitchell2011} to find values for chemokine production rates.
-We use initial populations and parameter values from the previous study while adding an extra equation to model chemokine production (final equation):
-
-\begin{equation*}
-\label{EqS1}
-\begin{aligned}
-\dot{T} &= - \beta T V \\
-\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
-\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
-\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
-\dot{F} &=  I_1[t-\tau_2] \\
-\dot{C} &= r I[t-\tau_3] - d C \\
-\end{aligned}
-\tag{\textbf{S}1}
-\end{equation*}
-\vspace{.05in}
-
-$C$ represents the chemokine population.
-Two parameters, $r$ and $d$, control the rate of chemokine production and the rate of chemokine decay.
-Chemokine decay, $d$, was set to $3.8508\cdot10^{-4}$/s, which equates to a 30 minute half-life.
-$r$, being the only free parameter, was fit to the empirical data shown in Fig.
-1 of the main paper using Matlab's \texttt{nlinfit} function.
-This process was repeated six times, once for each set of data.
-$\tau_3$, the chemokine production delay, was set to 8 hours for IP-10 fits and 16 hours for RANTES fits.
-
 \subsection{T Cell Production Rate}
 
-To find a value for $\sigma$ we use a differential equation model from \cite{Miao20101}:
+To find a value for $\sigma$, the rate of production of activated T cells (ATCs) we use a differential equation model from \cite{Miao2010}:
 
+{\footnotesize
 \begin{equation*}
@@
 \end{aligned}
-\tag{\textbf{S}2}
+\tag{\textbf{S}1}
 \end{equation*}
+}
 \vspace{.05in}
@@
 The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
+See Fig.
+1 for more details about the model.
 The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
 
-Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$.
-These circulating cells then transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$).
+Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
+These circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$) at a given speed, $v_{tcell}$.
 Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
@@
 This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
+The final value was found to be 777 per hour.
+
+\subsection{CyCells Implementation Details}
+
+The CyCells modeling environment splits its populations into two types: cells and molecules.
+Cells are considered to be unique agents and are modeled individually.
+On the other hand, molecules are considered to be too small and numerous to model explicitly and are thus represented as compartmentalized concentrations.
+These compartments are arranged as a grid that covers the environment.
+Each square in the grid contains a unique concentration of the given molecule.
+Multiple molecules are represented in separate grids.
+
+Molecular behavior is limited to diffusion and decay.
+At each time step, CyCells decreases the concentration of each square in the grid according to the decay rate specified for the given molecule.
+CyCells then diffuses the molecules by applying a discrete diffusion equation to each grid square, taking into account the specified diffusion constant and the concentrations in neighboring squares.
+
+New molecules can be secreted by cell objects, such as an infected cell secreting new virus.
+In this scenario CyCells calculates the amount of virus secreted per time step, based upon the cell's defined production rate, and then adds this quantity to the grid square that overlaps the secreting cell.
+If a cell wants to remove molecular concentration, it will subtract from the square it overlaps.
+If the concentration at that square is not enough, concentration will be removed from neighboring squares in an ever expanding diamond.
+
+When a cell's behavior depends on a local molecular concentration, the cell will measure the local concentration at its position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell.
+
+\subsection{Model Implementation Details}
+
+An overview of the model implementation is shown in Fig.
+2.
+The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (46,085 total cells).
+The sheet measures 10$mm$ per side, which is large enough to contain the infection (Fig.
+5).
+The molecular grid (S1.2) is initialized with a resolution of 50$\mu m$.
+
+Healthy cells transition to virus-incubating based on a probability that scales linearly with the local virus concentration.
+When a cell `becomes infected' it removes the amount of viral concentration equal to a single virion.
+
+Virus-incubating cells idle for 10 hours (with standard deviation $\sigma=1$ hour) and then transition to virus-secreting cells.
+Virus-incubating cells also begin to secrete chemokine after 8 hours (except for aH5N1 which only secretes RANTES after a 16 hour delay).
+
+Virus-secreting cells secrete both virus and chemokine (the RANTES portion of the chemokine production is added in 16 hours after the initial infection) and die after 1,000 minutes of production (16.7 hours).
+Virus-secreting cells will also probabilistically transition to apoptizing cells in the presence of T cells (defined by the existence of a T cell within 5$\mu m$).
+The number of T cells in the vicinity of the virus-secreting cell has no effect on the rate that apoptosis is induced in the model.
+
+Apoptizing cells continue to secrete virus and chemokine for one hour and then die.
+
+T cells are added to the model at a constant rate of 777 per hour (S1.1) through a global `oracle' cell that represents the lymph node.
+Because the model environment is only a small portion of the entire lung (1\%), most of the T cells miss the model window and are not visually represented.
+The T cells that miss enter a dormant searching state and transition to recirculating after a 10 minute delay.
+The cells that enter the model window are placed at a random position and begin a random walk.
+At each step, these searching cells check the local chemokine concentration.
+If the concentration is above the sensitivity threshold, the T cells immediately transition to the chemotaxing state.
+If not, searching T cells transition to the recirculating state and are removed from the model window.
+Searching and recirculating T cells have a probabilistic decay rate that corresponds to a 3 day lifespan.
+
+Recirculating T cells are not visually represented.
+These cells are individually tracked for 6 minutes, after which they are reintroduced as new searching cells with a new chance of entering the model window.
+
+Chemotaxing T cells move directly up the chemokine gradient until they find a local maximum.
+T cells probabilistically decay at a rate corresponding to a 2 hour lifespan.
+Chemotaxing T cells have no effect other than inducing apoptosis in virus-secreting cells by proximity.
 
@@
 
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Figure~\ref{fig:variance}).
-To test the strength of this effect, we ran each model fifty times using the default parameters given in Table~\ref{table:parameters}.
+Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~S\ref{fig:variance}).
+To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.
 $R^2$ values were calculated for individual runs versus the average of all runs.
@@
 
-The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~\ref{fig:variance}).
+The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~S\ref{fig:variance}).
 The rapid production of the pandemic H1N1 virus prevents the immune response from containing the infection.
 Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T Cells to gain ground is too small.
-aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH5N1 recovers and continues to expand.
+aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH1N1 recovers and continues to expand.
 
@@
 Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.
-We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table~\ref{table:strains}).
+We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).
 Because of this behavior IP-10 was not included in the aH5N1 model runs.
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
+
+T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.
+
+Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~S\ref{fig:chemokine}).
 The lack of both chemokines leads to runaway infections in all three strains.
 The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
-IP-10 alone proves to be as effective as the combination of IP-10 and RANTES, suggesting that RANTES does not play a significant role in infections that stimulate an IP-10 response.
-We assume receptor sensitivity is the same for both chemokines and that the chemokines in combination work additively.
+In sH1N1 and pH1N1 simulations, IP-10 alone proves to be as effective as the combination of IP-10 and RANTES.
+This suggests that RANTES does not play a significant role in infections that stimulate an IP-10 response due to the higher production rates of IP-10.
 
@@
 \bibliography{references}
+
+\pagebreak
 
@@
 \begin{center}
- \includegraphics[width=\textwidth]{variance}
+% \includegraphics[width=\textwidth]{variance}
  \end{center}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{chemokine}
+%	\includegraphics[width=\textwidth]{chemokine}
 	\caption{Effects of different chemokine combinations.
@@
 \end{figure}
+
+\begin{figure}[ht!]
+\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
+T cells show half way through the simulation.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
+ \label{video:cell_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{The second of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the virus concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:virus_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{The third of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the chemokine concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:chemokine_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{A closer look at the 2009 pandemic simulation.
+This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
+Note the high proportion of virus-secreting cells (orange) early on.
+As time passes, secreting cells are gradually contained to the point where they become very sparse.
+T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
+\end{figure}
+
+\pagebreak
 
@@
 
-\section*{Video}
-
-\begin{figure}
-\caption{First of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
-T cells show half way through the simulation.
-Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
- \label{video:cell_view}
-\end{figure}
-
-\begin{figure}
-\caption{Second of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the virus concentration.
-Notice the volatility when T cells arrive halfway through the simulation.
-Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).
-Refer to Figure 5 for the detailed legend.
-}
- \label{video:virus_view}
-\end{figure}
-
-\begin{figure}
-\caption{Third of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the chemokine concentration.
-Notice the volatility when T cells arrive halfway through the simulation.
-Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).
-Refer to Figure 5 for the detailed legend.
-}
- \label{video:chemokine_view}
-\end{figure}
-
-\begin{figure}
-\caption{A closer look at the 2009 pandemic simulation.
-This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
-Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
-Note the high proportion of virus-secreting cells (orange) early on.
-As time passes, secreting cells are gradually contained to the point where they become very sparse.
-T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
-\end{figure}
-
 \end{document}

### 2013-01-10 paper/response.tex
Final submitted to PLoS CB

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

\usepackage{multirow}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration (Response to Reviewers)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin, 
Stephanie Forrest, 
Soumya Banerjee,
Candice Clay,
Melanie Moses, 
Frederick Koster
\end{flushleft}
\vspace{1cm}

We thank the reviewers for their constructive and insightful comments.  The incorporation of the reviewers' comments has greatly improved the paper.  We include detailed responses to each comment below.

\section*{Reviewer 1}

\begin{enumerate}

\item \textbf{The title may be misleading as the model is very simplified from the immunological viewpoint and gives rather a qualitative rather than a quantitative view of the spatial dynamics.}

We changed the title by changing the phrase `quantifies constraints' to `reveals thresholds', which we believe more accurately reflects the paper's main focus. \\

\item \textbf{Introduction, p3: It would be interesting to add a general reference to agent-based models (ABM), in particular spatial ABM in immunology and briefly justify the use of an ABM rather than partial differential equations in the present context.}

Added reference to Beauchemin 2006 paper that discusses ABM advantages vs. differential equations as well as Bauer 2009.  We have added the following sentences to the introduction: \\

\textit{Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment. We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.} \\

\pagebreak

\item \textbf{Computational modeling section: there is no reference to Fig. 1 in the main text. As far as I understand the drawing in Fig. 1, I see 16 branches rather than 14, why?}

A reference to Fig. 1 was added to the Model section.  The number 14 refers to the depth of the bronchial tree, where each layer represents the bifurcation of a branch.  Thus, each level results in the doubling of the total number of pathways.  14 branches refers to $2^{14} = 16,384$ endpoints.  The figure simplifies this to 4 layers and thus $2^4 = 16$ endpoints.  The figure has been edited to read `14 bifurcations' rather than `14 branches' and the caption was changed to read `vascular bifurcations' for clarity. \\

\item \textbf{Acronyms in text and tables should be defined (for example, ODE, FOI, NHBE, ATC, ?).}

Acronyms in the text have been either defined or removed.  Acronyms in the table are now defined in the caption. \\

\item \textbf{Model definition:}
\begin{enumerate}
   \item \textbf{More details concerning the implementation of the ABM need to be given, at least in the supplement: a short description of the spatial arrangement (size of the simulation space, cell arrangement, number of epithelial cells, as well as how T cell movement vs. particle diffusion has been implemented.}

   \item \textbf{IgM addition is mentioned in section 2.1 of the supplement but not indicated in the main text.}

   \item \textbf{What happens in this ABM when a free virus particle infects a cell: is it absorbed and does it disappear from the pool of free viruses?}

\end{enumerate}

Two new sections were added to the supplement (S1.2, S1.3).  One describes details of the general CyCells model environment and the other describes details of the model specific to this paper.  These sections address all of the points raised here by the reviewer except IgM.     We added a sentence referencing the IgM implementation to the main paper: \\

\textit{IgM is modeled by increasing the viral decay rate by a factor of three after the third day.} \\

\item \textbf{Materials, p6: the sentence "treatment of the monolayer with protease" is not understandable without reading reference [19] and requires here more explanation.}

The description of viral measurement was expanded into two sentences and an explanation of why protease was used was added: \\

\textit{Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.  Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus (Mitchell 2011).} \\

\item \textbf{Chemokine production, p6:}

\begin{enumerate}
	\item \textbf{The definition of all the variables and parameters of the ODE model in the supplement (eqs S1) should be added in order to be more self-contained.}

The central differential equation was moved to the beginning of the model section of the main paper and a new table was added with descriptions of every population and parameter. \\

	\item \textbf{Fig. 3: "8500 pg/mL" is indicated as a measurement threshold for sH1N1.  Why is this not also the case for aH5N1 and pH1N1?}

The upper limit applied only to sH1N1 because there were insufficient sample dilutions made and the sample could not be repeated.  The sentence in the caption has been rewritten: \\

\textit{sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.} \\

\end{enumerate}

\pagebreak

\item \textbf{T cell sensitivity, p7, top: The sentence "Because multiple T cells ? we hypothesize ? more T cells are not more efficient above a critical threshold" is unclear. Is it not that the probability of apoptosis induction will increase with the T cell number even if the rate of transition from secreting to apoptotic cells does not scale with T cell number?}

It is {\em not} that the probability would increase with additional T cells at the infected epithelial cell in our model.  The section has been rearranged to improve clarity. The detail regarding multiple T cells was moved to other sections.  We also added two expanded model description sections to the supplement, which we hope clarifies this point. \\

\item \textbf{Spatial effects:}
\begin{enumerate}
	\item \textbf{p7, bottom: refer also to Fig. S1}

A reference to Figure S1 was added. \\

	\item \textbf{Figure 6, legend: Can the difference in behaviour for pH1N1 and sH1N1 infections not be explained by the difference in IP-10 production rates (even if small)?}

We performed a new model run combining pH1N1 virus and sH1N1 chemokine and found no difference in the infection profile.  A sentence was to the spatial effects results section: \\

\textit{Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.} \\

\end{enumerate}

\item \textbf{Discussion, p10, 2d §: reference should also be given to section 2.2 of the supplement where combinations of chemokines are considered.}

The Chemokine Directed T cell Search section of the discussion was significantly abbreviated for clarity and removal of information not directly relevant to the modeling exercise.  The paragraph in question was removed entirely. \\

\item \textbf{Equations:}
\begin{enumerate}
	\item \textbf{Eqs S1: add definition of variables and parameters}

	We  moved the first equation from the supplement to the main paper to  help address the point that the paper is not self-contained.  We have added a new table with descriptions and values for each population and parameter of this equation. \\

	\item \textbf{Eqs S2: add definition of $t_{rc}$ and $v_{tcell}$; reference to Fig. 1 of the main text should also be given in section 1.2 of the supplement as eqs. S2 relate to the bottom part of this figure.}

	These modifications have all been made. \\

	\item \textbf{Section 2.2: does the conclusion that RANTES does not play a significant role for H1N1 infections not simply follows from the fact that IP-10 has a much higher production rate (about 100 times) than RANTES in the H1N1 strains (Table 2, main text)?}

	Indeed, it does follow given the assumption that T cells are equally sensitive to each chemokine.  The text in S2.2 has been modified to make that assumption clear and to strengthen the conclusion.  The following two sentences in particular address the point. \\

	\textit{T cell sensitivity depends on receptor density (Desmetz 2006) and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.} \\

	\textit{This suggests that RANTES does not play a significant role in infections that stimulate an IP-10 response due to the higher production rates of IP-10.} \\

\end{enumerate}

\pagebreak

\item \textbf{Minor points:}
\begin{enumerate}
	\item \textbf{p7, Fig. 5, top row: T cells are barely visible. Could this be improved?}

	The figure has been enhanced to improve visibility and a note regarding the change has been added to the caption: \\

	\textit{Individual cell images for days 5.5 and 7 were enhanced (green dots are enlarged) manually to improve T cell visibility.  Thus, individual groups of T cells, which cover only an area of one or two epithelial cells in the model appear larger in the image.  Original images available upon request.} \\

	\item \textbf{p7, Fig. 6, legend, 5th line:  add "more expressing cells than sH1N1".}

	This has been done. \\

	\item \textbf{p8, line 4: taking the viral production indicated in Table 2, I find 2632 rather than 2643}

	This was calculated before we rounded to significant digits and has been fixed. (Impressive eye for detail!) \\

	\item \textbf{p9, line 12 from bottom: greater ? than}

	This sentence has been removed. \\

	\item \textbf{p9, line 3 from bottom: replace "into" by "in"}

	This has been done. \\

	\item \textbf{Note that some Table and reference numbers have been replaced by question marks in the supplement}

	These have been fixed.

\end{enumerate}

\end{enumerate}

\section*{Reviewer 2}

\begin{enumerate}

\item \textbf{Very little detail is given about the agent-based model and how chemokines influence cell migration behavior. The authors cite several papers (4,17,18) but as far as I can see these papers do not provide convincing evidence that chemokines direct migration of effector T cells to the sites of infection (foci of infected cells). In fact, I don't think that there is any evidence of that process. How chemokines direct migration of lymphocytes in tissues is not well understood and the cartoon knowledge of textbooks is mainly applicable to the process of extravasations and may not be used by T cells to find infected cells.}

We agree with the reviewer that for the influenza model the evidence for chemokine-directed migration of T cells to infected foci is incomplete.  To provide a justification for constructing a math model based on chemokine-directed migration, we have completely rewritten the Introduction in order to thoroughly review the available literature on chemokine function with respect to RNA virus infections.  We state that there is no evidence for T cells climbing chemokine gradients, and that this activity has been shown only for neutrophils.  However, we feel that the pieces of evidence, taken together, support a reasonable hypothesis for chemokine-directed migration, in addition to extravasation, providing a reasonable justification for the model-building. \\

\item \textbf{How you analyze experimental data is unclear. Which equations did you use?}

This was a common theme in our reviews.  We have made the paper more self-contained by moving the central differential equation to the main paper.  We added details of the inner workings of the model to the supplement in two new sections (S1.2 and S1.3).  This allows the paper to stand alone from our previous work in Mitchell 2011.  \\

\pagebreak

\item \textbf{Please check parameter values that you provided in Table 1. It appears to several of those are made up and are not really measured in the referenced work.}

Table 2 was split into a referenced parameter section and an estimated parameter section to make the distinction clear.  We reviewed all of our references and annotated two parameters in Table 2: Apoptosis Time and T Cell Speed as follows:

\begin{itemize}

\item Apoptosis time has been rewritten as an estimated parameter and a note has been made that this value is specific to low T cell densities.

\item T cell speed has been rewritten as an estimated parameter: \textit{The values were selected to be an order of magnitude faster than those calculated for inside the lymph node (Miller 2003).} \\

\end{itemize}

\end{enumerate}

\section*{Reviewer 3}

\textbf{A minor comment is that I think it would be interesting to comment in the discussion on the fact that most pathogenic isolates kill the host (in animal models, for sure) presumably before an large CTL infiltration could occur, but that's really not the focus of the manuscript.} \\

While it does not address early mortality in animal hosts, we added a note in the discussion addressing the fact that most humans did not die from pH1N1 infection: \\

in vivo \textit{pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation}

\end{document}

### 2013-01-10 paper/Text_S1.tex
Final submitted to PLoS CB

--- 2012-12-16 paper/chemokine_supplement.tex
@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration (SUPPLEMENT)}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration (SUPPLEMENT)}
 }
@@
 Stephanie Forrest$^{1}$, 
-Soumya Banerjee$^{1}$
+Soumya Banerjee$^{1}$,
 Candice Clay$^{2}$, 
 Melanie Moses$^{1}$, 
-Frederick Koster$^{1,2}$, 
+Frederick Koster$^{1,2}$
 \\
@@
 
-To find a value for $\sigma$, the rate of production of activated T cells (ATCs) we use a differential equation model from \cite{Miao2010}:
+We calculate the rate of production ($\sigma$) of CD8 T cells using a differential equation model from \cite{Miao2010}.
+The equations model T cell production and subsequent search over an area of infected lung tissue.
 
@@
 \begin{equation*}
-\label{EqS2}
+\label{EqS1}
 \begin{aligned}
@@
 \end{aligned}
-\tag{\textbf{S}1}
+\tag{S1}
 \end{equation*}
@@
 
-where $N_{c}$ is the number of circulating activated antigen-specific T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines and $N_{f}$ is the number of circulating T cells that have found an infected region.
-$T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
+$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region.
+$T$ is the number of uninfected \textit{target} cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
 The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
-The lung is assumed to be a circular region with radius $R$.
+The lung is modeled as a circular region with radius $R$.
 The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
@@
 The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
+$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
 
-Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
+Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
 These circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$) at a given speed, $v_{tcell}$.
+
+The average time an exiting CD8 T cell takes to migrate is proportional to the difference in the radii between the two regions ($r' - r$) at a given speed, $v_{tcell}$.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for that time.
 Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
@@
 The initial number of infected cells is assumed to be zero.
+Parameter values are listed in Table S1 in Text S1.
 This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
@@
 
-When a cell's behavior depends on a local molecular concentration, the cell will measure the local concentration at its position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell.
+To model cell dependence on the local molecular concentration, the cell will measure the local concentration at its position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell.
 
@@
 5).
-The molecular grid (S1.2) is initialized with a resolution of 50$\mu m$.
+The molecular grid (Text S1 1.2) is initialized with a resolution of 50$\mu m$.
+Each time step of the model represents 10 seconds.
 
@@
 
-Apoptizing cells continue to secrete virus and chemokine for one hour and then die.
+Apoptizing cells secrete virus and chemokine for one hour and then die.
 
-T cells are added to the model at a constant rate of 777 per hour (S1.1) through a global `oracle' cell that represents the lymph node.
+T cells are added to the model at a constant rate of 777 per hour (Text S1 1.1) through a global `oracle' cell that represents the lymph node.
 Because the model environment is only a small portion of the entire lung (1\%), most of the T cells miss the model window and are not visually represented.
@@
 $R^2$ values were calculated for individual runs versus the average of all runs.
-Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.
-sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.
-aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.
+The $R^2$ ($\pm$ one standard deviation) for pandemic, seasonal, and avian influenza are: (0.994 $\pm$ .003), (0.986 $\pm$ 0.011), and (0.917 $\pm$ 0.073) respectively.
 
@@
 
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~S\ref{fig:chemokine}).
+Four models runs (both, IP-10, RANTES, and none) were performed for sH1N1 and pH1N1 strains and two runs (RANTES and none) for  aH5N1.
+The runs show how the presence and/or absence of specific chemokines affect the simulated immune response (Fig.~S\ref{fig:chemokine}).
 The lack of both chemokines leads to runaway infections in all three strains.
@@
 
-\begin{figure}[ht!]
-\begin{center}
-% \includegraphics[width=\textwidth]{variance}
- \end{center}
-\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
+%\begin{figure}[ht!]
+%\begin{center}
+%% \includegraphics[width=\textwidth]{Figure_S1}
+% \end{center}
+%\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
 IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
@@
 The middle line shows the average while the outer lines show the 95\% confidence interval.}
- \label{fig:variance}
-\end{figure}
-
-\begin{figure}[ht!]
-\begin{center}
-%	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
+% \label{fig:variance}
+%\end{figure}
+%
+%
+%\begin{figure}[ht!]
+%\begin{center}
+%%	\includegraphics[width=\textwidth]{Figure_S2}
+%	\caption{Effects of different chemokine combinations.
 A) aH5N1 does not stimulate an IP-10 response.
 B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
+%	\label{fig:chemokine}
+%\end{center}
+%\end{figure}
+%
+%
+%\begin{figure}[ht!]
+%\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
 This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
@@
 Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
- \label{video:cell_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The second of three overlaid videos of a representative seasonal H1N1 infection.
+% \label{video:cell_view}
+%\end{figure}
+%
+%\begin{figure}[ht!]
+%\caption{The second of three overlaid videos of a representative seasonal H1N1 infection.
 This video spans the 10 day infection and shows the virus concentration.
@@
 }
- \label{video:virus_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The third of three overlaid videos of a representative seasonal H1N1 infection.
+% \label{video:virus_view}
+%\end{figure}
+%
+%\begin{figure}[ht!]
+%\caption{The third of three overlaid videos of a representative seasonal H1N1 infection.
 This video spans the 10 day infection and shows the chemokine concentration.
@@
 }
- \label{video:chemokine_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{A closer look at the 2009 pandemic simulation.
+% \label{video:chemokine_view}
+%\end{figure}
+%
+%\begin{figure}[ht!]
+%\caption{A closer look at the 2009 pandemic simulation.
 This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
@@
 T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
-\end{figure}
+%\end{figure}
 
@@
 %\label{tab:label}
+
 % \end{table}
+
+\renewcommand{\thetable}{S\arabic{table}}
 

### 2013-01-10 paper/chemokine_main.tex
Final submitted to PLoS CB

@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals constraints on chemokine directed migration}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
 }
@@
 Stephanie Forrest$^{1}$, 
-Soumya Banerjee$^{1}$
+Soumya Banerjee$^{1}$,
 Candice Clay$^{2}$, 
@@
 During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.
-T cell navigation to this region is undirected but migration is made possible by local cytokine and chemokine production from infected epithelial cells.
-We examined the efficiency of local chemokines to induce migration of activated CD8 T cells using an agent-based model (CyCells).
+T cell navigation to infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
+We focused on the efficiency of local chemokines to induce migration of activated CD8 T cells using a spatially explicit agent-based model (ABM) without constructing a model containing all elements of the immune response.
 
-Avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used to induce chemokine production \textit{in vitro}.
-Of the chemokines tested, secretion of CXCL10 (IP-10) and CCL5 (RANTES) were stimulated by infection, with IP-10 dominating the migration due to higher concentrations.
+To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.
 A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.
-We observe that T cell sensitivity to chemokine is high enough to maximize their efficiency.
 The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.
 
-The spatial nature of the model reveals unique challenges to T cell recruitment not visible in standard differential equation models.
-Infected cells can become isolated in expanding plaques, impeding infection control.
-In the model, T cells executed an efficient search with sensitivity to either chemokine.
-A key limitation imposed on T cells was illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci as revealed in cinematics.
-This spatial model is consistent with efficient T cell recruitment to infected lung.
-Without an apparent means to induce increased efficiency, vaccination remains the key to shortening the interval between initial infection and T cell activation.
+The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.
+Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.
+A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci.
+This spatially explicit model is consistent with efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.
 
@@
 
-The adaptive immune response induced during acute influenza A infection is a complex web of defense mechanisms that control viral replication of all but the most virulent strains.
+The adaptive immune response induced during acute influenza A infection is a complex combination of defense mechanisms.
 Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.
@@
 The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.
-On approximately day 5 after infection activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.
+On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.
 
@@
 The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.
-For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediated T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediated the opposite effect \cite{DeLemos2005}.
-In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of CXCL10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.
-In the herpesvirus model CXCL10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.
-In the parainfluenza virus model CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.
-In the influenza A model initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.
-The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency but viral clearance was not altered \cite{Fadel2008}.
+For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediates T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediates the opposite effect \cite{DeLemos2005}.
+In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of IP-10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.
+In the herpesvirus model, IP-10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.
+In the parainfluenza virus model, CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.
+In the influenza A model, initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.
+The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency, but viral clearance was not altered \cite{Fadel2008}.
 The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.
 
-Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, T cells have not be shown to climb chemokine gradients.
+Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
 Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.
@@
 
-A number of mathematical models have constructed the entire adaptive response and key components \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
-Focused math models can examine each event in detail and narrow the range of possible solutions.
-We sought to examine the CD8 T cell diaspora and recruitment to infected lung in detail.
-Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail.
-Activated T cells searching for and homing into infected tissue do so in a spatially complex environment.
-We therefore utilized the strategy of agent-based modeling (ABM) to represent the physical environment of the searching T cell.
-Thus T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
+There are many mathematical models of the entire adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
+Mathematical and computational models can examine each event in detail and narrow the range of possible behaviors.
+Here we examine the CD8 T cell diaspora and recruitment to infected lung in detail.
+Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.
+Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment.
+We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.
+In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
 
@@
 Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
-The model does not aim to predict control of viral replication but only compare simulated outcome among three influenza strains.
-To achieve this focused goal on the efficacy of the chemotactic signals alone, we use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.
+The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.
+We use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.
 Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.
@@
 % authors may use "Analysis" 
-\section*{Models}
+\section*{Materials and Methods}
 
@@
 
-We estimate chemokine production rates by adapting the delay differential equation model described in \cite{Mitchell2011} by adding one new equation to model chemokine production.
+We estimate chemokine production rates by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} by adding one new equation to model chemokine production.
 Initial population sizes and parameter values are taken from the previous study.
@@
 Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
+A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
 The lung was represented as a two-dimensional sheet of healthy epithelial cells.
-Vasculature was represented as a binary tree with fourteen branches originating at a single lymph node.
+Vasculature was represented as a binary tree with fourteen levels of bifurcations with the origin at the bronchial lymph node complex.
 Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
 T cells that do not encounter cytokine recirculate to the lymph node.
-At the site of infection when a T cell encounters an infected epithelial cell it induces apoptosis.
+When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.
 
 The simulation begins when a single cell in the center of the tissue is infected.
-After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine.
+After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type.
 Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-Chemokine diffuses from secreting cells, creating a ball of stimulation around the infected region.
+Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region.
 After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue.
-Because T  cells cannot choose their path through the branching network, we assume they arrive in tissue at random locations.
+Because T  cells cannot choose their path through the branching network, we assume they arrive in the capillary bed at random locations.
 
@@
 
-\section*{Materials}
+\subsection*{Materials}
 
@@
 Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
-Apical fluid for viral secretion, and basal media for chemokine secretion collected before treatment of the monolayer with protease, was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
+Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
+Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.
 Quantitative viral culture was performed by standard plaque assay.
@@
 
-The model simulates a chemokine gradient surrounding an infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).
+The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).
 T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.
 
-Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (single runs were used due to computational limitations, model variance is discussed in S2.1.).
+Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (model variance is discussed in Text S1 2.1.).
 We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
@@
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-Refer to the videos in Figures S3-S5 to see a representative simulation of seasonal influenza.
+Refer to the Videos S1-S3 to see a representative simulation of seasonal influenza.
 
@@
 Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.
+Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.
 
@@
 At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
-By day 7, this ratio is approximately 1:100 for both the pH1N1 and sH1N1 strains.
+By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.
 The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.
-The video in Figure S6 illustrates this event.
+Video S4 illustrates this event.
 
@@
 
-Limitations in the model are primarily due to a number of simplifications to and deletions of elements in the innate immune response, allowing us to build a tractable model where data was available.
-Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated T cells from regional lymph nodes.
+The model makes a number of simplifications to and deletions of elements in the innate and adaptive immune response, allowing us to build a tractable model where data was available.
+Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.
 Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
-The contribution of IgM antibody clearing free virus is represented at a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable as indicated by superior data-fitting models \cite{Wu2011}.
+The contribution of IgM antibody clearing free virus is represented as a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
+All of these rates may in fact be time-variable as indicated by data-fitting models \cite{Wu2011}.
 Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial in the control of lung infection \cite{Miao2010}.
-Thus, our model is not intended to predict clearance of virus from the lung.
-Rather our goal was to examine the features of the response that permitted T cells to sense and contact infected target cells.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.
+Thus, our model is not intended to predict clearance of virus from the lung, but to examine the response  features that permit CD8 T cells to sense and contact infected target cells.
 
@@
 An ODE model assumes that any virus particle is capable of infecting any healthy cell.
-Figure \ref{fig:cycells} shows that this is clearly not the case.
+Figure \ref{fig:cycells} shows that this is clearly not the case for viral adhesion and entry.
 In fact, most free virus exists on top of infected cells that are no longer candidates for viral binding and fusion.
@@
 
-Our ABM renders the model in OpenGL (Fig.~\ref{fig:cycells}).
-Seeing the model in action reveals spatial dynamics that are absent in ODEs and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems, for example, the spatial dynamics discussed earlier.
-First, because T cells find infected cells by climbing a gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
-Second, T cell clustering persists after all virus-secreting cells have been eliminated.
+Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.
+Because CD8 T cells find infected cells by presumably climbing a chemokine gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
+We also see that T cell clustering persists after all virus-secreting cells have been eliminated.
 The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.
@@
 Because T cells are clustered they cannot cover the increasing plaque effectively.
-These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
-The behavior of searching T cells described in this paper
-can enhance future global immune response models.
+These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM (\textit{in vivo} pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation).
+The behavior of searching T cells described in this paper can enhance future global immune response models.
 
@@
 % - removed - Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?
-The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (S2.2).
+The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (Text S1 2.2).
 Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity.
@@
 \begin{center}
-%\includegraphics[width=4in]{SystemChart}
+%\includegraphics[width=4in]{Figure_1}
 \end{center}
@@
 \begin{center}
-%\includegraphics[width=4in]{ModelChart}
+%\includegraphics[width=4in]{Figure_2}
 \end{center}
@@
 \begin{center}
-% \includegraphics[width=4in]{data}
+% \includegraphics[width=4in]{Figure_3}
  \end{center}
 \caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
-sH1N1 IP-10 registered three values, not included in model fitting, above the measurement threshold of 8,500 $pg/mL$ (empty red triangles).
+sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.
 Dashed lines show model fits to IP-10 and RANTES data.
@@
 \begin{center}
-% \includegraphics[width=4in]{sensitivity}
+% \includegraphics[width=4in]{Figure_4}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=4in]{cycells}
+% \includegraphics[width=4in]{Figure_5}
  \end{center}
@@
 Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL.
-The individual cell images for days 5.5 and 7 have been enhanced by hand for improved T cell visibility.
-Individual groups of T cells appear larger than they are and only cover an area of one or two epithelial cells in the model.
+Individual cell images for days 5.5 and 7 were enhanced (green dots are enlarged) manually to improve T cell visibility.
+Thus, individual groups of T cells, which cover only an area of one or two epithelial cells in the model appear larger in the image.
 Original images available upon request.
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{plaquesize}
+% \includegraphics[width=\textwidth]{Figure_6}
  \end{center}
@@
 
-% In the simulation, increasing chemokine concentration fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[ht!]
+\begin{center}
+% \includegraphics[width=\textwidth]{Figure_S1}
+ \end{center}
+\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
+IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
+Each run was initialized identically for each strain save for the random seed.
+The middle line shows the average while the outer lines show the 95\% confidence interval.}
+ \label{fig:variance}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+%	\includegraphics[width=\textwidth]{Figure_S2}
+	\caption{Effects of different chemokine combinations.
+A) aH5N1 does not stimulate an IP-10 response.
+B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
+	\label{fig:chemokine}
+\end{center}
+\end{figure}
+
+\setcounter{figure}{0}
+\renewcommand{\figurename}{Video}
+
+\begin{figure}[ht!]
+\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
+T cells show half way through the simulation.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
+ \label{video:cell_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{The second of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the virus concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:virus_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{The third of three overlaid videos of a representative seasonal H1N1 infection.
+This video spans the 10 day infection and shows the chemokine concentration.
+Notice the volatility when T cells arrive halfway through the simulation.
+Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).
+Refer to Figure 5 for the detailed legend.
+}
+ \label{video:chemokine_view}
+\end{figure}
+
+\begin{figure}[ht!]
+\caption{A closer look at the 2009 pandemic simulation.
+This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
+Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
+Note the high proportion of virus-secreting cells (orange) early on.
+As time passes, secreting cells are gradually contained to the point where they become very sparse.
+T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
+\end{figure}
 
@@
   \hline
-  $T$ & Healthy target eptithelial cells & 1,000,000 \\
+  $T$ & Healthy target epithelial cells & 1,000,000 \\
   $I_1$ & Virus-incubating cells & 0 \\
@@
   $C$ & Chemokine quantity (ng/mL) & 0 \\
+  \hline
   \hline
@@
   \hline                        
-  Paramter & Value & Source \\
+  Referenced Parameters & Value & Source \\
   \hline
-  Viral Diffusion in Airway & $.0318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Viral Decay in Airway &  $1/day$ & \cite{Lee2009} \\
-  Chemokine Diffusion Rate & $.318 \mu m^2/s$ & \cite{Beauchemin2006} \\
-  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & Selected$^1$\\
-  Infection Sensitivity Rate &  $2 hour/virion$ &  Selected$^2$ \\
-  Incubation Time &  $10 hours$ & \cite{Mitchell2011} \\
-  Expression Time &  $16.7 hours$ & \cite{Mitchell2011}$^3$ \\
-  Apoptosis Time & $1 hour$ & \cite{Ganusov2008} \\
-  T cell Production Rate & $777/hour$ & \cite{Miao2010} \\ 
-  Blood Circulation Time & $6 min$ & \cite{Banerjee2010b} \\
-  Search Time In Chemokine-Free Tissue & $10 min$ & \cite{Banerjee2010b} \\
-  T Cell Speed (Search) & $30 \mu m/s$ & \cite{Miller2003} \\
-  T Cell Speed (Chemotaxis) & $3 \mu m/s$ & \cite{Miller2003} \\
-  T Cell Sensitivity to Chemokine & $10 ng/mL$ & \cite{Gao2003} \\
-  T Cell Expected Kill Time & $10 min$ & Selected$^4$ \\
-  Epithelial Cell Radius & $25 \mu m$ & Selected$^5$ \\
-  T Cell Age (at FOI$^\ast$) & $2 hours$ & Selected$^6$ \\
-  T Cell Age (in Blood) & $3 days$ & Selected$^6$ \\
-  Onset of ATC$^\ddagger$ Lymph Node Exit & $5 days$ & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of $3/day$ & Selected$^7$ \\
+  Viral Diffusion in Airway & .0318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
+  Viral Decay in Airway &  1/day & \cite{Lee2009} \\
+  Chemokine Diffusion Rate & .318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
+  Incubation Time &  10 hours & \cite{Mitchell2011} \\
+  T cell Production Rate & 777/hour & \cite{Miao2010} \\ 
+  Blood Circulation Time & 6 minutes & \cite{Banerjee2010b} \\
+  Search Time In Chemokine-Free Tissue & 10 minutes & \cite{Banerjee2010b} \\
+  T Cell Sensitivity to Chemokine & 10 ng/mL & \cite{Gao2003} \\
+  Onset of ATC$^\ddagger$ Lymph Node Exit & 5 days & \cite{Banerjee2011} \\
+  \hline
+  \hline                        
+  Estimated Parameters & Value & Footnote \\
+  \hline
+  Chemokine Decay Rate &  $3.8508\cdot10^{-4}$/s & 1\\
+  Infection Sensitivity Rate &  2 hour/virion &  2 \\
+  Expression Time &  16.7 hours & \cite{Mitchell2011}$^3$ \\
+  T Cell Expected Kill Time & 10 min & 4 \\
+  Apoptosis Time & 60 minutes & \cite{Ganusov2008}$^5$ \\
+  Epithelial Cell Radius & 25 $\mu$m & 6 \\
+  T Cell Age (at FOI$^\ast$) & 2 hours & 7 \\
+  T Cell Age (in Blood) & 3 days & 7 \\
+  T Cell Speed (Search) & 30 $\mu$m/s & \cite{Miller2003}$^8$ \\
+  T Cell Speed (Chemotaxis) & 3 $\mu$m/s & \cite{Miller2003}$^8$ \\
+  IgM Strength & Viral decay of 3/day & 9 \\
   \hline  
 \end{tabular}
-\caption{1) Corresponds to a 30 minute half-life.
+\caption{Values are taken from literature if possible, otherwise values are estimated as follows.
+1) Corresponds to a 30 minute half-life.
 2) Epithelial cells are infected at a probabilistic rate such that the expected time for infection in the presence of a single virion is 2 hours.
@@
 This rate does not scale with T cell numbers.
-5)  The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
-6) Parameters used after discussions with David Woodland, Trudeau Institute.
-7) IgM presence is abstracted by increasing viral decay by a factor of three.
-$^\ast$Focus of Infection.
-$^\ddagger$Activated T Cell.}
+5) Calculated for low T cell densities.
+6) The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
+7) Chosen to be at the lower end of biologically plausible values because increased T cell counts are shown not to affect the model behavior.
+8) The values were selected to be an order of magnitude faster than those calculated for inside the lymph node.
+9) IgM presence is abstracted by increasing viral decay by a factor of three.
+$^\ddagger$Activated T Cell.
+$^\ast$Focus of Infection.}
 \label{tab:parameters}

### 2013-01-10 paper/cover_letter.tex
Final submitted to PLoS CB

@@
 
-We submit this manuscript entitled "Spatially explicit model of the lymphocyte diaspora in influenza-infected lung quantifies constraints of chemokine directed migration" for consideration of publication in PLoS Computational Biology.
-We feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
-
-Many insights have derived from modeling the immune response using differential equations (ODEs) and delay differential equations (DDEs) to simulate linear and non-linear components.
-We have taken a different approach, using agent-based spatially-explicit modeling to examine the recruitment of antigen-specific T cells to influenza-infected foci in the lung during a primary immune response.
-ODE and DDE models do not take spatial effects into account, but given that the search occurs within a very large uninfected tissue space, we felt the efficiency of this search, depending largely on chemokine signaling of infected regions, needed to be carefully examined.
-We found that, similar to the highly efficient recruitment of precursor T cells to the lymph nodes, activated T cell search is indeed efficient, with one caveat applicable to large infected foci.
-We wish to stimulate other investigators to use agent-based modeling to carefully examine other aspects of the immune response in a spatially explicit manner, thus improving the comprehensive models of the immune response.
+We resubmit this manuscript entitled "Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration" for consideration of publication in PLoS Computational Biology.
+We would like to thank each of you for giving us the opportunity to improve our paper and resubmit it.
+We would like to reiterate our thanks to our reviewers for giving us very constructive criticism that we believe has greatly improved the paper.
+We still feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
 \\
@@
 
-All authors have read and approved the manuscript.
+In review we have noticed that our bibliography does not use annotated journal names, which seem to be the standard for PLoS CB.
+We have {\em not} changed our bibliography as of yet as it was generated automatically using the Mendeley Desktop suite.
+If this is an issue, we would be happy to return and reconstruct the references manually.
+\\
+
+All authors have read and approved the revised manuscript.
 

### 2014-01-27 paper/chemokine_main.tex
Plos ONE Submission

@@
 
+\usepackage{color}
+\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage{ulem}
+
+\definecolor{dkred}{rgb}{0.75,0,0}
+\definecolor{dkgreen}{rgb}{0,0.5,0}
+\definecolor{dkblue}{rgb}{0,0,0.75}
+\definecolor{dkpurple}{rgb}{.375,0,.375}
+\definecolor{gray}{rgb}{0.5,0.5,0.5}
+
+\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
+\newcommand{\drew}[1]{{\color{dkgreen}#1}}
+\newcommand{\fred}[1]{{\color{dkblue}#1}}
+\newcommand{\steph}[1]{{\color{dkpurple}#1}}
+
 %% END MACROS SECTION
@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
+\textbf{Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
 }
@@
 
-During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a fraction of the total lung.
-T cell navigation to infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
-We focused on the efficiency of local chemokines to induce migration of activated CD8 T cells using a spatially explicit agent-based model (ABM) without constructing a model containing all elements of the immune response.
+During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a small fraction of the total lung.
+T cell navigation towards infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
+Weuse a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signalling.
 
 To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.
-A differential equation model was fit to empirical chemokine production rates and coupled with published T cell parameters to calibrate the spatial model to test inter-strain variation on T cell recruitment in the lungs.
-The modeled immune response is unable to clear the pandemic strain due to its high rate of viral production.
+These data were modeled to estimate chemokine production rates.
+The estimates were combined in the spatial model to test viral variation of T cell recruitment in the lungs.
 
@@
 Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.
-A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in large infected foci.
-This spatially explicit model is consistent with efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.
+This spatially explicit model describes efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.
+A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.
+We conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.
 
@@
 
-Clearance of influenza from the lung depends strongly on the efficiency with which T cells travel from the lymph node to the site of infection.
+Clearance of influenza from the lung depends critically on the efficiency with which T cells travel from the lymph node to the site of infection.
 However, T cells in the lung must navigate the large branching bronchial network.
@@
 We developed an agent-based model to quantify how these molecular signals affect search.
-Our experimental data show that two chemokines, IP-10 and RANTES, are important for T cell search.
-We incorporated the data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.
+Our experimental data show that two chemokines, IP-10 and RANTES, are secreted by infected human tracheal epithelial cells.
+We incorporated this data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.
 We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.
-The simulated immune response contains the first two infections but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.
-Direct interventions to improve this T cell search process are unlikely to be effective, and therefore vaccination is required to contain more virulent strains.
+The simulated immune response controls and eradicates the first two infections in 10 days but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.
+Interventions are unlikely to improve this chemokine-directed T cell search process in the lung but vaccination against more virulent strains will increase the number of searching cells earlier in the response.
 
@@
 
-The adaptive immune response induced during acute influenza A infection is a complex combination of defense mechanisms.
-Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control immunopathology.
+The adaptive immune response induced during acute influenza A infection is an integrated complex of combination of defense mechanisms.
+Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control the infection and immunopathology.
 The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.
@@
 
-The process of recruitment of activated CD8 effector cells to infected sites in the lung is not completely understood.
-Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into tissue \cite{Topham2001}.
-Effector cells must localize to sites of viral replication [4], but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
-Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005} where recruited T cells mediate either protection or immunopathology.
+CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in infected lung.
+Search problems in the lymph node have been simulated using live cell imaging data to provides reliable parameters of cell movement \cite{Vroomans2012}.
+The T precursor cell must encounter on average hundreds of irrelevant cells before it contacts the loaded DC, but the distance traveled is relatively short, and the direction is likely guided by chemokine signals.
+
+The second search, i.e.
+the process of recruiting activated CD8 effector cells to infected sites in the lung,  is different and not as well understood.
+In contrast to the dense lymph node volume, lung infection dynamics occur on the surface of epithelial cells, the size of the tissue layer being comparable to the size of a tennis court.
+Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
+Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.
+Effector cells must localize to sites of viral replication \cite{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
+Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005} where recruited T cells mediate either protection or immunopathology.
 Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
@@
 We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.
-Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, tertiary lymphoid structures, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
+Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
 The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.
-We use chemotactic protein data from in vitro cultures of bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} previously described in our laboratory.
-Using known parameters from the extensive literature on the murine model, we offer several insights on the control of viral replication in lung tissue derived from spatially explicit simulations.
+
+Previously we developed an agent-based model of the lymphocyte diaspora focusing on the production and diffusion of a generic inflammatory signal without experimental data to calibrate the model \cite{Banerjee2011}.
+Here we refine that model and incorporate chemotactic protein data from in vitro cultures of human bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} to focus on two features of searching in the large lung space.
+First, functional thresholds of the chemotactic proteins were surprisingly low in our model.
+Second there appears to be a threshold size of the infected focus above which the T cell search becomes inefficient.
 
@@
 
-\subsection*{Delay Differential Equation Model}
-
-We estimate chemokine production rates by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} by adding one new equation to model chemokine production.
-Initial population sizes and parameter values are taken from the previous study.
-{\footnotesize
-\begin{equation}
-\begin{aligned}
-\dot{T} &= - \beta T V \\
-\dot{I_1} &= \beta T V - \beta T[t-\tau_1]V[t-\tau_1] \\
-\dot{I_2} &= \beta T[t-\tau_1]V[t-\tau_1] - \delta I_2 \\
-\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
-\dot{F} &=  I_1[t-\tau_2] \\
-\dot{C} &= r I[t-\tau_3] - d C \\
-\end{aligned}
-\label{eq:dde}
-\end{equation}
-\vspace{.05in}
-}
-
-Table \ref{tab:dde} summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table \ref{tab:strains}).
-
 \subsection*{Computational Modeling}
@@
 Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}), and simulations measured efficiency of infection clearance under different environmental conditions.
-The lung was represented as a two-dimensional sheet of healthy epithelial cells.
-Vasculature was represented as a binary tree with fourteen levels of bifurcations with the origin at the bronchial lymph node complex.
-Activated T cells descend through the vascular tree until cytokine signal is detected on the local endothelium, at which point they exit the vasculature and follow the chemotactic gradient to the site of infection.
+A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart} and \ref{fig:variance}), and simulations measured efficiency of infection clearance for three different viruses.
+The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.4).
+Vasculature in a mouse is a binary tree with fourteen levels of bifurcation with the origin at the bronchial lymph node complex.
+Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).
+If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection.
 T cells that do not encounter cytokine recirculate to the lymph node.
 When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.
+Unimpeded, the infected secretes virus for 16.7 hours and then dies.
 
 The simulation begins when a single cell in the center of the tissue is infected.
-After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type.
+After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S1.
 Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region.
-After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node and circulate through the vasculature to the tissue.
-Because T  cells cannot choose their path through the branching network, we assume they arrive in the capillary bed at random locations.
+Chemokine diffuses from secreting cells according to kinetics described below, creating a circle of stimulation around the infected region.
+After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin to exit the lymph node at a steady rate and circulate through the vasculature to the tissue.
+Further details of the CyCells modeling environment are described in S1.3.
+
+Computational limitations force us to implement our model on the scale of a mouse.
+Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
+The surface area of a mouse lung is approximately 100$cm^2$.
 
@@
 
-In the model, epithelial cells are stationary and can be in one of five different states: \emph{healthy}, \emph{virus-incubating}, \emph{virus-expressing}, \emph{apoptotic}, and \emph{dead}.
-\emph{Healthy} cells remain unchanged unless infected by virus.
-Once infected, the cell transitions from {incubating} to \emph{expressing}.
-\emph{Expressing} cells secrete virus and chemokine for approximately 17 hours and then die.
-\emph{Expressing} cells become \emph{apoptotic} if they encounter activated T cells.
-Apoptoic cells continue to secrete virus until they die one hour after their transition.
-\emph{Dead} cells remain inert and do not regenerate over the course of an infection.
+In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \textit{healthy}, \textit{virus-incubating}, \textit{virus-expressing}, \textit{apoptotic}, and \textit{dead}.
+\textit{Healthy} cells remain unchanged unless infected by virus.
+Once infected, the cell transitions from \textit{incubating} to \textit{expressing} after a 10 hour incubation delay.
+\textit{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
+\textit{Expressing} cells initiate \textit{apoptosis} sooner if they are contacted by activated T cells.
+Apoptoic cells continue to secrete virus and chemokine and then die after one hour.
+\textit{Dead} cells take up space and do not regenerate over the course of an infection.
 
-T cells have three states.
-\emph{Circulating} T cells begin to emerge from the lymph node at five days post infection.
-\textit{Emigrating} T cells arrive at a random location on the lung's surface, wander randomly in the tissue for 10 minutes, and transition to \emph{circulating} in the absence of chemokine.
-\emph{Circulating} cells spend six minutes recirculating to the lymph node, transition to \emph{emigrating} and are reintroduced to a new random location in the lung.
-When a \emph{emigrating} T cell encounters chemokine, it converts to \emph{chemotaxing} and begins following the chemotactic gradient to the source of infection.
-\emph{Chemotaxing} T cells move continuously up the gradient, inducing apoptosis if they encounter \emph{expressing} epithelial cells.
-\emph{Chemotaxing} cells decay exponentially with an average lifespan of two hours.
+T cells are described by two states.
+\textit{Circulating} T cells begin to emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.
+If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.
+If a circulating cell encounters chemokine, it converts to \textit{chemotaxing} and begins following the chemotactic gradient to the infected region.
+\textit{Circulating} T cells decay exponentially with an average lifespan of three days.
+\textit{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells.
+\textit{Chemotaxing} cells decay exponentially with an average lifespan of two hours.
 
 The model contains two kinds of particles: virus and chemokine.
-Both are produced at constant rates by \emph{expressing} epithelial cells.
-Virus diffuses through the lung tissue, infecting healthy cells probabilistically according to the local virus concentration.
+Both are produced at constant rates by \textit{expressing} epithelial cells.
+Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration.
 Chemokine diffuses across the tissue but has no direct effect beyond activating T cells.
 Both particle types decay exponentially.
-IgM is modeled by increasing the viral decay rate by a factor of three after the third day.
+IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.
 
-Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.  Strain-specific values are shown in Table \ref{tab:strains}.
+Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.
+Strain-specific values are shown in Table \ref{tab:strains}.
+Further details regarding the model definition are included in S1.4.
 
@@
 Quantitative viral culture was performed by standard plaque assay.
-Quantitative chemokine levels were performed in 30 µl aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
+Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
 Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.
-TNF was ignored as its effects are outside the scope of this paper.
+TNF activity as not incorporated in the model.
+Data for other chemokines and cytokines is not shown.
 
@@
 IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
-To estimate per-cell production rates, we extended the ordinary differential equation (ODE) model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~\ref{eq:dde}).
-Model fits (Table \ref{tab:strains}) were computed for three strains (Fig.~\ref{fig:data}) using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
-The resulting chemokine production values were used in the CyCells ABM.
+To estimate per-cell production rates, we extended the delay differential equation (DDE) model of Ref.
+\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~S1).
+Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
 Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
-There is no positive correlation between viral production and induced chemokine production across the three strains.
+There was no positive correlation between viral production and induced chemokine production across the three strains.
 
-\subsection*{T cell sensitivity to chemokine}
+\subsection*{Stochastic modeling effects}
 
-The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table~\ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table~\ref{tab:parameters}).
-T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.
+Unlike ordinary differential equation (ODE) and DDE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~\ref{fig:variance}).
+To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.
+$R^2$ values were calculated for individual runs versus the average of all runs.
+The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: (0.957 $\pm$ 0.150), (0.992 $\pm$ 0.013), and (0.998 $\pm$ .002) respectively (i.e.
+there was little variation between runs).
+The three strains show different levels of replication as found in \cite{Mitchell2011}.
 
-Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude and found a threshold (Fig.~\ref{fig:sensitivity}) at a concentration of 100 $ng/ml$ (10 $ng/ml$ in aH5N1), beyond which there is no detectable effect on model behavior (model variance is discussed in Text S1 2.1.).
-We then set the sensitivity to 10 $ng/mL$ for all future runs (1 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
+Overlaying multiple runs on a single plot reveals a growth rate decline at 4 days post infection, reflecting the addition of IgM.
+Subsequently, for each infection the number of infected cells declines quickly after day five post infection due to the CD8 T cell response.
+aH5N1 is cleared completely, sH1N1 is more slowly cleared, and pH1N1 clearance at day 6 post infection is halted as replication overtakes clearance.
 
-% Because more than one T cell at each infected epithelial cell would not increasing effective target cell death, there may be a threshold of immigrant T cells above which no additional control of viral replication is possible.
+The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).
+Because pH1N1 replicates more rapidly, the number of infected cells outstrips the ability of the T cells to find them.
+Spatial effects reveal the reasons for this in simulation images and video (Fig.~\ref{fig:cycells}, Video S4).
 
@@
 
-Spatial effects of viral and chemokine diffusion play an important role in both the spreading and clearing of infections.
+Spatial effects of viral replication and chemokine diffusion play an important role in both the spreading and clearing of infections, imaged as plaque growth over time (Fig.~\ref{fig:cycells}).
 Free virus particles diffuse from virus secreting cells and infect healthy cells.
 Chemokine produced by infected cells attracts T cells to the infected cells.
-Although virus is produced at a higher rate than chemokine, its larger size diffuses much more slowly, while chemokine decays more quickly.
-These countervaling effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
-Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas dead cells are rare.
+Although larger size of virus particles causes virus to diffuse much more slowly than chemokine molecules, chemokine decays more quickly.
+These countervailing effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
+Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas there is a smaller percentage of dead cells.
 Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque.
@@
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-Refer to the Videos S1-S3 to see a representative simulation of seasonal influenza.
+Continuous change in plaque size and T cell number is demonstrated for seasonal H1N1 virus (Videos S1-S3).
 
-In aH5N1, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
-However, secreting cells were not eliminated in either H1N1 simulations (Fig.
-S1).
-Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i..
-T cells still accumulate, but they arrive more slowly than the plaque is growing, leading to lower average T cell killing rate.
-Further, the regions of concentrated chemokine lag behind the cell and virus spatial layout.
+In the aH5N1 infection, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
+However, secreting cells were not eliminated in either sH1N1 or pH1N1 simulations (Fig.~\ref{fig:variance}).
+Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i.
+(Fig.~\ref{fig:plaquesize}).
+T cells still accumulate, but they arrive at a lower rate than the number of infected cells that are added to the plaque.
+Further, the regions of concentrated chemokine diffuse slowly, causing them to be located behind the expanding infected cell and virus locations.
 It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.
-Thus, T cells can fail to detect cellular changes in the plaque.
+Thus, T cells can fail to locate new regions of infected cells within the plaque.
 Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.
-Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.
+Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate increased chemokine production rates.
 
-T cells are unable to kill cells that have not yet presented antigen.
-At day 5.5, the ratio of secreting cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
+\subsection*{Temporal effects}
+
+The notion that T cell can't find infected cells in large plaques was examined in more detail by comparing the ratios of T cells to infected cells in Fig.~\ref{fig:plaquesize}.
+T cells are unable to kill virus incubating epithelial cells that have not yet presented antigen.
+At day 5.5, the ratio of antigen presenting secreting epithelial cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
 By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.
-The high replication rate of pH1N1 enhances this effect (Fig.~\ref{fig:plaquesize}C) and the T cells can control (but not eliminate) the sH1N1 infection.
+The high replication rate of pH1N1 exaggerates this effect (Fig.~\ref{fig:plaquesize}C).
 Video S4 illustrates this event.
 
-A cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
-That is, in each secreting cell a new viral particle is produced approximately every 200 seconds.
+Because there is an interval between first secreted virus and the first arriving T cell, infection control cannot be completed by CD8 T cells alone.
+An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
 Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.
 Thus, T cells alone cannot contain the pH1N1 infection.
-In contrast, a sH1N1 virus-secreting cell produces a new virus particle every 2,632 seconds, allowing T cells to limit a single infected cell to 1.4 viral particles in the one-hour window.
+In contrast, in a sH1N1 virus infection, T cells would limit a single infected cell to producting 1.4 viral particles in the one-hour window.
 Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.
+Viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
 
 \section*{Discussion}
+
+Our aim was to examine the spatially determined restrictions on the efficiency of the T cell search.
+Due to the absence of critical data, the inability to image the search in the lung (as was done in the lymph node), and the need for a tractable computation, the model simplified the adaptive immune response.
+We focused instead on comparing the behavior of thee influenza viruses.
 
@@
 The model makes a number of simplifications to and deletions of elements in the innate and adaptive immune response, allowing us to build a tractable model where data was available.
-Antigen presentation T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.
+Antigen presentation and T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.
 Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
@@
 The actual quantitative chemokine environment in the infected lung is difficult to estimate for a math model.
-The hypercytokinemia documented in virulent influenza infections \cite{DeJong2006} likely does not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells in vitro depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate actual levels.
-Interestingly, the attenuation of the type I interferon response by H5N1 strains is not associated with attenuation of chemokine secretion \cite{Zeng2007}.
+Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells in vitro depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
 The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
@@
 The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 10 ng/mL and maximal levels of concentration in tissue of 10,000 pg/mL, we calculated the effective communication distance to be approximately 150 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
+For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
 This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-Spatially explicit modeling can provide support for not only communication distance but also the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
-% - removed - Can T cell recruitment to tissue during the primary immune response be manipulated to increase search efficiency, with particular implications for recovery of people prior vaccination?
-The high sensitivity to chemokine signals and lack of increased efficiency with an added chemokine do not appear to be avenues for manipulation (Text S1 2.2).
-Increasing inflammatory cells may raise the volume of tissue containing chemokine signal, thus amplifying T cell search success, but comes with the price of greater morbidity.
-T cell recruitment to regional lymph nodes is highly efficient, and numbers of responding clones is restricted by immunological dominance.
-We conclude that, with respect to the elements of recruitment studied in this spatially explicit model, recruitment is efficient and offers few if any options for manipulating its efficiency.
-Only vaccination provides the clonal expansion and tissue resident memory cells to accelerate the appearance of activated T cells in tissue.
+Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
@@
 Stelle (UNM Computer Science) for suggestions concerning the manuscript, cellular and chemokine behavior, and agent-based modeling.
+
+This publication was funded by NIH grants 1R21-AI-73607 (to F.K.), U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F.), and NSF grant EF 1038682 (to M.M.).
 
@@
 sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.
-Dashed lines show model fits to IP-10 and RANTES data.
-Human bronchial epithelial cells were infected at an MOI of 0.01 with one of the three strains of influenza.
+An extended differential equation model from \cite{Mitchell2011} (see Materials and Methods) was fit to IP-10 and RANTES data.
+These fits were used to obtain chemokine production values for use in the spatial CyCells model.
+Human bronchial epithelial cells were infected at an MOI of 0.01 (10,000 virions) with one of the three strains of influenza.
 Apical fluid for viral secretion and basal media for chemokine secretion was collected at the given time intervals post infection.
-Viral culture was performed by a standard plaque assay and chemokine levels were measured using 30 µl aliots for a panel of 17 chemokines and cytokines (not shown).}
+Viral culture was performed by a standard plaque assay and chemokine levels were measured using 30 $\mu$l aliots for a panel of 17 chemokines and cytokines (not shown).}
  \label{fig:data}
@@
 
-\begin{figure}[!ht]
+\begin{figure}[ht!]
 \begin{center}
-% \includegraphics[width=4in]{Figure_4}
+% \includegraphics[width=\textwidth]{Figure_4}
  \end{center}
-\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
-Total number of incubating, secreting and apoptotic cells are plotted for each infection.
-The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it.
-}
- \label{fig:sensitivity}
+\caption{\textbf{Model results:} Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
+Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input (Table \ref{tab:strains}) and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
+Therefore the figures approximate the rate of plaque growth over time.
+IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
+Each run was initialized identically for each strain save for the random seed.
+The middle line shows the average while the outer lines show the 95\% confidence interval.}
+ \label{fig:variance}
 \end{figure}
@@
 Free virus and chemokine particles are represented by compartmentalized concentrations of mols/mL and ng/mL.
-Individual cell images for days 5.5 and 7 were enhanced (green dots are enlarged) manually to improve T cell visibility.
-Thus, individual groups of T cells, which cover only an area of one or two epithelial cells in the model appear larger in the image.
-Original images available upon request.
-}
+Chemokine shown is an aggregate of total IP-10 and RANTES concentrations.
+See Videos S1-S3 for an animated visualization of each row.}
  \label{fig:cycells}
@@
  \end{center}
-\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1.} Plotted values: total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), and T cells (red).
+\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1.} Plotted values: total plaque size (blue), number of virus incubating cells (yellow), number of virus secreting cells (green), total number of T cells (black), and T cells at the focus of infection (FOI) (red).
 T cells clear secreting and incubating cells in aH5N1, fail to clear incubating cells in sH1N1, and fail to clear either type of infected cell in pH1N1.
-The number of expressing cells (green) after day 5 differs markedly among the three strains indicating that the T cells have differing success at controlling the infection.}
+The number of incubating cells (yellow) after day 5 differs markedly among the three strains indicating that the T cells have differing success at controlling the infection.}
  \label{fig:plaquesize}
@@
  \end{center}
-\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
-IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-Each run was initialized identically for each strain save for the random seed.
-The middle line shows the average while the outer lines show the 95\% confidence interval.}
- \label{fig:variance}
+\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+Total number of incubating, secreting and apoptotic cells are plotted for each infection.
+The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it.}
 \end{figure}
 
-\begin{figure}[ht!]
+\begin{figure}[!ht]
 \begin{center}
-%	\includegraphics[width=\textwidth]{Figure_S2}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
+% \includegraphics[width=4in]{Figure_S2}
+ \end{center}
+\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
 B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
+ \label{fig:sensitivity}
 \end{figure}
@@
 \caption{A closer look at the 2009 pandemic simulation.
-This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
+This video shows the infection from day 6 to day 7 with each frame spanning 60 simulated seconds.
 Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
@@
 \caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
-The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were take from the observed chemokine data (Fig.~\ref{fig:data}).
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
 Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
@@
   Incubation Time &  10 hours & \cite{Mitchell2011} \\
-  T cell Production Rate & 777/hour & \cite{Miao2010} \\ 
-  Blood Circulation Time & 6 minutes & \cite{Banerjee2010b} \\
-  Search Time In Chemokine-Free Tissue & 10 minutes & \cite{Banerjee2010b} \\
-  T Cell Sensitivity to Chemokine & 10 ng/mL & \cite{Gao2003} \\
-  Onset of ATC$^\ddagger$ Lymph Node Exit & 5 days & \cite{Banerjee2011} \\
+  Epithelial Cell Radius & 5 $\mu$m & \cite{Elbert1999} \\
+  T cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
+  T cell Production Rate & 1257/hour & \cite{Miao2010} \\ 
+  T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
+  Blood Circulation Time & 6 seconds & \cite{Banerjee2010b} \\
+  T Cell Sensitivity to Chemokine & 100 ng/mL & \cite{Nandagopal2011} \\
+  Onset of ATC$^\ddagger$ Lymph Node Exit & Day 5 & \cite{Banerjee2011} \\
+  IgM Strength & Viral decay of 10/day & \cite{Diamond2003} \\
+  IgM Onset & Day 4 & \cite{Diamond2003} \\
   \hline
@@
   Apoptosis Time & 60 minutes & \cite{Ganusov2008}$^5$ \\
-  Epithelial Cell Radius & 25 $\mu$m & 6 \\
-  T Cell Age (at FOI$^\ast$) & 2 hours & 7 \\
-  T Cell Age (in Blood) & 3 days & 7 \\
-  T Cell Speed (Search) & 30 $\mu$m/s & \cite{Miller2003}$^8$ \\
-  T Cell Speed (Chemotaxis) & 3 $\mu$m/s & \cite{Miller2003}$^8$ \\
-  IgM Strength & Viral decay of 3/day & 9 \\
+  T Cell Age (at FOI$^\ast$) & 2 hours & 6 \\
+  T Cell Age (in Blood) & 3 days & 6 \\
   \hline  
 \end{tabular}
-\caption{Values are taken from literature if possible, otherwise values are estimated as follows.
+\caption{Values are taken from experimental literature if possible and from earlier modeling papers if not.
+If we could not find a value in any literature we estimated parameters as follows.
 1) Corresponds to a 30 minute half-life.
@@
 5) Calculated for low T cell densities.
-6) The mean surface area of the epithelial cell available for virus contact and entry includes cilia and the radius is estimated to be 25 microns.
-7) Chosen to be at the lower end of biologically plausible values because increased T cell counts are shown not to affect the model behavior.
-8) The values were selected to be an order of magnitude faster than those calculated for inside the lymph node.
-9) IgM presence is abstracted by increasing viral decay by a factor of three.
-$^\ddagger$Activated T Cell.
-$^\ast$Focus of Infection.}
+6) Chosen to be at the lower end of biologically plausible values because increased T cell counts are shown not to affect the model behavior.
+}
 \label{tab:parameters}
@@
 \end{table}
+% 7) IgM presence is abstracted by increasing viral decay by a factor of ten. $^\ddagger$Activated T Cell.  $^\ast$Focus of Infection.
 
@@
 Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
-Viral production values are taken from \cite{Mitchell2011}.}
+Bootstrapping for the chemokine values was performed using the original fit of Eq.
+S1 to the data in Fig.~\ref{fig:data} to produce new data sets.
+Viral production values and confidence intervals are taken from \cite{Mitchell2011}.}
 \label{tab:strains}

### 2014-01-27 paper/chemokine_supplement.tex
Plos ONE Submission

@@
 
-\subsection{T Cell Production Rate}
+\subsection{Estimating chemokine production rates}
 
-To find a value for $\sigma$, the rate of production of activated T cells (ATCs) we use a differential equation model from \cite{Miao2010}:
+We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} \drew{Eq.
+1} by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
+Initial population sizes and parameter values are taken from the previous study.
+{\footnotesize
+\begin{equation}
+\begin{aligned}
+\dot{T} &= - \beta T V \\
+\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
+\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
+\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
+\dot{F} &=  I_{1 \tau_2} \\
+\dot{C} &= r I_{1 \tau_3} - d C \\
+\end{aligned}
+\label{eq:dde}
+\end{equation}
+\vspace{.05in}
+}
+
+$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
+Table 1 summarizes population and parameter values and descriptions.
+Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 3).
+
+\subsection{T cell production rate}
+
+We calculate the rate of production ($\sigma$) of CD8 T cells using a differential equation model from \cite{Miao2010}.
+The equations model T cell production and subsequent search over an area of infected lung tissue.
 
@@
 The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
+$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
 
-Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
-These circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$) at a given speed, $v_{tcell}$.
+Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
+We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
+
+The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time.
 Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
@@
 The initial number of infected cells is assumed to be zero.
+Parameter values are listed in Table S1 in Text S1.
 This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
-The final value was found to be 777 per hour.
+The final value was found to be 1,257 per hour.
 
-\subsection{CyCells Implementation Details}
+\subsection{CyCells implementation details}
 
@@
 Cells are considered to be unique agents and are modeled individually.
-On the other hand, molecules are considered to be too small and numerous to model explicitly and are thus represented as compartmentalized concentrations.
+Conversely, molecules are represented as compartmentalized concentrations.
 These compartments are arranged as a grid that covers the environment.
 Each square in the grid contains a unique concentration of the given molecule.
-Multiple molecules are represented in separate grids.
+Each type of  molecule is represented in a unique grid.
 
@@
 In this scenario CyCells calculates the amount of virus secreted per time step, based upon the cell's defined production rate, and then adds this quantity to the grid square that overlaps the secreting cell.
-If a cell wants to remove molecular concentration, it will subtract from the square it overlaps.
-If the concentration at that square is not enough, concentration will be removed from neighboring squares in an ever expanding diamond.
+If a cell ingests a molecule, it will subtract the appropriate concentration from the grid square it overlaps.
+If the concentration at that square is not enough to represent a full molecule, concentration will be removed from neighboring squares in an ever expanding diamond until the total concentration is equal to a single molecule.
 
-When a cell's behavior depends on a local molecular concentration, the cell will measure the local concentration at its position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell.
+Cells will measure the local concentration at their position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell in instances where cell behavior depends on the local molecular concentration.
 
-\subsection{Model Implementation Details}
+\subsection{Model implementation details}
 
@@
 2.
-The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (46,085 total cells).
-The sheet measures 10$mm$ per side, which is large enough to contain the infection (Fig.
+The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (288,212 total cells).
+The sheet measures 5$mm$ per side, which is large enough to contain the infection (Fig.
 5).
-The molecular grid (S1.2) is initialized with a resolution of 50$\mu m$.
+The molecular grid (Text S1 1.3) is initialized with a resolution of 20$\mu m$.
+Each time step of the model represents 6 seconds.
 
@@
 Virus-secreting cells secrete both virus and chemokine (the RANTES portion of the chemokine production is added in 16 hours after the initial infection) and die after 1,000 minutes of production (16.7 hours).
-Virus-secreting cells will also probabilistically transition to apoptizing cells in the presence of T cells (defined by the existence of a T cell within 5$\mu m$).
+Virus-secreting cells will also probabilistically transition to apoptizing cells in the presence of T cells (defined by the existence of a T cell within 2$\mu m$).
 The number of T cells in the vicinity of the virus-secreting cell has no effect on the rate that apoptosis is induced in the model.
 
-Apoptizing cells continue to secrete virus and chemokine for one hour and then die.
+Apoptizing cells secrete virus and chemokine for one hour and then die.
 
-T cells are added to the model at a constant rate of 777 per hour (S1.1) through a global `oracle' cell that represents the lymph node.
-Because the model environment is only a small portion of the entire lung (1\%), most of the T cells miss the model window and are not visually represented.
-The T cells that miss enter a dormant searching state and transition to recirculating after a 10 minute delay.
-The cells that enter the model window are placed at a random position and begin a random walk.
-At each step, these searching cells check the local chemokine concentration.
+T cells are added to the model at a constant rate of 1,257 per hour (Text S1 1.1).
+Because the model environment is only a small portion of the entire lung (0.25\%), most of the T cells miss the model window and are not visually represented.
+The T cells that miss recirculate and enter the lung at a new random location after a delay of six seconds.
+
+The cells that enter the model window are placed at a random position and immediately check the local chemokine concentration.
 If the concentration is above the sensitivity threshold, the T cells immediately transition to the chemotaxing state.
-If not, searching T cells transition to the recirculating state and are removed from the model window.
-Searching and recirculating T cells have a probabilistic decay rate that corresponds to a 3 day lifespan.
+If not, the T cells recirculate and reenter the lung in a new location after a six second delay.
+Recirculating T cells have a probabilistic decay rate that corresponds to a 3 day lifespan.
+Recirculating T cells are not visually represented.
 
-Recirculating T cells are not visually represented.
-These cells are individually tracked for 6 minutes, after which they are reintroduced as new searching cells with a new chance of entering the model window.
-
-Chemotaxing T cells move directly up the chemokine gradient until they find a local maximum.
+Chemotaxing T cells move directly up the chemokine gradient until they find a local concentration maximum.
 T cells probabilistically decay at a rate corresponding to a 2 hour lifespan.
 Chemotaxing T cells have no effect other than inducing apoptosis in virus-secreting cells by proximity.
+
+\subsection{Modeling Decisions}
+
+\subsubsection{Two dimensional lung}
+
+We model the lung as a two-dimensional system for several reasons.
+Unlike the lymph node where dendritic cells and T cells navigate a three-dimensional volume, lung infection dynamics are confined to a thin tissue between capillary endothelial cells and alveolar epithelial cells.
+Although these alveoli are segregated by the lung's acinar structure on a small scale, we do not represent this in the model because the spreading of influenza eventually ignores the boundaries between acinii.
+Therefore our model does not incorporate this small-scale level of segregation.
+
+\subsubsection{Uniform blood flow}
+
+Our model assumes that blood flows uniformly through the lung vascular network.
+It is possible that blood flow is increased in the direction of an infected region of the lung by local inflammation.
+Thus our model may underestimate the recruitment efficiency of local inflammation and hyperemia.
+If this is indeed the case, our model underestimates the effect of the immune response.
+
+%\subsubsection{T cell velocity}
+%\fred{T cell velocity in the lung is extremely difficult to measure.} \removed{Little is known of T cell chemotaxis speed in the lung.}  We give chemotaxing T cells a velocity of $3\mu m/s$ and T cells that are performing a random walk a velocity of $30\mu m/s$.
+\removed{These values are faster that speeds found in other organs, such as the lymph node.
+This is reasonable because T cells in the lung travel on top of a packed surface without needing to navigate through blood flow.
+Furthermore, T cell speed has little effect on the model behavior as detailed below (S2.3).}  \fred{These rapid values estimate velocity of capillary blood flow in the extensive network of capillaries beyond the 14th bifurcation of the lung arterial system.
+T cell velocity would be expected to be slower within interstitial tissue, but this was not modeled in most simulations.
+In runs where velocity was varied over 5 orders of magnitude, including slow velocities measured in lymph node and skin by other authors, T cell velocity had little effect on model behavior and infection control as shown below (S2.3).}
+%
+
+%See Table~\ref{table:assumptions}.
 
@@
 
-\subsection{Stochastic modeling effects}
+\subsection*{T cell sensitivity to chemokine}
 
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~S\ref{fig:variance}).
-To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.
-$R^2$ values were calculated for individual runs versus the average of all runs.
-Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.
-sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.
-aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.
+The model simulates a chemokine gradient surrounding the infected focus (Fig.
+5), based on the calculated per-infected cell secretion rate (Table 3) and known chemical parameters for a 10 kDa protein (Table 2).
+T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.
 
-Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
-Therefore the figures approximate plaque growth over time.
+Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude.
+Concentration-dependent behavior of cells responding to chemokine was the same over a wide range of concentrations from 0.01 pg/mL to 100 ng/mL (model variance is discussed in Text S1 2.1.).
+Decreased recruitment was predicted only by supra-natural concentration sensitivity levels at and above 1 $\mu$g/mL.
+We then set the sensitivity to 100 $ng/mL$ for all future runs (10 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
 
-Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the addition of IgM.
-In addition, for each infection the number of infected cells declines quickly at day five due to the T cell response.
-
-The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~S\ref{fig:variance}).
-The rapid production of the pandemic H1N1 virus prevents the immune response from containing the infection.
-Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T Cells to gain ground is too small.
-aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH1N1 recovers and continues to expand.
+% Because more than one T cell at each infected epithelial cell would not increasing effective target cell death, there may be a threshold of immigrant T cells above which no additional control of viral replication is possible.
 
@@
 This suggests that RANTES does not play a significant role in infections that stimulate an IP-10 response due to the higher production rates of IP-10.
+
+%\drew{
+%\subsection{T cell speed}
+%
+%\removed{T cell speed over bronchial epithelial cells is unknown.}\fred{T cell velocity within the capillary network and after diapedesis with the lung interstitial tissue is unknown.}  We tested values for T cell speed to measure the effect it had on the general infection dynamics (Fig.~S\ref{fig:speed}).
+The results show that the difference in T cell speeds is small, although there does appear to be an discernible effect on the infection during \removed{the early phase of the T cell response} \fred{the initial appearance of activated T cell migration between days 5 and 7 post infection.}.
+%}
 
@@
 
-\begin{figure}[ht!]
+\setcounter{figure}{0}
+\renewcommand{\thefigure}{S\arabic{figure}}
+
+\begin{figure}[!ht]
 \begin{center}
-% \includegraphics[width=\textwidth]{variance}
+% \includegraphics[width=\textwidth]{Figure_S1}
  \end{center}
-\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
-IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-Each run was initialized identically for each strain save for the random seed.
-The middle line shows the average while the outer lines show the 95\% confidence interval.}
- \label{fig:variance}
+\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+Total number of incubating, secreting and apoptotic cells are plotted for each infection.
+The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it.
+}
+ \label{fig:sensitivity}
 \end{figure}
@@
 \begin{center}
-%	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
+%	\includegraphics[width=\textwidth]{Figure_S2}
+	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
 B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
@@
 \end{figure}
+
+%\begin{figure}[ht!]
+%\begin{center}
+%	\includegraphics[width=\textwidth]{Figure_S3}
+%	\caption{\drew{\textbf{Effects of different T cell speeds.}  With the exception of sH1N1, T cell speeds show a small effect during the early phase of the T cell response.
+After a time of approximately two days, the infection dynamics converge to a similar rate of growth across all speed values.}}
+%	\label{fig:speed}
+%\end{center}
+%\end{figure}
+
+\setcounter{figure}{0}
+\renewcommand{\figurename}{Video}
 
@@
 \caption{A closer look at the 2009 pandemic simulation.
-This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
+This video shows the infection from day 6 to day 7 with each frame spanning 1 simulated minute.
 Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
@@
   V(0) &  $1e4$ & \cite{Mitchell2011} \\
-  $r_{cell}$ &  $10 \mu m$ & \cite{Miao2010} \\
+  $r_{cell}$ &  $5 \mu m$ & \cite{Miao2010} \\
   $t_{rc}$ & $6s$ & \cite{Peters1983} \\
-  $v_{tcell}$ & $0.3 \mu m/s$ & \cite{Miller2003} \\
   $k_e$ & $6.4e-5/cell/day$ & \cite{Miao2010} \\
@@
   $\delta$ & $16.7 hours$ & \cite{Mitchell2011} \\
-  $a$ & $20.9$ & ABM fits \\
-  $b$ & $70.8$ & ABM fits \\
+  $a$ & $67.7$ & ABM fits \\
+  $b$ & $433.5$ & ABM fits \\
   \hline  
@@
 
+%\begin{table}[!ht]
+%\begin{center}
+%\begin{tabular}{ | c | c | }
+%  \hline                        
+%  Assumption & Description \\
+%  \hline
+%  2D hexagonal layout & The lung can be stretched flat.  T cells travel on the cellular surface. \\
+%  Even blood flow to all capillaries & We assume that there is negligible increased blood flow to the infected region. \\
+%  Rapid T cell circulation & Activated T cells in circulation do not spend time in new lymph nodes as they circulate. \\
+%  T cell speed & T cells move at a rate of XXX in the absence of chemokine and XXXX once it detects chemokine. \\
+%  T cell search time & T cells search in the lung for 6 minutes before they recirculate. \\
+%  \hline  
+%\end{tabular}
+%\caption{Supplemental model parameters.
+Most values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
+Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
+%\label{table:assumptions}
+%\end{center}
+%\end{table}
+
 \end{document}

### 2014-01-27 paper/cover_letter.tex
Plos ONE Submission

@@
 
-We resubmit this manuscript entitled "Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration" for consideration of publication in PLoS Computational Biology.
-We would like to thank each of you for giving us the opportunity to improve our paper and resubmit it.
-We would like to reiterate our thanks to our reviewers for giving us very constructive criticism that we believe has greatly improved the paper.
-We still feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
+We submit this manuscript entitled "Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration" for consideration of publication in PLOS ONE.
+Our paper implements an agent-based model of bronchial influenza infection and the subsequent T cell response using new empirical values of chemokine secretion.
+The paper shows that chemokine levels have a limited yet important role in directing T cell migration and that the spatial effects of chemokine diffusion can actually hinder T cell search in rapidly spreading viruses.
+We feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
  \\
@@
 
-In review we have noticed that our bibliography does not use annotated journal names, which seem to be the standard for PLoS CB.
+In review we have noticed that our bibliography does not use annotated journal names, which seem to be the standard for PLOS ONE.
 We have {\em not} changed our bibliography as of yet as it was generated automatically using the Mendeley Desktop suite.
@@
 \\
+
+% This work is a resubmission of a rejected PLOS CB manuscript (PCOMPBIOL-D-12-01516R1).
 

### 2014-06-07 paper/chemokine_main_revised.tex
R4 in progress

--- 2014-01-27 paper/chemokine_main.tex
@@
 {\Large
-\textbf{Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
+\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
 }
@@
 During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a small fraction of the total lung.
-T cell navigation towards infected foci is undirected but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
-Weuse a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signalling.
+T cell navigation towards infected foci is undirected, but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
+We use a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signaling.
 
 To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.
-These data were modeled to estimate chemokine production rates.
+These data were used to estimate chemokine production rates.
 The estimates were combined in the spatial model to test viral variation of T cell recruitment in the lungs.
@@
 A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.
-We conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.
+Finally, a sensitivity analysis highlights the viral parameters as the most significant in influencing the result of the infection.
+Thus, we conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.
 
@@
 % Author Summary not valid for PLoS ONE submissions.   
-\section*{Author Summary}
-
-Clearance of influenza from the lung depends critically on the efficiency with which T cells travel from the lymph node to the site of infection.
-However, T cells in the lung must navigate the large branching bronchial network.
-This maze-like structure is a challenge for T cell search, and the vast majority of T cells exit the lung without encountering infection.
-Infected cells produce cytokines and chemokines, which signal T cells to exit the vasculature and enter tissue.
-We developed an agent-based model to quantify how these molecular signals affect search.
-Our experimental data show that two chemokines, IP-10 and RANTES, are secreted by infected human tracheal epithelial cells.
-We incorporated this data in our spatially explicit model, which allowed us to identify challenges to the search process not obvious in conventional differential equation models.
-We ran three sets of simulations, using parameters from seasonal influenza, avian influenza, and 2009 pandemic influenza.
-The simulated immune response controls and eradicates the first two infections in 10 days but is unable to contain the highly virulent pandemic strain due to the rapidly expanding plaque size.
-Interventions are unlikely to improve this chemokine-directed T cell search process in the lung but vaccination against more virulent strains will increase the number of searching cells earlier in the response.
+% \section*{Author Summary}
 
@@
 
-The adaptive immune response induced during acute influenza A infection is an integrated complex of combination of defense mechanisms.
+%The adaptive immune response to influenza A infection in the lung requires a complex combination of different defensive mechanisms.
+CD8 T cells must navigate from the lymph node through the vascular system to the site of infection.
+%
+%
+%The high replication rate of 2009 pandemic H1N1 influenza presents unique challenges to the immune system when compared to other strains of influenza.
+Studies have attributed the high mortality rate of hosts infected with pH1N1 to the virus' high replication rate.
+%
+%We suggest pH1N1 virulence may only tell part of the story.
+Static analysis of viral production rates and plaque sizes ignore the spatial structure of the infection.
+We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain appropriate numbers of activated CD8 T cells.
+%
+%There are many mathematical models of the adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
+Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.
+Here we examine spatial aspects of the the CD8 T cell diaspora and recruitment to infected lung.
+Activated CD8 T cells searching for and homing into infected tissue do so in a complex environment.
+We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.
+In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
+%
+%%Therefore, we constructed a spatial agent-based model to examine how T cells are recruited to the site of infection, and once there, how they navigate over the plaque to contain the growing infection.
+Because our model strives to answer the specific questions regarding T cell recruitment and movement in the lung, we ignore other important aspects of the immune response such as...
+%
+%% Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
+The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.
+%
+%Previous work has shown that T cell search is likely aided by chemokine secretion by infected cells \cite{Banerjee2011}, but details of this process are unknown.
+To find secretion rates of chemokines, we infected human epithelial cells \textit{in virto} with one of three strains of influenza: avian aH5N1, seasonal sH1N1, and pandemic 2009 pH1N1.
+Wells were measured  for specific chemokines at specific time points over a period of two days.
+Two chemokines: IP-10 (CXCL10) and RANTES (CCL5) displayed elevated concentrations over the course of the infections.
+We extended a previous ODE model \cite{Mitchell2011} and fit it to the chemokine data to deduce per cell chemokine production rates.
+We then used the resulting values to inform our spatial model of T cell recruitment.
+%
+%Our results reveal unique spatial challenges to T cell search that arise from chemokine recruitment of T cells to the site of infection.
+Molecular diffusion of chemokines creates the chemical gradients that allow T cells to find the infection.
+Conversely, the coarse nature of the diffusion process creates pockets of high chemokine concentration.
+These pockets, or local maxima, have the unfortunate effect of recruiting and trapping T cells that navigate within their basin of attraction.
+This effect hinders the ability of T cells to properly cover the entire infection.
+We find that viruses with very high replication rates, such as pandemic H1N1, can create infections that spread across tissue faster than the chemokine signals can diffuse.
+In these instances, the chemical signal that is so crucial to T cell recruitment will actually hinder the immune system's search process.
+%
+%These results show interventions that increase chemokine concentrations and T cell counts may not help clear an infection.
+Furthermore, clearing the pH1N1 infection may be limited by the rate at which T cells can kill infected epithelial cells.
+This suggests that early interventions to infection, such as vaccination, may be the only way to avoid runaway infection.
+
+The adaptive immune response induced during acute influenza A infection is an integrated complex combination of defense mechanisms.
 Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control the infection and immunopathology.
@@
 
-\subsection*{Computational Modeling}
-
-Computational modeling used CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart} and \ref{fig:variance}), and simulations measured efficiency of infection clearance for three different viruses.
-The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.4).
-Vasculature in a mouse is a binary tree with fourteen levels of bifurcation with the origin at the bronchial lymph node complex.
-Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).
-If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection.
-T cells that do not encounter cytokine recirculate to the lymph node.
-When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.
-Unimpeded, the infected secretes virus for 16.7 hours and then dies.
-
-The simulation begins when a single cell in the center of the tissue is infected.
-After the eclipse phase (incubation), the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S1.
-Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-Chemokine diffuses from secreting cells according to kinetics described below, creating a circle of stimulation around the infected region.
-After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells begin to exit the lymph node at a steady rate and circulate through the vasculature to the tissue.
-Further details of the CyCells modeling environment are described in S1.3.
-
-Computational limitations force us to implement our model on the scale of a mouse.
-Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
-The surface area of a mouse lung is approximately 100$cm^2$.
-
-\subsection*{Model Definition}
-
-In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \textit{healthy}, \textit{virus-incubating}, \textit{virus-expressing}, \textit{apoptotic}, and \textit{dead}.
-\textit{Healthy} cells remain unchanged unless infected by virus.
-Once infected, the cell transitions from \textit{incubating} to \textit{expressing} after a 10 hour incubation delay.
-\textit{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
-\textit{Expressing} cells initiate \textit{apoptosis} sooner if they are contacted by activated T cells.
-Apoptoic cells continue to secrete virus and chemokine and then die after one hour.
-\textit{Dead} cells take up space and do not regenerate over the course of an infection.
-
-T cells are described by two states.
-\textit{Circulating} T cells begin to emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.
-If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.
-If a circulating cell encounters chemokine, it converts to \textit{chemotaxing} and begins following the chemotactic gradient to the infected region.
-\textit{Circulating} T cells decay exponentially with an average lifespan of three days.
-\textit{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells.
-\textit{Chemotaxing} cells decay exponentially with an average lifespan of two hours.
-
-The model contains two kinds of particles: virus and chemokine.
-Both are produced at constant rates by \textit{expressing} epithelial cells.
-Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration.
-Chemokine diffuses across the tissue but has no direct effect beyond activating T cells.
-Both particle types decay exponentially.
-IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.
-
-Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.
-Strain-specific values are shown in Table \ref{tab:strains}.
-Further details regarding the model definition are included in S1.4.
-
 \subsection*{Materials}
 
-Chemokine secretion:  Epithelial cell culture and supernatant collection was performed as described \cite{Mitchell2011}.
+Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.
 Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
@@
 
-% Results and Discussion can be combined.
+\subsection*{Computational Modeling}
+
+We implemented our spatial model using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
+A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations measured efficiency of infection clearance for three different viruses.
+The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.2).
+Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).
+If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection.
+T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.
+When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.
+Unimpeded, infected cells secrete virus for 16.7 hours and then die.
+
+The simulation begins when a single cell in the center of the tissue is infected.
+After an incubation period, the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S2.1 and Table \ref{tab:strains}.
+Virus diffuses locally, infecting nearby cells, and continuing the cycle.
+Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region.
+After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a steady rate and travel through the vasculature to the tissue.
+Further details of the CyCells modeling environment are described in S1.1.
+
+Computational limitations force us to implement our model on the scale of a mouse.
+Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
+The surface area of a mouse lung is approximately 100$cm^2$.
+
+\subsection*{Model Definition}
+
+In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}.
+\texttt{Healthy} cells remain unchanged unless infected by virus.
+Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay.
+\texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
+\texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells.
+\texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour.
+\texttt{Dead} cells take up space and do not regenerate over the course of an infection.
+
+T cells are described by two states.
+\texttt{Circulating} T cells emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.
+If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.
+If a circulating cell encounters chemokine, it converts to \texttt{chemotaxing} and begins following the chemotactic gradient to the infected region.
+\texttt{Circulating} T cells decay exponentially with an average lifespan of three days.
+\texttt{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells.
+\texttt{Chemotaxing} T cells decay exponentially with an average lifespan of two hours.
+
+The model contains two kinds of particles: virus and chemokine.
+Both are produced at constant rates by \texttt{expressing} epithelial cells.
+Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration.
+Chemokine diffuses across the tissue but has no direct effect beyond activating T cells.
+Both particle types decay exponentially.
+IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.
+
+Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.
+Strain-specific values are shown in Table \ref{tab:strains}.
+Further details regarding the model definition are included in S1.2.
+
 \section*{Results}
 
-\subsection*{Chemokine production}
-
-To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
-The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.
-IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
-To estimate per-cell production rates, we extended the delay differential equation (DDE) model of Ref.
-\cite{Mitchell2011} to represent chemokine production from infected cells (Eq.~S1).
-Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
-The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
-Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
-There was no positive correlation between viral production and induced chemokine production across the three strains.
-
-\subsection*{Stochastic modeling effects}
+\subsection*{Model Results}
 
 Unlike ordinary differential equation (ODE) and DDE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~\ref{fig:variance}).
-To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.
-$R^2$ values were calculated for individual runs versus the average of all runs.
-The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: (0.957 $\pm$ 0.150), (0.992 $\pm$ 0.013), and (0.998 $\pm$ .002) respectively (i.e.
+Therefore, we ran th eCyCells model 50 times for each of the three strains of influenza: avian, seasonal, and 2009 pandemic (Fig.~\ref{fig:variance}).
+In every case, the model resulted in viral clearance by day 10 in both the avian and seasonal influenza strains.
+Conversely, every simulation of the 2009 pandemic influenza resulted in a runaway infection.
+
+We used $R^2$ values of individual runs versus the average of all runs in order to evaluate the variance of the model.
+The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: 0.957 $\pm$ 0.150, 0.992 $\pm$ 0.013, and 0.998 $\pm$ .002 respectively (i.e.
 there was little variation between runs).
@@
 
-The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).
+% The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).
 Because pH1N1 replicates more rapidly, the number of infected cells outstrips the ability of the T cells to find them.
 Spatial effects reveal the reasons for this in simulation images and video (Fig.~\ref{fig:cycells}, Video S4).
+
+\subsection*{Sensitivity Analysis}
+
+We performed a sensitivity analysis of most of the model parameters due to the large number of free parameters in the model (Fig.
+S?).
+Each parameter included in the analysis was varied over several values while the rest of the parameters were held constant.
+This analysis allowed us to qualitatively classify parameters as sensitive or insensitive depending on how the model outcome changed as each parameter was varied.
+A more detailed description of how we classified the parameters is included in section S2.4.
+
+Five parameters stand out as having a strong effect on the model results.
+They are: IgM strength, viral decay rate, infection sensitivity, viral diffusion, and the T cell secretion rate.
+We also know from the results of the three virus strains that the viral production rate is also sensitive parameter.
+It is interesting to note that four of the parameters are directly related to the efficacy of the virus in isolation.
+Furthermore, IgM strength, as coded in the model, also directly affects the decay rate of the virus and can also be included in this group.
+Only the T cell secretion rate appears to have a strong effect on the model while not being directly related to the virus kinetics.
+Also of note, none of the other T cell parameters nor any of the chemokine parameters seem to have as strong of an effect on the model as the virus parameters.
+This supports the hypothesis that the virus dynamics dominate the course of an infection.
+
+\section*{Discussion}
+
+Our aim was to examine the spatially determined restrictions on the efficiency of the T cell search.
+Due to the absence of critical data, the inability to image the search in the lung (as was done in the lymph node), and the need for a tractable computation, the model simplified the adaptive immune response.
+We focused instead on comparing the behavior of thee influenza viruses.
 
@@
 
-\section*{Discussion}
-
-Our aim was to examine the spatially determined restrictions on the efficiency of the T cell search.
-Due to the absence of critical data, the inability to image the search in the lung (as was done in the lymph node), and the need for a tractable computation, the model simplified the adaptive immune response.
-We focused instead on comparing the behavior of thee influenza viruses.
-
 \subsection*{Modeling Methodology}
@@
 Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.
-Thus, our model is not intended to predict clearance of virus from the lung, but to examine the response  features that permit CD8 T cells to sense and contact infected target cells.
+Thus, our model is not intended to predict the exact infection size over time.
+Rather, our model allows us to qualitatively compare the infection kinetics of the three strains of influenza and to examine the response features that permit CD8 T cells to sense and contact infected target cells.
 
-The use of an ABM has certain advantages over a spatially homogeneous ODE model.
+The use of an ABM has certain advantages over a spatially homogeneous differential equation model.
 An ODE model assumes that any virus particle is capable of infecting any healthy cell.
@@
 
-Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.
-Because CD8 T cells find infected cells by presumably climbing a chemokine gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
+\subsection*{Conclusions}
+
+Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and are difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.
+Because CD8 T cells presumably find infected cells by climbing a chemokine gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
 We also see that T cell clustering persists after all virus-secreting cells have been eliminated.
 The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.
-Finally, we can see that infected cells are more disperse as infection size grows.
+Furthermore, we can see that infected cells are more disperse as infection size grows.
 Because T cells are clustered they cannot cover the increasing plaque effectively.
-These spatial observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM (\textit{in vivo} pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation).
+Finally, our sensitivity analysis suggests that viral dynamics are the dominant factor in determining the course of an infection.
+Changes in the chemokine and T cell parameters do not have a strong effect on the infection result.
+
+These model observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
+\textit{In vivo} pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation.
 The behavior of searching T cells described in this paper can enhance future global immune response models.
-
-\subsection*{Chemokine Directed T Cell Search}
-
-The actual quantitative chemokine environment in the infected lung is difficult to estimate for a math model.
-Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells in vitro depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
-Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
-The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
-
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
-The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
-This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-
-Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
@@
 Bezerra, B.
-Edwards, and G.
+Edwards, ThanhVu Nguyen, and G.
 Stelle (UNM Computer Science) for suggestions concerning the manuscript, cellular and chemokine behavior, and agent-based modeling.
@@
 \begin{center}
-%\includegraphics[width=4in]{Figure_1}
+\includegraphics[width=4in]{Figure_1}
 \end{center}
@@
 \begin{center}
-%\includegraphics[width=4in]{Figure_2}
+\includegraphics[width=4in]{Figure_2}
 \end{center}
@@
 \begin{center}
-% \includegraphics[width=4in]{Figure_3}
+ \includegraphics[width=4in]{Figure_3}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{Figure_4}
+\includegraphics[width=\textwidth]{Figure_4}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=4in]{Figure_5}
+\includegraphics[width=4in]{Figure_5}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{Figure_6}
+\includegraphics[width=\textwidth]{Figure_6}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{Figure_S1}
+\includegraphics[width=\textwidth]{Figure_S1}
  \end{center}
@@
 \begin{center}
-% \includegraphics[width=4in]{Figure_S2}
+\includegraphics[width=4in]{Figure_S2}
  \end{center}
@@
 \begin{table}
-\begin{center}
-\begin{tabular}{|  l  l  l  |}
+\centering
+\begin{tabular}{ | r | c | c | c | }
   \hline                        
-  Population & Description & Initial Value\\
+  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
+   & \footnotesize{$(pg/s\cdot cell)$}  & \footnotesize{$(pg/s\cdot cell)$} &  \footnotesize{$(PFU/s\cdot cell)$} \\
   \hline
-  $T$ & Healthy target epithelial cells & 1,000,000 \\
-  $I_1$ & Virus-incubating cells & 0 \\
-  $I_2$ & Virus-secreting cells & 0 \\
-  $V$ & Free virus particles & 10,000 \\
-  $F$ & Interferon quantity & 0 \\
-  $C$ & Chemokine quantity (ng/mL) & 0 \\
+  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
+   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
    \hline
+  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
+   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
    \hline
-  Parameter & Description & Value \\
-  \hline
-  $\beta$ & Viral infection rates &  \\
-    & \hspace{2em} Avian & 5.3e-7 (PFU$\cdot$h)$^{-1}$ \\
-    & \hspace{2em} Seasonal & 6.1e-7 (PFU$\cdot$h)$^{-1}$ \\
-    & \hspace{2em} Pandemic &  2.7e-6 (PFU$\cdot$h)$^{-1}$\\
-  $p$ & Viral production rates & \\
-    & \hspace{2em} Avian & 0.20 (PFU/h) \\
-    & \hspace{2em} Seasonal & 1.4 (PFU/h) \\
-    & \hspace{2em} Pandemic & 18.3 (PFU/h) \\
-  $e$ & Interferon strengths & \\
-    & \hspace{2em} Avian & 1.0e-8 \\
-    & \hspace{2em} Seasonal & 1.6e-6 \\
-    & \hspace{2em} Pandemic & 3.4e-3 \\
-  $\tau_2$ & Interferon production delays &  \\
-    & \hspace{2em} Avian & 21.5 (h) \\
-    & \hspace{2em} Seasonal & 23.6 (h) \\
-    & \hspace{2em} Pandemic & 21.0 (h) \\
-  $\delta$ & Virus-secreting cell decay rate & 0.6 (h$^{-1}$) \\
-  $d$ & Chemokine decay rate & 1.386 (h$^{-1}$) \\
-  $\tau_1$ & Viral incubation time & 10 (h) \\
-  $\tau_3$ & Chemokine production delays & \\
-    & \hspace{2em} IP-10 & 8 (h)\\
-    & \hspace{2em} RANTES & 16 (h)\\
+  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
+   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
   \hline
 \end{tabular}
-\caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
-The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
-Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
-\label{tab:dde}
-\end{center}
+\caption{Strain-specific parameters.
+Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
+Bootstrapping for the chemokine values was performed using the original fit of Eq.
+S1 to the data in Fig.~\ref{fig:data} to produce new data sets.
+Viral production values and confidence intervals are taken from \cite{Mitchell2011}.}
+\label{tab:strains}
 \end{table}
@@
 
-\begin{table}
-\centering
-\begin{tabular}{ | r | c | c | c | }
-  \hline                        
-  \multicolumn{1}{|c|}{\multirow{2}{*}{Strain}} & IP-10 Production & RANTES Production & Viral Production \\
-   & \footnotesize{$(pg/s\cdot cell)$}  & \footnotesize{$(pg/s\cdot cell)$} &  \footnotesize{$(PFU/s\cdot cell)$} \\
-  \hline
-  \multirow{2}{*}{Avian H5N1} & 2.0e-4 &  1.3e-5 & 5.4e-5 \\
-   &  \footnotesize{8.4e-5 --- 4.2e-4} & \footnotesize{7.9e-6 --- 1.9e-5} & \footnotesize{4.4e-5 --- 3.7e-4}\\ 
-   \hline
-  \multirow{2}{*}{Seasonal H1N1} & 1.8e-4 &  8.9e-7 & 3.8e-4 \\
-   & \footnotesize{1.2e-4 --- 3.0e-4} & \footnotesize{4.8e-7 --- 1.6e-6} & \footnotesize{2.8e-4 --- 1.5e-3}\\
-   \hline
-  \multirow{2}{*}{Pandemic H1N1} & 8.7e-5 &  4.3e-6 & 5.1e-3 \\
-   & \footnotesize{1.7e-5 --- 7.1e-4} & \footnotesize{5.0e-7 --- 3.5e-5} & \footnotesize{2.8e-3 --- 5.3e-3} \\
-  \hline
-\end{tabular}
-\caption{Strain-specific model parameters.
-Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
-Bootstrapping for the chemokine values was performed using the original fit of Eq.
-S1 to the data in Fig.~\ref{fig:data} to produce new data sets.
-Viral production values and confidence intervals are taken from \cite{Mitchell2011}.}
-\label{tab:strains}
-\end{table}
-
 \end{document}

### 2014-06-07 paper/chemokine_supplement_revised.tex
R4 in progress

--- 2014-01-27 paper/chemokine_supplement.tex
@@
 
-\subsection{Estimating chemokine production rates}
-
-We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} \drew{Eq.
-1} by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
-Initial population sizes and parameter values are taken from the previous study.
-{\footnotesize
-\begin{equation}
-\begin{aligned}
-\dot{T} &= - \beta T V \\
-\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
-\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
-\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
-\dot{F} &=  I_{1 \tau_2} \\
-\dot{C} &= r I_{1 \tau_3} - d C \\
-\end{aligned}
-\label{eq:dde}
-\end{equation}
-\vspace{.05in}
-}
-
-$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
-Table 1 summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 3).
-
-\subsection{T cell production rate}
-
-We calculate the rate of production ($\sigma$) of CD8 T cells using a differential equation model from \cite{Miao2010}.
-The equations model T cell production and subsequent search over an area of infected lung tissue.
-
-{\footnotesize
-\begin{equation*}
-\label{EqS1}
-\begin{aligned}
-\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
-\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
-\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
-\dot{T} &= \rho T -\beta TV \\
-\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
-\dot{V} &= pI - \beta TV - \gamma (t) V \\
-\gamma (t) &= \left\{ \begin{array}{rcl}
-	1/\mbox{day} & \mbox{,}  & t < 5  \\
-	3/\mbox{day} & \mbox{,} & t \geq 5  
-	\end{array}\right. \\
-\end{aligned}
-\tag{S1}
-\end{equation*}
-}
-\vspace{.05in}
-
-$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region.
-$T$ is the number of uninfected \textit{target} cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
-The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
-The lung is modeled as a circular region with radius $R$.
-The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
-See Fig.
-1 for more details about the model.
-The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
-$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
-
-Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
-We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-
-The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time.
-Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
-Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
-Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
-Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
-The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
-The initial number of infected cells is assumed to be zero.
-Parameter values are listed in Table S1 in Text S1.
-This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
-The final value was found to be 1,257 per hour.
-
 \subsection{CyCells implementation details}
@@
 
-\subsection*{T cell sensitivity to chemokine}
+\subsection{Estimating chemokine production rates}
+
+We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq.
+1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
+Initial population sizes and parameter values are taken from the previous study.
+{\footnotesize
+\begin{equation}
+\begin{aligned}
+\dot{T} &= - \beta T V \\
+\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
+\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
+\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
+\dot{F} &=  I_{1 \tau_2} \\
+\dot{C} &= r I_{1 \tau_3} - d C \\
+\end{aligned}
+\label{eq:dde}
+\end{equation}
+\vspace{.05in}
+}
+
+$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
+Table S1 summarizes population and parameter values and descriptions.
+Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 1).
+
+%\subsection*{Chemokine Secretion}
+
+%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
+The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
+
+%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref.
+\cite{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).
+Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
+Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
+There was no positive correlation between viral production and induced chemokine production across the three strains.
+
+\subsection{T cell production rate}
+
+We calculate the rate of production ($\sigma$) of CD8 T cells using a differential equation model from \cite{Miao2010}.
+The equations model T cell production and subsequent search over an area of infected lung tissue.
+
+{\footnotesize
+\begin{equation*}
+\label{EqS1}
+\begin{aligned}
+\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
+\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
+\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
+\dot{T} &= \rho T -\beta TV \\
+\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
+\dot{V} &= pI - \beta TV - \gamma (t) V \\
+\gamma (t) &= \left\{ \begin{array}{rcl}
+	1/\mbox{day} & \mbox{,}  & t < 5  \\
+	3/\mbox{day} & \mbox{,} & t \geq 5  
+	\end{array}\right. \\
+\end{aligned}
+\tag{S1}
+\end{equation*}
+}
+\vspace{0.5in}
+
+$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region.
+$T$ is the number of uninfected \textit{target} cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
+The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
+The lung is modeled as a circular region with radius $R$.
+The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
+See Fig.
+1 for more details about the model.
+The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
+$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
+
+Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
+We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
+
+The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time.
+Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
+Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
+Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
+Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
+The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
+The initial number of infected cells is assumed to be zero.
+Parameter values are listed in Table S1 in Text S1.
+This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
+The final value was found to be 1,257 per hour.
+
+\subsection{T cell sensitivity to chemokine}
 
@@
 
+\subsection{General sensitivity analysis}
+
+Model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Table~\ref{tab:parameters}).
+Due to the difficulty of estimating critical, unmeasurable values, we performed a sensitivity analysis for 16 parameters in order to test whether variation over an order of magnitude or more altered the infection outcome.
+Individual parameters were varied over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
+Each parameter was varied over all three influenza strains, creating three sets of sensitivity plots (Figures~\ref{fig:asensitivity}-\ref{fig:psensitivity}).
+
+We then categorize the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection.
+When evaluating the results, we consider model runs that asymptotically converge to be qualitatively stable and model runs that do not converge to be qualitatively divergent.
+
+\subsubsection{Stable Parameters}
+
+The parameters in the first grouping did not affect the outcome of the infection unless they were adjusted to values outside the realm of possibility.
+Specifically, each parameter in this group seems to lie within a range that does not significantly affect the asymptotic behavior of the model.
+Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).
+Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.
+The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal effects.
+
+The stability of these parameters makes sense in the context of our model.
+Chemokine exists in our model to provide a chemical gradient that T cells may follow to the focus of infection.
+The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.
+Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells may follow.
+
+Similarly, T cells affect the model by clearing cells expressing virus.
+As discussed in the main paper, the chemokine gradient creates areas of maximal concentration that attract all the T cells inside its basin of attraction.
+Thus, most T cells are attracted to the same areas of the infection and overlap considerably.
+Increasing T cell numbers and efficiency will not help clear the infection beyond a certain point.
+
+While we have claimed that these parameters are qualitatively stable within a certain range, many of the parameters were set to values that did lead to a divergent model behavior.
+We have deemed these deviations acceptable on an individual basis as described below.
+
+Apoptosis time diverges slightly in the sH1N1 strain.
+As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper.
+
+Varying the chemokine decay rate results in stable model behavior for values one order of magnitude larger and smaller than the default.
+A value two orders of magnitude smaller results in divergent behavior for both the aH5N1 strain and the sH1N1 strain and a value two orders of magnitude larger results in divergent behavior for only the sH1N1 strain.
+This maximum decay rate used for all three strains corresponds to an implausible 18 second half-life.
+The minimal value corresponds to a similarly implausible 50 hour half-life.
+The fact that an extremely low decay rate can hinder clearance is interesting.
+In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.
+This confirms that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.
+Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection.
+
+The chemokine diffusion rate only diverges in the aH5N1 strain for the highest value.
+If anything, our estimate for the diffusion rate is already large as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.
+Thus the diffusion rate we used based on Stokes-Einstein and water viscosity was likely within a physiologic range.
+
+Chemokine secretion values differ between strains.
+It is important to node that aH5N1 and sH1N1 show a threshold at the same concentration: near 1e-6 $pg/s\cdot cell$.
+While this may seem contradictory, it is actually an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.
+Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence.
+
+T cell circulation times were tested over a very large range and only diverge at the very edge of that range.
+Because vascular and lymph circulation times are unknown, parameter values of 1 minute and 3 minutes in mice may be plausible and warrant further study.
+
+The T cell kill rate only diverges on the lower end in sH1N1.
+The extreme value corresponds to a T cell needing 100 minutes to induce the apoptosis of a single infected cell and is not biologically reasonable.
+The intermediate value corresponds to a time of 33.3 minutes and is also unlikely.
+
+T cell movement in tissue has been observed \cite{Egen2011}.  Thus, we consider the extreme values biologically implausible.
+
+T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.
+
+\subsubsection{Difference in Peak Only}
+
+Two parameters, viral incubation time and viral expression time, do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.
+This effect occurs because the T cells short circuit the lifespan of infected cells.
+Once T cells arrive, virus secreting cells no longer survive for their full lifespan.
+Rather, the secretion time becomes limited by the apoptosis time parameter.
+Because these terms refer to how and when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response.
+
+\subsubsection{Unstable Parameters}
+
+This leaves five parameters shown that appear to directly affect the result of the infection: IgM strength, infection sensitivity, T cell secretion rate, viral decay, and viral diffusion.
+By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.
+Of interest, nearly all of these parameters are directly related to the behavior of the virus.
+Only the T cell secretion rate and IgM parameters are not.
+
+IgM strength is not inserted until day 4 of the simulation, but once it does it directly affects the decay rate of the virus, and does so for the rest of the simulation.
+The higher the strength of this parameter, the less virus there is in the system.
+Because the true effective strength of IgM is unknown in the context of this model, it is not a target of our investigation.
+
+Viral decay has the exact same effect as the IgM strength parameter.  Its value directly determines how much virus remains in the system over the course of the infection.
+
+Infection sensitivity describes the ability of the virus to infect healthy cells, yet its strength is directly related to how much virus is present in the area of the healthy cells.
+Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.
+Thus, it behaves similarly to the IgM and viral decay parameters.
+
+Shown elsewhere, the viral secretion rate has a strong effect on the outcome of the infection.
+Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.
+Thus, it's affect is similar to the previous parameters.
+
+Viral diffusion does not change how much virus there is in the system.
+Rather, it determines how fast the virus may spread across the alveoli.
+While its mechanism is different, its effect may be even stronger than the previous parameters.
+By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may outpace the body's ability to generate a chemical gradient.
+Thus T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.
+
+Finally, the T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
+Thus it is reasonable to assume that there is a threshold of T cell secretion, beyond which the dynamics of the infection do not change.
+This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
+Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.
+
 %\drew{
@@
 %}
+
+\section{Discussion}
+
+\subsection{Chemokine directed T cell search}
+
+The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
+Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+
+A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+
+Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
@@
 \begin{center}
-% \includegraphics[width=\textwidth]{Figure_S1}
+ \includegraphics[width=\textwidth]{Figure_S1}
  \end{center}
@@
 \begin{center}
-%	\includegraphics[width=\textwidth]{Figure_S2}
+	\includegraphics[width=\textwidth]{Figure_S2}
 	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{|  l  l  l  |}
+  \hline
+  Population & Description & Initial Value\\
+  \hline
+  $T$ & Healthy target epithelial cells & 1,000,000 \\
+  $I_1$ & Virus-incubating cells & 0 \\
+  $I_2$ & Virus-secreting cells & 0 \\
+  $V$ & Free virus particles & 10,000 \\
+  $F$ & Interferon quantity & 0 \\
+  $C$ & Chemokine quantity (ng/mL) & 0 \\
+  \hline
+  \hline
+  Parameter & Description & Value \\
+  \hline
+  $\beta$ & Viral infection rates &  \\
+    & \hspace{2em} Avian & 5.3e-7 (PFU$\cdot$h)$^{-1}$ \\
+    & \hspace{2em} Seasonal & 6.1e-7 (PFU$\cdot$h)$^{-1}$ \\
+    & \hspace{2em} Pandemic &  2.7e-6 (PFU$\cdot$h)$^{-1}$\\
+  $p$ & Viral production rates & \\
+    & \hspace{2em} Avian & 0.20 (PFU/h) \\
+    & \hspace{2em} Seasonal & 1.4 (PFU/h) \\
+    & \hspace{2em} Pandemic & 18.3 (PFU/h) \\
+  $e$ & Interferon strengths & \\
+    & \hspace{2em} Avian & 1.0e-8 \\
+    & \hspace{2em} Seasonal & 1.6e-6 \\
+    & \hspace{2em} Pandemic & 3.4e-3 \\
+  $\tau_2$ & Interferon production delays &  \\
+    & \hspace{2em} Avian & 21.5 (h) \\
+    & \hspace{2em} Seasonal & 23.6 (h) \\
+    & \hspace{2em} Pandemic & 21.0 (h) \\
+  $\delta$ & Virus-secreting cell decay rate & 0.6 (h$^{-1}$) \\
+  $d$ & Chemokine decay rate & 1.386 (h$^{-1}$) \\
+  $\tau_1$ & Viral incubation time & 10 (h) \\
+  $\tau_3$ & Chemokine production delays & \\
+    & \hspace{2em} IP-10 & 8 (h)\\
+    & \hspace{2em} RANTES & 16 (h)\\
+  \hline
+\end{tabular}
+\caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
+Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
+\label{tab:dde}
+\end{center}
+\end{table}
+
 \begin{table}[!ht]

### 2014-06-07 paper/sensitivity_section.tex
R4 in progress

--- 2012-12-16 paper/chemokine_supplement.tex
@@
 % Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
+%\bibliographystyle{plos2009}
 
@@
 
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section{Models}
+% Results and Discussion can be combined.
+\section*{Results}
 
-\subsection{T Cell Production Rate}
+\subsection*{General sensitivity analysis}
 
-To find a value for $\sigma$, the rate of production of activated T cells (ATCs) we use a differential equation model from \cite{Miao2010}:
+Model parameters were chosen from literature when available and estimated otherwise (Table~\ref{tab:parameters}).
+Due to the possible variability in the chosen values, we performed a rudimentary sensitivity analysis on a majority of the parameters.
+Individual parameters were varied over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
+Each parameter was varied over all three influenza strains, creating three sets of sensitivity plots (Figures~\ref{fig:asensitivity}-\ref{fig:psensitivity}).
 
-{\footnotesize
-\begin{equation*}
-\label{EqS2}
-\begin{aligned}
-\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}}    & \pi r'^{2} &= a \sqrt{I} + b \\
-\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
-\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
-\dot{T} &= \rho T -\beta TV \\
-\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
-\dot{V} &= pI - \beta TV - \gamma (t) V \\
-\gamma (t) &= \left\{ \begin{array}{rcl}
-	1/\mbox{day} & \mbox{,}  & t < 5  \\
-	3/\mbox{day} & \mbox{,} & t \geq 5  
-	\end{array}\right. \\
-\end{aligned}
-\tag{\textbf{S}1}
-\end{equation*}
-}
-\vspace{.05in}
+We then categorize the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection.
 
-where $N_{c}$ is the number of circulating activated antigen-specific T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines and $N_{f}$ is the number of circulating T cells that have found an infected region.
-$T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
-The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
-The lung is assumed to be a circular region with radius $R$.
-The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
-See Fig.
-1 for more details about the model.
-The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
+\subsubsection*{Stable Parameters}
 
-Circulating activated antigen-specific T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
-These circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) transition to the infected region after walking randomly for an average time proportional to the difference in the radii between the two regions ($r' - r$) at a given speed, $v_{tcell}$.
-Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
-Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
-Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
-Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
-The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
-The initial number of infected cells is assumed to be zero.
-This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
-The final value was found to be 777 per hour.
+The parameters in the first grouping do not affect the outcome of the infection unless adjusted to values that are outside the realm of possibility.
+Specifically, each parameter in this group seems to lie within a range that does not affect the behavior of the model at all.
+Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).
+Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.
+The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal effects.
 
-\subsection{CyCells Implementation Details}
+The stability of these parameters makes sense in the context of our model.
+Chemokine exists in our model to provide a chemical gradient that T cells may follow to the focus of infection.
+The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.
+Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells may follow.
 
-The CyCells modeling environment splits its populations into two types: cells and molecules.
-Cells are considered to be unique agents and are modeled individually.
-On the other hand, molecules are considered to be too small and numerous to model explicitly and are thus represented as compartmentalized concentrations.
-These compartments are arranged as a grid that covers the environment.
-Each square in the grid contains a unique concentration of the given molecule.
-Multiple molecules are represented in separate grids.
+Similarly, T cells affect the model by clearing cells expressing virus.
+As discussed in the main paper, the chemokine gradient creates areas of maximal concentration that attract all the T cells inside its basin of attraction.
+Thus, most T cells are attracted to the same areas of the infection and overlap considerably.
+Increasing T cell numbers and efficiency will not help clear the infection beyond a certain point.
 
-Molecular behavior is limited to diffusion and decay.
-At each time step, CyCells decreases the concentration of each square in the grid according to the decay rate specified for the given molecule.
-CyCells then diffuses the molecules by applying a discrete diffusion equation to each grid square, taking into account the specified diffusion constant and the concentrations in neighboring squares.
+While we have classified these parameters as stable within a certain range, many of the parameters have values that do lead to a different model behavior.
+We have deemed these deviations acceptable as on an individual basis.
 
-New molecules can be secreted by cell objects, such as an infected cell secreting new virus.
-In this scenario CyCells calculates the amount of virus secreted per time step, based upon the cell's defined production rate, and then adds this quantity to the grid square that overlaps the secreting cell.
-If a cell wants to remove molecular concentration, it will subtract from the square it overlaps.
-If the concentration at that square is not enough, concentration will be removed from neighboring squares in an ever expanding diamond.
+Apoptosis time diverges slightly in the sH1N1 strain.
+As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper.
 
-When a cell's behavior depends on a local molecular concentration, the cell will measure the local concentration at its position by calculating a linearly interpolated combination of the concentrations in the grid squares surrounding the cell.
+The chemokine decay rate creates a divergence for its maximum value in the aH5N1 strain and both of its extreme values in the seasonal strain, but is stable over its inner values.
+The maximum decay rate corresponds to an implausible 18 second half-life.
+The minimal value corresponds to a similarly implausible 50 hour half-life.
+The fact that an extremely low decay rate can hinder clearance is interesting.
+In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.
+This confirms that the quantity of chemokine is immaterial as long as there is enough for T cells to be able to detect it.
 
-\subsection{Model Implementation Details}
+The chemokine diffusion rate only diverges for the maximum value in the sH1N1 strain.
+If anything, our estimate for the diffusion rate is already large as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.
+Thus, it is also of note that lower values do not affect the model.
 
-An overview of the model implementation is shown in Fig.
-2.
-The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (46,085 total cells).
-The sheet measures 10$mm$ per side, which is large enough to contain the infection (Fig.
-5).
-The molecular grid (S1.2) is initialized with a resolution of 50$\mu m$.
+Chemokine secretion values differ between strains.
+It is important to node that aH5N1 and sH1N1 show a threshold at the same place: near 1e-6 $pg/s\cdot cell$.
+While this may seem suspicious, it is actually an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.
+Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of instability.
 
-Healthy cells transition to virus-incubating based on a probability that scales linearly with the local virus concentration.
-When a cell `becomes infected' it removes the amount of viral concentration equal to a single virion.
+T cell circulation times were tested over a very large range and only diverge at the very edge of that range.
+Vascular and lymph circulation times of 1 minute and 3 minutes are implausible in a mouse model.
 
-Virus-incubating cells idle for 10 hours (with standard deviation $\sigma=1$ hour) and then transition to virus-secreting cells.
-Virus-incubating cells also begin to secrete chemokine after 8 hours (except for aH5N1 which only secretes RANTES after a 16 hour delay).
+The T cell kill rate only diverges on the lower end in sH1N1.
+The extreme value corresponds to a T cell needing 100 minutes to induce the apoptosis of a single infected cell and is not biologically reasonable.
+The intermediate value corresponds to a time of 33.3 minutes and is also unlikely.
 
-Virus-secreting cells secrete both virus and chemokine (the RANTES portion of the chemokine production is added in 16 hours after the initial infection) and die after 1,000 minutes of production (16.7 hours).
-Virus-secreting cells will also probabilistically transition to apoptizing cells in the presence of T cells (defined by the existence of a T cell within 5$\mu m$).
-The number of T cells in the vicinity of the virus-secreting cell has no effect on the rate that apoptosis is induced in the model.
+T cell movement in tissue has been observed \cite{Egen2011}.  Thus, we consider the extreme values biologically implausible.
 
-Apoptizing cells continue to secrete virus and chemokine for one hour and then die.
+T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.
 
-T cells are added to the model at a constant rate of 777 per hour (S1.1) through a global `oracle' cell that represents the lymph node.
-Because the model environment is only a small portion of the entire lung (1\%), most of the T cells miss the model window and are not visually represented.
-The T cells that miss enter a dormant searching state and transition to recirculating after a 10 minute delay.
-The cells that enter the model window are placed at a random position and begin a random walk.
-At each step, these searching cells check the local chemokine concentration.
-If the concentration is above the sensitivity threshold, the T cells immediately transition to the chemotaxing state.
-If not, searching T cells transition to the recirculating state and are removed from the model window.
-Searching and recirculating T cells have a probabilistic decay rate that corresponds to a 3 day lifespan.
+\subsubsection*{Difference in Peak Only}
 
-Recirculating T cells are not visually represented.
-These cells are individually tracked for 6 minutes, after which they are reintroduced as new searching cells with a new chance of entering the model window.
+Two parameters, viral incubation time and viral expression time, do seem to affect the model's behavior up until the introduction of the T cell response, after which the results converge back to the same point.
+This suggests that the secondary response dominates the effect of these two parameters.
+This effect occurs because the T cells short circuit the lifespan of infected cells.
+Once T cells arrive, virus secreting cells no longer survive for their full lifespan.
+Rather, the secretion time becomes limited by the apoptosis time parameter.
+Because these are both delay terms, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response.
 
-Chemotaxing T cells move directly up the chemokine gradient until they find a local maximum.
-T cells probabilistically decay at a rate corresponding to a 2 hour lifespan.
-Chemotaxing T cells have no effect other than inducing apoptosis in virus-secreting cells by proximity.
+\subsubsection*{Unstable Parameters}
 
-% Results and Discussion can be combined.
-\section{Results}
+This leaves five parameters shown that appear to directly affect the result of the infection: IgM strength, infection sensitivity, T cell secretion rate, viral decay, and viral diffusion.
+By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.
+Of interest, nearly all of these parameters are directly related to the behavior of the virus.
+Only the T cell secretion rate parameter is not.
 
-\subsection{Stochastic modeling effects}
+IgM strength does not take effect until day 4 of the simulation, but once it does it directly affects the decay rate of the virus, and does so for the rest of the simulation.
+The higher the strength of this parameter, the less virus there is in the system.
 
-Unlike ODE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~S\ref{fig:variance}).
-To test the strength of this effect, we ran each model fifty times using the default parameters given in Table 2.
-$R^2$ values were calculated for individual runs versus the average of all runs.
-Pandemic H1N1 showed the least inter-run variance with an mean $R^2$ value of 0.9935 and standard deviation of 0.0029.
-sH1N1 had a mean $R^2$ 0.9859 with a standard deviation of 0.0109.
-aH5N1 had a mean $R^2$ of 0.9167 with a standard deviation of 0.0726.
+Viral decay has the exact same effect as the IgM strength parameter.  Its value directly determines how much virus remains in the system over the course of the infection.
 
-Each run took the calculated viral production and chemokine production rates for the three different strains of influenza as input and reported the total number of infected cells, including incubating, virus secreting and apoptotic, but not including dead cells.
-Therefore the figures approximate plaque growth over time.
+Infection sensitivity describes the ability of the virus to infect healthy cells, yet its strength is directly related to how much virus is present in the area of the healthy cells.
+Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.
+Thus, it behaves similarly to the IgM and viral decay parameters.
 
-Overlaying multiple runs on a single plot reveals a slight growth rate transition 3 days p.i., which reflects the addition of IgM.
-In addition, for each infection the number of infected cells declines quickly at day five due to the T cell response.
+Shown elsewhere, the viral secretion rate has a strong effect on the outcome of the infection.
+Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.
+Thus, it's affect is similar to the previous parameters.
 
-The three strains show different levels of virulence, consistent with the results found in \cite{Mitchell2011} (Fig.~S\ref{fig:variance}).
-The rapid production of the pandemic H1N1 virus prevents the immune response from containing the infection.
-Because pH1N1 is replicates at such a rapid rate, the window of opportunity for T Cells to gain ground is too small.
-aH5N1 is cleared completely, sH1N1 is contained but not fully cleared, and pH1N1 recovers and continues to expand.
+Viral diffusion does not change how much virus there is in the system.
+Rather, it determines how fast the virus may spread across the alveoli.
+While its mechanism is different, its effect may be even stronger than the previous parameters.
+By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may outpace the body's ability to generate a chemical gradient.
+Thus T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.
 
-\subsection{Chemokine combinations}
-
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.
-We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).
-Because of this behavior IP-10 was not included in the aH5N1 model runs.
-
-T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.
-
-Four models runs were performed for each strain (two for aH5N1) to look at how the presence and/or lack of specific chemokines affect the simulated immune response (Fig.~S\ref{fig:chemokine}).
-The lack of both chemokines leads to runaway infections in all three strains.
-The presence of only RANTES is enough to contain the aH5N1 infection, but is weaker than IP-10 in both H1N1 strains.
-In sH1N1 and pH1N1 simulations, IP-10 alone proves to be as effective as the combination of IP-10 and RANTES.
-This suggests that RANTES does not play a significant role in infections that stimulate an IP-10 response due to the higher production rates of IP-10.
-
-% The bibtex filename
-\bibliography{references}
-
-\pagebreak
-
-\section*{Figure Legends}
-
-%\begin{figure}[!ht]
-%\begin{center}
-%%\includegraphics[width=4in]{figure_name.2.eps}
-%\end{center}
-%\caption{
-%{\bf Bold the first sentence.}  Rest of figure 2  caption.  Caption 
-%should be left justified, as specified by the options to the caption 
-%package.
-%}
-%\label{Figure_label}
-%\end{figure}
-
-\begin{figure}[ht!]
-\begin{center}
-% \includegraphics[width=\textwidth]{variance}
- \end{center}
-\caption{Model results: Time series plots of fifty runs of aH5N1 (A), sH1N1 (B), and pH1N1 (C) infections (gray).
-IP-10 and RANTES were simulated in each run, except for aH5N1, which  produced only RANTES.
-Each run was initialized identically for each strain save for the random seed.
-The middle line shows the average while the outer lines show the 95\% confidence interval.}
- \label{fig:variance}
-\end{figure}
-
-\begin{figure}[ht!]
-\begin{center}
-%	\includegraphics[width=\textwidth]{chemokine}
-	\caption{Effects of different chemokine combinations.
-A) aH5N1 does not stimulate an IP-10 response.
-B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
-	\label{fig:chemokine}
-\end{center}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
-T cells show half way through the simulation.
-Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.}
- \label{video:cell_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The second of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the virus concentration.
-Notice the volatility when T cells arrive halfway through the simulation.
-Virus concentration ranges from 1e-13 mols/mL (white) to 1e-27 mols/mL (black).
-Refer to Figure 5 for the detailed legend.
-}
- \label{video:virus_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{The third of three overlaid videos of a representative seasonal H1N1 infection.
-This video spans the 10 day infection and shows the chemokine concentration.
-Notice the volatility when T cells arrive halfway through the simulation.
-Chemokine concentration ranges from 1e8 ng/mL (white) to 1e-6 ng/mL (black).
-Refer to Figure 5 for the detailed legend.
-}
- \label{video:chemokine_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{A closer look at the 2009 pandemic simulation.
-This video shows the infection from day 6 to day 7 with each frame spanning 100 simulated seconds.
-Healthy cells are gray, virus-incubating cells are yellow, virus-secreting cells are orange, apoptotic cells are red, and T cells are green.
-Note the high proportion of virus-secreting cells (orange) early on.
-As time passes, secreting cells are gradually contained to the point where they become very sparse.
-T cell clumping often prevents the T cells from quick discovery of new secreting cells.}
-\end{figure}
-
-\pagebreak
-
-\section*{Tables}
-
-%\begin{table}[!ht]
-%\caption{
-%\bf{Table title}}
-%\begin{tabular}{|c|c|c|}
-%table information
-%\end{tabular}
-%\begin{flushleft}Table caption
-%\end{flushleft}
-%\label{tab:label}
-% \end{table}
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{ | c | c | c | }
-  \hline                        
-  Paramter & Value & Source \\
-  \hline
-  T(0) & $1e6$ & \cite{Mitchell2011} \\
-  V(0) &  $1e4$ & \cite{Mitchell2011} \\
-  $r_{cell}$ &  $10 \mu m$ & \cite{Miao2010} \\
-  $t_{rc}$ & $6s$ & \cite{Peters1983} \\
-  $v_{tcell}$ & $0.3 \mu m/s$ & \cite{Miller2003} \\
-  $k_e$ & $6.4e-5/cell/day$ & \cite{Miao2010} \\
-  $\beta$ & $4.8e-7/cell/PFU$ & \cite{Mitchell2011} \\
-  $p$ & $0.18 PFU/h$ & \cite{Mitchell2011} \\
-  $\delta$ & $16.7 hours$ & \cite{Mitchell2011} \\
-  $a$ & $20.9$ & ABM fits \\
-  $b$ & $70.8$ & ABM fits \\
-  \hline  
-\end{tabular}
-\caption{Supplemental model parameters.
-Most values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
-Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
-\label{table:supplement}
-\end{center}
-\end{table}
+Finally, the T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
+Thus it is reasonable to assume that there is a threshold of T cell secretion, beyond which the dynamics of the infection do not change.
+This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
+Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.
 

### 2014-10-06 paper/chemokine_main_revised.tex
middle of Steph's edits

@@
 \usepackage{color}
-\usepackage[usenames,dvipsnames]{xcolor}
+\usepackage[usenames,dvipsnames,table]{xcolor}
 \usepackage{ulem}
@@
 Candice Clay$^{2}$, 
+Judy Cannon$^{3}$,
 Melanie Moses$^{1}$, 
@@
 \\
+\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico, Albuquerque, NM, USA
+\\
 $\ast$ E-mail: Corresponding drew@cs.unm.edu
@@
 
-During the primary immune response, clearance of influenza in the lung requires the homing of activated CD8 T cells from regional lymph nodes to infected foci comprising a small fraction of the total lung.
-T cell navigation towards infected foci is undirected, but emigration into tissue is directed by local cytokine and chemokine production from infected epithelial cells.
-We use a spatially explicit agent-based model (ABM) to examine the efficiency of the recruitment of activated CD8 T cells by local chemokine signaling.
+Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
+Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.
+One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.
+Here, we focus on T cell navigating from the lymph nodes, through the vascular system to the foci, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.
 
-To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.
-These data were used to estimate chemokine production rates.
-The estimates were combined in the spatial model to test viral variation of T cell recruitment in the lungs.
+We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.
+The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.
+As plaques expand in size, infected regions become more isolated, making it difficult for T cells to cover the infection efficiently.
+Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.
+Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.
+However, the model parameters that collectively determine how efficiently T cells find and clear infected cells are also significant in the model.
+These model observations provide explanation of the pandemic H1N1 dynamics that have not been well explained by previous studies.
 
-The spatial nature of the ABM model reveals unique challenges to T cell recruitment not appreciated in standard differential equation models.
-Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.
+% To determine local chemokine production, avian H5N1, seasonal H1N1, and 2009 H1N1 pandemic influenza strains were used \textit{in vitro} to induce the secretion of CXCL10 (IP-10) and CCL5 (RANTES) in human bronchial epithelial cells.
+These data were fit to a differential equation model in order estimate viral and chemokine production rates.
+
+% Infected cells can become isolated in expanding plaques, impeding T cell search, even though the T cells could directionally migrate to low levels of chemokine.
 This spatially explicit model describes efficient T cell recruitment to small infected foci in the lung, but as foci become large T cell search becomes inefficient, thus emphasizing the importance of enhanced control early in infection.
 A key limitation imposed on T cells is illustrated by their failure to clear the pandemic H1N1 virus after day 6, when T cells became inefficient in finding infected cells in large foci.
-Finally, a sensitivity analysis highlights the viral parameters as the most significant in influencing the result of the infection.
 Thus, we conclude that viral dynamics, rather than chemokine dynamics, dominate the efficiency of T cell recruitment in determining whether the influenza infection is controlled.
@@
 
-%The adaptive immune response to influenza A infection in the lung requires a complex combination of different defensive mechanisms.
-CD8 T cells must navigate from the lymph node through the vascular system to the site of infection.
-%
-%
-%The high replication rate of 2009 pandemic H1N1 influenza presents unique challenges to the immune system when compared to other strains of influenza.
-Studies have attributed the high mortality rate of hosts infected with pH1N1 to the virus' high replication rate.
-%
-%We suggest pH1N1 virulence may only tell part of the story.
-Static analysis of viral production rates and plaque sizes ignore the spatial structure of the infection.
-We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain appropriate numbers of activated CD8 T cells.
-%
-%There are many mathematical models of the adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
-Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.
-Here we examine spatial aspects of the the CD8 T cell diaspora and recruitment to infected lung.
-Activated CD8 T cells searching for and homing into infected tissue do so in a complex environment.
-We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.
-In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
-%
-%%Therefore, we constructed a spatial agent-based model to examine how T cells are recruited to the site of infection, and once there, how they navigate over the plaque to contain the growing infection.
-Because our model strives to answer the specific questions regarding T cell recruitment and movement in the lung, we ignore other important aspects of the immune response such as...
-%
-%% Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
-The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.
-%
-%Previous work has shown that T cell search is likely aided by chemokine secretion by infected cells \cite{Banerjee2011}, but details of this process are unknown.
-To find secretion rates of chemokines, we infected human epithelial cells \textit{in virto} with one of three strains of influenza: avian aH5N1, seasonal sH1N1, and pandemic 2009 pH1N1.
-Wells were measured  for specific chemokines at specific time points over a period of two days.
-Two chemokines: IP-10 (CXCL10) and RANTES (CCL5) displayed elevated concentrations over the course of the infections.
-We extended a previous ODE model \cite{Mitchell2011} and fit it to the chemokine data to deduce per cell chemokine production rates.
-We then used the resulting values to inform our spatial model of T cell recruitment.
-%
-%Our results reveal unique spatial challenges to T cell search that arise from chemokine recruitment of T cells to the site of infection.
-Molecular diffusion of chemokines creates the chemical gradients that allow T cells to find the infection.
-Conversely, the coarse nature of the diffusion process creates pockets of high chemokine concentration.
-These pockets, or local maxima, have the unfortunate effect of recruiting and trapping T cells that navigate within their basin of attraction.
-This effect hinders the ability of T cells to properly cover the entire infection.
-We find that viruses with very high replication rates, such as pandemic H1N1, can create infections that spread across tissue faster than the chemokine signals can diffuse.
-In these instances, the chemical signal that is so crucial to T cell recruitment will actually hinder the immune system's search process.
-%
-%These results show interventions that increase chemokine concentrations and T cell counts may not help clear an infection.
-Furthermore, clearing the pH1N1 infection may be limited by the rate at which T cells can kill infected epithelial cells.
-This suggests that early interventions to infection, such as vaccination, may be the only way to avoid runaway infection.
+Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.
+Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood..Pathogenicity is a function of the virus and its interaction with the host immune response.
 
-The adaptive immune response induced during acute influenza A infection is an integrated complex combination of defense mechanisms.
-Understanding the behavior and interactions of each cellular component may lead to improved vaccines and strategies to control the infection and immunopathology.
-The innate immune response is critical for early control of viral replication, followed by antibody-mediated viral neutralization \cite{Kohlmeier2009a, Joo2008, Oslund2011}.
-Complete resolution of influenza pneumonia depends on antigen-specific CD8+ T effector cells \cite{Cerwenka1999, Kim2011}.
-The murine model has been critical in the reductive analysis of each phase and component, permitting temporally explicit descriptions of the induction phase in the regional lymph node \cite{Miller2003, Allan2006, Ingulli2009} where recruitment of clonal precursor CD8 T cells is highly efficient \cite{VanHeijst2009}.
-On approximately day 5 after infection, activated CD8 T cells are released from the secondary lymphoid tissue \cite{Miao2010a} and distributed throughout the body in a process known as the ‘lymphocyte diaspora’ \cite{Marshall2001}.
+Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.
+However, determining the relative contribution of these myriad experimental components is challenging.
+We address this challenege with computational modeling, which allows us to study the relative contributions of different aspects of the host immune response, and to account for strain-specific viral dynamics in the model.
 
-CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in infected lung.
+Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.
+We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.
+Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.
+Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
+Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.
+Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.
+We extend this work to examine how the unique properties of the different strains of influenza may affect this process.
+
+Here, we present a comprehensive agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
+Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models of to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008}.
+We use a combination of in vitro generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection to parameterize our model.
+Once parameterized with these data, the model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.
+These three strains provide a range of documented pathogenicities abd \textit{in vitro} replication efficiencies [TODO citation(s) with pathogenicity data of 3 strains] \cite{Mitchell2011}, which we study in the model.
+The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.
+Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.
+
+[TODO rewrite to be more specific] The model design allows us to compare between influenza strains how the combination of viral and immune system characteristics affect the strength of the given infection.
+The results give us specific insights into how these features [TODO list some] determine the pathogenicity of the different viruses.
+
+The spatial nature of the model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
+Because T cells use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
+The model suggests that the host cells' chemokine response may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.
+Finally, the sensitivity analysis isolated the effects of each parameter in the model.
+Based on this analysis, each parameter was classified as critical or not.
+We find that most parameters can be individually varied across a broad range of values without strong effect on model results.
+Further, the parameter classification is consistent across each broad parameter type.
+Specifically, chemokine-related parameters are stable within biologically plausible ranges.
+In contrast, parameters related to viral kinetics have a strong effect on model results.
+We conclude that influenza infection dynamics are dominated by the replication properties of the virus, confirming that the higher replication rate of 2009 pH1N1 influenza overwhelmed contributions of the chemotactic signals, leading to uncontrolled \textit{in silico} infection.
+
+% fred used 'obscurred' instead of overwhelmed
+
+% The model predicts the control and clearance of the aH5N1 and sH1N1 strains and predicts a runaway infection for the pH1N1 strain.
+These results are consistent with the previously reported replication rates of the three strains of influenza.
+
+%  Infected cells secrete chemokine molecules which then diffuse across the infected region.  This process creates isolated regions, or local maxima, of high chemokine concentrations.
+
+% While the local maxima do move and flatten, their speed is limited by their diffusion and decay rates.
+
+% You may title this section "Methods" or "Models". 
+% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
+% authors may use "Analysis" 
+\section*{Materials and Methods}
+
+% We are interested in how the interaction between the virus and the host's secondary immune response affects the virulence of the infection.
+Because these interactions are difficult and expensive to observe \textit{in vivo}, we create a model \textit{in silico} to represent the system in question.
+Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \cite{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).
+
+Our modeling methodology follows three steps.
+First, we define the system we are going to model.
+Second, we build the model with this definition in mind.
+Finally, we parameterize the model using empirical data, values from literature, and computational techniques.
+
+\subsection*{Model Definition}
+
+%Our model examines the secondary immune response to an influenza infection in a human lung.
+Specifically, we are interested in how antigen activated T cells efficiently navigate from the lymph node, through vascular system, and into the lung at the site of infection.
+
+The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).
+The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).
+Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.
+If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.
+T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.
+Once inside the endothelium, T cells follow the chemokine gradient to localized areas of maximum chemokine concentration.
+When a T cell contacts an infected epithelial cell it induces apoptosis (S1.4).
+Otherwise, infected cells continue to secrete virus for a fixed time and then die.
+
+The model is initialized with a single infected cell.
+After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (S2.1 and Table \ref{tab:strains}).
+Virus from expressing cells diffuses locally, infecting neighboring cells.
+Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI.
+After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.
+These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.
+
+\subsection*{Model Implementation}
+
+The model is implemented using the CyCells software \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).
+Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models [TODO add Catherine ref?] \cite{Miller2003, Allan2006, Ingulli2009}.
+Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
+
+Parameters that are independent of the chosen influenza strain are shown in Table \ref{tab:parameters}.
+Strain-specific values are shown in Table \ref{tab:strains}.
+Further details regarding the model definition are included in S1.2.
+
+[TODO cite Catherine or ourselves] In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}.
+\texttt{Healthy} cells remain unchanged unless infected by virus.
+Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay.
+\texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
+\texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells.
+\texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour.
+\texttt{Dead} cells take up space and do not regenerate over the course of an infection.
+Cell regrowth is not implemented as the rate of regrowth is unknown.
+
+T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.
+T cells emerge from the lymph node at five days post infection at a rate of 1,257 cells per hour.
+T cell travel time from the lymph node to a random location on the lung's surface is six seconds.
+If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.
+If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI.
+Circulating T cells decay exponentially with an average lifespan of three days.
+Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells.
+Chemotaxing T cells decay exponentially with an average lifespan of two hours.
+
+In addition to T cells, the model contains two types of particles: virus and chemokine.
+Both are produced at constant rates by expressing epithelial cells.
+Virus diffuses through the lung tissue, infecting healthy cells at a rate proportional to the virus concentration at the location of the cell.
+Chemokine diffuses across the tissue but has no direct effect beyond recruiting T cells.
+Both particle types decay exponentially.
+IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.
+
+\subsection*{Model Parameters}
+
+Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}.
+These values were taken from the literature when available.
+Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.
+All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.
+[TODO number of model parameters?]
+
+Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
+We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).
+T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2).
+T cell production is assumed to be constant after day 5 [TODO reference or justification for constant rate].
+
+\subsection*{Materials}
+
+In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.
+Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.
+Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
+Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
+Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.
+Quantitative viral culture was performed by standard plaque assay.
+Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
+Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).
+TNF activity is not incorporated in the model.
+Data for other chemokines and cytokines is not shown.
+
+\section*{Results}
+
+\subsection*{Model Results}
+
+Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
+For the avian and seasonal strains, all model runs cleared the virus by day 10.
+Conversely, each simulation of the 2009 pandemic influenza led to runaway infection.
+
+We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).  
+The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs.
+Consistent with earlier results \cite{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).
+
+In all studied strains, virus expansion declines at four days post infection (Fig.~\ref{fig:variance}), due to IgM activation.
+Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five post infection [TODO p.i.
+for post infection].
+aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six post infection, pH1N1 replication dominates the IgM and T cell response.
+These results agree with [TODO "previously reported" + citation] \textit{in vivo} differences among the three different strains.
+
+% The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).
+Because pH1N1 replicates more rapidly, the number of infected cells outstrips the ability of the T cells to find them.
+Spatial effects reveal the reasons for this in simulation images and video (Fig.~\ref{fig:cycells}, Video S4).
+
+\subsection*{Spatial Effects}
+
+% Spatial effects of viral replication and chemokine diffusion play an important role in both the spread and clearance of infections.  [TODO maybe place in intro]
+
+% Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.
+
+Plaque growth in the model is illustrated in Figure \ref{fig:cycells}.
+Because the virus particles are an order of magnitude larger than the chemokine moledules, they diffuse more slowly, according to the Stokes Einstein equation.
+However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).
+In the model these countervailing forces produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}).
+Before day 4 the plaque is dominated by active (incubating and secreting) cells, with a small percentage of dead cells (Panel A).
+Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque.
+T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B).
+By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells (Panel C).
+Alhtough not visible in the figure, plaque size and T cell number changes coninuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).
+
+% [TODO label fig. 5 with A B C...]
+
+In the aH5N1 infection, the plaque is densely populated with infected cells, allowing T cells to find secreting cells easily and clear the infection (Figs.~\ref{fig:plaquesize}A).
+However, secreting cells were not cleared in either the sH1N1 or pH1N1 simulations (Fig.~\ref{fig:variance}).
+In these runs, secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i.
+(Fig.~\ref{fig:plaquesize}).
+T cells continue to accumulate, but they arrive at a slower rate than the rate at which new cells become infected.
+Further, the newly arriving T cells are less efficient because the epithelial cells are more sparsely distributed in the plaque.
+
+Finally, the spatial nature of our model reveals that regions of concentrated chemokine diffuse more slowly than the ever-expanding infected cell and virus locations, thus misdirecting the chemotaxing T cells.
+It takes time for infected cells to begin producing chemokine, while the preexisting pockets of high chemokine density are slow to decay.
+Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
+This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse..
+
+% Taken together, the delayed response and low proportion of virus secreting cells contribute strongly to the failure of T cells clearing the infection.
+Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (Fig.
+S5), suggesting that the effect of high viral production rates dominate increased chemokine production rates.
+
+\subsection*{Sensitivity Analysis}
+
+% One advantage of computational modeling is the ability to manipulate the model in ways not available in the real system.
+Here, we have taken advantage of our model by adjusting each parameter in question while holding the rest constant.
+This sensitivity analysis allows us to make two main observations.
+First, we can directly examine how each component of the model affects the final behavior of the model.
+Second, we can determine which model parameters are sensitive (those that affect the model's behavior) and which are stable (those which do not affect the model's behavior).[TODO move somewhere or delete]
+
+We conducted a sensitivity analysis of sixteen model parameters to identify those that are most important for determining model outcomes.
+We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.
+T cell radius is known \cite{abbas2011cellular} and does not factor strongly into the spatial model's implementation.
+T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine ([TODO reference supplement]).
+The two onset parameters are known \cite{Diamond2003}.
+Viral production is an independent variable, as discussed.above, and determined by each of the three strains \cite{Mitchell2011}.
+
+% The results were examined individually and are discussed in depth in the supplement.
+
+Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.
+Results were compared for the size of the peak infection and the number of ucleared infected cells at day 10 p.i..
+Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{mostly stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold, \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.
+
+Table \ref{tab:sensitivity} shows that five parameters are sensitive: IgM strength, infection sensitivity, viral decay rate, viral diffusion, and the T cell secretion rate.
+In addition, section [TOOD reference] shows that viral production rate is a sensitive parameter.
+Four of the parameters (infection sensitivity, viral decay rate, viral diffusion rate, and viral production rate) are directly related to viral efficacy in isolation.
+Further, IgM strength, as coded in the model, directly affects viral decay rate and can be included in this group.
+The T cell secretion rate is the only sensitive parameter that does not directly control virus kinetics.
+None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model.
+This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.
+
+\subsection*{Window of Vulnerability}
+
+An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s (Table~\ref{tab:strains}).
+Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes.
+viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles, which can infect new cells.
+Thus, T cells alone cannot contain the pH1N1 infection in our model.
+By contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the Avian strain this number is 0.2.
+% Thus, viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
+
+Even if apoptosis was instantaneous (apoptosis time equals 0) the cell would still secrete virus for 10 minutes, producing 3 new virions (pH1N1).
+We tested this effect in the model by setting both relevant parameters to zero, and indeed with this assumption all three strains were cleared by the T cell response (Fig.~\ref{fig:instantkill}).
+We now ask how much T cell delay the model can tolerate and still clear a given viral strain.
+To answer this, we formulate the basic equations of the model in terms of $R=0$, the viral replication rate, and solve for the case where $R_0 < 1$.
+
+\begin{equation}
+\lim_{t \to \infty} R_0 = \frac{p}{2} * (T_{arr} + T_{kill} + T_{apop})
+\end{equation}
+
+% $R_0$ can simplistically be defined as the viral production rate multiplied by the time of production.
+Using this definition, the claculated $R_0$ of the sH1N1 virus is 1.6, suggesting growth, yet the actual infection is eventually cleared  by day 10 in our model.
+This can be attributed to the spatial effects of the ever-expanding ring of infection.
+
+%[TODO justify using the limit of p/2]
+
+This equation accounts for spatial effects of viral diffusion in the following way.
+Some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.
+This effect is minimal in the beginning when the edge of the plaque is tightly curved.
+As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus diffuses over healthy cells and 50\% over infected or dead cells.
+We correct for this in the equation by dividing $p$ by 2.
+[TODO This correction ignores the viral gradient in the plaque which pushes the new virus particles slightly outwards.]  Equation 1 explains how sH1N1 can be cleared in the model even though it produces 1.6 viral particles during the vulnerability window.
+Equation 1 calculates the effective $R_0$ for sH1N1 to be 0.8.
+
+------------ STOP HERE, FIX BELOW  ----------------
+
+Where $T_{arr}$ is the time it takes for a T cell to arrive at the location of an infected cell, $T_{kill}$ is the time it takes for the T cell to induce apoptosis, and $T_{apop}$ is the apoptosis time.
+Note that this equation calculates the number of virions that may be produced by an infected cell, but omits the number of cells that can subsequently be infected by the virus.
+A quick calculation of a virion's expected life span in the presence of IgM (2.4 hours) combined with the expected time it takes to infect a healthy cell (2 hours), results in a value greater than one.
+Given that virus is removed from the model when a cell becomes infected, each virus is limited to infecting one target cell, and thus this quantity does not affect the equation for $R_0$.
+Given these assumptions, we can then solve for $T_{arr}$ where $R_0 < 1$:
+
+\begin{equation*}
+\lim_{t \to \infty} T_{arr} < 2 / p - (T_{kill} + T_{apop})   \hspace{0.5cm} given  \hspace{0.5cm}  T_{kill} + T_{apop} < 2/p
+\end{equation*}
+
+Solving for $T_{arr}$ using our parameter set, we find that $T_{arr} = 9.1h$ for aH5N1, $T_{arr} = 18m$ for sH1N1, and that T cells alone will not be able to clear the infection in pH1N1 as  $T_{kill} + T_{apop} > 2/p$ for pandemic influenza given the established parameter values for $T_{kill}$ and $T_{apop}$.
+Given the result of Figure \ref{fig:instantkill}, we do know that the pH1N1 infection can be cleared if the apoptosis induction time and the apoptosis time are short enough.
+Solving for $T_{kill} + T_{apop} > 2/p$  given the viral production rate of pH1N1, we can hypothesize that to clear the pH1N1 infection, the two delay terms must total no more than 393 seconds if we assume instant T cell arrival.
+
+\section*{Discussion}
+
+We asked whether features of the T cell search affect the host's ability to clear the virus.
+We used a spatial model, developed in CyCells, to simulate the T cell search process.
+The model allowed us to observe how T cells find and then navigate over the infected region of the lung.
+The results of the model allow us to make several conclusions.
+
+First, our model confirms previous results \cite{Mitchell2011} showing that in the context of the T cell response, pandemic H1N1 influenza expands more quickly than seasonal H1N1 influenza, which in turn is expands more quickly than avian H5N1 influenza.
+Further, our model shows consistent clearance of the aH5N1 infection by day 10 post-infection, near clearance of the sH1N1 infection by day 10, and the complete failure to clear the pH1N1 infection by day 10.
+
+Second, the spatial nature of our model allowed us to observe unique challenges to T cell search in the epithelium caused by an infection that spreads more quickly than the chemokine gradient can diffuse.
+In these cases, the T cells become trapped in pockets of high chemokine concentration that lag behind the expanding infection.
+These pockets of chemokine also attract and trap newly arriving T cells, thus limiting the effectiveness of increasing the number of T cells in the system.
+
+Finally, we used our model to perform a sensitivity analysis on a majority of the free parameters.
+This allowed us to test the individual effects of each parameter on the system as a whole.
+The analysis shows that the virus kinetics determine the course of the infection, while the parameters related to the T cells and the chemokine do not have a strong effect on the result.
+This strongly suggests that flu therapeutic strategies may be most effective focusing on viral control rather than enhancing T cell responses.
+
+\subsection*{T Cell Recruitment}
+
+CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in the infected lung.
 Search problems in the lymph node have been simulated using live cell imaging data to provides reliable parameters of cell movement \cite{Vroomans2012}.
@@
 
-The second search, i.e.
-the process of recruiting activated CD8 effector cells to infected sites in the lung,  is different and not as well understood.
-In contrast to the dense lymph node volume, lung infection dynamics occur on the surface of epithelial cells, the size of the tissue layer being comparable to the size of a tennis court.
-Initial localization and extravasation may depend on a number of inflammatory signals but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
+Here, we have focused on the second search, i.e.
+the process of recruiting activated CD8 effector cells to infected sites in the lung, which is different and not as well understood.
+In contrast to the dense 3D lymph node volume, our model treats the lung as a 2D surface where the epithelial cell monolayer is approximately the size a tennis court.
+T cells must somehow rapidly locate the infection, which at day 5 post-infection is approximately the size of a United States dime.
+
+The initial localization and extravasation of T cells into the bronchial tissue may depend on a number of inflammatory signals, but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
 Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.
 Effector cells must localize to sites of viral replication \cite{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
-Chemotactic proteins play a major role in recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005} where recruited T cells mediate either protection or immunopathology.
-Infected epithelial cells secrete chemokines \cite{Chan2005} and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
-Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.
 
+In this paper, we assume T cells clime the chemokine gradient though ligand-receptor binding.
 The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.
@@
 
+The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
+Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+
 Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
 Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.
-Finally the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.
+Finally, the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.
+While T cell chemotaxis may not be proven, previous work has shown severe limitations to viral clearance in the absence of this ability \cite{Banerjee2011}.
+Furthermore, our sensitivity analysis shows that while classified as \textit{mostly stable}, reducing T cell velocity (and subsequently, the ability of T cells to climb gradients) hinders clearance of the non-trivial infections of sH1N1 and pH1N1 influenzas.
 
-There are many mathematical models of the entire adaptive response \cite{Miao2010, Saenz2010, Beltman2007, Handel2008, Zheng2008, Lee2009}, filling gaps with reasonable assumptions of cellular behavior in the absence of experimental support.
-Mathematical and computational models can examine each event in detail and narrow the range of possible behaviors.
-Here we examine the CD8 T cell diaspora and recruitment to infected lung in detail.
-Models using ordinary differential equations are able to predict events with temporal fidelity, but lack the ability to examine the spatial detail \cite{Beauchemin2006a, Bauer2009}.
-Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment.
-We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.
-In the model, CD8 T cells travel in vascular and lymphatic channels during part of their search, and respond to localized inflammatory signals to achieve compartmentalized cell-cell interactions.
+Another key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
 
-We ask how small foci of infected tissue, scattered through a very large space of uninfected tissue, can attract and retain limited numbers of activated CD8 T cells.
-Our model does not consider important features of tissue immunity such as T cell proliferation, resident T cells, and regulatory interactions with other cell phenotypes \cite{McGill2009}.
-The model does not aim to predict control of viral replication but only to compare simulated outcomes among three influenza strains arising from the efficacy of the chemotactic signals.
-
-Previously we developed an agent-based model of the lymphocyte diaspora focusing on the production and diffusion of a generic inflammatory signal without experimental data to calibrate the model \cite{Banerjee2011}.
-Here we refine that model and incorporate chemotactic protein data from in vitro cultures of human bronchial epithelial cells infected with three strains of influenza virus with different replication rates \cite{Mitchell2011} to focus on two features of searching in the large lung space.
-First, functional thresholds of the chemotactic proteins were surprisingly low in our model.
-Second there appears to be a threshold size of the infected focus above which the T cell search becomes inefficient.
-
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Materials and Methods}
-
-\subsection*{Materials}
-
-Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.
-Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
-Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
-Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.
-Quantitative viral culture was performed by standard plaque assay.
-Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
-Only IP-10, RANTES, and TNF$\alpha$ showed increases in production.
-TNF activity as not incorporated in the model.
-Data for other chemokines and cytokines is not shown.
-
-\subsection*{Computational Modeling}
-
-We implemented our spatial model using CyCells \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response.
-A simplified model of CD8 T cell activation and recirculation (Fig.~\ref{fig:systemchart}) was implemented in CyCells (Fig.~\ref{fig:modelchart}) and simulations measured efficiency of infection clearance for three different viruses.
-The lung was represented as a two-dimensional sheet of healthy epithelial cells (S1.2).
-Because we assume activated T cells descend at random into branches of the vascular network, T cells are introduced uniformly at random across the entire lung surface (Fig.~\ref{fig:systemchart}).
-If T cells detect cytokine signal on the local endothelium they remain in tissue and follow the chemotactic gradient to the site of infection.
-T cells that do not encounter cytokine recirculate to the lymph node where they reenter the bronchial vascular network.
-When a T cell contacts an infected epithelial cell at the site of infection it induces apoptosis.
-Unimpeded, infected cells secrete virus for 16.7 hours and then die.
-
-The simulation begins when a single cell in the center of the tissue is infected.
-After an incubation period, the infected cell begins secreting virus and chemokine according to the virus strain and chemokine type as determined by simulations reported in S2.1 and Table \ref{tab:strains}.
-Virus diffuses locally, infecting nearby cells, and continuing the cycle.
-Chemokine diffuses from secreting cells, creating a circle of stimulation around the infected region.
-After a five day delay to simulate lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a steady rate and travel through the vasculature to the tissue.
-Further details of the CyCells modeling environment are described in S1.1.
-
-Computational limitations force us to implement our model on the scale of a mouse.
-Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
-The surface area of a mouse lung is approximately 100$cm^2$.
-
-\subsection*{Model Definition}
-
-In the model (Fig.~\ref{fig:modelchart}), epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}.
-\texttt{Healthy} cells remain unchanged unless infected by virus.
-Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay.
-\texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
-\texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells.
-\texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour.
-\texttt{Dead} cells take up space and do not regenerate over the course of an infection.
-
-T cells are described by two states.
-\texttt{Circulating} T cells emerge from the lymph node at five days post infection and arrive at a random location on the lung's surface after six seconds of travel.
-If chemokine is not encountered, the T cell remains a circulating cell, returning to a new location in the lung after another six seconds.
-If a circulating cell encounters chemokine, it converts to \texttt{chemotaxing} and begins following the chemotactic gradient to the infected region.
-\texttt{Circulating} T cells decay exponentially with an average lifespan of three days.
-\texttt{Chemotaxing} T cells move continuously up the gradient through the two dimensional lung endothelium, inducing apoptosis in the \textit{expressing} epithelial cells.
-\texttt{Chemotaxing} T cells decay exponentially with an average lifespan of two hours.
-
-The model contains two kinds of particles: virus and chemokine.
-Both are produced at constant rates by \texttt{expressing} epithelial cells.
-Virus diffuses through the lung tissue, infecting healthy cells at a rate according to the local virus concentration.
-Chemokine diffuses across the tissue but has no direct effect beyond activating T cells.
-Both particle types decay exponentially.
-IgM is modeled by increasing the viral decay rate by a factor of ten after the third day.
-
-Parameters that are consistent between every model are shown in Table \ref{tab:parameters}.
-Strain-specific values are shown in Table \ref{tab:strains}.
-Further details regarding the model definition are included in S1.2.
-
-\section*{Results}
-
-\subsection*{Model Results}
-
-Unlike ordinary differential equation (ODE) and DDE models, which are deterministic, stochastic models such as CyCells can produce different results on different runs (Fig.~\ref{fig:variance}).
-Therefore, we ran th eCyCells model 50 times for each of the three strains of influenza: avian, seasonal, and 2009 pandemic (Fig.~\ref{fig:variance}).
-In every case, the model resulted in viral clearance by day 10 in both the avian and seasonal influenza strains.
-Conversely, every simulation of the 2009 pandemic influenza resulted in a runaway infection.
-
-We used $R^2$ values of individual runs versus the average of all runs in order to evaluate the variance of the model.
-The $R^2$ ($\pm$ one standard deviation) for avian, seasonal, and pandemic influenza are: 0.957 $\pm$ 0.150, 0.992 $\pm$ 0.013, and 0.998 $\pm$ .002 respectively (i.e.
-there was little variation between runs).
-The three strains show different levels of replication as found in \cite{Mitchell2011}.
-
-Overlaying multiple runs on a single plot reveals a growth rate decline at 4 days post infection, reflecting the addition of IgM.
-Subsequently, for each infection the number of infected cells declines quickly after day five post infection due to the CD8 T cell response.
-aH5N1 is cleared completely, sH1N1 is more slowly cleared, and pH1N1 clearance at day 6 post infection is halted as replication overtakes clearance.
-
-% The rapid production of the pH1N1 virus prevents the immune response from containing the infection (Fig.~\ref{fig:variance}C).
-Because pH1N1 replicates more rapidly, the number of infected cells outstrips the ability of the T cells to find them.
-Spatial effects reveal the reasons for this in simulation images and video (Fig.~\ref{fig:cycells}, Video S4).
-
-\subsection*{Sensitivity Analysis}
-
-We performed a sensitivity analysis of most of the model parameters due to the large number of free parameters in the model (Fig.
-S?).
-Each parameter included in the analysis was varied over several values while the rest of the parameters were held constant.
-This analysis allowed us to qualitatively classify parameters as sensitive or insensitive depending on how the model outcome changed as each parameter was varied.
-A more detailed description of how we classified the parameters is included in section S2.4.
-
-Five parameters stand out as having a strong effect on the model results.
-They are: IgM strength, viral decay rate, infection sensitivity, viral diffusion, and the T cell secretion rate.
-We also know from the results of the three virus strains that the viral production rate is also sensitive parameter.
-It is interesting to note that four of the parameters are directly related to the efficacy of the virus in isolation.
-Furthermore, IgM strength, as coded in the model, also directly affects the decay rate of the virus and can also be included in this group.
-Only the T cell secretion rate appears to have a strong effect on the model while not being directly related to the virus kinetics.
-Also of note, none of the other T cell parameters nor any of the chemokine parameters seem to have as strong of an effect on the model as the virus parameters.
-This supports the hypothesis that the virus dynamics dominate the course of an infection.
-
-\section*{Discussion}
-
-Our aim was to examine the spatially determined restrictions on the efficiency of the T cell search.
-Due to the absence of critical data, the inability to image the search in the lung (as was done in the lymph node), and the need for a tractable computation, the model simplified the adaptive immune response.
-We focused instead on comparing the behavior of thee influenza viruses.
-
-\subsection*{Spatial effects}
-
-Spatial effects of viral replication and chemokine diffusion play an important role in both the spreading and clearing of infections, imaged as plaque growth over time (Fig.~\ref{fig:cycells}).
-Free virus particles diffuse from virus secreting cells and infect healthy cells.
-Chemokine produced by infected cells attracts T cells to the infected cells.
-Although larger size of virus particles causes virus to diffuse much more slowly than chemokine molecules, chemokine decays more quickly.
-These countervailing effects result in similar spatial profiles for the two particle types (Fig.~\ref{fig:cycells}).
-Until day 4 the plaque is dominated by active (incubating and secreting) cells, whereas there is a smaller percentage of dead cells.
-Over time, cells in the plaque's interior die, and active cells form a decreasing proportion of the plaque.
-T cells arrive at day 5 and begin killing the virus-secreting cells.
-By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells.
-Continuous change in plaque size and T cell number is demonstrated for seasonal H1N1 virus (Videos S1-S3).
-
-In the aH5N1 infection, the plaque is dense, allowing T cells to find secreting cells easily, and infection is eliminated.
-However, secreting cells were not eliminated in either sH1N1 or pH1N1 simulations (Fig.~\ref{fig:variance}).
-Secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i.
-(Fig.~\ref{fig:plaquesize}).
-T cells still accumulate, but they arrive at a lower rate than the number of infected cells that are added to the plaque.
-Further, the regions of concentrated chemokine diffuse slowly, causing them to be located behind the expanding infected cell and virus locations.
-It takes time for newly secreting cells to produce chemokine while pockets of high chemokine density are slow to decay.
-Thus, T cells can fail to locate new regions of infected cells within the plaque.
-Taken together, the delayed response and low proportion of virus secreting cells prevent T cells from completely clearing the infection.
-Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate increased chemokine production rates.
-
-\subsection*{Temporal effects}
-
-The notion that T cell can't find infected cells in large plaques was examined in more detail by comparing the ratios of T cells to infected cells in Fig.~\ref{fig:plaquesize}.
-T cells are unable to kill virus incubating epithelial cells that have not yet presented antigen.
-At day 5.5, the ratio of antigen presenting secreting epithelial cells to the total plaque size is high (Fig.~\ref{fig:cycells}-\ref{fig:plaquesize}).
-By day 7, this ratio is very small for both the pH1N1 and sH1N1 strains, making it much more difficult for searching T cells to find the secreting cells.
-The high replication rate of pH1N1 exaggerates this effect (Fig.~\ref{fig:plaquesize}C).
-Video S4 illustrates this event.
-
-Because there is an interval between first secreted virus and the first arriving T cell, infection control cannot be completed by CD8 T cells alone.
-An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s \cite{Mitchell2011}.
-Assuming that secretion continues for one hour after apoptosis is initiated, the best a T cell could do is limit production to 18 new viral particles.
-Thus, T cells alone cannot contain the pH1N1 infection.
-In contrast, in a sH1N1 virus infection, T cells would limit a single infected cell to producting 1.4 viral particles in the one-hour window.
-Avian H5N1 virus-secreting cells produce only 0.2 viral particles in the interval after induced apoptosis.
-Viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
+Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, T cell chemotaxis, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
@@
 
-The model makes a number of simplifications to and deletions of elements in the innate and adaptive immune response, allowing us to build a tractable model where data was available.
-Antigen presentation and T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.
-Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
-The contribution of IgM antibody clearing free virus is represented as a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable as indicated by data-fitting models \cite{Wu2011}.
-Immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.
-Thus, our model is not intended to predict the exact infection size over time.
-Rather, our model allows us to qualitatively compare the infection kinetics of the three strains of influenza and to examine the response features that permit CD8 T cells to sense and contact infected target cells.
+Previous works, including our own, have successfully used ODE population models to great effect.
+Fitting the free paramters of ODE models to experimental data gives insights into the inner-workings of the system in question that are often not possible to obtain through direct observation and measurement.
+Maio et al.
+uses ODEs to examine how the T cell response coupled with IgG and IgM affect influenza infection in mice \cite{Miao2010}.
+Saenz et.
+al.
+examines the effect of the interferon response on influenza infection in equines \cite{Saenz2010}.
+Handel et.
+al.
+similarly uses ODE models to examine which immune responses dominate during the course of a influenza infection in mice \cite{Handel2008}.
+Lee et.
+al.
+create a very comprehensive model of the adaptive immune response, including dendritic cells, CD4+ T cells, CD8+ T cells, and B cells \cite{Lee2009}.
+Our previous work uses an ODE model oinfluenza infection matched to \textit{in vitro} experimental data to estimate per-cell viral production rates for the three different strains of influenza discussed in this paper: aH5N1, sH1N1, and pH1N1 \cite{Mitchell2011}.
 
-The use of an ABM has certain advantages over a spatially homogeneous differential equation model.
+These ODE models are effective approaches to estimate unmeasurable parameters of influenza infection and the subsequent immune response.
+Unfortunately, ODE models are not adept at modeling spatial effects of the system in question.
+Thus, other papers have successfully used spatial agent-based approaches to explore how spatial constraints affect the immune response to influenza infection.
+Both Beltman et.
+al.
+\cite{Beltman2007} and Zheng et.
+al.
+\cite{Zheng2008} successfully use unique spatial models to examine how viral peptides are presented to unactivated T cells in the lymph node.
+
+Here, we use spatial agent-based model (ABM) to examine the dynamics of how T cells locate and eradicate infected cells in the lung after they are activated in the lymph node.
+Our use of an ABM has certain advantages over a spatially homogeneous differential equation model.
 An ODE model assumes that any virus particle is capable of infecting any healthy cell.
@@
 This is limiting as can be seen in Figure~\ref{fig:cycells} where the early infection has a higher proportion of virus overlapping healthy cells.
+Furthermore, explicitly modeling T cell chemotaxis reveals unique challenges to viral clearance brought about by infections that spread more rapidly than the chemokine gradient can decay and diffuse.
+
+The main downside of an agent-based model is the increased computational complexity when compared to a population model.
+Rather than representing the agents as population totals, our model represents each one explicitly.
+To account for this, our model necessarily makes a number of simplifications to and deletions of elements in the innate and adaptive system, including antigen presentation, clonal expansion, antibody affinity maturation, and other non-T cell involvement.
+While these parameters remain to be tested in other models, our current model allow us to build a tractable model of the aspects of the immune response necessary to explore specific aspects of virus and immune system interactions.
+
+% Antigen presentation and T cell clonal expansion in secondary lymphoid organs is represented solely by the constant rate of emigration of activated CD8 T cells from regional lymph nodes.
+Virus-specific strain replication rates are represented as constant rates, and virus clearance is also constant.
+The contribution of IgM antibody clearing free virus is represented as a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
+All of these rates may in fact be time-variable as indicated by data-fitting models \cite{Wu2011}.
+The immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.
+These assumptions allow us to build a tractable model of the aspects of the immune response necessary to explore specific aspects of virus and immune system interactions.
+
+%\subsection*{Chemokine Directed T Cell Search}
+%
+%The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
+Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+%
+%A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
+For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+%
+%Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
 
@@
 
-We thank J.
-Cannon (UNM Pathology and Molecular Genetics \& Microbiology), F.
+We thank F.
 Asperti-Boursin (UNM Pathology and Computer Science), S.
@@
 \begin{center}
-\includegraphics[width=4in]{Figure_5}
+\includegraphics[width=\textwidth]{Figure_5}
  \end{center}
@@
  \label{fig:plaquesize}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=4in]{Figure_7}
+ \end{center}
+\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1 with instant cell death.} The model results }
+ \label{fig:instantkill}
 \end{figure}
@@
   Epithelial Cell Radius & 5 $\mu$m & \cite{Elbert1999} \\
-  T cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
-  T cell Production Rate & 1257/hour & \cite{Miao2010} \\ 
+  T Cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
+  T Cell Secretion Rate & 1257 cells/hour & \cite{Miao2010} \\ 
   T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
@@
   Chemokine Decay Rate & $3.8508\cdot10^{-4}$/s & 1\\
-  Infection Sensitivity Rate & 2 hour/virion &  2 \\
+  Infection Sensitivity Rate & 0.5 cell/hour$\cdot$virion &  2 \\
   Expression Time & 16.7 hours & \cite{Mitchell2011}$^3$ \\
@@
 
+\begin{table}
+\begin{center}
+\begin{tabular}{| c | l | c c c |}
+  \hline                        
+  Category & Parameter & Avian H5N1 & Seasonal H1N1 & Pandemic H1N1 \\
+  \hline
+  \multirow{3}{*}{Chemokine} & Chemokine Decay Rate & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable \\
+  & Chemokine Diffusion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable& \cellcolor{green!50}stable \\
+  & Chemokine Secretion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
+  \hline
+  \multirow{6}{*}{T Cell} & Circulation Time & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
+  & T Cell Kill Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
+  & T Cell Velocity & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
+  & Circulating Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable& \cellcolor{blue!30}mostly stable \\
+  & Found Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
+  & T Cell Secretion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  \hline
+  \multirow{3}{*}{Delay} & Apoptosis Time & \cellcolor{green!50}stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
+  & Expression Time & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change \\
+  & Incubation Time &  \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change \\
+  \hline 
+  \multirow{4}{*}{Virus} & IgM Strength & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  & Infection Sensitivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  & Viral Decay Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  & Viral Diffusion Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  \hline  
+\end{tabular}
+\caption{\textbf{Sensitivity Results:} The above parameters were varied over predetermined ranges in isolation, resulting in new model runs for every new value tested.
+The results of the sensitivity analysis were then qualitatively evaluated for each individual parameter.
+A model run's behavior was determined by examining the height of the peak of the infection at day 5 post-infection and the number of infected cells at day 10 post-infection.
+Each combination of influenza strain and free parameter was classified as belonging to one of four categories.
+Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{mostly stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.}
+\label{tab:sensitivity}
+\end{center}
+\end{table}
+% 7) IgM presence is abstracted by increasing viral decay by a factor of ten. $^\ddagger$Activated T Cell.  $^\ast$Focus of Infection.
+
 \end{document}

### 2014-10-06 paper/chemokine_supplement_revised.tex
middle of Steph's edits

@@
 
-T cells are added to the model at a constant rate of 1,257 per hour (Text S1 1.1).
+T cells are added to the model at a constant rate of 1,257 per hour (Text S1 2.1).
 Because the model environment is only a small portion of the entire lung (0.25\%), most of the T cells miss the model window and are not visually represented.
@@
 
-We estimate chemokine production rates, `r', by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq.
+We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq.
 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
@@
 $\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
-Table S1 summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to experimental data (see Results and Table 1).
+Table S2 summarizes population and parameter values and descriptions.
+Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
+The results of these fits are discussed in the main paper and shown in Table 1.
 
@@
 
-We calculate the rate of production ($\sigma$) of CD8 T cells using a differential equation model from \cite{Miao2010}.
+We calculate the rate of production, $\sigma$, of CD8 T cells using a differential equation model from \cite{Miao2010}.
 The equations model T cell production and subsequent search over an area of infected lung tissue.
@@
 Because this parameter is unknown, we simulated T cell sensitivity levels ranging over 10 orders of magnitude.
-Concentration-dependent behavior of cells responding to chemokine was the same over a wide range of concentrations from 0.01 pg/mL to 100 ng/mL (model variance is discussed in Text S1 2.1.).
+Concentration-dependent behavior of cells responding to chemokine was the same over a wide range of concentrations from 0.01 pg/mL to 100 ng/mL (model variance is discussed in the main text).
 Decreased recruitment was predicted only by supra-natural concentration sensitivity levels at and above 1 $\mu$g/mL.
@@
 Four models runs (both, IP-10, RANTES, and none) were performed for sH1N1 and pH1N1 strains and two runs (RANTES and none) for  aH5N1.
-The runs show how the presence and/or absence of specific chemokines affect the simulated immune response (Fig.~S\ref{fig:chemokine}).
+The runs show how the presence and/or absence of specific chemokines affect the simulated immune response (Fig.~\ref{fig:chemokine}).
 The lack of both chemokines leads to runaway infections in all three strains.
@@
 
-Model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Table~\ref{tab:parameters}).
-Due to the difficulty of estimating critical, unmeasurable values, we performed a sensitivity analysis for 16 parameters in order to test whether variation over an order of magnitude or more altered the infection outcome.
+A general sensitivity analysis serves two purposes.
+First, it allows us to observe how the model behavior changes as a single parameter is varied.
+Second, it allows us to examine which parameters affect the model strongly and which do not.
+Here we describe the sensitivity analysis of 16 of our parameters in detail.
+
+The model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Tables 1 and 2).
 Individual parameters were varied over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
@@
 
-We then categorize the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection.
-When evaluating the results, we consider model runs that asymptotically converge to be qualitatively stable and model runs that do not converge to be qualitatively divergent.
+We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection (Table 3).
+Parameters are presented along with their range using the following format:  \\
+
+\textbf{Parameter Name:} [ \textit{min | default | max} ].
+
+% When evaluating the results, we consider model runs that asymptotically converge to be qualitatively stable and model runs that do not converge to be qualitatively divergent.
 
@@
 Specifically, each parameter in this group seems to lie within a range that does not significantly affect the asymptotic behavior of the model.
+We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{mostly stable} parameters.
+The difference between the two being that \textit{mostly stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.
+
 Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).
 Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.
-The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal effects.
+The importance of the apoptosis time parameter is discussed in more detail in the main results section: Temporal Effects.
 
@@
 We have deemed these deviations acceptable on an individual basis as described below.
+\\
 
-Apoptosis time diverges slightly in the sH1N1 strain.
+\textbf{Apoptosis Time:} [ \textit{0 | 1h | 2h} ] - Apoptosis time describes how long it takes for a cell to complete apoptosis and transition to an inert dead state.
+
+We chose to limit the increase of the apoptosis to a factor of two because larger values would not be biologically relevant.  In the other direction we lowered the value all the way to zero.
+
+Adjusting the apoptosis time causes the model results to diverge slightly in the sH1N1 strain.
+Also of interest, reducing the apoptosis time to zero still did not allow the immune response to clear the pandemic infection.
 As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper.
+\\
+
+\textbf{Chemokine Decay Rate:} [ \textit{3e-6 Hz |  3e-4 Hz | 3e-2 Hz} ] - The chemokine decay rate defines how quickly the chemokine is removed from the system.
+A higher rate of decay corresponds to a quicker rate of removal.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.  These extremes are not biologically plausible but were tested anyway.
 
@@
 In this case, lack of decay allows the chemokine to diffuse homogeneously across the entire infection, removing the concentration gradient required by T cells to find the active areas of the infection.
-This confirms that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.
+This suggests that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.
 Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection.
+\\
+
+\textbf{Chemokine Diffusion Rate:}  [ \textit{3e-3 $\mu m^2/s$ | 0.3 $\mu m^2/s$ | 30 $\mu m^2/s$} ] - The chemokine diffusion rate regulates how quickly the chemokine molecules spread out over the infected region.
+A larger diffusion coefficient corresponds to a more rapid rate of spread.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
 
@@
 Thus the diffusion rate we used based on Stokes-Einstein and water viscosity was likely within a physiologic range.
+\\
+
+\textbf{Chemokine Secretion Rate:} [ \textit{1e-7 $pg/s\cdot cell$ |  1e-5 $pg/s\cdot cell$ | 1e-3 $pg/s\cdot cell$} ] - The chemokine secretion rate defines how much chemokine each infected cell secretes per second.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
 
@@
 Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence.
+\\
 
-T cell circulation times were tested over a very large range and only diverge at the very edge of that range.
-Because vascular and lymph circulation times are unknown, parameter values of 1 minute and 3 minutes in mice may be plausible and warrant further study.
+\textbf{T Cell Circulation Time:} [ \textit{1s | 6s | 1h} ] - T cell circulation time defines how long a T cell takes to recirculate through the vascular system when it does not encounter the infection as it passes through the lung.
+This value should correspond to the vascular circulation time of a mouse as we used a mouse-sized model for computational reasons.
 
-The T cell kill rate only diverges on the lower end in sH1N1.
+A value of 1s was chosen as the minimum due to computational constraints, yet this value is likely already too small to be biologically plausible.
+We extended the range to up to 600 times the default value to examine what happens when the circulation time is so long that the T cells are effectively removed from the simulation if they miss the focus of infection on their first pass through the lung.
+
+T cell circulation times were tested over a very large range and does diverge for high values.
+Classifying this parameter as \textit{mostly stable} may seem counterintuitive when examining the sensitivity figures.
+It is important to note that in this case, we tested extra values on the higher end of the baseline than other plots.
+While each strain does show divergence beyond a certain value, the intermediate values of 1s and 20s remain qualitatively similar to the main result, demonstrating a stable range surrounding the default value.
+Because vascular and lymph circulation times are unknown, parameter values of up to 1 minute and 3 minutes in mice may be plausible and warrant further study.
+\\
+
+\textbf{T Cell Expected Kill Time:} [ \textit{0m | 10m | 100m} ]  - The T cell expected kill time defines the T cell kill rate, the rate at which T cells probabilistically induce apoptosis of  virus-expressing epithelial cells in their immediate vicinity.
+A lower kill time corresponds to a higher kill rate, and vice-versa.
+
+We tested values only one order of magnitude higher than the default as it does not seem reasonable to expect a T cell to take over an hour to induce apoptosis of a virus-expressing cell.
+On the other side, we allowed T cells to induce apoptosis instantly to see if any delay in the induction of apoptosis was responsible for the inability for the immune response to clear the pandemic infection.
+
+The parameter only diverges on the higher end in sH1N1.
 The extreme value corresponds to a T cell needing 100 minutes to induce the apoptosis of a single infected cell and is not biologically reasonable.
 The intermediate value corresponds to a time of 33.3 minutes and is also unlikely.
+Of interest, removing the parameter entirely by setting the kill time to instant was still not enough to allow the immune response to clear the pandemic infection.
+\\
 
-T cell movement in tissue has been observed \cite{Egen2011}.  Thus, we consider the extreme values biologically implausible.
+\textbf{T Cell Speed:} [ \textit{3.6 $\mu m/h$ | 6 $\mu m/m$ | 10 $\mu m/s$} ] - T cell speed determines how fast T cells move over the epithelial monolayer.
 
-T cell decay parameters allow the model to diverge only in the most extreme cases.  Neither of these values are reasonable and the parameters show stable behavior otherwise.
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+
+Divergence does occur in sH1N1 for the extreme values on both ends (but not for the intermediate ones).
+T cell movement in tissue has been observed \cite{Egen2011}.
+Thus, we consider the extreme values biologically implausible.
+\\
+
+\textbf{Circulating T Cell Expected Age:} [ \textit{1h |  4d | 400d} ] - The expected age determines the decay rate of T cells in blood.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+
+T cell decay parameters allow the model to diverge only in the most extreme cases.
+Neither of these values are reasonable and the parameters show stable behavior otherwise.
+The default parameter appears to define a model as effective at clearing the infection as models with much larger values.
+\\
+
+\textbf{Chemotaxing T Cell Expected Age:} [ \textit{12m | 2h | 20h} ] - The expected age determines the decay rate of T cells in the epithelium.
+
+We tested values up to one order of magnitude larger and smaller than the default parameter value.
+
+T cell decay parameters allow the model to diverge only in the most extreme cases.
+Neither of these values are reasonable and the parameters show stable behavior otherwise.
+While the default of 2 hours may seem short, it appears to determine similar behavior as models with larger values.
 
@@
 
-Two parameters, viral incubation time and viral expression time, do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.
-This effect occurs because the T cells short circuit the lifespan of infected cells.
-Once T cells arrive, virus secreting cells no longer survive for their full lifespan.
-Rather, the secretion time becomes limited by the apoptosis time parameter.
-Because these terms refer to how and when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response.
+The two parameters in this grouping do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.
+Interestingly, these are the two parameters that define the transition delay between the different stages of epithelial cell infection.
+Because these terms refer to the timing of when infected cells release the virus particles, neither one directly affects the kinetics of the virus itself and thus the infection's behavior is still determined by the properties of the virus and the T cell response.
+\\
+
+\textbf{Incubation Delay:} [ \textit{5h | 10h | 20h} ] - Incubation delay describes how long an epithelial cell takes to transition to the virus-expressing state after it is initially infected.
+During this period, the cell, while infected, does not secrete virus.
+
+We tested values up to a factor of two larger and smaller than the original amount.
+
+While a longer incubation time does not change the number of virions released by each infected cell, it does slow down the spread of the infection.
+Conversely, a shorter incubation time allows the infection to spread more quickly by allowing the infected cell to release the new virus in to the system earlier.
+Thus, changes in the incubation time do affect the infection profile before the introduction of T cells.
+The infection dynamics change dramatically upon the arrival of T cells.
+As discussed, one of the main factors in determining whether or not the infection is cleared is the amount of new virus produced per infected cell.
+While this does depend somewhat on the overall size of the plaque, this value depends heavily on shutting down the expression of new virus, something T cells are unable to do during the cellular incubation phase as the virus-incubating cell has not expressed virus for T cells to detect.
+Thus, assuming the original plaque has not grown so large that it can no longer be effectively covered by the T cell response, the overall course of the infection will be unaffected by the length of the incubation period.
+\\
+
+\textbf{Expression Delay:} [ \textit{100m | 1,000m | 50h} ] - The expression delay is the amount of time a virus secreting cell will secrete virus in the absence of a T cell intervention.
+At the end of this time period, the expressing cell will transition to the \text{dead} state and remain inert for the remainder of the simulation.
+
+The parameter was increased up by a factor of 3 and down by a factor of 10.  In each case, the change is extreme enough such that the resulting values are not biologically likely.
+
+Adjusting this parameter seems to affect the model behavior up until the arrival of the T cell response, after which the model converges back to the baseline behavior.
+This effect occurs because once T cells arrive, virus secreting cells no longer survive for their full lifespan.
+Rather, the delay becomes limited by the apoptosis time parameter.
 
@@
 Only the T cell secretion rate and IgM parameters are not.
+\\
 
-IgM strength is not inserted until day 4 of the simulation, but once it does it directly affects the decay rate of the virus, and does so for the rest of the simulation.
+\textbf{IgM Strength:} [ \textit{1 | 10 | 1,000} ] - IgM strength simulates the presence of IgM at day 4 post-infection by increasing the decay rate of free virus particles by the given factor.
+
+We examined a very large range of values, from the lowest possible value of 1 (effectively removing IgM), to an increased decay rate of three orders of magnitude.
+
+While it is true that the extreme values have a large effect on the model behavior, it is more of interest that the intermediate values also seem to change the model behavior.
+This shows that there is no stable area around the default as is seen in the \textit{mostly stable} parameters.
 The higher the strength of this parameter, the less virus there is in the system.
 Because the true effective strength of IgM is unknown in the context of this model, it is not a target of our investigation.
+\\
 
-Viral decay has the exact same effect as the IgM strength parameter.  Its value directly determines how much virus remains in the system over the course of the infection.
+\textbf{Viral Decay Rate:} [ \textit{1e-7 Hz | 1e-5 Hz | 1e-3 Hz} ] - Viral decay determines how quickly free virus is removed form the system.  A larger decay rate corresponds to faster removal.
 
-Infection sensitivity describes the ability of the virus to infect healthy cells, yet its strength is directly related to how much virus is present in the area of the healthy cells.
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+
+Viral decay has the exact same effect as the IgM strength parameter.
+Its value directly determines how much virus remains in the system over the course of the infection.
+There is a strong relationship between the decay rate and the resulting infection profile across all values tested.
+\\
+
+\textbf{Infection Sensitivity:} [ \textit{12 m/virion | 2 h/virion | 20 h/virion} ] - The infection sensitivity describes the ability of the virus to infect healthy cells.
+A larger value corresponds to a virus particle needing less time to infect a nearby healthy epithelial cell.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+
+The strength of the infection sensitivity parameter is directly related to how much virus is present in the area of the healthy cells.
 Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.
-Thus, it behaves similarly to the IgM and viral decay parameters.
+Thus, it behaves similarly to the IgM and viral decay parameters as it directly determines the virulence of the virus.
+\\
+
+\textbf{Viral Secretion Rate:} [ \textit{5.4e-5 $PFU/s\cdot cel$ | 3.8e-4 $PFU/s\cdot cel$ | 5.1e-3 $PFU/s\cdot cel$} ] - The viral secretion rate determines how quickly virus-secreting epithelial cells secrete new virus particles.
+
+An explicit sensitivity analysis was not directly performed.  The range of the parameters are taken from the respective secretion rates of the aH5N1, sH1N1, and pH1N1 influenzas (Table 1).
 
@@
 Thus, it's affect is similar to the previous parameters.
+\\
 
-Viral diffusion does not change how much virus there is in the system.
+\textbf{Viral Diffustion Rate:} [ \textit{3e-4 $\mu m^2/s$ | 3e-2 $\mu m^2/s$ | 3 $\mu m^2/s$} ] - The viral diffusion rate controls how quickly the virus spreads out over the monolayer.
+
+We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+
+The viral diffusion rate does not change how much virus there is in the system.
 Rather, it determines how fast the virus may spread across the alveoli.
 While its mechanism is different, its effect may be even stronger than the previous parameters.
-By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may outpace the body's ability to generate a chemical gradient.
-Thus T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.
+By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may out-pace the body's ability to generate a chemical gradient.
+Thus, T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.
+\\
 
-Finally, the T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
-Thus it is reasonable to assume that there is a threshold of T cell secretion, beyond which the dynamics of the infection do not change.
+\textbf{T Cell Secretion Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell secretion rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.
+
+We tested values up to three times larger and ten times smaller than the default value.  Values beyond a three-fold increase were not computationally tractable.
+
+The T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
+Thus, it is reasonable to assume that there is a threshold of T cell secretion beyond which the dynamics of the infection do not change.
 This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
 Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.
+The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.
 
@@
 
-\section{Discussion}
-
-\subsection{Chemokine directed T cell search}
-
-The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
-Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
-Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
-The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
-
-A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
-The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
-For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
-This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
-
-Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, contribution of resident memory T cells and B cells, and effector cell lifespan.
+%\section{Discussion}
 
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figure_S1}
+ \includegraphics[width=\textwidth]{Figure_S-1}
  \end{center}
@@
 
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{Figure_S3}
+	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.}
+	\label{fig:asensitivity}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{Figure_S4}
+	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.}
+	\label{fig:ssensitivity}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{Figure_S5}
+	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.}
+	\label{fig:psensitivity}
+\end{center}
+\end{figure}
+
 %\begin{figure}[ht!]
 %\begin{center}
-%	\includegraphics[width=\textwidth]{Figure_S3}
+%	\includegraphics[width=\textwidth]{Figure_SX}
 %	\caption{\drew{\textbf{Effects of different T cell speeds.}  With the exception of sH1N1, T cell speeds show a small effect during the early phase of the T cell response.
@@
 \begin{center}
+\begin{tabular}{| r r r | c | r r r | c | r r r |}
+  \multicolumn{3}{c}{aH5N1} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{sH1N1} & \multicolumn{1}{c}{} & \multicolumn{3}{c}{pH1N1}\\
+  \cline{1-3} \cline{5-7} \cline{9-11}
+  Time & CXCL10 & CCL5 & & Time & CXCL10 & CCL5 & & Time & CXCL10 & CCL5 \\
+   \multicolumn{1}{|c}{\footnotesize{$hours$}} & \multicolumn{1}{c}{\footnotesize{$pg/mL$}} & \multicolumn{1}{c|}{\footnotesize{$pg/mL$}} & & \multicolumn{1}{c}{\footnotesize{$hours$}} & \multicolumn{1}{c}{\footnotesize{$pg/mL$}} & \multicolumn{1}{c|}{\footnotesize{$pg/mL$}} & & \multicolumn{1}{c}{\footnotesize{$hours$}} & \multicolumn{1}{c}{\footnotesize{$pg/mL$}} & \multicolumn{1}{c|}{\footnotesize{$pg/mL$}}\\
+  \cline{1-3} \cline{5-7} \cline{9-11}
+  0 & 440 & 1 & & 0 & 94 & 62 & & 0 & 179 & 45  \\
+  2 & 391 & 0 & & 2 & 24 & 45 & & 12 & 1,835 & 55 \\
+  8 & 1,723 & 1 & & 8 & 87 & 50 & & 16 & 1,349 & 46 \\
+  12 & 3,462 & 1 & & 10 & 296 & 57 & & 20 & 23,458 & 150 \\
+  16 & 4,618 & 2 & & 16 & 770 & 87 & & 24 & 12,073 & 93 \\
+  20 & 6,807 & 6 & & 20 & 3,048 & 134 & & 32 & 30,700 & 380 \\
+  24 & 8,164 & 16 & & 24 & 3,901 & 151 & & 48 & 19,814 & 1,224  \\
+  30 & $>$ 8,500 & 68 & & 28 & 8,261 & 150 &  &  &  &  \\
+  38 & $>$ 8,500 & 239 & & 32 & 12,935 & 245 &  &  &  &  \\
+  48 & $>$ 8,500 & 655 & & 38 & 26,970 & 780 &  &  &  &  \\
+  72 & $>$ 8,500 & 415 & & 42 & 24,120 & 695 &  &  &  &  \\
+  & & & & 72 & 21,441 & 1,665 &  &  &  &  \\
+    \cline{1-3} \cline{5-7} \cline{9-11}
+\end{tabular}
+\caption{Experimental data. }
+\label{tab:data}
+\end{center}
+\end{table}
+
+\begin{table}
+\begin{center}
 \begin{tabular}{|  l  l  l  |}
@@
 \caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
-The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.
+3).
 Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}

### 2014-10-06 apop_killrate.m

@@ -1,0 +1,27 @@
+% Instant kill rate and apoptosis plot
+
+figure('Position',[100 100 640 360])
+
+colors = [.625  0  .2 ;
+          1 .75  0 ;
+          .28  .625 0 ;
+          .1 .1 .9 ;
+          .75 .25 .75];
+
+plot(time, infect(a_apopandkill), 'color', colors(3,:), 'LineWidth', 2);
+hold on
+plot(time, infect(s_apopandkill), 'color', colors(4,:), 'LineWidth', 2);
+plot(time, infect(w_apopandkill), 'color', colors(1,:), 'LineWidth', 2);
+hold off
+
+% set(gca, 'YLim', [ymin ymax])
+set(gca, 'XLim', [0 10])
+set(gca, 'FontSize', 20)
+set(gca, 'FontWeight', 'bold')
+
+ylabel('Infected Cells')
+xlabel('Time (days)')
+lh = legend('aH5N1', 'sH1N1', 'pH1N1');
+set(lh, 'FontSize', 14)
+
+title('Instant Kill and Apoptosis');

### 2014-10-06 chemokine.m

@@ -1,3 +1,5 @@
+cd('/nfs/adaptive/drew/chemokine');
+
 % Chemokine Effects% Sensitivity Plot
 
 figure('Position',[100 500 1800 700])
@@ -11,15 +13,15 @@ colors = [.75  0  0 ;
           0  .5 0 ;
           0 0 .75];
       
-plot(time, infect(an20), 'color', colors(1,:), 'LineWidth', 2)
+plot(time, infect(an19), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-plot(time, infect(ar20), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(ar19), 'color', colors(2,:), 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 16)
 set(gca, 'YColor', 'k')
-set(gca, 'YLim', [1 1200])
-set(gca, 'YTick', [0 200 400 600 800 1000 1200])
+set(gca, 'YLim', [1 10000])
+set(gca, 'YTick', [0 2000 4000 6000 8000 10000])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
@@ -31,17 +33,17 @@ legend('None', 'RANTES', 2)
 
 subplot(1, 3, 2)
 
-plot(time, infect(sn20), 'color', colors(1,:), 'LineWidth', 2)
+plot(time, infect(sn19), 'color', colors(1,:), 'LineWidth', 2)
 hold on
-plot(time, infect(sr20), 'color', colors(2,:), 'LineWidth', 2)
-plot(time, infect(si20), 'color', colors(3,:), 'LineWidth', 2)
-plot(time, infect(sb20), 'color', colors(4,:), 'LineWidth', 2) 
+plot(time, infect(sr19), 'color', colors(2,:), 'LineWidth', 2)
+plot(time, infect(si19), 'color', colors(3,:), 'LineWidth', 2)
+plot(time, infect(sb19), 'color', colors(4,:), 'LineWidth', 2) 
 hold off
 
 set(gca, 'FontSize', 16)
 set(gca, 'YColor', 'k')
-set(gca, 'YLim', [1 1200])
-set(gca, 'YTick', [0 200 400 600 800 1000 1200])
+set(gca, 'YLim', [1 10000])
+set(gca, 'YTick', [0 2000 4000 6000 8000 10000])
[…]

### 2014-10-06 data.m

@@ -4,7 +4,8 @@ figure('Position',[100 500 1800 600])
 
 % --- Avian Data ---
 avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
-av  = [ 63, 125, 825, 2563,  988, 1650, 2838, 17375, 5775];
+av  = [ 100, 150, 825, 2563, 988, 1650, 2838, 17375, 5775];
+ave = [   0,  29, 229,  348, 166,  390,  112,  3062, 1261];
 
 ait = [120, 480, 600, 960, 1200, 1440, 1680,  1920,  2280,  2880]/60;
 ai  = [ 24,  87, 296, 770, 3048, 3901, 8261, 12935, 26970, 24120];
@@ -15,6 +16,7 @@ ar  = [ 45,  50,  57,  87,  134,  151,  150,  245,  780,  695];
 % --- Seasonal Data ---
 svt = [600,  720,  960,  1200,  1440,  1800,  2280,  2880]/60;
 sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
+sve = [  0, 1676, 1578, 15695, 17271,  2992,  2886,  3837];
 
 sit = [120,  480,  720,  960, 1200, 1440]/60;
 si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
@@ -29,6 +31,7 @@ sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
 % --- Swine Data ---
 wvt = [120,   720,   960,   1200,   1440,   1800,   2880]/60;
 wv  = [ 20, 64620, 84053, 254587, 665559, 814586, 654773];
+wve = [  0, 18294, 36080,  43752, 156340, 229980,  47662];
 
 wit = [480,  720,  960,  1200,  1440,  1920,  2880]/60;
 wi  = [ 75+1, 1835, 1349, 23458, 12073, 30700, 19814] - 75;
@@ -52,8 +55,9 @@ subplot(1, 3, 1)
 [haxes, hline1, hline2] = plotyy(avt, av, art, ar, 'semilogy', 'semilogy');
 hold on
 hline3 = plot(ait, ai);
-hline4 = plot(aft/60, aif);
-hline5 = plot(aft/60, arf);
+hline4 = errorbar(avt, av, ave*2, 'ob');
[…]

### 2014-10-06 plaquesize.m

@@ -5,46 +5,49 @@ figure('Position',[100 500 1800 500])
 colors = [.625  0  .2 ;
           1 .75  0 ;
           .28  .625 0 ;
-          .1 .1 .9];
+          .1 .1 .9 ;
+          .75 .25 .75];
 
 % Avian
 
 subplot(1, 3, 1)
-semilogy(time, infect(ar20)+ar20(:,5), 'color', colors(4,:), 'LineWidth', 2)
+semilogy(time, infect(ar19)+ar19(:,5), 'color', colors(4,:), 'LineWidth', 2)
 hold on
-semilogy(time, ar20(:,3), 'color', colors(2,:), 'LineWidth', 2)
-semilogy(time, ar20(:,4), 'color', colors(3,:), 'LineWidth', 2)
-semilogy(time, ar20(:,7), 'color', colors(1,:), 'LineWidth', 2)
+semilogy(time, ar19(:,3), 'color', colors(2,:), 'LineWidth', 2)
+semilogy(time, ar19(:,4), 'color', colors(3,:), 'LineWidth', 2)
+semilogy(time, ar19(:,7), 'color', colors(1,:), 'LineWidth', 2)
+semilogy(time, sb19(:,7)+sb19(:,8)+sb19(:,9), 'color', 'black', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 20)
 set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
-set(gca, 'YLim', [0 6000])
-set(gca, 'YTick', [1 10 100 1000])
+set(gca, 'YLim', [0 100000])
+set(gca, 'YTick', [1 10 100 1000 10000 100000])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
-ylabel('Infected Cell Count (log10)')
+ylabel('Cell Count (log10)')
 title('aH5N1 Influenza')
-lh = legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2);
-set(lh, 'FontSize', 16)
-set(lh, 'FontWeight', 'normal')
+%lh = legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T cells at FOI', 'Total T cells', 2);
+%set(lh, 'FontSize', 16)
[…]

### 2014-10-06 sensitivity.m

@@ -1,16 +1,19 @@
+cd('/nfs/adaptive/drew/chemokine');
+
 % Sensitivity Plot
 
-figure('Position',[100 500 1800 1200])
+figure('Position',[100 500 1800 600])
 
 % Avian
 
-subplot(2, 3, 1)
+subplot(1, 3, 1)
 
 %colors = winter(10);
 colors = [.625  0  .2 ;
           1  .75  0 ;
-          .28 .625  0 ;
-          .1  .1 .9 ;
+          .1  .1  .9 ;
+%          .28 .625  0 ;
+          .5  .5 .5 ;
           .5  .5 .5 ;
           .5  .5 .5 ;
           .5  .5 .5 ;
@@ -19,214 +22,214 @@ colors = [.625  0  .2 ;
           .5  .5 .5 ;
           .5  .5 .5 ];
 
-hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);      
+hline4 = plot(time, infect(ar20), 'color', colors(4,:), 'LineWidth', 2);
 hold on
-hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
-hline3 = plot(time, infect(ar19), 'color', colors(3,:), 'LineWidth', 2);
-hline4 = plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2); 
+plot(time, infect(ar21), 'color', colors(5,:), 'LineWidth', 2); 
 plot(time, infect(ar22), 'color', colors(6,:), 'LineWidth', 2); 
 plot(time, infect(ar23), 'color', colors(7,:), 'LineWidth', 2);
 plot(time, infect(ar24), 'color', colors(8,:), 'LineWidth', 2);
 plot(time, infect(ar25), 'color', colors(9,:), 'LineWidth', 2); 
 plot(time, infect(ar26), 'color', colors(10,:), 'LineWidth', 2);
-hline5 = plot(time, infect(ar20), 'color', colors(4,:), 'LineWidth', 2);
+hline2 = plot(time, infect(ar18), 'color', colors(2,:), 'LineWidth', 2);      
[…]

### 2014-10-06 speed.m

@@ -1,0 +1,95 @@
+% Sensitivity Plot
+
+figure('Position',[100 500 1800 600])
+
+% Avian
+
+subplot(1, 3, 1)
+
+%colors = winter(10);
+colors = [.2  .2   .8  ;
+          .35 .35   .725 ;
+          .50   .50  .65 ;
+          .65 .50   .50 ;
+          .725  .35  .35 ;
+          .8  .2   .2 ;
+          .1 .1 .1];
+
+plot(time, infect(speeda100), 'color', colors(1,:), 'LineWidth', 2);
+hold on
+plot(time, infect(speeda30), 'color', colors(2,:), 'LineWidth', 2);      
+plot(time, infect(speeda10), 'color', colors(3,:), 'LineWidth', 2);
+plot(time, infect(speeda3), 'color', colors(4,:), 'LineWidth', 2); 
+plot(time, infect(speeda1), 'color', colors(5,:), 'LineWidth', 2); 
+plot(time, infect(speeda03), 'color', colors(6,:), 'LineWidth', 2);
+plot(time, infect(speeda03_1), 'color', colors(7,:), 'LineWidth', 2);
+hold off
+
+set(gca, 'FontSize', 18)
+set(gca, 'FontWeight', 'bold')
+set(gca, 'YColor', 'k')
+set(gca, 'YLim', [1 600])
+set(gca, 'YTick', [0 100 200 300 400 500 600])
+set(gca, 'XLim', [0 10])
+set(gca, 'XTick', 0:2:10)
+xlabel('Days Post-Infection')
+ylabel('Infected Cell Count')
+title('aH5N1 Influenza')
+lh = legend('10 \mum/s', '3 \mum/s', '1 \mum/s', '0.3 \mum/s', '0.1 \mum/s',  '0.03 \mum/s', '0.03 \mum/s', 2);
+set(lh, 'FontSize', 14)
+
+% Seasonal
+
+subplot(1, 3, 2)
+
+plot(time, infect(speeds100), 'color', colors(1,:), 'LineWidth', 2);
+hold on
+plot(time, infect(speeds30), 'color', colors(2,:), 'LineWidth', 2);      
+plot(time, infect(speeds10), 'color', colors(3,:), 'LineWidth', 2);
+plot(time, infect(speeds3), 'color', colors(4,:), 'LineWidth', 2); 
[…]

### 2014-10-06 variance.m

@@ -32,14 +32,14 @@ hold on
 %plot(time, infect(vara33), 'color', [.75 0 0]) % .7988
 %plot(time, infect(vara35), 'color', [.75 0 0]) % .5881
 plot(time, aavg, '-k', 'LineWidth', 2)
-hline3 = plot(time, alow, 'Color', 'k', 'LineStyle', '--');
-plot(time, ahigh, 'Color', 'k', 'LineStyle', '--')
+hline3 = plot(time, alow, 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2);
+plot(time, ahigh, 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 16)
 set(gca, 'YColor', 'k')
-set(gca, 'YLim', [1 600])
-set(gca, 'YTick', [0 100 200 300 400 500 600])
+set(gca, 'YLim', [0 8000])
+set(gca, 'YTick', [0 2000 4000 6000 8000])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
@@ -76,14 +76,14 @@ plot(time, [infect(vars00) infect(vars01) infect(vars02) infect(vars03) infect(v
 hold on
 %plot(time, infect(vars16), 'color', [.75 0 0])
 plot(time, savg, '-k', 'LineWidth', 2)
-plot(time, slow, 'Color', 'k', 'LineStyle', '--')
-plot(time, shigh, 'Color', 'k', 'LineStyle', '--')
+plot(time, slow, 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2)
+plot(time, shigh, 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 16)
 set(gca, 'YColor', 'k')
-set(gca, 'YLim', [0 600])
-set(gca, 'YTick', [100 200 300 400 500 600])
+set(gca, 'YLim', [0 8000])
+set(gca, 'YTick', [2000 4000 6000 8000])
 set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
[…]

### 2015-03-31 paper/chemokine_main_revised.tex
RSI Submission

@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration}
+\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung}
 }
@@
 \\
-\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico, Albuquerque, NM, USA
+\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
 \\
@@
 Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
+Viral clearance depends in part on the ability of T cells to locate the infection in time to prevent uncontrolled spread of the virus.
+We use an agent-based spatial model to quantify how efficiently T cells locate and control different strains of influenza in the lung.
+We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature.
+The spatial model reveals challenges for T cell recruitment not apparent in standard differential equation models.
+The model shows that virus kinetics largely determine the course of the infection, while T cell chemokine parameters have less effect.
+
+%Further analysis of the model identifies the viral parameters as most significant and suggests possible mitigation strategies.
+
+% ---- 300 word abstract
+
+% Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
 Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.
 One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.
-Here, we focus on T cell navigating from the lymph nodes, through the vascular system to the foci, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.
+Here, we focus on T cell navigating from the lymph nodes through the vascular system to the infection, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.
 
-We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.
+%We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.
 The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.
-As plaques expand in size, infected regions become more isolated, making it difficult for T cells to cover the infection efficiently.
+As plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.
 Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.
 Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.
-However, the model parameters that collectively determine how efficiently T cells find and clear infected cells are also significant in the model.
-These model observations provide explanation of the pandemic H1N1 dynamics that have not been well explained by previous studies.
+Finally, we use the model parameters to determine an upper bound on T cell search time in the model.
+
+% ---- end 300 word abstract
 
@@
 
+\section*{Keywords}
+
+Agent-Based Model, Systems Biology, Immunology, Computational Biology, Virology
+
 \section*{Introduction}
@@
 Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.
-Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood..Pathogenicity is a function of the virus and its interaction with the host immune response.
+Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.
+Pathogenicity is a function of the virus and its interaction with the host immune response.
 
@@
 However, determining the relative contribution of these myriad experimental components is challenging.
-We address this challenege with computational modeling, which allows us to study the relative contributions of different aspects of the host immune response, and to account for strain-specific viral dynamics in the model.
+We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.
 
-Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.
+Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.
 We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.
@@
 
-Here, we present a comprehensive agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
-Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models of to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008}.
-We use a combination of in vitro generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection to parameterize our model.
-Once parameterized with these data, the model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.
-These three strains provide a range of documented pathogenicities abd \textit{in vitro} replication efficiencies [TODO citation(s) with pathogenicity data of 3 strains] \cite{Mitchell2011}, which we study in the model.
+Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
+Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.
+We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.
+The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.
+These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.
 The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.
@@
 
-[TODO rewrite to be more specific] The model design allows us to compare between influenza strains how the combination of viral and immune system characteristics affect the strength of the given infection.
-The results give us specific insights into how these features [TODO list some] determine the pathogenicity of the different viruses.
+% The model design allows us to compare between influenza strains how the T cell response is affected by the viral and immune system characteristics.
+The results show that efficient T cell search is crucial to viral clearance in the model and that the search process is dependent on more than just the replication rate of the virus, including the spatial landscape of the chemokine gradient and the location and density of virus-secreting cells.
 
-The spatial nature of the model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
-Because T cells use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
-The model suggests that the host cells' chemokine response may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.
-Finally, the sensitivity analysis isolated the effects of each parameter in the model.
-Based on this analysis, each parameter was classified as critical or not.
-We find that most parameters can be individually varied across a broad range of values without strong effect on model results.
+The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
+Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
+The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.
+Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.
+Based on this analysis, each parameter was classified as sensitive or not.
+We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.
 Further, the parameter classification is consistent across each broad parameter type.
@@
 In contrast, parameters related to viral kinetics have a strong effect on model results.
-We conclude that influenza infection dynamics are dominated by the replication properties of the virus, confirming that the higher replication rate of 2009 pH1N1 influenza overwhelmed contributions of the chemotactic signals, leading to uncontrolled \textit{in silico} infection.
+We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.
 
-% fred used 'obscurred' instead of overwhelmed
+% fred used 'obscured' instead of overwhelmed
 
@@
 
-Our modeling methodology follows three steps.
+% Our modeling methodology follows three steps.
 First, we define the system we are going to model.
@@
 
-The model is implemented using the CyCells software \cite{Warrender2006}, a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).
-Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models [TODO add Catherine ref?] \cite{Miller2003, Allan2006, Ingulli2009}.
+The model is implemented using the CyCells software \cite{Warrender2006} (\texttt{https://github.com/drewlevin/cycells}), a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).
+Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \cite{Miller2003, Allan2006, Ingulli2009}.
 Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
@@
 
-[TODO cite Catherine or ourselves] In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead}.
+In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \cite{bachem1996simulated, Beauchemin2005, Mitchell2011}.
 \texttt{Healthy} cells remain unchanged unless infected by virus.
@@
 T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.
-T cells emerge from the lymph node at five days post infection at a rate of 1,257 cells per hour.
+T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour.
 T cell travel time from the lymph node to a random location on the lung's surface is six seconds.
@@
 All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.
-[TODO number of model parameters?]
 
@@
 T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2).
-T cell production is assumed to be constant after day 5 [TODO reference or justification for constant rate].
+T cell production is assumed to be constant after day 5 \cite{MartIn-Fontecha2003}.
 
@@
 
-We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).  
+% We estimated variance across the runs using the $R^2$ statistic, $\pm$ one standard deviation: 0.957 $\pm$ 0.150 (avian), 0.992 $\pm$ 0.013 (seasonal), and 0.998 $\pm$ .002 (pandemic).
 The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs.
@@
 
-In all studied strains, virus expansion declines at four days post infection (Fig.~\ref{fig:variance}), due to IgM activation.
-Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five post infection [TODO p.i.
-for post infection].
-aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six post infection, pH1N1 replication dominates the IgM and T cell response.
-These results agree with [TODO "previously reported" + citation] \textit{in vivo} differences among the three different strains.
+We estimated variance across runs by calculating the standard deviation at each time point.
+For aH5N1, the maximum standard deviation was 37, which occurred at day 5.05 where the mean value was 259.
+For sH1N1, the maximum standard deviation was 242, which occurred at day 4.93 where the mean value was 2,607.
+For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018.
+
+In all studied strains, virus expansion declines at four days p.i.
+(Fig.~\ref{fig:variance}), due to IgM appearance.
+Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five p.i.
+aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six p.i., pH1N1 replication dominates the IgM and T cell response.
+These results agree with previously reported \textit{in vivo} differences among the three different strains \cite{Mitchell2011}.
+
+% Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
 
@@
 
-% Spatial effects of viral replication and chemokine diffusion play an important role in both the spread and clearance of infections.  [TODO maybe place in intro]
-
 % Free virus particles diffuse from virus secreting cells and infect healthy cells.  Chemokine produced by infected cells attracts T cells to the infected cells.
@@
 Plaque growth in the model is illustrated in Figure \ref{fig:cycells}.
-Because the virus particles are an order of magnitude larger than the chemokine moledules, they diffuse more slowly, according to the Stokes Einstein equation.
+Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly, according to the Stokes-Einstein equation.
 However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).
@@
 By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells (Panel C).
-Alhtough not visible in the figure, plaque size and T cell number changes coninuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).
-
-% [TODO label fig. 5 with A B C...]
+Although not visible in the figure, plaque size and T cell number changes continuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).
 
@@
 Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
-This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse..
+This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.
 
@@
 T cell radius is known \cite{abbas2011cellular} and does not factor strongly into the spatial model's implementation.
-T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine ([TODO reference supplement]).
+T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine (S2.3).
 The two onset parameters are known \cite{Diamond2003}.
-Viral production is an independent variable, as discussed.above, and determined by each of the three strains \cite{Mitchell2011}.
-
-% The results were examined individually and are discussed in depth in the supplement.
+Viral production is an independent variable, as discussed above, and determined by each of the three strains \cite{Mitchell2011}.
 
 Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.
-Results were compared for the size of the peak infection and the number of ucleared infected cells at day 10 p.i..
-Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{mostly stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold, \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.
+Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.
+Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.
 
-Table \ref{tab:sensitivity} shows that five parameters are sensitive: IgM strength, infection sensitivity, viral decay rate, viral diffusion, and the T cell secretion rate.
-In addition, section [TOOD reference] shows that viral production rate is a sensitive parameter.
-Four of the parameters (infection sensitivity, viral decay rate, viral diffusion rate, and viral production rate) are directly related to viral efficacy in isolation.
-Further, IgM strength, as coded in the model, directly affects viral decay rate and can be included in this group.
-The T cell secretion rate is the only sensitive parameter that does not directly control virus kinetics.
-None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model.
+Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.
+In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.
+The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.
+Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.
+None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
 This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.
 
-\subsection*{Window of Vulnerability}
+\subsection*{Window of Control}
+
+Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).
+We asked whether T cells could play a role by assessing multiple factors that control the overall T cell response.
+Here, we define the window of control as the maximum time virus-secreting cells may produce virus while still allowing the infection to be cleared.
+This window can be defined as the combination of the the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).
 
 An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s (Table~\ref{tab:strains}).
-Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes.
-viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles, which can infect new cells.
-Thus, T cells alone cannot contain the pH1N1 infection in our model.
-By contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the Avian strain this number is 0.2.
+Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes of T cell arrival, viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles that can infect new cells.
+In contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the aH5N1 strain this number is 0.2.
 % Thus, viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
 
-Even if apoptosis was instantaneous (apoptosis time equals 0) the cell would still secrete virus for 10 minutes, producing 3 new virions (pH1N1).
-We tested this effect in the model by setting both relevant parameters to zero, and indeed with this assumption all three strains were cleared by the T cell response (Fig.~\ref{fig:instantkill}).
+Even if T cell arrival and apoptosis time were instantaneous, a cell would still secrete virus for 10 minutes ($T_{kill}$), producing 3 new virions in the case of pH1N1.
+Only in the case where both $T_{kill}$ and $T_{apop}$ are simultaneously reduced can the pandemic infection be cleared.
+We confirmed this effect in the model by first setting both relevant parameters (apoptosis time and T cell kill time) to zero, and as expected all three strains were cleared (Fig.~\ref{fig:instantkill}).
 We now ask how much T cell delay the model can tolerate and still clear a given viral strain.
-To answer this, we formulate the basic equations of the model in terms of $R=0$, the viral replication rate, and solve for the case where $R_0 < 1$.
+To answer this, we formulate the basic equations of the model in terms of $R_0$, the viral replication rate, and solve for the case where $R_0 < 1$.
 
 \begin{equation}
-\lim_{t \to \infty} R_0 = \frac{p}{2} * (T_{arr} + T_{kill} + T_{apop})
+R_0 = p \times (T_{arr} + T_{kill} + T_{apop}) \times E
 \end{equation}
+
+\noindent where $p$ is the secretion rate of the virus and $E$ is viral effectiveness, i.e.
+the proportion of virus that infects cells.
+Equation 1 assumes each free virus particle can infect no more than one healthy epithelial cell.
+
+%%%
+% Note that this equation calculates the number of virions that may be produced by an infected cell, but omits the number of cells that can subsequently be infected by the virus.
+A quick calculation of a virion's expected life span in the presence of IgM (2.4 hours) combined with the expected time it takes to infect a healthy cell (2 hours), results in a value greater than one.
+Given that virus is removed from the model when a cell becomes infected, each virus is limited to infecting one target cell, and thus this quantity does not affect the equation for $R_0$.
 
@@
 This can be attributed to the spatial effects of the ever-expanding ring of infection.
+%%%
 
-%[TODO justify using the limit of p/2]
+$p$, $T_{kill}$, and $T_{apop}$ are model parameters (Table \ref{tab:parameters}).
+$E$ incorporates both spatial effects of viral diffusion and temporal effects of viral decay.
+As the infected plaque expands, some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.
+This effect is minimal in the beginning when the edge of the plaque is small and tightly curved.
+As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus could be expected to diffuse over healthy cells and 50\% over infected or dead cells.
+In the presence of healthy cells, virions infect the cells at a rate of $0.5/h$ and decay at a rate of $0.42/h$ in the presence of IgM after day 4 p.i.
+(Table \ref{tab:parameters}).
+This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (S2.6).
+Solving for $T_{arr}$ where $R_0 < 1$ gives:
 
-This equation accounts for spatial effects of viral diffusion in the following way.
-Some secreted virus diffuses to cells that are already infected or dead (the existing plaque) and do not contribute to new infections.
-This effect is minimal in the beginning when the edge of the plaque is tightly curved.
-As the plaque expands, its radius grows and the edge approaches the limiting case of a straight line when 50\% of the virus diffuses over healthy cells and 50\% over infected or dead cells.
-We correct for this in the equation by dividing $p$ by 2.
-[TODO This correction ignores the viral gradient in the plaque which pushes the new virus particles slightly outwards.]  Equation 1 explains how sH1N1 can be cleared in the model even though it produces 1.6 viral particles during the vulnerability window.
-Equation 1 calculates the effective $R_0$ for sH1N1 to be 0.8.
+\begin{equation}
+\lim_{t \to \infty} T_{arr} < \frac{1}{pE} - (T_{kill} + T_{apop})
+\end{equation}
 
------------- STOP HERE, FIX BELOW  ----------------
+Using the model parameters, this equation suggests that $T_{arr} < 18h$ for aH5N1 and $T_{arr} < 93m$ for sH1N1.  pH1N1, however, cannot be cleared for any value of $T_{arr}$ in this scenario.
 
-Where $T_{arr}$ is the time it takes for a T cell to arrive at the location of an infected cell, $T_{kill}$ is the time it takes for the T cell to induce apoptosis, and $T_{apop}$ is the apoptosis time.
-Note that this equation calculates the number of virions that may be produced by an infected cell, but omits the number of cells that can subsequently be infected by the virus.
-A quick calculation of a virion's expected life span in the presence of IgM (2.4 hours) combined with the expected time it takes to infect a healthy cell (2 hours), results in a value greater than one.
-Given that virus is removed from the model when a cell becomes infected, each virus is limited to infecting one target cell, and thus this quantity does not affect the equation for $R_0$.
-Given these assumptions, we can then solve for $T_{arr}$ where $R_0 < 1$:
-
-\begin{equation*}
-\lim_{t \to \infty} T_{arr} < 2 / p - (T_{kill} + T_{apop})   \hspace{0.5cm} given  \hspace{0.5cm}  T_{kill} + T_{apop} < 2/p
-\end{equation*}
-
-Solving for $T_{arr}$ using our parameter set, we find that $T_{arr} = 9.1h$ for aH5N1, $T_{arr} = 18m$ for sH1N1, and that T cells alone will not be able to clear the infection in pH1N1 as  $T_{kill} + T_{apop} > 2/p$ for pandemic influenza given the established parameter values for $T_{kill}$ and $T_{apop}$.
-Given the result of Figure \ref{fig:instantkill}, we do know that the pH1N1 infection can be cleared if the apoptosis induction time and the apoptosis time are short enough.
-Solving for $T_{kill} + T_{apop} > 2/p$  given the viral production rate of pH1N1, we can hypothesize that to clear the pH1N1 infection, the two delay terms must total no more than 393 seconds if we assume instant T cell arrival.
+The above calculation suggests a combined transition point for sH1N1 at 163 minutes ($R_0 = 1$ for $T_{arr}+T_{kill}+T_{apop}=163m$), which is consistent with the sensitivity analysis (Fig.~S4: Apoptosis Time: 90m - 120m) and suggests that the modeled $T_{arr}$ is between 33 and 63 minutes for sH1N1.
+These data show that the balance between viral production and T cell response is a key factor in clearance of infection: pH1N1 can theoretically be cleared with instantaneous T cell responses, but in actuality viral production exceeds T cell response time and this leads to uncontrolled infection.
 
@@
 
-We asked whether features of the T cell search affect the host's ability to clear the virus.
-We used a spatial model, developed in CyCells, to simulate the T cell search process.
-The model allowed us to observe how T cells find and then navigate over the infected region of the lung.
-The results of the model allow us to make several conclusions.
+We used a spatial model to study how T cell search in the lung affects the host's ability to clear virus.
+CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in the infected lung.
+Search problems in the lymph node have been simulated using live cell imaging data to provide reliable parameters of cell movement \cite{Vroomans2012}.
+Our model focused on the second search, i.e.
+the process of recruiting activated CD8 effector cells to infected sites in the lung, which is not as well understood.
+In contrast, the dense 3D lymph node volume, our model approximates a mouse lung as a 2D surface where the epithelial cell monolayer is 100 $cm^2$ (only the area containing the FOI is modeled explicitly).
+T cells must somehow rapidly locate the infection, which at day 5 p.i.
+is approximately 0.5 $mm$ in diameter.
+Our model revealed four main conclusions regarding chemokine aided T cell search in the lung.
 
-First, our model confirms previous results \cite{Mitchell2011} showing that in the context of the T cell response, pandemic H1N1 influenza expands more quickly than seasonal H1N1 influenza, which in turn is expands more quickly than avian H5N1 influenza.
-Further, our model shows consistent clearance of the aH5N1 infection by day 10 post-infection, near clearance of the sH1N1 infection by day 10, and the complete failure to clear the pH1N1 infection by day 10.
+First, confirming earlier results \cite{Mitchell2011}, the model consistently clears the aH5N1 strain by day 10 p.i., contains sH1N1 by day 10 p.i., and fails to clear the pH1N1 infection.
+Second, the model revealed spatial constraints on T cell search when the infection spreads more quickly than the chemokine gradient can diffuse.
+In these cases, T cells become trapped in areas of high chemokine concentration which lag behind the expanding infection.
+High concentrations of chemokine also attract and trap arriving T cells, thus limiting the direct benefit of increasing numbers of T cells.
+Third, the sensitivity analysis tested the individual effects of each parameter on the model's behavior.
+The analysis shows that virus kinetics largely determine the course of the infection, while T cell and chemokine parameters do not have a strong effect.
+Finally, we examined the window of control which suggests that pandemic influenza could not be cleared even if apoptosis were instantaneous.
 
-Second, the spatial nature of our model allowed us to observe unique challenges to T cell search in the epithelium caused by an infection that spreads more quickly than the chemokine gradient can diffuse.
-In these cases, the T cells become trapped in pockets of high chemokine concentration that lag behind the expanding infection.
-These pockets of chemokine also attract and trap newly arriving T cells, thus limiting the effectiveness of increasing the number of T cells in the system.
+% \subsection*{T Cell Recruitment}
 
-Finally, we used our model to perform a sensitivity analysis on a majority of the free parameters.
-This allowed us to test the individual effects of each parameter on the system as a whole.
-The analysis shows that the virus kinetics determine the course of the infection, while the parameters related to the T cells and the chemokine do not have a strong effect on the result.
-This strongly suggests that flu therapeutic strategies may be most effective focusing on viral control rather than enhancing T cell responses.
+% The naive T cell must encounter on average hundreds of irrelevant cells before it contacts the loaded DC, but the distance traveled is relatively short.
 
-\subsection*{T Cell Recruitment}
-
-CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in the infected lung.
-Search problems in the lymph node have been simulated using live cell imaging data to provides reliable parameters of cell movement \cite{Vroomans2012}.
-The T precursor cell must encounter on average hundreds of irrelevant cells before it contacts the loaded DC, but the distance traveled is relatively short, and the direction is likely guided by chemokine signals.
-
-Here, we have focused on the second search, i.e.
-the process of recruiting activated CD8 effector cells to infected sites in the lung, which is different and not as well understood.
-In contrast to the dense 3D lymph node volume, our model treats the lung as a 2D surface where the epithelial cell monolayer is approximately the size a tennis court.
-T cells must somehow rapidly locate the infection, which at day 5 post-infection is approximately the size of a United States dime.
-
-The initial localization and extravasation of T cells into the bronchial tissue may depend on a number of inflammatory signals, but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
+% The initial localization and extravasation of T cells into the bronchial tissue may depend on a number of inflammatory signals, but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
 Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.
@@
 
-In this paper, we assume T cells clime the chemokine gradient though ligand-receptor binding.
+% In this paper, we assume T cells climb the chemokine gradient though ligand-receptor binding.
 The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.
@@
 The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
+There is clear evidence from multiple models that chemokines and chemokine receptors are required for effector T cell localization to infected tissues \cite{Christensen2004, Christensen2006, Dawson2000, DeLemos2005, Fadel2008, Gadhamsetty2014, Groom2011a, Groom2011, Hsieh2006, Klein2005, Kohlmeier2009, Kohlmeier2011, Thapa2008, Wareing2004, Wuest2008, Pawelek2012}.
+However, the actual effect of chemokines on effector T cells in tissues is still largely unknown.
 Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
-Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
-The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} and may better approximate chemokine levels in tissue.
+We have determined the level of chemokine released by infected bronchial epithelial cells (Table~\ref{tab:strains}) and used these as the best approximation for chemokine levels in tissues.
+Our model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+The sensitivity analysis classifies the chemokine secretion rate as stable, suggesting the total amount of chemokine is not a significant factor in determining viral clearance in our model.
 
 Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
-Following arrival at the infected site, CD8 T cells secrete cytokines when mononuclear cells are present but in their absence kill infected epithelial cells only by contact cytolysis \cite{Hufford2011}.
-Finally, the exit of T cells from tissue is tightly regulated to keep effector cells from leaving or dying prematurely \cite{Jennrich2012, Richter2007} but effective life-span is unknown.
 While T cell chemotaxis may not be proven, previous work has shown severe limitations to viral clearance in the absence of this ability \cite{Banerjee2011}.
-Furthermore, our sensitivity analysis shows that while classified as \textit{mostly stable}, reducing T cell velocity (and subsequently, the ability of T cells to climb gradients) hinders clearance of the non-trivial infections of sH1N1 and pH1N1 influenzas.
+Our sensitivity analysis shows that reducing T cell velocity (and consequentially, the ability of T cells to climb gradients) hinders clearance of sH1N1 and pH1N1 strains.
+These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.
 
-Another key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+%Our model of T cell search is consistent with the literature and sheds light on T cell movement in the epithelium.
+
+% Another key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
 The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
@@
 
-Future work with spatially explicit modeling should explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, T cell chemotaxis, contribution of resident memory T cells and B cells, and effector cell lifespan.
+% \subsection*{Modeling Methodology}
 
-\subsection*{Modeling Methodology}
-
-Previous works, including our own, have successfully used ODE population models to great effect.
+% Previous works, including our own, have successfully used ODE population models to great effect.
 Fitting the free paramters of ODE models to experimental data gives insights into the inner-workings of the system in question that are often not possible to obtain through direct observation and measurement.
@@
 create a very comprehensive model of the adaptive immune response, including dendritic cells, CD4+ T cells, CD8+ T cells, and B cells \cite{Lee2009}.
-Our previous work uses an ODE model oinfluenza infection matched to \textit{in vitro} experimental data to estimate per-cell viral production rates for the three different strains of influenza discussed in this paper: aH5N1, sH1N1, and pH1N1 \cite{Mitchell2011}.
+Our previous work uses an ODE model influenza infection matched to \textit{in vitro} experimental data to estimate per-cell viral production rates for the three different strains of influenza discussed in this paper: aH5N1, sH1N1, and pH1N1 \cite{Mitchell2011}.
 
-These ODE models are effective approaches to estimate unmeasurable parameters of influenza infection and the subsequent immune response.
+% These ODE models are effective approaches to estimate unmeasurable parameters of influenza infection and the subsequent immune response.
 Unfortunately, ODE models are not adept at modeling spatial effects of the system in question.
@@
 
-Here, we use spatial agent-based model (ABM) to examine the dynamics of how T cells locate and eradicate infected cells in the lung after they are activated in the lymph node.
+% [MOVED UP] Here, we use spatial agent-based model (ABM) to examine the dynamics of how T cells locate and eradicate infected cells in the lung after they are activated in the lymph node.
 Our use of an ABM has certain advantages over a spatially homogeneous differential equation model.
@@
 
-The main downside of an agent-based model is the increased computational complexity when compared to a population model.
+% A drawback of an agent-based model is the increased computational complexity when compared to a population model.
 Rather than representing the agents as population totals, our model represents each one explicitly.
@@
 While these parameters remain to be tested in other models, our current model allow us to build a tractable model of the aspects of the immune response necessary to explore specific aspects of virus and immune system interactions.
+
+% Finally, while our model is not meant to be biologically correct in the sense it does not attempt to replicate the entire immune response, the results of our sensitivity analysis are important for future biological modeling and experimental design.
+Empirical verification of the model's sensitive parameters (T cell production rate, response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
 
@@
 
-\subsection*{Conclusions}
+% \subsection*{Conclusions}
 
-Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and are difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.
-Because CD8 T cells presumably find infected cells by climbing a chemokine gradient, we see T cell clustering at local maxima of chemokine concentration, a possible explanation for why T cells do not increase in effectiveness as their numbers increase.
-We also see that T cell clustering persists after all virus-secreting cells have been eliminated.
-The local chemokine maxima takes time to diffuse and decay so that T cells can climb the gradient to a new maximum.
-Furthermore, we can see that infected cells are more disperse as infection size grows.
-Because T cells are clustered they cannot cover the increasing plaque effectively.
-Finally, our sensitivity analysis suggests that viral dynamics are the dominant factor in determining the course of an infection.
-Changes in the chemokine and T cell parameters do not have a strong effect on the infection result.
+% Our ABM reveals spatial patterns and dynamics that are absent in differential equation models and are difficult or impossible to observe in \textit{in vitro} and \textit{in vivo} systems discussed earlier.
 
-These model observations provide explanations for the pH1N1 resurgence that would be obscured without the visualization tools provided by the ABM.
-\textit{In vivo} pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation.
-The behavior of searching T cells described in this paper can enhance future global immune response models.
+Most earlier studies of T cell motility and search have focused on the lymph node.
+Yet, circulation through the peripheral tissue is critical to clearance \cite{Pawelek2012, Ganusov2014}.
+We studied the dynamics of T cell movement in tissue, accounting for observed phenotypic differences in three strains of influenza.
+The spatially explicit model was parameterized using \textit{in vitro} chemokine data.
+
+Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \cite{Handel2008, Lee2009, Miao2010, Mitchell2011, Saenz2010}.
+The use of an ABM can complement spatially homogeneous differential equation models \cite{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.
+In this study, we examined how chemokine-directed T cell search contributes to infection clearance.
+Explicitly modeling T cell chemotaxis revealed unique challenges to viral clearance brought about by infections that spread more rapidly than the chemokine gradient could decay and diffuse.
+Furthermore, our model was also able to account for spatial constraints of viral diffusion and infection.
+Figure \ref{fig:cycells} illustrates how new virus diffuses over cells that are no longer able to be infected.
+ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
+Our model shows how the early infection has a higher proportion of virus overlapping healthy cells (Figure~\ref{fig:cycells}).
+
+In the model, CD8 T cells climb a chemokine gradient to find infected epithelial cells and cluster at local maxima of chemokine concentration.
+Because T cells are clustered, they cannot cover the expanding plaque effectively, where infected cells on the periphery become more highly dispersed as the plaque grows.
+Thus, T cells in the model saturate at a relatively low threshold, beyond which additional T cells do not improve clearance rates.
+These observations provide an explanation for pH1N1 dynamics that would be obscured without the visualizations provided by spatial modeling.
+These results, together with the sensitivity analysis, suggest that viral dynamics are the dominant influence in the course of influenza infection.
+Changes in the chemokine and T cell parameters did not have a strong effect in the model (Table \ref{tab:sensitivity}).
+
+The window of control describes the limit on the amount of time an infected cell may secrete virus while still allowing the infection to be cleared.
+Studying the delay in the model revealed one reason why T cells fail to clear pH1N1 infections.
+We formalized this effect in an equation which calculates an upper bound on the search time for T cells to find new virus-secreting cells, and we showed that this value is consistent with the results of our sensitivity analysis.
+
+Our model captures important aspects of T cell response to influenza infection which have not been addressed by earlier T cell models.
+The results of our sensitivity analysis are important for future biological modeling and experimental design.
+Empirical verification of the model's sensitive parameters (T cell production rate, viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
+Finally, both qualitative and quantitative analysis of our model suggests therapeutic interventions of influenza infection in the lung will likely be most effective by focusing on viral control rather than enhancing T cell responses.
+
+% The results point to future experimental studies which could help elucidate the key components that characterize phenotypic differences among different strains of influenza.
+
+% Finally,
+
+% Future work with spatially explicit models could explore the communication distance in more detail, as well as the role of immigrant CD8 T cell proliferation, T cell chemotaxis, contribution of resident memory T cells and B cells, and effector cell lifespan.
+
+% Last pp on impact
+
+% The behavior of searching T cells described in this paper can enhance future global immune response models.
 
@@
 
-This publication was funded by NIH grants 1R21-AI-73607 (to F.K.), U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F.), and NSF grant EF 1038682 (to M.M.).
+This publication was funded by NIH grants 1R21-AI-73607 (to F.K.), U01-AI-074561 (to F.K.), DARPA grant P-1070-113237 (to S.F.), and NSF grant EF 1038682 (to M.M.) and a James S.
+McDonnell Foundation grant for the study of Complex Systems (to M.M.).
 
@@
 \begin{center}
- \includegraphics[width=4in]{Figure_3}
+ \includegraphics[width=\textwidth]{Figure_3}
  \end{center}
-\caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in ng/mL.
+\caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in pg/mL.
 sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.
-An extended differential equation model from \cite{Mitchell2011} (see Materials and Methods) was fit to IP-10 and RANTES data.
+An extended differential equation model from \cite{Mitchell2011} was fit to IP-10 and RANTES data (Section S2.1 Equation S1) .
 These fits were used to obtain chemokine production values for use in the spatial CyCells model.
@@
 Each run was initialized identically for each strain save for the random seed.
-The middle line shows the average while the outer lines show the 95\% confidence interval.}
+The middle line shows the mean while the red dashed lines show the 96\% credible confidence interval.}
  \label{fig:variance}
@@
  \end{center}
-\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1 with instant cell death.} The model results }
+\caption{{\bf Simulated infections of aH5N1, sH1N1, and pH1N1 with instant cell death.} The model results how that the combined delay of the T cell kill time and apoptosis time form a barrier to infection clearance.
+Removing both delays results in infection clearance for all strains.}
  \label{fig:instantkill}
@@
 
-\begin{table}
+\begin{table}[!ht]
 \centering
@@
 
-\begin{table}
+\begin{table}[!ht]
 \begin{center}
@@
   T Cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
-  T Cell Secretion Rate & 1257 cells/hour & \cite{Miao2010} \\ 
+  T Cell Production Rate & 1257 cells/hour & \cite{Miao2010} \\ 
   T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
@@
   Onset of ATC$^\ddagger$ Lymph Node Exit & Day 5 & \cite{Banerjee2011} \\
-  IgM Strength & Viral decay of 10/day & \cite{Diamond2003} \\
+  Viral Response to IgM & Viral decay of 10/day & \cite{Diamond2003} \\
   IgM Onset & Day 4 & \cite{Diamond2003} \\
@@
   Chemokine Decay Rate & $3.8508\cdot10^{-4}$/s & 1\\
-  Infection Sensitivity Rate & 0.5 cell/hour$\cdot$virion &  2 \\
+  Infectivity & 0.5 cell/hour$\cdot$virion &  2 \\
   Expression Time & 16.7 hours & \cite{Mitchell2011}$^3$ \\
@@
   \hline
-  \multirow{3}{*}{Chemokine} & Chemokine Decay Rate & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable \\
-  & Chemokine Diffusion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable& \cellcolor{green!50}stable \\
-  & Chemokine Secretion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
+  \multirow{3}{*}{Chemokine} & Chemokine Decay Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable \\
+  & Chemokine Diffusion Rate & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable& \cellcolor{green!50}stable \\
+  & Chemokine Secretion Rate & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
   \hline
-  \multirow{6}{*}{T Cell} & Circulation Time & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
-  & T Cell Kill Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
-  & T Cell Velocity & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
-  & Circulating Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable& \cellcolor{blue!30}mostly stable \\
-  & Found Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}mostly stable & \cellcolor{blue!30}mostly stable \\
-  & T Cell Secretion Rate & \cellcolor{blue!30}mostly stable & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  \multirow{6}{*}{T Cell} & Circulation Time & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Kill Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Velocity & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & Circulating Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable& \cellcolor{blue!30}bounded stable \\
+  & Found Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Production Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
   \hline
@@
   \hline 
-  \multirow{4}{*}{Virus} & IgM Strength & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
-  & Infection Sensitivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  \multirow{4}{*}{Virus} & Viral Response to IgM & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  & Infectivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
   & Viral Decay Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
@@
 Each combination of influenza strain and free parameter was classified as belonging to one of four categories.
-Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{mostly stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.}
+Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{bounded stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.}
 \label{tab:sensitivity}

### 2015-03-31 paper/chemokine_supplement_revised.tex
RSI Submission

@@
 \doublespacing
+
+% Use lscape packages to allow for specific pages to be displayed as landscape
+% \usepackage{lscape}
+\usepackage{pdflscape}
 
@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration (SUPPLEMENT)}
+\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung (SUPPLEMENT)}
 }
@@
 \bf{2} Lovelace Respiratory Research Institute, Albuquerque, NM, USA
+\\
+\bf{3} Department of Molecular Genetics \& Microbiology Department of Pathology, University of New Mexico Health Sciences Center, Albuquerque, NM, USA
 \\
@@
 {\footnotesize
-\begin{equation}
+\begin{equation*}
 \begin{aligned}
@@
 \end{aligned}
+\tag{S1}
 \label{eq:dde}
-\end{equation}
+\end{equation*}
 \vspace{.05in}
@@
 \begin{equation*}
-\label{EqS1}
+\label{eq:miao}
 \begin{aligned}
@@
 \end{aligned}
-\tag{S1}
+\tag{S2}
 \end{equation*}
@@
 The initial number of infected cells is assumed to be zero.
-Parameter values are listed in Table S1 in Text S1.
+Parameter values are listed in Table S3.
 This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
@@
 
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we hypothesize that it renders IP-10 ineffectual.
+Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.
+Thus, IP-10 production was not included in aH5N1 simulations.
+
+%Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+
+% we hypothesize that it renders IP-10 ineffectual.
 We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).
 Because of this behavior IP-10 was not included in the aH5N1 model runs.
+
+% we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.
+Thus, IP-10 production was not included in aH5N1 simulations.
 
@@
 
-We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior, parameters that affect the model's peak infection size but do not affect final clearance, and those parameters that affect the final clearance of the infection (Table 3).
+We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table 3).
 Parameters are presented along with their range using the following format:  \\
@@
 
-The parameters in the first grouping did not affect the outcome of the infection unless they were adjusted to values outside the realm of possibility.
-Specifically, each parameter in this group seems to lie within a range that does not significantly affect the asymptotic behavior of the model.
-We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{mostly stable} parameters.
-The difference between the two being that \textit{mostly stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.
+Stable parameters do not affect the outcome of the infection unless they are adjusted to values outside the realm of possibility.
+Specifically, the plausible parameter values in this group are within a range that does not significantly affect the asymptotic behavior of the model.
+We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{bounded stable} parameters.
+\textit{Bounded stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.
 
-Of interest, this group can be mostly split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).
+Of interest, this group can be split into two types of parameters: those governing chemokine behavior (chemokine decay, chemokine diffusion, and chemokine secretion) and those governing T cell behavior (T cell circulation time, T cell kill rate, T cell velocity, and the two T cell decay rates).
 Only the apoptosis time parameter does not fit into one of these two groups, and its inclusion as a stable parameter may be suspect due to the limited range of the values tested.
@@
 The total quantity of the chemokine in the lung does not have a strong effect on the location and size of the gradient.
-Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells may follow.
+Thus, the model will be stable for any values of chemokine secretion, diffusion, and decay that provide a gradient that T cells are able to follow.
 
@@
 While we have claimed that these parameters are qualitatively stable within a certain range, many of the parameters were set to values that did lead to a divergent model behavior.
+Testing these extreme values provides bracketing information regarding the range of stability of the parameter in question.
 We have deemed these deviations acceptable on an individual basis as described below.
@@
 
+Compartmental modeling design suggests the removal of such stable parameters.
+Due to the nature of the ABM, inclusion of these parameters is mechanistically necessary.
+Thus, we include and parameterize them using the methods described in the main paper and Table 2.
+
 \textbf{Apoptosis Time:} [ \textit{0 | 1h | 2h} ] - Apoptosis time describes how long it takes for a cell to complete apoptosis and transition to an inert dead state.
 
-We chose to limit the increase of the apoptosis to a factor of two because larger values would not be biologically relevant.  In the other direction we lowered the value all the way to zero.
+We chose to limit the increase of the apoptosis to a factor of two because larger values would not be biologically relevant.  We considered values as low as zero.
 
-Adjusting the apoptosis time causes the model results to diverge slightly in the sH1N1 strain.
+Adjusting the apoptosis time causes the model results to diverge slightly in the sH1N1 strain (Fig.~\ref{fig:ssensitivity}).
 Also of interest, reducing the apoptosis time to zero still did not allow the immune response to clear the pandemic infection.
-As stated earlier, its inclusion in this group is already suspect and its effects are described in more detail in the results section of the main paper.
+As stated earlier, its inclusion in this group is marginal and its effects are described in more detail in the results section of the main paper.
 \\
@@
 
-We tested values up to two orders of magnitude larger and smaller than the default parameter value.  These extremes are not biologically plausible but were tested anyway.
+We tested values up to two orders of magnitude larger and smaller than the default parameter value even though these extremes are not biologically plausible.
 
@@
 This suggests that the quantity of chemokine is immaterial so long as there is enough for T cells to be able to detect it.
-Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection.
+% Rather, the presence of chemokine is useful only if its presence results in a chemical gradient surrounding the infection.
 \\
@@
 The chemokine diffusion rate only diverges in the aH5N1 strain for the highest value.
-If anything, our estimate for the diffusion rate is already large as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.
-Thus the diffusion rate we used based on Stokes-Einstein and water viscosity was likely within a physiologic range.
+Our estimate for the diffusion rate is already elevated as it is optimistically based off of the Stokes-Einstein equation using viscosity of water.
 \\
@@
 It is important to node that aH5N1 and sH1N1 show a threshold at the same concentration: near 1e-6 $pg/s\cdot cell$.
-While this may seem contradictory, it is actually an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.
+This is an artifact of our artificially selected chemokine detection threshold, detailed in section S2.1 and Figure~\ref{fig:sensitivity}.
 Because we initially picked a sensitivity threshold near the edge of the stable range of possible values, decreasing the total concentration of chemokine inadvertently crosses that arbitrary threshold and does not necessarily suggest an actual region of divergence.
@@
 
-A value of 1s was chosen as the minimum due to computational constraints, yet this value is likely already too small to be biologically plausible.
+A value of 1 second was chosen as the minimum due to computational constraints, yet this value is likely already too small to be biologically plausible.
 We extended the range to up to 600 times the default value to examine what happens when the circulation time is so long that the T cells are effectively removed from the simulation if they miss the focus of infection on their first pass through the lung.
@@
 T cell circulation times were tested over a very large range and does diverge for high values.
-Classifying this parameter as \textit{mostly stable} may seem counterintuitive when examining the sensitivity figures.
+Classifying this parameter as \textit{bounded stable} may seem counterintuitive when examining the sensitivity figures.
 It is important to note that in this case, we tested extra values on the higher end of the baseline than other plots.
-While each strain does show divergence beyond a certain value, the intermediate values of 1s and 20s remain qualitatively similar to the main result, demonstrating a stable range surrounding the default value.
+While each strain does show divergence beyond a certain value, the intermediate values of 1 second and 20 seconds remain qualitatively similar to the main result, demonstrating a stable range surrounding the default value.
 Because vascular and lymph circulation times are unknown, parameter values of up to 1 minute and 3 minutes in mice may be plausible and warrant further study.
@@
 
-\textbf{T Cell Expected Kill Time:} [ \textit{0m | 10m | 100m} ]  - The T cell expected kill time defines the T cell kill rate, the rate at which T cells probabilistically induce apoptosis of  virus-expressing epithelial cells in their immediate vicinity.
-A lower kill time corresponds to a higher kill rate, and vice-versa.
+\textbf{T Cell Expected Kill Time:} [ \textit{0m | 10m | 100m} ]  - The T cell expected kill time defines the rate at which T cells probabilistically induce apoptosis of  virus-expressing epithelial cells in their immediate vicinity.
+A lower kill time corresponds to a higher kill rate.
 
-We tested values only one order of magnitude higher than the default as it does not seem reasonable to expect a T cell to take over an hour to induce apoptosis of a virus-expressing cell.
+We tested values only one order of magnitude higher than the default limiting the time to induce apoptosis of a virus-expressing cell to one hour.
 On the other side, we allowed T cells to induce apoptosis instantly to see if any delay in the induction of apoptosis was responsible for the inability for the immune response to clear the pandemic infection.
@@
 Neither of these values are reasonable and the parameters show stable behavior otherwise.
-While the default of 2 hours may seem short, it appears to determine similar behavior as models with larger values.
+While the default of 2 hours may seem short, it appears to generate similar behavior to models with larger values.
 
@@
 
-The two parameters in this grouping do seem to affect the model's behavior up until the introduction of the T cell response, followed by convergence of infected cell numbers back to the same levels.
+Difference in peak only parameters affect the model's behavior up until the introduction of the T cell response, followed by a return to similar day 10 outcomes.
 Interestingly, these are the two parameters that define the transition delay between the different stages of epithelial cell infection.
@@
 
-We tested values up to a factor of two larger and smaller than the original amount.
+We tested values up to a factor of two larger and smaller than the original parameter setting.
 
@@
 The infection dynamics change dramatically upon the arrival of T cells.
-As discussed, one of the main factors in determining whether or not the infection is cleared is the amount of new virus produced per infected cell.
+One of the main factors in determining whether or not the infection is cleared is the amount of new virus produced per infected cell.
 While this does depend somewhat on the overall size of the plaque, this value depends heavily on shutting down the expression of new virus, something T cells are unable to do during the cellular incubation phase as the virus-incubating cell has not expressed virus for T cells to detect.
@@
 
-The parameter was increased up by a factor of 3 and down by a factor of 10.  In each case, the change is extreme enough such that the resulting values are not biologically likely.
+The parameter was increased up by a factor of 3 and down by a factor of 10.  In each case, the change is extreme enough that the resulting values are not biologically likely.
 
@@
 
-\subsubsection{Unstable Parameters}
+\subsubsection{Sensitive Parameters}
 
-This leaves five parameters shown that appear to directly affect the result of the infection: IgM strength, infection sensitivity, T cell secretion rate, viral decay, and viral diffusion.
+Sensitive parameters directly affect the result of the infection.
+There are five parameters classified as sensitive: viral response to IgM, infectivity, T cell production rate, viral decay, and viral diffusion.
 By comparing the three strains of influenza, we also know that the viral secretion rate affects the result.
-Of interest, nearly all of these parameters are directly related to the behavior of the virus.
-Only the T cell secretion rate and IgM parameters are not.
+Of interest, a majority of these parameters are related to the behavior of the virus.
+The only sensitive parameter not related to the behavior or the virus is the T cell production rate.
 \\
 
-\textbf{IgM Strength:} [ \textit{1 | 10 | 1,000} ] - IgM strength simulates the presence of IgM at day 4 post-infection by increasing the decay rate of free virus particles by the given factor.
+\textbf{Viral Response to IgM:} [ \textit{1 | 10 | 1,000} ] - Viral response to IgM simulates the presence of IgM at day 4 post-infection by increasing the decay rate of free virus particles by the given factor.
 
@@
 While it is true that the extreme values have a large effect on the model behavior, it is more of interest that the intermediate values also seem to change the model behavior.
-This shows that there is no stable area around the default as is seen in the \textit{mostly stable} parameters.
+This shows that there is no stable area around the default as is seen in the \textit{bounded stable} parameters.
 The higher the strength of this parameter, the less virus there is in the system.
@@
 
-Viral decay has the exact same effect as the IgM strength parameter.
+Viral decay has the exact same effect as the viral response to IgM parameter.
 Its value directly determines how much virus remains in the system over the course of the infection.
@@
 
-\textbf{Infection Sensitivity:} [ \textit{12 m/virion | 2 h/virion | 20 h/virion} ] - The infection sensitivity describes the ability of the virus to infect healthy cells.
+\textbf{Infectivity:} [ \textit{12 m/virion | 2 h/virion | 20 h/virion} ] - Infectivity describes the ability of the virus to infect healthy cells.
 A larger value corresponds to a virus particle needing less time to infect a nearby healthy epithelial cell.
@@
 
-The strength of the infection sensitivity parameter is directly related to how much virus is present in the area of the healthy cells.
-Thus, increasing its value by any factor is equivalent to increasing the amount of virus by the same factor.
+The strength of the infectivity parameter is directly related to how much virus is present in the area of the healthy cells.
+Thus, viral density is linearly proportional to infectivity.
 Thus, it behaves similarly to the IgM and viral decay parameters as it directly determines the virulence of the virus.
@@
 
-Shown elsewhere, the viral secretion rate has a strong effect on the outcome of the infection.
+The viral secretion rate has a strong effect on the outcome of the infection.
 Similar to the previously discussed parameters, its value directly affects how much virus there is in the system.
-Thus, it's affect is similar to the previous parameters.
+Thus, its effect is similar to the previous parameters.
 \\
 
-\textbf{Viral Diffustion Rate:} [ \textit{3e-4 $\mu m^2/s$ | 3e-2 $\mu m^2/s$ | 3 $\mu m^2/s$} ] - The viral diffusion rate controls how quickly the virus spreads out over the monolayer.
+\textbf{Viral Diffusion Rate:} [ \textit{3e-4 $\mu m^2/s$ | 3e-2 $\mu m^2/s$ | 3 $\mu m^2/s$} ] - The viral diffusion rate controls how quickly the virus spreads across the monolayer.
 
@@
 While its mechanism is different, its effect may be even stronger than the previous parameters.
-By allowing the virus to diffuse faster than the chemokine (an unlikely scenario), the virus spread may out-pace the body's ability to generate a chemical gradient.
-Thus, T cells will constantly be directed to locations behind the front edge rapidly spreading viral cloud and will be unable to 'get ahead' of the spread of the infection.
+Increasing the viral diffusion rate so that the virus diffuses faster than the chemokine (an unlikely scenario) allows the virus to out-pace the body's ability to establish a chemical gradient around the infected cells.
+In this scenario T cells are constantly directed to locations behind the rapidly spreading viral cloud and would be unable to `get ahead' of the infection.
 \\
 
-\textbf{T Cell Secretion Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell secretion rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.
+\textbf{T Cell Production Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell production rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.
 
@@
 
-The T cell secretion rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
-Thus, it is reasonable to assume that there is a threshold of T cell secretion beyond which the dynamics of the infection do not change.
+The T cell production rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
+Thus, it is reasonable to assume that there is a threshold of T cell production beyond which the dynamics of the infection do not change.
 This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
-Increasing the number of T cells in the system does not seem to help beyond a certain point because the T cells overlap in space and become redundant.
+Increasing the number of T cells in the system does not control the infection beyond a certain point because the T cells overlap in space and become redundant.
 The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.
+
+\subsection{Window of Control}
+
+We predict viral control in the model by approximating $R_0$ and determining if the value is above or below a value of one.
+$R_0$ is calculated as the number of virions produced by each cell multiplied by the number of cells infected by each virion.
+The first half of this equation is the rate of viral production multiplied by the time the cell produces virus.
+The second part of the equation is termed the viral efficiency rate, $E$.
+The equation here is repeated from the main paper:
+
+\begin{equation}
+R_0 = p \times (T_{arr} + T_{kill} + T_{apop}) \times E
+\end{equation}
+
+In the main paper, $E(t)$ is approximated as 0.27, and this value is accurate enough to provide useful predictions.
+Calculating the true value for $E(t)$ is more difficult as it depends on many factors, including the area of the plaque, the depth of the ring of infected (but not dead cells) around the plaque, the local viral density, and the arrival rate and effectiveness of T cells.
+Each of these factors change during the infection.
+
+To find values of $E(t)$ for each of the three strains we reran the model in the absence of T cells and calculated the amount of virus that infects cells, $I(t)$, and the amount that decays $D(t)$, at each time step.
+The viral effectiveness rate is then calculated to be
+
+\begin{equation}
+E(t) = \frac{I(t)}{I(t) + D(t)}
+\end{equation}
+
+This approach accounts for the factors listed above and provides a unique measure of $E(t)$ at each point in time.
+We focus on results in the model after day 5 p.i.
+corresponding to the arrival of T cells in the main model.
+In each simulation $I(t)$ remains relatively constant, while $D(t)$ increases linearly with time.
+We perform least-squared regressions on $I(t)$ and $D(t)$ for each strain and combine the results to give an estimate for $E(t)$ (Table~\ref{tab:effectiveness}).
+
+We use the estimated values of $E(t)$ to calculate $R_0$ values for each strain in the absence of T cell mediated apoptosis.
+Because all three $E(t)$ values all change by less than 10\% over the five day window, we estimate $R_0$ values using the average $E(t)$ value for each strain (Table~\ref{tab:effectiveness}).
+The three strains show similar values for $R_0$, suggesting a saturation of viral effectiveness as the plaque grows.
+This is not the case early in the infection as pH1N1 growth clearly outpaces sH1N1 which in turn expands more rapidly than aH5N1.
+In the main model, the T cell response limits viral production time, which in turn increases the effectiveness of the virus that is produced.
+Thus, it is reasonable to assume viral effectiveness remains high in the presence of the T cell response.
+By accounting for basic spatial and temporal effects as described in the main paper, we approximate $E(t)$ as 0.27 over the course of the infection.
 
@@
 %\section{Discussion}
+
+\section*{Videos}
+
+Videos can be found at the following web address: http://cs.unm.edu\textbackslash\%7Edrew/chemokine
 
@@
 
+\pagebreak
+\begin{landscape}
+
 \begin{figure}[ht!]
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S3}
-	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.}
+	\includegraphics[width=9in]{Figure_S3}
+	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.
+In each subgraph, the thicker purple line shows the results of the model using the default parameter value.}
 	\label{fig:asensitivity}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S4}
-	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.}
+	\includegraphics[width=9in]{Figure_S4}
+	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.
+In each subgraph, the thicker purple line shows the results of the model using the default parameter value.
+The divergence between Apoptosis Time values of 90m and 120m (upper-left plot) highlights an important threshold of infection control.}
 	\label{fig:ssensitivity}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S5}
-	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.}
+	\includegraphics[width=9in]{Figure_S5}
+	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.
+In each subgraph, the thicker purple line shows the results of the model using the default parameter value.}
 	\label{fig:psensitivity}
@@
 \end{figure}
+
+\end{landscape}
+
+%\begin{figure}[ht!]
+%\begin{center}
+%	\includegraphics[width=\textwidth]{Figure_S6}
+%	\caption{\textbf{pH1N1 sensitivity analysis.} Model results and fits of the viral effectiveness rate, $E$.}
+%	\label{fig:effectiveness}
+%\end{center}
+%\end{figure}
 
@@
 %\label{tab:label}
-
 % \end{table}
@@
 \end{tabular}
-\caption{Experimental data. }
+\caption{\textbf{Empirical cytokine titers for three strains of influenza.}  Measured CXCL10 (IP-10) and CCL5 (RANTES) shown in pg/mL for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.
+sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL.
+Human bronchial epithelial cells were infected at an MOI of 0.01 (10,000 virions) with one of the three strains of influenza.
+Basal media for chemokine secretion was collected at the given time intervals post infection.
+Chemokine levels were measured using 30 $\mu$L aliots for a panel of 17 chemokines and cytokines (not shown).}
 \label{tab:data}
@@
 \end{tabular}
-\caption{All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
+\caption{\textbf{Parameters for Supplemental Model 1 (S2.1 Equation S1)}.
+All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
 The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.
@@
 \end{tabular}
-\caption{Supplemental model parameters.
-Most values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
+\caption{\textbf{Parameters for Supplemental Model 2 (S2.2 Equation S2).}  Values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
 Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
-\label{table:supplement}
+\label{tab:supplement}
 \end{center}
+\end{table}
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{ | c  r | c  c | }
+  \hline                        
+  \multicolumn{2}{|c|}{\multirow{2}{*}{Strain}} & Slope & Intercept \\
+  \multicolumn{2}{|c|}{} & \footnotesize{$(PFU/s)$}  & \footnotesize{$(PFU)$} \\
+  \hline
+  \multirow{2}{*}{Avian H5N1} & I(t) & 5.86e-8 & -0.0133 \\
+   & D(t) &  1.03e-7 & -0.0223 \\ 
+   \hline
+  \multirow{2}{*}{Seasonal H1N1} & I(t) & 3.25e-7 &  0.0225\\
+  & D(t) &  6.93e-6 & 0.0439 \\
+   \hline
+  \multirow{2}{*}{Pandemic H1N1} & I(t) & 8.79e-7 & 0.093 \\
+   & D(t) & 2.71e-4 & 5.28 \\
+  \hline
+  \hline
+  \multicolumn{2}{|c|}{Strain} & Avg. $E(t)$ & Estimated $R_0$ \\
+  \hline
+  \multicolumn{2}{|c|}{Avian H5N1} & 0.355 & 1.16 \\
+  \multicolumn{2}{|c|}{Seasonal H1N1} & 0.049 & 1.11 \\
+  \multicolumn{2}{|c|}{Pandemic H1N1} & 0.0037 & 1.12 \\
+  \hline
+\end{tabular}
+\caption{\textbf{Linear fits to model runs for $I(t)$ and $D(t)$ across all three strains.}  Every parameter was significant at the $p < 0.05$ level.
+$E(t)$ is estimated to be $I(t) / [I(t) + D(t)]$.
+$E(t)$ does not vary by more than 10\% from day 5 to day 10, thus the average of each is calculated and used to estimate $R_0$ for all three strains in the absence of a T cell response.}
+\label{tab:effectiveness}
 \end{table}

### 2015-03-31 paper/cover_letter.tex
RSI Submission

@@
 \usepackage{setspace} 
-\doublespacing
+%\doublespacing
 
@@
 
-We submit this manuscript entitled "Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration" for consideration of publication in PLOS ONE.
+We submit this manuscript entitled "Spatial model of lymphocyte search in influenza-infected lung reveals constraints on chemokine directed migration" for consideration of publication in Royal Society Interface.
 Our paper implements an agent-based model of bronchial influenza infection and the subsequent T cell response using new empirical values of chemokine secretion.
@@
 
-This work has not been published elsewhere, and is complementary to a recent publication from our group (H Mitchell, D Levin, et al. J Virol 2011;85(2):1125-1135). \\
+We recommend the following reviewers for their expertise in influenza and/or immunological modeling:
+\begin{enumerate}
+\item Rob de Boer - University of Utrecht
+\item Fred Adler - University of Utah
+\item Judy Day - University of Tennessee
+\item Hulin Wu - Univ. of Rochester
+\end{enumerate}
 
-In review we have noticed that our bibliography does not use annotated journal names, which seem to be the standard for PLOS ONE.
-We have {\em not} changed our bibliography as of yet as it was generated automatically using the Mendeley Desktop suite.
-If this is an issue, we would be happy to return and reconstruct the references manually.
-\\
+We ask that you excuse the following people from review due to involvement in and discussion of earlier iterations of this work:
+\begin{enumerate}
+\item Alan Perelson - Los Alamos National Laboratories 
+\item Catherine Beauchemin - Ryerson University
+\item Rustom Antia - Emory University
+\end{enumerate}
+
+This work has not been published elsewhere, and is complementary to a previous publication from our group (H Mitchell, D Levin, et al. J Virol 2011;85(2):1125-1135). \\
 
@@
 
-All authors have read and approved the revised manuscript.
+All authors have read and approved the revised manuscript and report no conflicts of interest.
 
 \begin{flushleft}Respectfully yours,\end{flushleft}
-Drew Levin \\
-Stephanie Forrest \\
-Soumya Banerjee \\
-Candice Clay \\
-Melanie Moses \\
-Frederick Koster
+
+\noindent Drew Levin, et. al. \\
+University of New Mexico Department of Computer Science \\
+MSC01 1130 \\
+1 University of New Mexico \\
+Albuquerque, NM 87131 \\
+drew@cs.unm.edu
 

### 2015-03-31 paper/final/chemokine_RSI_Levin.tex
RSI Submission

--- 2015-03-31 paper/chemokine_main_revised.tex
@@
 % Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
+\bibliographystyle{bibstyle}
 
@@
 {\Large
-\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung}
+\textbf{Spatial model of lymphocyte search in influenza-infected lung reveals constraints on chemokine directed migration}
 }
@@
 However, determining the relative contribution of these myriad experimental components is challenging.
-We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.
+We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response, and to account for strain-specific viral dynamics in the model.
 
@@
 
-Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.
-In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.
-The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.
+Table \ref{tab:sensitivity} shows that five parameters are sensitive: viral response to IgM, infectivity, viral decay rate, viral diffusion, and the T cell production rate.
+In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter.
+Four of the parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.
 Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.
+The T cell production rate is the only sensitive parameter that does not directly control virus kinetics.
 None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
@@
 
+\section*{Competing Interests}
+
+We have no competing interests.
+
+\section*{Contributions}
+
+\begin{itemize}
+\item DL designed and implemented the modeling and analysis and wrote the paper.
+\item SF aided modeling design and implementation and helped write the paper.
+\item SB aided modeling design and implementation and helped write the paper.
+\item CC collected the data and helped design the study.
+\item JC aided  modeling analysis and helped write the paper.
+\item MM aided modeling design and implementation and helped write the paper.
+\item FK collected the data, helped design the study, and helped write the paper.
+\end{itemize}
+
+All authors approve of the manuscript.
+
 % Do NOT remove this, even if you are not including acknowledgments
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figure_3}
+ \includegraphics[width=6in]{Figure_3}
  \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{Figure_4}
+\includegraphics[width=6in]{Figure_4}
  \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{Figure_5}
+\includegraphics[width=6in]{Figure_5}
  \end{center}
@@
 \begin{center}
-\includegraphics[width=\textwidth]{Figure_6}
+\includegraphics[width=6in]{Figure_6}
  \end{center}
@@
 
-\setcounter{figure}{0}
-\renewcommand{\thefigure}{S\arabic{figure}}
-
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics[width=\textwidth]{Figure_S1}
- \end{center}
-\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
+%\setcounter{figure}{0}
+%\renewcommand{\thefigure}{S\arabic{figure}}
+%
+%\begin{figure}[ht!]
+%\begin{center}
+%\includegraphics[width=\textwidth]{Figure_S1}
+% \end{center}
+%\caption{{\bf Varying T cell sensitivity to chemokine.}  H5N1 model results use RANTES  only, and sH1N1 and pH1N1 use both IP-10 and RANTES.
 Total number of incubating, secreting and apoptotic cells are plotted for each infection.
 The sensitivity value specifies the minimum level of chemokine concentration required for T cells to detect it.}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
-\includegraphics[width=4in]{Figure_S2}
- \end{center}
-\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
+%\end{figure}
+%
+%
+%\begin{figure}[!ht]
+%\begin{center}
+%\includegraphics[width=4in]{Figure_S2}
+% \end{center}
+%\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
 B-C) sH1N1 and pH1N1 show no significant difference between IP-10 alone versus IP-10 and RANTES combined.}
- \label{fig:sensitivity}
-\end{figure}
-
-\setcounter{figure}{0}
-\renewcommand{\figurename}{Video}
-
-\begin{figure}[ht!]
-\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
+% \label{fig:sensitivity}
+%\end{figure}
+%
+%\setcounter{figure}{0}
+%\renewcommand{\figurename}{Video}
+%
+%\begin{figure}[ht!]
+%\caption{The first of three overlaid videos of a representative seasonal H1N1 infection.
 This video spans the 10 day infection and shows the cells as they transition from healthy to infected to dead.
@@
 }
- \label{video:chemokine_view}
-\end{figure}
-
-\begin{figure}[ht!]
-\caption{A closer look at the 2009 pandemic simulation.
+% \label{video:chemokine_view}
+%\end{figure}
+%
+%\begin{figure}[ht!]
+%\caption{A closer look at the 2009 pandemic simulation.
 This video shows the infection from day 6 to day 7 with each frame spanning 60 simulated seconds.
@@
   & Found Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
-  & T Cell Production Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Production Rate & \cellcolor{blue!30}bounded stable & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
   \hline

### 2015-03-31 apop_killrate.m

@@ -1,6 +1,6 @@
 % Instant kill rate and apoptosis plot
 
-figure('Position',[100 100 640 360])
+figure('Position',[100 100 960 540])
 
 colors = [.625  0  .2 ;
           1 .75  0 ;
@@ -17,11 +17,11 @@ hold off
 % set(gca, 'YLim', [ymin ymax])
 set(gca, 'XLim', [0 10])
 set(gca, 'FontSize', 20)
-set(gca, 'FontWeight', 'bold')
+%set(gca, 'FontWeight', 'bold')
 
 ylabel('Infected Cells')
 xlabel('Time (days)')
 lh = legend('aH5N1', 'sH1N1', 'pH1N1');
 set(lh, 'FontSize', 14)
 
-title('Instant Kill and Apoptosis');
+title('Instant Kill and Apoptosis', 'FontSize', 24);

### 2015-03-31 chemokine.m

@@ -1,8 +1,21 @@
 cd('/nfs/adaptive/drew/chemokine');
 
-% Chemokine Effects% Sensitivity Plot
+width = 20;
+height = 8;
 
-figure('Position',[100 500 1800 700])
+font_size = 18;
+title_font_size = 24;
+legend_font_size = 14;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
+
+% Chemokine Effects
 
 % Avian
 
@@ -18,7 +31,7 @@ hold on
 plot(time, infect(ar19), 'color', colors(2,:), 'LineWidth', 2)
 hold off
 
-set(gca, 'FontSize', 16)
+set(gca, 'FontSize', font_size)
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [1 10000])
 set(gca, 'YTick', [0 2000 4000 6000 8000 10000])
@@ -26,8 +39,9 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Infected Cell Count')
-title('aH5N1 Influenza')
-legend('None', 'RANTES', 2)
+title('aH5N1 Influenza', 'FontSize', title_font_size)
+lh = legend('None', 'RANTES', 2);
+set(lh, 'FontSize', legend_font_size);
 
 % Seasonal
 
@@ -40,7 +54,7 @@ plot(time, infect(si19), 'color', colors(3,:), 'LineWidth', 2)
 plot(time, infect(sb19), 'color', colors(4,:), 'LineWidth', 2) 
 hold off
 
-set(gca, 'FontSize', 16)
+set(gca, 'FontSize', font_size)
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [1 10000])
 set(gca, 'YTick', [0 2000 4000 6000 8000 10000])
@@ -48,8 +62,9 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 %ylabel('Infected Cell Count')
-title('sH1N1 Influenza')
-legend('None', 'RANTES', 'IP-10', 'Both', 2)
[…]

### 2015-03-31 plaquesize.m

@@ -20,7 +20,7 @@ semilogy(time, sb19(:,7)+sb19(:,8)+sb19(:,9), 'color', 'black', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 20)
-set(gca, 'FontWeight', 'bold')
+%set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [0 100000])
 set(gca, 'YTick', [1 10 100 1000 10000 100000])
@@ -28,7 +28,7 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Cell Count (log10)')
-title('aH5N1 Influenza')
+title('aH5N1 Influenza', 'FontSize', 24)
 %lh = legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T cells at FOI', 'Total T cells', 2);
 %set(lh, 'FontSize', 16)
 %set(lh, 'FontWeight', 'normal')
@@ -44,7 +44,7 @@ semilogy(time, sb19(:,7)+sb19(:,8)+sb19(:,9), 'color', 'black', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 20)
-set(gca, 'FontWeight', 'bold')
+%set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [0 100000])
 set(gca, 'YTick', [1 10 100 1000 10000 100000])
@@ -52,7 +52,7 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 %ylabel('Infected Cell Count (log10)')
-title('sH1N1 Influenza')
+title('sH1N1 Influenza', 'FontSize', 24)
 %set(gca, 'FontSize', 12)
 % legend('Plaque size (cells)', 'Virus-incubating cells', 'Virus-secreting cells', 'T-cells at FOI', 2)
 
@@ -68,7 +68,7 @@ semilogy(time, sb19(:,7)+sb19(:,8)+sb19(:,9), 'color', 'black', 'LineWidth', 2)
 hold off
 
 set(gca, 'FontSize', 20)
-set(gca, 'FontWeight', 'bold')
+%set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
[…]

### 2015-03-31 plot_data.m

@@ -1,0 +1,305 @@
+% Creates Data Plot
+
+width = 20;
+height = 8;
+
+font_size = 18;
+title_font_size = 24;
+legend_font_size = 14;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
+
+% --- Avian Data ---
+avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
+av  = [ 100, 150, 825, 2563, 988, 1650, 2838, 17375, 5775];
+ave = [   0,  29, 229,  348, 166,  390,  112,  3062, 1261];
+
+ait = [120, 480, 600, 960, 1200, 1440, 1680,  1920,  2280,  2880]/60;
+ai  = [ 24,  87, 296, 770, 3048, 3901, 8261, 12935, 26970, 24120];
+
+art = [120, 480, 600, 960, 1200, 1440, 1680, 1920, 2280, 2880]/60;
+ar  = [ 45,  50,  57,  87,  134,  151,  150,  245,  780,  695];
+
+% --- Seasonal Data ---
+svt = [600,  720,  960,  1200,  1440,  1800,  2280,  2880]/60;
+sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
+sve = [  0, 1676, 1578, 15695, 17271,  2992,  2886,  3837];
+
+sit = [120,  480,  720,  960, 1200, 1440]/60;
+si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
+
+siit  = [1440, 1800, 2280, 2880]/60;
+%sii   = [8164, 10200, 12913, 16306] - 300;
+sii   = [8164, 8500, 8500, 8500] - 300;
+
+srt = [120, 480, 720, 960, 1200, 1440, 1800, 2280, 2880]/60;
+sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
+
+% --- Swine Data ---24
+wvt = [120,   720,   960,   1200,   1440,   1800,   2880]/60;
+wv  = [ 20, 64620, 84053, 254587, 665559, 814586, 654773];
[…]

### 2015-03-31 sensitivity.m

@@ -2,7 +2,20 @@ cd('/nfs/adaptive/drew/chemokine');
 
 % Sensitivity Plot
 
-figure('Position',[100 500 1800 600])
+width = 20;
+height = 7;
+
+font_size = 18;
+title_font_size = 24;
+legend_font_size = 14;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
 
 % Avian
 
@@ -35,8 +48,8 @@ hline1 = plot(time, infect(ar17), 'color', colors(1,:), 'LineWidth', 2);
 hline3 = plot(time, infect(ar19), 'color', colors(3,:), 'LineWidth', 2);
 hold off
 
-set(gca, 'FontSize', 18)
-set(gca, 'FontWeight', 'bold')
+set(gca, 'FontSize', font_size)
+%set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
 set(gca, 'YLim', [1 8000])
 set(gca, 'YTick', [0 2000 4000 6000 8000])
@@ -44,9 +57,9 @@ set(gca, 'XLim', [0 10])
 set(gca, 'XTick', 0:2:10)
 xlabel('Days Post-Infection')
 ylabel('Infected Cell Count')
-title('aH5N1 Influenza')
+title('aH5N1 Influenza', 'FontSize', title_font_size)
 lh = legend([hline1 hline2 hline3 hline4], '10 \mug/mL', '1 \mug/mL', '100 ng/mL', '10 ng/mL to .01 pg/mL', 2);
-set(lh, 'FontSize', 14)
+set(lh, 'FontSize', legend_font_size)
 
 % Seasonal
 
@@ -65,8 +78,8 @@ hline1 = plot(time, infect(sb17), 'color', colors(1,:), 'LineWidth', 2);
 hline3 = plot(time, infect(sb19), 'color', colors(3,:), 'LineWidth', 2);
 hold off
 
-set(gca, 'FontSize', 18)
-set(gca, 'FontWeight', 'bold')
+set(gca, 'FontSize', font_size)
+%set(gca, 'FontWeight', 'bold')
 set(gca, 'YColor', 'k')
[…]

### 2015-03-31 stderror.m

@@ -31,8 +31,8 @@ for i=1:rows
     astderr(i) = std(vara(i,:));% / sqrt(cols);
 end
 
-alow = aavg - 2 * astderr;
-ahigh = aavg + 2 * astderr;
+% alow = aavg - 2 * astderr;
+% ahigh = aavg + 2 * astderr;
 
 
 % Seasonal

### 2015-03-31 variance.m

@@ -1,42 +1,44 @@
-figure('Position',[100 500 1800 700])
+width = 20;
+height = 8;
+
+font_size = 18;
+title_font_size = 24;
+legend_font_size = 14;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
 
 % Avian
 
 subplot(1, 3, 1)
 
-aavg = (infect(vara00) + infect(vara01) + infect(vara02) + infect(vara03) + infect(vara04) + ...
-        infect(vara05) + infect(vara06) + infect(vara07) + infect(vara08) + infect(vara09) + ...
-        infect(vara10) + infect(vara11) + infect(vara12) + infect(vara13) + infect(vara14) + ...
-        infect(vara15) + infect(vara16) + infect(vara17) + infect(vara18) + infect(vara19) + ...
-        infect(vara20) + infect(vara21) + infect(vara22) + infect(vara23) + infect(vara24) + ...
-        infect(vara25) + infect(vara26) + infect(vara27) + infect(vara28) + infect(vara29) + ...
-        infect(vara30) + infect(vara31) + infect(vara32) + infect(vara33) + infect(vara34) + ...
-        infect(vara35) + infect(vara36) + infect(vara37) + infect(vara38) + infect(vara39) + ...        
-        infect(vara40) + infect(vara41) + infect(vara42) + infect(vara43) + infect(vara44) + ...
-        infect(vara45) + infect(vara46) + infect(vara47) + infect(vara48) + infect(vara49)) / 50;
-    
-hline1 = plot(time, [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
-            infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
-            infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
-            infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
-            infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
[…]

### 2015-04-07 paper/chemokine_main_revised.tex
RSI revisions

@@
 
-Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.
+{\color{dkblue}Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.
 However, determining the relative contribution of these myriad experimental components is challenging.
-We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.
+In this study, human epithelial cells were infected with Avian, seasonal, and pandemic influenza \textit{in vitro} and were measured for virus and various cytokine concentrations over the course of 48 hours.
+The cytokine profile revealed that the only difference between influenza strains in epithelial cells is in the production of chemokines CXCL10 (IP-10) and CCL5 (RANTES).
+Normally chemokines attract T cells to infection sites, and because \textit{in vivo} cell movement in lung tissue cannot be observed directly, we developed a model to examine this effect.}
 
-Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.
+%We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.
+
+Rather than developing a comprehensive immune system model, we focus on the role of chemokines in directing antigen-specific CD8 T cells from the lymph node to the focus of infection (FOI) in the human lung.
 We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.
@@
 Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
-Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.
+Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands IP-10 and RANTES respectively.
 Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.
@@
 
-Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
+Because the recruitment of T cells through chemotaxis is a spatial process, we present a spatial agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
 Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.
 We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.
-The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.
+The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1 (sH1N1), Avian H5N1 (aH5N1), and pandemic H1N1 (pH1N1).
 These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.
 The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.
-Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.
+Sensitivity analysis shows most chosen model parameters have a limited effect on the qualitative results.
+Those that have a strong effect are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.
 
@@
 
-The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
+{\color{dkblue}The main contributions of this paper are as follows.
+First, \textit{in vitro} influenza infection of human epithelial cells stimulated production of chemokines IP-10 and RANTES, suggesting that these two chemokines play a direct role in recruiting T cells to the focus of infection.
+Second, the spatial nature of the model allows us to identify physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
 Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
@@
 In contrast, parameters related to viral kinetics have a strong effect on model results.
-We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.
+We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.}
 
@@
 The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).
-The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).
-Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.
+{\color{dkblue}While the volume of the lung consists of spherical alveoli, the spread of influenza throughout the lung occurs through the interconnected bronchiole-alveoli complex, which can be viewed as a connected monolayer.
+Thus, we represent the lung as a two-dimensional sheet of healthy epithelial cells (S1.3).}   Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.
 If cytokine signal is detected on the local endothelium the T cell remains in tissue and follows the chemotactic gradient to the FOI.
@@
 The results of our sensitivity analysis are important for future biological modeling and experimental design.
-Empirical verification of the model's sensitive parameters (T cell production rate, viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
+Empirical verification of the model's sensitive parameters (viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
 Finally, both qualitative and quantitative analysis of our model suggests therapeutic interventions of influenza infection in the lung will likely be most effective by focusing on viral control rather than enhancing T cell responses.
@@
 \begin{center}
-\begin{tabular}{| l l l |}
+\begin{tabular}{| l | c | l l l | c |}
   \hline                        
-  Referenced Parameters & Value & Source \\
+  Referenced Parameters & Units & Min & Value & Max & Source \\
   \hline
-  Viral Diffusion in Airway & .0318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
-  Viral Decay in Airway &  1/day & \cite{Lee2009} \\
-  Chemokine Diffusion Rate & .318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
-  Incubation Time &  10 hours & \cite{Mitchell2011} \\
-  Epithelial Cell Radius & 5 $\mu$m & \cite{Elbert1999} \\
-  T Cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
-  T Cell Production Rate & 1257 cells/hour & \cite{Miao2010} \\ 
-  T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
-  Blood Circulation Time & 6 seconds & \cite{Banerjee2010b} \\
-  T Cell Sensitivity to Chemokine & 100 ng/mL & \cite{Nandagopal2011} \\
-  Onset of ATC$^\ddagger$ Lymph Node Exit & Day 5 & \cite{Banerjee2011} \\
-  Viral Response to IgM & Viral decay of 10/day & \cite{Diamond2003} \\
-  IgM Onset & Day 4 & \cite{Diamond2003} \\
+  $^{\dagger}$Viral Diffusion in Airway & \textit{$\mu$m$^2$/s} & 3.18e-4 & \textbf{3.18e-2} & 3.18  & \cite{Beauchemin2006} \\
+  $^{\dagger}$Viral Decay in Airway & \textit{day$^{-1}$} & 0.01 & \textbf{1} & 100 & \cite{Lee2009} \\
+  Chemokine Diffusion Rate & \textit{$\mu$m$^2$/s} & 3.18e-3 & \textbf{.318} & 318 & \cite{Beauchemin2006} \\
+  Incubation Time & \textit{hours} & 5 & \textbf{10} & 20 & \cite{Mitchell2011} \\
+  Epithelial Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \cite{Elbert1999} \\
+  T Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \cite{abbas2011cellular} \\
+  T Cell Production Rate & \textit{cells/h} & 125 & \textbf{1,257} & 3,750 & \cite{Miao2010} \\ 
+  T Cell Speed & \textit{$\mu$m/min} & 6e-2 & \textbf{6} & 600 & \cite{Egen2011} \\
+  Blood Circulation Time & \textit{seconds} & 1 & \textbf{6} & 3,600 & \cite{Banerjee2010b} \\
+  T Cell Sensitivity to Chemokine & \textit{ng/mL} & | & \textbf{100} & | & \cite{Nandagopal2011} \\
+  Onset of T Cell Lymph Node Exit & \textit{days }& | & \textbf{5} & | & \cite{Banerjee2011} \\
+  $^{\dagger}$IgM Viral Decay Factor & | & 1 & \textbf{10} & 1,000 & \cite{Diamond2003} \\
+  IgM Onset & \textit{days} & | & \textbf{4} & | & \cite{Diamond2003} \\
   \hline
   \hline                        
-  Estimated Parameters & Value & Footnote \\
+  Estimated Parameters & Units & Min & Value & Max & Footnote \\
   \hline
-  Chemokine Decay Rate & $3.8508\cdot10^{-4}$/s & 1\\
-  Infectivity & 0.5 cell/hour$\cdot$virion &  2 \\
-  Expression Time & 16.7 hours & \cite{Mitchell2011}$^3$ \\
-  T Cell Expected Kill Time & 10 min & 4 \\
-  Apoptosis Time & 60 minutes & \cite{Ganusov2008}$^5$ \\
-  T Cell Age (at FOI$^\ast$) & 2 hours & 6 \\
-  T Cell Age (in Blood) & 3 days & 6 \\
+  Chemokine Decay Rate & \textit{Hz} & 3.8e-6 & \textbf{3.8e-4} & 3.8e-2 & 1\\
+  $^{\dagger}$Infectivity & \textit{min/virion} & 12 & \textbf{120} & 1,200   &  2 \\
+  Expression Time & \textit{min} & 100 & \textbf{1,000} & 3,000 & \cite{Mitchell2011}$^3$ \\
+  T Cell Expected Kill Time & \textit{min} & 0 & \textbf{10} & 100 & 4 \\
+  Apoptosis Time & \textit{hours} & 0 & \textbf{1} & 2 & \cite{Ganusov2008}$^5$ \\
+  T Cell Age (at FOI) & \textit{min} & 12 & \textbf{120} & 1,200 & 6 \\
+  T Cell Age (in Blood) & \textit{days} & 0.04 & \textbf{4} & 400 & 6 \\
   \hline  
 \end{tabular}
-\caption{Values are taken from experimental literature if possible and from earlier modeling papers if not.
-If we could not find a value in any literature we estimated parameters as follows.
+\caption{Default values used for the model in bold.
+Min and max represent the extreme values tested in the sensitivity analysis (Table~\ref{tab:sensitivity} and Fig.~S3-5).
+$^{\dagger}$ denotes parameters determined to be sensitive by the sensitivity analysis.
+Values were taken from experimental literature if possible and from earlier modeling papers if not.
+Parameters not found in literature were estimated as follows.
 1) Corresponds to a 30 minute half-life.

### 2015-09-07 paper/chemokine_main_revised.tex
clean-up to JTB inital state

--- 2015-03-31 paper/chemokine_main_revised.tex
@@
-\documentclass[10pt]{article}
+\documentclass[preprint,10pt,authoryear]{elsarticle}
 
@@
 % cite package, to clean up citations in the main text. Do not remove.
-\usepackage{cite}
+%\usepackage{cite}
+\usepackage{natbib}
 
@@
 
-% Use the PLoS provided bibtex style
-\bibliographystyle{plos2009}
+\bibliographystyle{elsarticle-num-names}
 
@@
 \usepackage[usenames,dvipsnames,table]{xcolor}
-\usepackage{ulem}
+%\usepackage{ulem}
 
@@
 
+\journal{Theoretical Biology}
+
 %% END MACROS SECTION
@@
 {\Large
-\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung}
+\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung}
 }
@@
 
-Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
+%Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
 Viral clearance depends in part on the ability of T cells to locate the infection in time to prevent uncontrolled spread of the virus.
@@
 The spatial model reveals challenges for T cell recruitment not apparent in standard differential equation models.
-The model shows that virus kinetics largely determine the course of the infection, while T cell chemokine parameters have less effect.
+The model shows that virus kinetics largely determine the course of the infection, especially in the case of the pandemic flu strain, while T cell control of the seasonal flu strain is more important.
+Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.
+
+%while T cell chemokine parameters have less effect.
 
@@
 
-% ---- 300 word abstract
+% ---- 300-ish word abstract
 
-% Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
+Emerging strains of influenza, such as avian H5N1 and 2009 pandemic H1N1, are more virulent than seasonal H1N1 influenza, yet the underlying mechanisms for these differences are not well understood.
 Subtle differences in how a given strain interacts with the immune system are likely a key factor in determining virulence.
 One aspect of the interaction is the ability of T cells to locate the foci of the infection in time to prevent uncontrolled expansion.
-Here, we focus on T cell navigating from the lymph nodes through the vascular system to the infection, using an agent-based spatial model to investigate how different strains of influenza influence the efficiency of this process.
+Here, we develop an agent based spatial model to focus on T cell migration from lymph nodes through the vascular system to sites of infection.
+We use our model to investigate how different strains of influenza modulate this process.
 
-%We calibrate the model using viral and chemokine secretion rates measured \textit{in vitro} for avian H5N1, seasonal H1N1, and 2009 pandemic H1N1 influenza, together with values taken from literature when available.
+We calibrate the model using viral and chemokine secretion rates we measure \textit{in vitro} together with values taken from literature.
 The spatial nature of the model reveals unique challenges for T cell recruitment that are not apparent in standard differential equation models.
-As plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.
-Further, differences in diffusion rates between chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.
-Sensitivity analysis of the model identifies the viral parameters as the most significant factor in determining the course of the resulting infection.
+% -- Fred --
+%We find that as plaques expand in size, the ratio of infected cells to dead cells decreases, making it difficult for T cells to encounter infected cells efficiently.
+Further, differences in the diffusion rates of chemokines and virions can misdirect T cells to areas that no longer contain infected epithelial cells.
+While our sensitivity analysis identifies viral parameters as the most significant factor in determining the course of the infection, we are also able to determine an upper bound of T cell search time using the model.
 Finally, we use the model parameters to determine an upper bound on T cell search time in the model.
+Our results shed light on the key factors differentiating immune response against seasonal, avian, and pandemic flu.
+% -- Fred , after p2s2 -- 
+In this model comparing three influenza viruses, plaque expansion is governed primarily by the replication rate of the virus strain, and the efficiency of the T cell search-and-kill is limited by the density of infected epithelial cells in each plaque.
+Thus for each virus there is a different threshold of T cell search time after which recruited T cells are unable to control further expansion.
+Future models could use this relationship to more accurately predict control of the infection.
 
-% ---- end 300 word abstract
+% ---- end 300-ish word abstract
 
@@
 
-% Please keep the Author Summary between 150 and 200 words
-% Use first person. PLoS ONE authors please skip this step. 
-% Author Summary not valid for PLoS ONE submissions.   
-% \section*{Author Summary}
-
 \section*{Keywords}
@@
 
-Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \cite{Who2009}.
+Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \citep{Who2009}.
 Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.
@@
 
-Identifying the critical components of this complex process is key to understanding viral pathogenicity and designing therapeutic interventions.
-However, determining the relative contribution of these myriad experimental components is challenging.
-We address this challenge with computational modeling, which gives us the advantage of studying relative contributions of different aspects of the host immune response and accounts for strain-specific viral dynamics in the model.
+Identifying the critical components of this complex process, and how they interact, is key to understanding viral pathogenicity and designing therapeutic interventions.
+We address this challenge through computational modeling, which allows us to study the relative contributions of different aspects of the host immune response and to account for strain-specific viral dynamics in the model.
+Several earlier theoretical models have been proposed to study specific aspects of influenza infections.
+Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but ignore localized spatial effects.
+Spatial models, on the other hand, have been developed to examine interactions between dendritic cells and T cells in the lymph node \citep{Beauchemin2005, Beltman2007, Zheng2008, Mirsky2011, Celli2012, Vroomans2012, Textor2014}, but have not been extended to consider local conditions in the lung.
+In this paper, we examine how the unique properties of three different strains of influenza, in the presence of chemokines, affect T-cell search in the human lung.
 
-Rather than developing a comprehensive immune system model we focus on one aspect of the adaptive response specifically, studying how activated antigen-specific CD8 T cells migrate from the lymph node to the focus of infection (FOI) in the human lung.
-We assume that chemotactic gradients play an important role in guiding T cells to the FOI and that T cells resist extravasation and recirculate in the absence of inflammatory signals.
-Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \cite{Bromley2008, Medoff2005, Castellino2006 ,Gunn1998, Okada2005}, where recruited T cells mediate either protection or immunopathology.
-Infected epithelial cells secrete chemokines \cite{Chan2005}, and contact with CD8 T cells also triggers secretion \cite{Zhao2000}.
-Activated CD8 T effector cells upregulate chemotactic protein receptors, particularly CXCR3 and CCR5 \cite{Hoji2005, Groom2011a}, responding to their primary ligands CXCL10 (IP-10) and CCL5 (RANTES) respectively.
-Previous work showed that efficient recruitment of T cells to the FOI is crucial for the eventual clearance of the virus \cite{Cerwenka1999, Kim2011}.
-We extend this work to examine how the unique properties of the different strains of influenza may affect this process.
+Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, cytokines, and replicating influenza virus.
+Chemotactic proteins are known to enhance T cell recruitment in both acute infections and chronic inflammatory diseases \citep{Gunn1998, Medoff2005, Okada2005, Castellino2006, Bromley2008}.
+Infected epithelial cells secrete chemokines, especially upon contact with CD8 T cells  \citep{Zhao2000, Chan2005}.
+Previous work showed that efficient recruitment of T cells to the Focus Of Infection (FOI) is crucial for the eventual clearance of the virus\citep{Cerwenka1999, Kim2011}.
+As measuring real-time T cell movement in \textit{in vivo} lung tissue is not currently viable, study these localized processes is aided by spatial modeling of how T cells interact with the FOI and the chemotactic environment.
+Therefore, instead of developing a comprehensive immune system model, we present a spatially explicit agent-based model (ABM) to describe T cell interactions with chemotactic signals and a dynamically growing plaque.
+Using the model, we investigate the pathogenic potential of the three  influenza strains: seasonal H1N1 (sH1N1), Avian H5N1 (aH5N1), and pandemic H1N1 (pH1N1).
+These three strains were selected for their differences in \textit{in vitro} replication rates \citep{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.
 
-Here, we present an agent-based model (ABM) of the secondary immune response to an influenza infection in the lung.
-Previous models have shed light on the details of the primary and secondary immune response, some by applying ordinary differential equation (ODE) models to empirical data \cite{Miao2010, Saenz2010, Handel2008, Lee2009, Murillo2013}, and some by using spatial models to examine the interactions between dendritic cells and T cells in the lymph node \cite{Beltman2007, Zheng2008, Vroomans2012, Textor2014}.
-We parameterize our model with \textit{in vitro} generated measurements of viral expansion dynamics as well as published data of host immune responses to flu infection.
-The model was used to determine the pathogenic potential of the three individual flu strains: seasonal H1N1, avian H5N1, and pandemic H1N1.
-These three strains were selected for their differences in \textit{in vitro} replication rates \cite{Mitchell2011} as well as differences in \textit{in vivo} severity of human and animal pneumonia and mortality.
-The model describes the dynamic, spatially explicit evolution of viral infections in the context of the developing CD8 T cell response.
-Sensitivity analysis identifies parameters that are candidates for explaining the dramatic observed differences in pathogenicity among these different flu strains.
+The model is parameterized with values taken from the literature when available.
+Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.
+To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.
+Of the cytokines tested, only the chemokines CXCL10 (IP-10) and CCL5 (RANTES) were expressed differently across the three influenza strains.
+IP-10 and RANTES are both chemokines that normally attract T cells to the FOI \citep{Hoji2005, Groom2011a}.
+These results suggest that the two chemokines may contribute to the immune system's varying ability to control different strains of influenza, and were thus included in our model.
 
-% The model design allows us to compare between influenza strains how the T cell response is affected by the viral and immune system characteristics.
-The results show that efficient T cell search is crucial to viral clearance in the model and that the search process is dependent on more than just the replication rate of the virus, including the spatial landscape of the chemokine gradient and the location and density of virus-secreting cells.
+The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.
+The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
+Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
+The model suggests that the chemokine response of the host cells can hinder T cells’ ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.
+A sensitivity analysis of the model parameters isolated the effects of each parameter in the model.
+The analysis shows that most model parameters can vary across a broad range of values without strong effect on predicted outcomes.
+Further analysis shows model predictions remain unchanged within broad categories of parameter type (e.g., all parameters related to chemokines, parameters related to viral kinetics, etc.).
+For example, the chemokine-related parameters are stable in that they do not affect the model behavior within biologically plausible ranges.
+In contrast, parameters related to viral kinetics significantly alter model predictions.
+Finally we examine the `window of control' which establishes an upper bound on the time it takes T cells to arrive at the FOI while still clearing the infection.
+We find that this value is consistent with our sensitivity analysis and helps explain the challenge of controlling the highly virulent pH1N1 strain.
 
-The spatial nature of the model identifies physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
-Because T cells in the model use the chemokine gradient to locate the FOI, they tend to cluster in areas of high chemokine concentration.
-The model suggests that the chemokine response of the host cells may actually hinder T cells' ability to find and clear infected cells in the presence of the rapidly replicating pandemic strain.
-Finally, the sensitivity analysis shows the isolated effects of each parameter in the model.
-Based on this analysis, each parameter was classified as sensitive or not.
-We find that most parameters can be individually varied across a broad range of values without strong effect on predicted outcomes.
-Further, the parameter classification is consistent across each broad parameter type.
-Specifically, chemokine-related parameters are stable within biologically plausible ranges.
-In contrast, parameters related to viral kinetics have a strong effect on model results.
-We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection, confirming that the higher replication rate of 2009 pH1N1 influenza obscured contributions of the chemotactic signals, leading to uncontrolled infection \textit{in silico}.
+%We conclude that influenza infection dynamics are dominated by the replication properties of the virus and spatial properties of the infection.
+The higher replication rate of 2009 pH1N1 influenza dominated the effect of chemotactic signals, leading to uncontrolled infection \textit{in silico}.
 
-% fred used 'obscured' instead of overwhelmed
-
-% The model predicts the control and clearance of the aH5N1 and sH1N1 strains and predicts a runaway infection for the pH1N1 strain.
-These results are consistent with the previously reported replication rates of the three strains of influenza.
-
-%  Infected cells secrete chemokine molecules which then diffuse across the infected region.  This process creates isolated regions, or local maxima, of high chemokine concentrations.
-
-% While the local maxima do move and flatten, their speed is limited by their diffusion and decay rates.
-
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
-\section*{Materials and Methods}
+\section*{Methods and Models}
 
@@
 Because these interactions are difficult and expensive to observe \textit{in vivo}, we create a model \textit{in silico} to represent the system in question.
-Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \cite{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).
+Because we are interested in individual interactions between the virus and the immune response, as well as the spatial nature of the spreading infection and the subsequent T cell search, we eschew traditional differential equation models \citep{Beauchemin2006a, Bauer2009} in favor of a spatial agent-based model (ABM).
 
@@
 The model focuses on activated CD8 T cells migrating through the vascular and lymph networks, including their movement over tissue after extravasation (Fig.~\ref{fig:systemchart}).
-The lung is represented as a two-dimensional sheet of healthy epithelial cells (S1.3).
+While the volume of the lung consists of spherical alveoli, the spread of influenza throughout the lung occurs through the interconnected bronchiole-alveoli complex, which can be viewed as a connected monolayer.
+Thus, we represent the lung as a two-dimensional sheet of healthy epithelial cells (S1.3).
 Because we assume activated T cells descend at random into branches of the vascular network, T cells in the model are introduced uniformly at random across the modeled lung surface.
@@
 The model is initialized with a single infected cell.
-After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (S2.1 and Table \ref{tab:strains}).
+After the incubation period, the infected cell begins secreting virus and chemokine at rates determined by the virus strain and chemokine type (Table \ref{tab:strains}).
 Virus from expressing cells diffuses locally, infecting neighboring cells.
@@
 
-The model is implemented using the CyCells software \cite{Warrender2006} (\texttt{https://github.com/drewlevin/cycells}), a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).
-Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \cite{Miller2003, Allan2006, Ingulli2009}.
+The model is implemented using an updated version of the CyCells software \citep{Warrender2006},
+(\texttt{github.com/drewlevin/cycells}),
+a modeling platform for two- or three-dimensional agent-based simulations of the immune response (S1.1).
+Due to computational constrains our model is scaled to the size of a mouse lung (approximately 100$cm^2$), a well-established practice in spatially explicit influenza models \citep{Miller2003, Allan2006, Ingulli2009}.
 Parameters affected by this choice include the T cell production rate, T cell circulation time, and the total size of the lung (Table \ref{tab:parameters}).
@@
 
-In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \cite{bachem1996simulated, Beauchemin2005, Mitchell2011}.
+In the model, epithelial cells are stationary and are described by one of five sequential states: \texttt{healthy}, \texttt{virus-incubating}, \texttt{virus-expressing}, \texttt{apoptotic}, and \texttt{dead} \citep{bachem1996simulated, Beauchemin2005, Mitchell2011}.
 \texttt{Healthy} cells remain unchanged unless infected by virus.
@@
 These values were taken from the literature when available.
-Other parameters were determined experimentally, and some parameters were estimated using biologically plausible ranges.
-All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior.
+Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.
+All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).
 
 Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
-We extended a previously published differential equation model \cite{Mitchell2011} to obtain values for per-cell chemokine production rates (S2.1).
-T cell production rates were derived from measured replication rates \textit{in vitro} \cite{Miao2010} using another differential equation model (S2.2).
-T cell production is assumed to be constant after day 5 \cite{MartIn-Fontecha2003}.
+We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
+T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
+T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.
+
+\subsection*{Models for Parameter Estimation}
+
+We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \citep{Mitchell2011} Eq.
+1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
+Initial population sizes and parameter values are taken from the previous study.
+{\footnotesize
+\begin{equation*}
+\begin{aligned}
+\dot{T} &= - \beta T V \\
+\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
+\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
+\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
+\dot{F} &=  I_{1 \tau_2} \\
+\dot{C} &= r I_{1 \tau_3} - d C \\
+\end{aligned}
+\tag{Eq. 1}
+\label{eq:dde}
+\end{equation*}
+\vspace{.05in}
+}
+
+$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
+Table S2 summarizes population and parameter values and descriptions.
+Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
+The results of these fits are discussed in the main paper and shown in Table 1.
+IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
+
+%\subsection*{Chemokine Secretion}
+
+%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
+The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \citep{Mitchell2011}.
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
+
+%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref.
+\citep{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).
+Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
+Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
+There was no positive correlation between viral production and induced chemokine production across the three strains.
+
+% \subsection{Estimating the T cell production rate}
+
+We calculate the rate of production, $\sigma$, of CD8 T cells using a differential equation model from \citep{Miao2010a}.
+The equations model T cell production and subsequent search over an area of infected lung tissue.
+
+{\footnotesize
+\begin{equation*}
+\label{eq:miao}
+\begin{aligned}
+\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
+\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
+\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
+\dot{T} &= \rho T -\beta TV \\
+\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
+\dot{V} &= pI - \beta TV - \gamma (t) V \\
+\gamma (t) &= \left\{ \begin{array}{rcl}
+	1/\mbox{day} & \mbox{,}  & t < 5  \\
+	3/\mbox{day} & \mbox{,} & t \geq 5  
+	\end{array}\right. \\
+\end{aligned}
+\tag{Eq. 2}
+\end{equation*}
+}
+\vspace{0.5in}
+
+$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region.
+$T$ is the number of uninfected target cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
+The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
+The lung is modeled as a circular region with radius $R$.
+The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
+See Fig.
+1 for more details about the model.
+The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
+$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
+
+Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
+We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
+
+The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.
+Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time.
+Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
+Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
+Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
+Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
+The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
+The initial number of infected cells is assumed to be zero.
+Parameter values are listed in Table S3.
+This ODE was fit to data taken from \citep{Miao2010a} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
+The final value was found to be 1,257 per hour.
 
@@
 In an effort to obtain per-cell chemokine and cytokine production rates, human epithelial cells were infected with influenza \textit{in vitro}.
-Epithelial cell culture and supernatant collection was performed as described in \cite{Mitchell2011}.
+Epithelial cell culture and supernatant collection was performed as described in \citep{Mitchell2011}.
 Briefly, undifferentiated human tracheal epithelial cells (University of Miami) were cultured for 4 weeks to achieve fully differentiated confluent monolayers on collagen-coated transwell inserts, or commercial differentiated human bronchial epithelial cells (EpiAirway Tissue, MatTek Corp., Ashland, MA) used immediately upon receipt, were infected at an MOI of 0.01 with either seasonal H1N1 virus A/New Caledonia/20/99 (sH1N1), the 2009 H1N1 pandemic strain A/California/04/09 (pH1N1), or avian H5N1 virus A/Hong Kong/483/97 (aH5N1) derived from a fatal human infection.
 Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
-Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \cite{Mitchell2011}.
+Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus \citep{Mitchell2011}.
 Quantitative viral culture was performed by standard plaque assay.
@@
 
-Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
+Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
 For the avian and seasonal strains, all model runs cleared the virus by day 10.
-Conversely, each simulation of the 2009 pandemic influenza led to runaway infection.
+Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.
 
@@
 The $R^2$ was computed with respect to the mean number of infected cells at each simluated time point across all 50 runs.
-Consistent with earlier results \cite{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).
+Consistent with earlier results \citep{Mitchell2011}, the time course of the infection varies significantly across the three strains (Fig.~\ref{fig:variance}).
 
@@
 For sH1N1, the maximum standard deviation was 242, which occurred at day 4.93 where the mean value was 2,607.
-For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018.
+For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018 (Fig.~\ref{fig:variance}).
 
-In all studied strains, virus expansion declines at four days p.i.
+In all studied strains, virus population growth slows at four days p.i.
 (Fig.~\ref{fig:variance}), due to IgM appearance.
 Subsequently, the simulated CD8 T cell response causes the number of infected cells to decline quickly after day five p.i.
-aH5N1 is cleared completely, sH1N1 is cleared more slowly, and at day six p.i., pH1N1 replication dominates the IgM and T cell response.
-These results agree with previously reported \textit{in vivo} differences among the three different strains \cite{Mitchell2011}.
+aH5N1 is cleared completely, sH1N1 is cleared more slowly.
+In contrast, at day six p.i., pH1N1 replication overwhelms the IgM and T cell response.
+These results are consistent with previously reported \textit{in vitro} differences among the three different strains \citep{Mitchell2011}.
 
-% Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+% Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \citep{Zeng2007}.
 
@@
 
-Plaque growth in the model is illustrated in Figure \ref{fig:cycells}.
+Plaque growth in the model is illustrated in Figure \ref{fig:cycells} and Videos S1-S3.
 Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly, according to the Stokes-Einstein equation.
 However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).
-In the model these countervailing forces produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}).
-Before day 4 the plaque is dominated by active (incubating and secreting) cells, with a small percentage of dead cells (Panel A).
+In the model these countervailing characteristics produce similar spatio-temporal profiles for the two particle types (Fig.~\ref{fig:cycells}).
+Before day 4 the plaque is populated primarily by virus-incubating and virus-secreting cells, with a small percentage of dead cells (Panel A).
 Over time, cells in the plaque's interior die, and active cells comprise a decreasing proportion of the plaque.
 T cells arrive at day 5 and begin killing the virus-secreting cells (Panel B).
-By day 6 many expressing cells have been eliminated and the plaque is dominated by dead cells (Panel C).
-Although not visible in the figure, plaque size and T cell number changes continuously for all strains and is demonstrated for seasonal H1N1 virus (Videos S1-S3).
+By day 6 many expressing cells have been eliminated and the plaque is populated by dead cells (Panel C).
 
@@
 However, secreting cells were not cleared in either the sH1N1 or pH1N1 simulations (Fig.~\ref{fig:variance}).
-In these runs, secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i.
+In these runs, virus-secreting cells accounted for at most 10\% of the active cell population and less than 1\% of the total plaque at 6 days p.i.
 (Fig.~\ref{fig:plaquesize}).
 T cells continue to accumulate, but they arrive at a slower rate than the rate at which new cells become infected.
-Further, the newly arriving T cells are less efficient because the epithelial cells are more sparsely distributed in the plaque.
+Further, the newly arriving T cells are less efficient because the infected epithelial cells are more sparsely distributed in the plaque.
 
-Finally, the spatial nature of our model reveals that regions of concentrated chemokine diffuse more slowly than the ever-expanding infected cell and virus locations, thus misdirecting the chemotaxing T cells.
-It takes time for infected cells to begin producing chemokine, while the preexisting pockets of high chemokine density are slow to decay.
+Finally, the spatial nature of our model reveals that chemokines can diffuse more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.
+It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.
 Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
@@
 
-We conducted a sensitivity analysis of sixteen model parameters to identify those that are most important for determining model outcomes.
+We conducted a sensitivity analysis of sixteen model parameters to identify those that have an effect on determining model outcomes (S2.3).
 We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.
-T cell radius is known \cite{abbas2011cellular} and does not factor strongly into the spatial model's implementation.
-T cell sensitivity was set arbitrarily to mimic a realistic threshold of response to cytokine (S2.3).
-The two onset parameters are known \cite{Diamond2003}.
-Viral production is an independent variable, as discussed above, and determined by each of the three strains \cite{Mitchell2011}.
+T cell radius is known \citep{abbas2011cellular} and does not factor strongly into the spatial model's implementation.
+T cell sensitivity was set to mimic a realistic threshold of response to cytokine (see S2.1 for a more complete explanation of how T cell sensitivity was determined).
+The two onset parameters are known \citep{Diamond2003}.
+Viral production is an independent variable, as discussed above, and determined by each of the three strains \citep{Mitchell2011}.
 
@@
 Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.
-Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters affect the behavior of the model whenever they are changed.
+Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters alter the outcome of the simulation for any variation within a biologically plausible range.
 
@@
 Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.
-None of the remaining T cell parameters or the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
+None of the T cell parameters nor the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
 This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.
@@
 Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).
-We asked whether T cells could play a role by assessing multiple factors that control the overall T cell response.
-Here, we define the window of control as the maximum time virus-secreting cells may produce virus while still allowing the infection to be cleared.
+Here we examine the role of T cells by assessing multiple factors that control the overall T cell response.
+We seek to identify the maximum amount of time virus-secreting cells may produce virus while still allowing the infection to be cleared.
+We name this threshold the `window of control'.
 This window can be defined as the combination of the the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).
+We examine how much virus a virus-secreting cell can produce inside this window.
+Values above one suggest a growing infection, while values below one suggest infection control.
 
-An epithelial cell infected with pH1N1 produces new virus at the rate of 5.08e-3 PFU/s (Table~\ref{tab:strains}).
-Using the parameters from Table \ref{tab:parameters}, T cell-induced apoptosis occurs within 10 minutes of T cell arrival, viral secretion continues for one hour after apoptosis is initiated, and under these circumstances the cell produces 21 new viral particles that can infect new cells.
-In contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles in the 70 minute window, and for the aH5N1 strain this number is 0.2.
+An epithelial cell infected with pH1N1 produces new virus at the rate of 5.1e-3 PFU/s (Table~\ref{tab:strains}).
+Assuming instantaneous T cell arrival at a new virus-secreting cell, T cell-induced apoptosis occurs within 10 minutes ($T_{kill}$) and viral secretion continues for 60 minutes after that ($T_{apop}$) according to our model (Table \ref{tab:parameters}).
+Under these circumstances the cell produces 21 new viral particles that can infect new cells in this 70 minute window that occurs after T cell arrival.
+In contrast, with the sH1N1 strain, a single infected cell produces 1.6 viral particles, and for the aH5N1 strain this number is 0.2.
 % Thus, viral clearance depends on the interaction of the rate of apoptosis, viral decay, viral production, and the viral infection probability.
@@
 Even if T cell arrival and apoptosis time were instantaneous, a cell would still secrete virus for 10 minutes ($T_{kill}$), producing 3 new virions in the case of pH1N1.
-Only in the case where both $T_{kill}$ and $T_{apop}$ are simultaneously reduced can the pandemic infection be cleared.
-We confirmed this effect in the model by first setting both relevant parameters (apoptosis time and T cell kill time) to zero, and as expected all three strains were cleared (Fig.~\ref{fig:instantkill}).
-We now ask how much T cell delay the model can tolerate and still clear a given viral strain.
+Only in the case where both $T_{kill}$ and $T_{apop}$ are simultaneously reduced can the pandemic infection be limited to less than one new virion per infected cell.
+We confirm this effect in the model by first setting both relevant parameters (apoptosis time and T cell kill time) to zero, and as expected all three strains were cleared (Fig.~\ref{fig:instantkill}).
+We now ask how much T cell arrival delay the model can tolerate and still clear a given viral strain.
 To answer this, we formulate the basic equations of the model in terms of $R_0$, the viral replication rate, and solve for the case where $R_0 < 1$.
 
-\begin{equation}
+\begin{equation*}
 R_0 = p \times (T_{arr} + T_{kill} + T_{apop}) \times E
-\end{equation}
+\tag{Eq. 3}
+\end{equation*}
 
@@
 (Table \ref{tab:parameters}).
-This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (S2.6).
+This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (see S2.4 for a more complete explanation of how $E$ was determined).
 Solving for $T_{arr}$ where $R_0 < 1$ gives:
 
-\begin{equation}
+\begin{equation*}
 \lim_{t \to \infty} T_{arr} < \frac{1}{pE} - (T_{kill} + T_{apop})
-\end{equation}
+\tag{Eq. 4}
+\end{equation*}
 
@@
 The above calculation suggests a combined transition point for sH1N1 at 163 minutes ($R_0 = 1$ for $T_{arr}+T_{kill}+T_{apop}=163m$), which is consistent with the sensitivity analysis (Fig.~S4: Apoptosis Time: 90m - 120m) and suggests that the modeled $T_{arr}$ is between 33 and 63 minutes for sH1N1.
-These data show that the balance between viral production and T cell response is a key factor in clearance of infection: pH1N1 can theoretically be cleared with instantaneous T cell responses, but in actuality viral production exceeds T cell response time and this leads to uncontrolled infection.
+These data show that the balance between viral production and T cell response is a key factor in clearance of infection: pH1N1 can theoretically be cleared with instantaneous T cell effectiveness (Fig.~\ref{fig:instantkill}), but in actuality viral production exceeds T cell response time and this leads to uncontrolled infection.
 
@@
 CD8 T cells conduct two `searches' in two different tissue environments, first to encounter antigen-loaded dendritic cells in the lymph node, and second to encounter local inflammatory signals in the infected lung.
-Search problems in the lymph node have been simulated using live cell imaging data to provide reliable parameters of cell movement \cite{Vroomans2012}.
+Search problems in the lymph node have been simulated using live cell imaging data to provide reliable parameters of cell movement \citep{Mirsky2011, Vroomans2012}.
 Our model focused on the second search, i.e.
 the process of recruiting activated CD8 effector cells to infected sites in the lung, which is not as well understood.
-In contrast, the dense 3D lymph node volume, our model approximates a mouse lung as a 2D surface where the epithelial cell monolayer is 100 $cm^2$ (only the area containing the FOI is modeled explicitly).
+In contrast to the dense 3D lymph node volume, our model approximates a mouse lung as a 2D surface where the epithelial cell monolayer is 100 $cm^2$ (only the area containing the FOI is modeled explicitly).
 T cells must somehow rapidly locate the infection, which at day 5 p.i.
 is approximately 0.5 $mm$ in diameter.
-Our model revealed four main conclusions regarding chemokine aided T cell search in the lung.
+Our model revealed four main insights regarding chemokine aided T cell search in the lung.
 
-First, confirming earlier results \cite{Mitchell2011}, the model consistently clears the aH5N1 strain by day 10 p.i., contains sH1N1 by day 10 p.i., and fails to clear the pH1N1 infection.
+First, %confirming earlier results \citep{Mitchell2011}, 
+the model consistently clears the aH5N1 strain by day 10 p.i., contains sH1N1 by day 10 p.i., and fails to clear the pH1N1 infection.
 Second, the model revealed spatial constraints on T cell search when the infection spreads more quickly than the chemokine gradient can diffuse.
@@
 Third, the sensitivity analysis tested the individual effects of each parameter on the model's behavior.
-The analysis shows that virus kinetics largely determine the course of the infection, while T cell and chemokine parameters do not have a strong effect.
-Finally, we examined the window of control which suggests that pandemic influenza could not be cleared even if apoptosis were instantaneous.
+The analysis shows that infection outcomes are highly sensitive to the properties of the specific virus.
+% virus kinetics largely determine the course of the infection, while T cell and chemokine parameters do not have a strong effect.  
+Finally, we examined the window of control which suggests that pandemic influenza could not be cleared even if T cell arrival was instantaneous.
 
@@
 
-% The initial localization and extravasation of T cells into the bronchial tissue may depend on a number of inflammatory signals, but viral antigen does not appear to induce recruitment into this tissue \cite{Topham2001}.
+% The initial localization and extravasation of T cells into the bronchial tissue may depend on a number of inflammatory signals, but viral antigen does not appear to induce recruitment into this tissue \citep{Topham2001}.
 Moreover, the T cells leaving the lymph node are released into the very large systemic vascular `space' with no initial guiding signals.
-Effector cells must localize to sites of viral replication \cite{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
+Effector cells must localize to sites of viral replication \citep{Cerwenka1999}, but it is not clear whether cells passing through uninfected lung tissue leave through the pulmonary vein or exit the capillary bed and leave the lung through lymphatic channels.
 
 % In this paper, we assume T cells climb the chemokine gradient though ligand-receptor binding.
-The consequences of chemotactic ligand-receptor interactions, however, are complex \cite{Groom2011a, Groom2011} and variable in the models studied.
-For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediates T cell recruitment to infected brain and subsequent immunopathology \cite{Christensen2004, Christensen2006}, while the CCR5 mediates the opposite effect \cite{DeLemos2005}.
-In contrast, in the West Nile virus model \cite{Klein2005} and the dengue model \cite{Hsieh2006}, deficiency of IP-10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.
-In the herpesvirus model, IP-10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \cite{Wuest2008, Thapa2008}.
-In the parainfluenza virus model, CXCR3 receptor is critical in CD4+ T cell migration to the lung \cite{Kohlmeier2009}.
-In the influenza A model, initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \cite{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.
-The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency, but viral clearance was not altered \cite{Fadel2008}.
-The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \cite{Kohlmeier2011}.
+The consequences of chemotactic ligand-receptor interactions, however, are complex \citep{Groom2011a, Groom2011} and variable in the models studied.
+For example, in the lymphochoriomeningitis virus model the CXCR3 receptor mediates T cell recruitment to infected brain and subsequent immunopathology \citep{Christensen2004, Christensen2006}, while the CCR5 mediates the opposite effect \citep{DeLemos2005}.
+In contrast, in the West Nile virus model \citep{Klein2005} and the dengue model \citep{Hsieh2006}, deficiency of IP-10 reduced T cell recruitment to the brain resulting in higher viral burden and increased mortality.
+In the herpesvirus model, IP-10 was critical in T cell recruitment and disease control in the HSV-2-infected brain \citep{Wuest2008, Thapa2008}.
+In the parainfluenza virus model, CXCR3 receptor is critical in CD4+ T cell migration to the lung \citep{Kohlmeier2009}.
+In the influenza A model, initial studies with chemokine receptor knock-out mice obtained mixed results with respect to changing the course of disease \citep{Dawson2000, Wareing2004}, concluding that redundancy in chemokine signals may confound interpretation.
+The CXCR3 receptor mediated T cell localization and spared the increased mortality of CCR5 deficiency, but viral clearance was not altered \citep{Fadel2008}.
+The CXCR3 receptor also mediates the balance between effector versus memory cell differentiation among recruited CD8 T cells in the lung \citep{Kohlmeier2011}.
 
 The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
-There is clear evidence from multiple models that chemokines and chemokine receptors are required for effector T cell localization to infected tissues \cite{Christensen2004, Christensen2006, Dawson2000, DeLemos2005, Fadel2008, Gadhamsetty2014, Groom2011a, Groom2011, Hsieh2006, Klein2005, Kohlmeier2009, Kohlmeier2011, Thapa2008, Wareing2004, Wuest2008, Pawelek2012}.
+There is clear evidence from multiple models that chemokines and chemokine receptors are required for effector T cell localization to infected tissues \citep{Christensen2004, Christensen2006, Dawson2000, DeLemos2005, Fadel2008, Gadhamsetty2014, Groom2011a, Groom2011, Hsieh2006, Klein2005, Kohlmeier2009, Kohlmeier2011, Thapa2008, Wareing2004, Wuest2008, Pawelek2012}.
 However, the actual effect of chemokines on effector T cells in tissues is still largely unknown.
-Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} and may better approximate chemokine levels in tissue.
+Blood levels documented in virulent influenza infections \citep{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \citep{Mitchell2011, Chan2010, Chan2005, Zeng2011} and may better approximate chemokine levels in tissue.
 We have determined the level of chemokine released by infected bronchial epithelial cells (Table~\ref{tab:strains}) and used these as the best approximation for chemokine levels in tissues.
-Our model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+Our model did not incorporate  the potential contributions from other cytokines that did not show different levels in the physical experiment.
+This includes chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \citep{Matsukura1996, Arndt2002} as well as chemokines secreted by immigrant macrophages \citep{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \citep{Zhao2000}.
 The sensitivity analysis classifies the chemokine secretion rate as stable, suggesting the total amount of chemokine is not a significant factor in determining viral clearance in our model.
 
-Although leukocytes exhibit directional behavior to chemokines \cite{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
-While T cell chemotaxis may not be proven, previous work has shown severe limitations to viral clearance in the absence of this ability \cite{Banerjee2011}.
-Our sensitivity analysis shows that reducing T cell velocity (and consequentially, the ability of T cells to climb gradients) hinders clearance of sH1N1 and pH1N1 strains.
+Although leukocytes exhibit directional behavior to chemokines \citep{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
+While T cell chemotaxis may not be proven, our previous work has shown severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
+Our sensitivity analysis shows that slowing T cell velocity (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
 These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.
@@
 
-% Another key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+% Another key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \citep{Thelen2008}.
 The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
 For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
-This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \citep{Francis1997}.
 
@@
 Maio et al.
-uses ODEs to examine how the T cell response coupled with IgG and IgM affect influenza infection in mice \cite{Miao2010}.
+uses ODEs to examine how the T cell response coupled with IgG and IgM affect influenza infection in mice \citep{Miao2010a}.
 Saenz et.
 al.
-examines the effect of the interferon response on influenza infection in equines \cite{Saenz2010}.
+examines the effect of the interferon response on influenza infection in equines \citep{Saenz2010}.
 Handel et.
 al.
-similarly uses ODE models to examine which immune responses dominate during the course of a influenza infection in mice \cite{Handel2008}.
+similarly uses ODE models to examine which immune responses dominate during the course of a influenza infection in mice \citep{Handel2008}.
 Lee et.
 al.
-create a very comprehensive model of the adaptive immune response, including dendritic cells, CD4+ T cells, CD8+ T cells, and B cells \cite{Lee2009}.
-Our previous work uses an ODE model influenza infection matched to \textit{in vitro} experimental data to estimate per-cell viral production rates for the three different strains of influenza discussed in this paper: aH5N1, sH1N1, and pH1N1 \cite{Mitchell2011}.
+create a very comprehensive model of the adaptive immune response, including dendritic cells, CD4+ T cells, CD8+ T cells, and B cells \citep{Lee2009}.
+Our previous work uses an ODE model influenza infection matched to \textit{in vitro} experimental data to estimate per-cell viral production rates for the three different strains of influenza discussed in this paper: aH5N1, sH1N1, and pH1N1 \citep{Mitchell2011}.
 
@@
 al.
-\cite{Beltman2007} and Zheng et.
+\citep{Beltman2007} and Zheng et.
 al.
-\cite{Zheng2008} successfully use unique spatial models to examine how viral peptides are presented to unactivated T cells in the lymph node.
+\citep{Zheng2008} successfully use unique spatial models to examine how viral peptides are presented to unactivated T cells in the lymph node.
 
@@
 The contribution of IgM antibody clearing free virus is represented as a constant rate, and the class switch to higher affinity antibody mediated by CD4+ T cells is not represented.
-All of these rates may in fact be time-variable as indicated by data-fitting models \cite{Wu2011}.
+All of these rates may in fact be time-variable as indicated by data-fitting models \citep{Wu2011}.
 The immigration and contributions of virus-nonspecific immune cells such as macrophages and/or dendritic cells are not represented in our model.
-Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \cite{Miao2010}.
+Finally, the proliferation of activated T cells in lung tissue is not represented, but is thought to be crucial to the control of lung infection \citep{Miao2010a}.
 These assumptions allow us to build a tractable model of the aspects of the immune response necessary to explore specific aspects of virus and immune system interactions.
@@
 %The \textit{in vivo} quantitative chemokine parameters in the infected lung are difficult to estimate.
-Blood levels documented in virulent influenza infections \cite{DeJong2006} may not reflect lung tissue concentrations.
-Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \cite{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
-Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
-The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \cite{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \cite{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \cite{Zhao2000}.
+Blood levels documented in virulent influenza infections \citep{DeJong2006} may not reflect lung tissue concentrations.
+Dynamic chemokine concentrations secreted by bronchial epithelial cells \textit{in vitro} depend on infection intensity and cell maturation state \citep{Mitchell2011, Chan2010, Chan2005, Zeng2011} but may better approximate tissue levels.
+Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \citep{Zeng2007}.
+The model did not incorporate the potential contributions from other chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \citep{Matsukura1996, Arndt2002}, nor did the model account for chemokines secreted by immigrant macrophages \citep{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \citep{Zhao2000}.
 %
-%A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \cite{Thelen2008}.
+%A key determinant in the efficiency of chemokine-directed T cell migration towards virus-secreting epithelial cells is the communication distance, defined by the threshold of sufficient chemokine signal required to induce directed motion of the cell up the chemical gradient \citep{Thelen2008}.
 The diameter of this gradient generated by a single cell is a function of production rate, decay rate, protein diffusion and the sensitivity threshold.
 For the threshold of 100 ng/mL and maximal levels of concentration in tissue of 10,000 ng/mL, we calculated the effective communication distance to be approximately 100 microns in our model by simulating a single chemokine producing cell and observing the radius of the resulting chemokine gradient.
-This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \cite{Francis1997}.
+This calculation is similar to the distance calculated for generic cytokines secreted by a suspended solitary cell \citep{Francis1997}.
 %
@@
 
-Most earlier studies of T cell motility and search have focused on the lymph node.
-Yet, circulation through the peripheral tissue is critical to clearance \cite{Pawelek2012, Ganusov2014}.
+% Most earlier studies of T cell motility and search have focused on the lymph node.
+Yet, circulation through the peripheral tissue is critical to clearance \citep{Pawelek2012, Ganusov2014}.
 We studied the dynamics of T cell movement in tissue, accounting for observed phenotypic differences in three strains of influenza.
@@
 
-Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \cite{Handel2008, Lee2009, Miao2010, Mitchell2011, Saenz2010}.
-The use of an ABM can complement spatially homogeneous differential equation models \cite{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.
-In this study, we examined how chemokine-directed T cell search contributes to infection clearance.
+Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.
+The use of an ABM can complement spatially homogeneous differential equation models \citep{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.
+In this study, we tested how chemokine-directed T cell search contributes to infection clearance.
 Explicitly modeling T cell chemotaxis revealed unique challenges to viral clearance brought about by infections that spread more rapidly than the chemokine gradient could decay and diffuse.
@@
 ODE models account for this discrepancy by lowering rates of infection by a constant amount, but this assumes that the proportion of unsuitable virions will always be the same.
-Our model shows how the early infection has a higher proportion of virus overlapping healthy cells (Figure~\ref{fig:cycells}).
+Our model visualizes how the early infection has a higher proportion of virus overlapping healthy cells (Figure~\ref{fig:cycells}).
 
-In the model, CD8 T cells climb a chemokine gradient to find infected epithelial cells and cluster at local maxima of chemokine concentration.
+In our spatial model, CD8 T cells climb a chemokine gradient to find infected epithelial cells and cluster at local maxima of chemokine concentration.
 Because T cells are clustered, they cannot cover the expanding plaque effectively, where infected cells on the periphery become more highly dispersed as the plaque grows.
-Thus, T cells in the model saturate at a relatively low threshold, beyond which additional T cells do not improve clearance rates.
+Thus, T cells in the model become redundant at a relatively low threshold, beyond which additional T cells do not improve clearance rates.
 These observations provide an explanation for pH1N1 dynamics that would be obscured without the visualizations provided by spatial modeling.
-These results, together with the sensitivity analysis, suggest that viral dynamics are the dominant influence in the course of influenza infection.
+%These results, together with the sensitivity analysis, suggest that viral dynamics are the dominant influence in the course of influenza infection.
 Changes in the chemokine and T cell parameters did not have a strong effect in the model (Table \ref{tab:sensitivity}).
@@
 The window of control describes the limit on the amount of time an infected cell may secrete virus while still allowing the infection to be cleared.
-Studying the delay in the model revealed one reason why T cells fail to clear pH1N1 infections.
+Studying the time of T cell arrival at virus-secreting cells in the model revealed one reason why T cells fail to clear pH1N1 infections.
 We formalized this effect in an equation which calculates an upper bound on the search time for T cells to find new virus-secreting cells, and we showed that this value is consistent with the results of our sensitivity analysis.
 
-Our model captures important aspects of T cell response to influenza infection which have not been addressed by earlier T cell models.
+Our model captures important quantitative and spatial aspects of T cell response to influenza infection which have not been addressed by earlier T cell models.
 The results of our sensitivity analysis are important for future biological modeling and experimental design.
-Empirical verification of the model's sensitive parameters (T cell production rate, viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
-Finally, both qualitative and quantitative analysis of our model suggests therapeutic interventions of influenza infection in the lung will likely be most effective by focusing on viral control rather than enhancing T cell responses.
+Empirical verification of the model's sensitive parameters (viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies.
+Finally, knowledge of the spatial nature of the interaction between T cells and chemokines can aid in the design of future immune system models.
 
@@
 
-%\section*{References}
+\section*{References}
 % The bibtex filename
+\singlespacing
 \bibliography{references}
@@
 sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.
-An extended differential equation model from \cite{Mitchell2011} was fit to IP-10 and RANTES data (Section S2.1 Equation S1) .
+An extended differential equation model from \citep{Mitchell2011} was fit to IP-10 and RANTES data (Equation 1) .
 These fits were used to obtain chemokine production values for use in the spatial CyCells model.
@@
 \caption{Strain-specific parameters.
-Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \cite{Wu1986}.
+Small text values show 95\% confidence intervals resulting from 1,000 bootstrapping runs for each parameter \citep{Wu1986}.
 Bootstrapping for the chemokine values was performed using the original fit of Eq.
 S1 to the data in Fig.~\ref{fig:data} to produce new data sets.
-Viral production values and confidence intervals are taken from \cite{Mitchell2011}.}
+Viral production values and confidence intervals are taken from \citep{Mitchell2011}.}
 \label{tab:strains}
@@
 \begin{center}
-\begin{tabular}{| l l l |}
+\begin{tabular}{| l | c | l l l | c |}
   \hline                        
-  Referenced Parameters & Value & Source \\
+  Referenced Parameters & Units & Min & Value & Max & Source \\
   \hline
-  Viral Diffusion in Airway & .0318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
-  Viral Decay in Airway &  1/day & \cite{Lee2009} \\
-  Chemokine Diffusion Rate & .318 $\mu$m$^2$/s & \cite{Beauchemin2006} \\
-  Incubation Time &  10 hours & \cite{Mitchell2011} \\
-  Epithelial Cell Radius & 5 $\mu$m & \cite{Elbert1999} \\
-  T Cell Radius & 5 $\mu$m & \cite{abbas2011cellular} \\
-  T Cell Production Rate & 1257 cells/hour & \cite{Miao2010} \\ 
-  T Cell Speed & 6 $\mu$m/min & \cite{Egen2011} \\
-  Blood Circulation Time & 6 seconds & \cite{Banerjee2010b} \\
-  T Cell Sensitivity to Chemokine & 100 ng/mL & \cite{Nandagopal2011} \\
-  Onset of ATC$^\ddagger$ Lymph Node Exit & Day 5 & \cite{Banerjee2011} \\
-  Viral Response to IgM & Viral decay of 10/day & \cite{Diamond2003} \\
-  IgM Onset & Day 4 & \cite{Diamond2003} \\
+  $^{\dagger}$Viral Diffusion in Airway & \textit{$\mu$m$^2$/s} & 3.18e-4 & \textbf{3.18e-2} & 3.18  & \citep{Beauchemin2006} \\
+  $^{\dagger}$Viral Decay in Airway & \textit{day$^{-1}$} & 0.01 & \textbf{1} & 100 & \citep{Lee2009} \\
+  Chemokine Diffusion Rate & \textit{$\mu$m$^2$/s} & 3.18e-3 & \textbf{.318} & 318 & \citep{Beauchemin2006} \\
+  Incubation Time & \textit{hours} & 5 & \textbf{10} & 20 & \citep{Mitchell2011} \\
+  Epithelial Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \citep{Elbert1999} \\
+  T Cell Radius & \textit{$\mu$m} & | & \textbf{5} & | & \citep{abbas2011cellular} \\
+  T Cell Production Rate & \textit{cells/h} & 125 & \textbf{1,257} & 3,750 & \citep{Miao2010a} \\ 
+  T Cell Speed & \textit{$\mu$m/min} & 6e-2 & \textbf{6} & 600 & \citep{Egen2011} \\
+  Blood Circulation Time & \textit{seconds} & 1 & \textbf{6} & 3,600 & \citep{Banerjee2010b} \\
+  T Cell Sensitivity to Chemokine & \textit{ng/mL} & | & \textbf{100} & | & \citep{Nandagopal2011} \\
+  Onset of T Cell Lymph Node Exit & \textit{days }& | & \textbf{5} & | & \citep{Banerjee2011} \\
+  $^{\dagger}$IgM Viral Decay Factor & | & 1 & \textbf{10} & 1,000 & \citep{Diamond2003} \\
+  IgM Onset & \textit{days} & | & \textbf{4} & | & \citep{Diamond2003} \\
   \hline
   \hline                        
-  Estimated Parameters & Value & Footnote \\
+  Estimated Parameters$^*$ & Units & Min & Value & Max & Footnote \\
   \hline
-  Chemokine Decay Rate & $3.8508\cdot10^{-4}$/s & 1\\
-  Infectivity & 0.5 cell/hour$\cdot$virion &  2 \\
-  Expression Time & 16.7 hours & \cite{Mitchell2011}$^3$ \\
-  T Cell Expected Kill Time & 10 min & 4 \\
-  Apoptosis Time & 60 minutes & \cite{Ganusov2008}$^5$ \\
-  T Cell Age (at FOI$^\ast$) & 2 hours & 6 \\
-  T Cell Age (in Blood) & 3 days & 6 \\
+  Chemokine Decay Rate & \textit{Hz} & 3.8e-6 & \textbf{3.8e-4} & 3.8e-2 & 1\\
+  $^{\dagger}$Infectivity & \textit{min/virion} & 12 & \textbf{120} & 1,200   &  2 \\
+  Expression Time & \textit{min} & 100 & \textbf{1,000} & 3,000 & \citep{Mitchell2011}$^3$ \\
+  T Cell Expected Kill Time & \textit{min} & 0 & \textbf{10} & 100 & 4 \\
+  Apoptosis Time & \textit{hours} & 0 & \textbf{1} & 2 & \citep{Ganusov2008}$^5$ \\
+  T Cell Age (at FOI) & \textit{min} & 12 & \textbf{120} & 1,200 & 6 \\
+  T Cell Age (in Blood) & \textit{days} & 0.04 & \textbf{4} & 400 & 6 \\
   \hline  
 \end{tabular}
-\caption{Values are taken from experimental literature if possible and from earlier modeling papers if not.
-If we could not find a value in any literature we estimated parameters as follows.
+\caption{Default values used for the model in bold.
+Min and max represent the extreme values tested in the sensitivity analysis (Table~\ref{tab:sensitivity} and Fig.~S3-5).
+$^*$ All estimated parameter values are examined in the sensitivity analysis (Table~\ref{tab:sensitivity}, S2.3).
+$^{\dagger}$ denotes parameters determined to be sensitive by the sensitivity analysis.
+Values were taken from experimental literature if possible and from earlier modeling papers if not.
+Parameters not found in literature were estimated as follows.
 1) Corresponds to a 30 minute half-life.
@@
 \end{tabular}
-\caption{\textbf{Sensitivity Results:} The above parameters were varied over predetermined ranges in isolation, resulting in new model runs for every new value tested.
+\caption{\textbf{Sensitivity Results:} The above parameters were varied over predetermined ranges in isolation, resulting in new model runs for every new value tested (Figs.
+S3-S5).
 The results of the sensitivity analysis were then qualitatively evaluated for each individual parameter.

### 2015-09-07 paper/chemokine_supplement_revised.tex
RSI revisions
clean-up to JTB inital state

@@
 
-% Title must be 150 characters or less
 \begin{flushleft}
 {\Large
-\textbf{Chemokines versus viral replication: a spatial of model of lymphocyte search in the influenza infected lung (SUPPLEMENT)}
+\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung (SUPPLEMENT)}
 }
@@
 
-% You may title this section "Methods" or "Models". 
-% "Models" is not a valid title for PLoS ONE authors. However, PLoS ONE
-% authors may use "Analysis" 
 \section{Models}
@@
 
-% Results and Discussion can be combined.
 \section{Results}
-
-\subsection{Estimating chemokine production rates}
-
-We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in \cite{Mitchell2011} Eq.
-1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
-Initial population sizes and parameter values are taken from the previous study.
-{\footnotesize
-\begin{equation*}
-\begin{aligned}
-\dot{T} &= - \beta T V \\
-\dot{I_1} &= \beta T V - \beta T_{\tau_1}V_{\tau_1} \\
-\dot{I_2} &= \beta T_{\tau_1}V_{\tau_1} - \delta I_2 \\
-\dot{V} &= \frac{p}{1+eF} I_2  - \beta T V  \\
-\dot{F} &=  I_{1 \tau_2} \\
-\dot{C} &= r I_{1 \tau_3} - d C \\
-\end{aligned}
-\tag{S1}
-\label{eq:dde}
-\end{equation*}
-\vspace{.05in}
-}
-
-$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
-Table S2 summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
-The results of these fits are discussed in the main paper and shown in Table 1.
-
-%\subsection*{Chemokine Secretion}
-
-%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
-The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \cite{Mitchell2011}.
-IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
-
-%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref.
-\cite{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).
-Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
-The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
-Best-fit expression rates were similar for all strains except for significantly higher RANTES production in aH5N1.
-There was no positive correlation between viral production and induced chemokine production across the three strains.
-
-\subsection{T cell production rate}
-
-We calculate the rate of production, $\sigma$, of CD8 T cells using a differential equation model from \cite{Miao2010}.
-The equations model T cell production and subsequent search over an area of infected lung tissue.
-
-{\footnotesize
-\begin{equation*}
-\label{eq:miao}
-\begin{aligned}
-\dot{N_{c}} &= \sigma - \frac{r'^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} & \pi r'^{2} &= a \sqrt{I} + b \\
-\dot{N'_{f}} &= \frac{(r'^{2} - r^{2}) \cdot N_{c}}{R^{2} \cdot t_{rc}} - \frac{v_{tcell}}{(r' - r) / 2} \hspace*{2cm}  & \pi r^{2} &= \pi r_{cell}^{2} \cdot I \\
-\dot{N_{f}} &= \frac{r^{2} \cdot N_{c}}{R^{2} \cdot t_{rc}} + \frac{v_{tcell}}{(r' - r) / 2} \\
-\dot{T} &= \rho T -\beta TV \\
-\dot{I} &= \beta TV - \delta I - k_{e} N_{f} I \\
-\dot{V} &= pI - \beta TV - \gamma (t) V \\
-\gamma (t) &= \left\{ \begin{array}{rcl}
-	1/\mbox{day} & \mbox{,}  & t < 5  \\
-	3/\mbox{day} & \mbox{,} & t \geq 5  
-	\end{array}\right. \\
-\end{aligned}
-\tag{S2}
-\end{equation*}
-}
-\vspace{0.5in}
-
-$N_{c}$ is the number of circulating activated antigen-specific CD8 T cells, $N'_{f}$ is the number of circulating T cells that have found and exit into a region of lung tissue expressing chemokines, $N_{f}$ is the number of circulating T cells that have found an infected region.
-$T$ is the number of uninfected \textit{target} cells, $I$ is the number of productively infected cells, and $V$ is the viral titer in serum.
-The infected region is assumed to be of radius $r$ and is within a region expressing chemokines of radius $r'$ ($r  < r'$).
-The lung is modeled as a circular region with radius $R$.
-The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
-See Fig.
-1 for more details about the model.
-The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
-$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
-
-Circulating CD8 T cells ($N_{c}$) are assumed to be released from lymph nodes at a constant rate $\sigma$ and circulate to the $N'_{f}$ population over a time defined by $t_{rc}$.
-We assume these circulating cells transition into $N'_{f}$ and $N_{f}$ at rates proportional to the areas of the chemokine expressing and infected regions relative to the whole lung area.
-
-The average time an exiting CD8 T cell takes to migrate to the infected region is equal to the difference in the radii between the two regions ($r' - r$) times the T cell speed, $v_{tcell}$.
-Circulating cells that are in the chemokine expressing region ($N'_{f}$) move into the infected region after performing chemotaxis for that time.
-Target cells ($T$) become infected by virus at rate $\beta TV$, where $\beta$ is the rate constant characterizing infection.
-Infected cells ($I$) die at rate $\delta$ in addition to being lysed by T cells ($N_{f}$) at a rate $k_{e}$.
-Finally the viral titers ($V$) increase due to production of virus at rate $p$ by infected cells.
-Virus is also cleared due to uptake by infected cells (at a rate $- \beta TV$) and due to antibody (at a rate $\gamma (t)$ that changes after 5 days post infection).
-The initial viral titer was initialized to 10,000 PFUs and the initial number of target cells was one million.
-The initial number of infected cells is assumed to be zero.
-Parameter values are listed in Table S3.
-This ODE was fit to data taken from \cite{Miao2010} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
-The final value was found to be 1,257 per hour.
 
@@
 5), based on the calculated per-infected cell secretion rate (Table 3) and known chemical parameters for a 10 kDa protein (Table 2).
-T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.
+T cell sensitivity depends on receptor density \citep{Desmetz2006} and this was assumed to be constant.
 
@@
 Decreased recruitment was predicted only by supra-natural concentration sensitivity levels at and above 1 $\mu$g/mL.
-We then set the sensitivity to 100 $ng/mL$ for all future runs (10 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \cite{Gao2003}.
+We then set the sensitivity to 100 $ng/mL$ for all future runs (10 $nM$ concentration assuming a chemokine molecular weight of 10 $kDa$) \citep{Gao2003}.
 
@@
 
-Because aH5N1 has been shown to suppress the production of interferon \cite{Mitchell2011}, we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.
+Because aH5N1 has been shown to suppress the production of interferon \citep{Mitchell2011}, we consider that IP-10 production is markedly reduced in comparison with that of RANTES and this may lead to elevated RANTES secretion measured in aH5N1 \textit{in vitro} infection when compared to the other strains.
 Thus, IP-10 production was not included in aH5N1 simulations.
 
-%Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \cite{Zeng2007}.
+%Interestingly, the attenuation of the type I interferon response by H5N1 viruses is not associated with attenuation of chemokine secretion in our results and in others \citep{Zeng2007}.
 
@@
 
-T cell sensitivity depends on receptor density \cite{Desmetz2006} and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.
+T cell sensitivity depends on receptor density \citep{Desmetz2006} and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.
 
@@
 Divergence does occur in sH1N1 for the extreme values on both ends (but not for the intermediate ones).
-T cell movement in tissue has been observed \cite{Egen2011}.
+T cell movement in tissue has been observed \citep{Egen2011}.
 Thus, we consider the extreme values biologically implausible.
@@
 While the default of 2 hours may seem short, it appears to generate similar behavior to models with larger values.
+\\
+
+\textbf{T Cell Production Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell production rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.
+
+We tested values up to three times larger and ten times smaller than the default value.  Values beyond a three-fold increase were not computationally tractable.
+
+The T cell production rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
+Thus, it is reasonable to assume that there is a threshold of T cell production beyond which the dynamics of the infection do not change.
+This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
+Increasing the number of T cells in the system does not control the infection beyond a certain point because the T cells overlap in space and become redundant.
+The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.
 
@@
 
-We tested values up to two orders of magnitude larger and smaller than the default parameter value.
+We tested values up to one order of magnitude larger and smaller than the default parameter value.
 
@@
 \\
-
-\textbf{T Cell Production Rate:} [ \textit{125 cells/h | 1257 cells/h | 3750 cells/h} ] - The T cell production rate determines the rate at which T cells enter the blood stream starting at day 5 post-infection.
-
-We tested values up to three times larger and ten times smaller than the default value.  Values beyond a three-fold increase were not computationally tractable.
-
-The T cell production rate does have a consistent response over its different values, but this effect is minimized above a certain rate.
-Thus, it is reasonable to assume that there is a threshold of T cell production beyond which the dynamics of the infection do not change.
-This is consistent with our observations of T cell clumping in areas of high chemokine concentration.
-Increasing the number of T cells in the system does not control the infection beyond a certain point because the T cells overlap in space and become redundant.
-The rapid decline of the infection size for the maximum value of $3 \mu m^2/s$ is simply an artifact of the infection exhausting the entire model space by day 3.
 
@@
 \end{tabular}
-\caption{\textbf{Parameters for Supplemental Model 1 (S2.1 Equation S1)}.
-All parameters and populations are taken from \cite{Mitchell2011} except for $C$, $d$, and $\tau_3$.
+\caption{\textbf{Parameters for Preliminary Model 1 (Equation 1)}.
+All parameters and populations are taken from \citep{Mitchell2011} except for $C$, $d$, and $\tau_3$.
 The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.
@@
   \hline
-  T(0) & $1e6$ & \cite{Mitchell2011} \\
-  V(0) &  $1e4$ & \cite{Mitchell2011} \\
-  $r_{cell}$ &  $5 \mu m$ & \cite{Miao2010} \\
-  $t_{rc}$ & $6s$ & \cite{Peters1983} \\
-  $k_e$ & $6.4e-5/cell/day$ & \cite{Miao2010} \\
-  $\beta$ & $4.8e-7/cell/PFU$ & \cite{Mitchell2011} \\
-  $p$ & $0.18 PFU/h$ & \cite{Mitchell2011} \\
-  $\delta$ & $16.7 hours$ & \cite{Mitchell2011} \\
+  T(0) & $1e6$ & \citep{Mitchell2011} \\
+  V(0) &  $1e4$ & \citep{Mitchell2011} \\
+  $r_{cell}$ &  $5 \mu m$ & \citep{Miao2010a} \\
+  $t_{rc}$ & $6s$ & \citep{Peters1983} \\
+  $k_e$ & $6.4e-5/cell/day$ & \citep{Miao2010a} \\
+  $\beta$ & $4.8e-7/cell/PFU$ & \citep{Mitchell2011} \\
+  $p$ & $0.18 PFU/h$ & \citep{Mitchell2011} \\
+  $\delta$ & $16.7 hours$ & \citep{Mitchell2011} \\
   $a$ & $67.7$ & ABM fits \\
@@
 \end{tabular}
-\caption{\textbf{Parameters for Supplemental Model 2 (S2.2 Equation S2).}  Values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
+\caption{\textbf{Parameters for Preliminary Model 2 (Equation 2).}  Values are chosen to match the equations borrowed from both \citep{Mitchell2011} and \citep{Miao2010a}.
 Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
@@
 %\caption{Supplemental model parameters.
-Most values are chosen to match the equations borrowed from both \cite{Mitchell2011} and \cite{Miao2010}.
+Most values are chosen to match the equations borrowed from both \citep{Mitchell2011} and \citep{Miao2010a}.
 Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}

### 2015-09-07 paper/cover_letter.tex
clean-up to JTB inital state

@@
 
+\pagenumbering{gobble}
+
 \begin{flushleft}Dear Editors, \end{flushleft}
 
-We submit this manuscript entitled "Spatial model of lymphocyte search in influenza-infected lung reveals constraints on chemokine directed migration" for consideration of publication in Royal Society Interface.
+We submit this manuscript entitled "A spatial model of the efficiency of T cell search in the influenza-infected lung" for consideration of publication in Royal Society Interface.
 Our paper implements an agent-based model of bronchial influenza infection and the subsequent T cell response using new empirical values of chemokine secretion.
 The paper shows that chemokine levels have a limited yet important role in directing T cell migration and that the spatial effects of chemokine diffusion can actually hinder T cell search in rapidly spreading viruses.
+Our sensitivity analysis reveals the individual contributions of the model parameters on the course of the infection.
+Finally, we calculate a maximum threshold for T cell search time, beyond which leads to uncontrolled infection.
 We feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
@@
 \begin{enumerate}
-\item Rob de Boer - University of Utrecht
-\item Fred Adler - University of Utah
 \item Judy Day - University of Tennessee 
-\item Hulin Wu - Univ. of Rochester
+\item Grant Lythe - Leeds University
+\item Carmen Molina-Paris - Leeds University
+\item Johannes Textor - University of Utrecht
 \end{enumerate}
 
-We ask that you excuse the following people from review due to involvement in and discussion of earlier iterations of this work:
+We suggest that you excuse the following people from review due to involvement in and discussion of earlier iterations of this work:
 \begin{enumerate}

### 2015-09-07 paper/highlights.tex
clean-up to JTB inital state

\documentclass{article}

\begin{document}

\noindent Highlights:, \\

\vspace{0.25in}

\noindent A spatial agent-based model examines T cell search in an influenza infected lung. \\

\noindent Experimental data show chemokines CXCL10 and CCL5 are stimulated by infection. \\

\noindent Rapidly expanding plaques of infection create challenges for chemotaxing T cells. \\

\noindent Sensitivity analysis isolates the effects of each parameter in the model. \\

\noindent Model parameters are used to determine an upper bound on T cell search time. \\

\end{document}

### 2015-09-07 data.m

@@ -1,289 +1,0 @@
-% Creates Data Plot
-
-figure('Position',[100 500 1800 600])
-
-% --- Avian Data ---
-avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
-av  = [ 100, 150, 825, 2563, 988, 1650, 2838, 17375, 5775];
-ave = [   0,  29, 229,  348, 166,  390,  112,  3062, 1261];
-
-ait = [120, 480, 600, 960, 1200, 1440, 1680,  1920,  2280,  2880]/60;
-ai  = [ 24,  87, 296, 770, 3048, 3901, 8261, 12935, 26970, 24120];
-
-art = [120, 480, 600, 960, 1200, 1440, 1680, 1920, 2280, 2880]/60;
-ar  = [ 45,  50,  57,  87,  134,  151,  150,  245,  780,  695];
-
-% --- Seasonal Data ---
-svt = [600,  720,  960,  1200,  1440,  1800,  2280,  2880]/60;
-sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
-sve = [  0, 1676, 1578, 15695, 17271,  2992,  2886,  3837];
-
-sit = [120,  480,  720,  960, 1200, 1440]/60;
-si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
-
-siit  = [1440, 1800, 2280, 2880]/60;
-%sii   = [8164, 10200, 12913, 16306] - 300;
-sii   = [8164, 8500, 8500, 8500] - 300;
-
-srt = [120, 480, 720, 960, 1200, 1440, 1800, 2280, 2880]/60;
-sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
-
-% --- Swine Data ---
-wvt = [120,   720,   960,   1200,   1440,   1800,   2880]/60;
-wv  = [ 20, 64620, 84053, 254587, 665559, 814586, 654773];
-wve = [  0, 18294, 36080,  43752, 156340, 229980,  47662];
-
-wit = [480,  720,  960,  1200,  1440,  1920,  2880]/60;
-wi  = [ 75+1, 1835, 1349, 23458, 12073, 30700, 19814] - 75;
-
-wrt = [480, 720,  960, 1200, 1440, 1920, 2880]/60;
-wr  = [ 36,  55,   46,  150,   93,  380, 1224];
[…]

### 2016-01-28 paper/response.tex
In process of drafting Response 3

@@
 \usepackage{color} 
+\usepackage[usenames,dvipsnames,table]{xcolor}
 
@@
 
+\usepackage{color}
+\usepackage{csquotes}
+\usepackage{ulem}
+
 %% ** EDIT HERE **
 %% PLEASE INCLUDE ALL MACROS BELOW
+
+\definecolor{dkred}{rgb}{0.75,0,0}
+\definecolor{dkgreen}{rgb}{0,0.5,0}
+\definecolor{dkblue}{rgb}{0,0,0.75}
+\definecolor{dkpurple}{rgb}{.375,0,.375}
+\definecolor{gray}{rgb}{0.5,0.5,0.5}
+
+\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
+%\newcommand{\removed}[1]{\textcolor{dkred}{#1}}
+\newcommand{\new}[1]{{\color{dkgreen}#1}}
+
+\newenvironment{response}{\fontfamily{cms}\selectfont\small}{\par}
+
+\renewcommand{\rmdefault}{cmr}
+\renewcommand{\sfdefault}{lmss}
 
@@
 {\Large
-\textbf{Spatially explicit model of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration (Response to Reviewers)}
+\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung (Response to Reviewers)}
 }
@@
 Candice Clay,
+Judy Cannon, 
 Melanie Moses, 
@@
 We thank the reviewers for their constructive and insightful comments.
-The incorporation of the reviewers' comments has greatly improved the paper.
+We have done our best to address each point raised and feel the incorporation of the reviewers' comments has greatly improved the paper.
 We include detailed responses to each comment below.
@@
 
-\item \textbf{The title may be misleading as the model is very simplified from the immunological viewpoint and gives rather a qualitative rather than a quantitative view of the spatial dynamics.}
+\item \textbf{To me, the most interesting result would be estimating parameters for the dynamics of the virus and production of chemokines but the authors provide little to no details of how this was done.
+This is not sufficient as the model has so many parameters I would like to see how you estimated those.
+What are the model fits?
+Which parameters were fixed in fitting and why?
+What if Mitchell et al.
+study has parameters which are not compartible with your experiments?
+you need to explore this much deeper than the current ms.
+does.}
 
-We changed the title by changing the phrase `quantifies constraints' to `reveals thresholds', which we believe more accurately reflects the paper's main focus. \\
+\begin{response}
+We added one new paragraph and modified another in the Models for Parameter Estimation subsection to address the reviewer's three main questions, as well as others the reviewer did not explicitly ask for.
+In summary:
+\begin{itemize}
+\item The model was fit using a genetic algorithm to minimize the sum of the squared log residuals between the model and the data.
+In an earlier analysis the GA fits were superior to non-linear regression methods in both the Berkeley Madonna and Matlab software suites.
+\item All parameters were held constant during the fitting procedure aside from the chemokine secretion rate.  
+\item The model was fit to experimental data measured during the experiments reported in Mitchell 2011.  However, the chemokine data (Fig. 3, Table S1) have not been published previously.
+\item 1,000 bootstrapping runs were performed to generate confidence intervals over each estimated parameter value.
+\end{itemize}
 
-\item \textbf{Introduction, p3: It would be interesting to add a general reference to agent-based models (ABM), in particular spatial ABM in immunology and briefly justify the use of an ABM rather than partial differential equations in the present context.}
+The new text to address the comment is:
 
-Added reference to Beauchemin 2006 paper that discusses ABM advantages vs. differential equations as well as Bauer 2009.  We have added the following sentences to the introduction: \\
+\begin{displayquote}
+\new{To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.
+3, Table S1).
+The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus (Mitchell et al., 2011).
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..}
 
-\textit{Activated CD8 T cells searching for and homing into infected tissue do so in a spatially complex environment.
-We therefore used agent-based modeling (ABM) to represent the physical environment of the searching T cell.} \\
-
-\pagebreak
-
-\item \textbf{Computational modeling section: there is no reference to Fig. 1 in the main text. As far as I understand the drawing in Fig. 1, I see 16 branches rather than 14, why?}
-
-A reference to Fig.
-1 was added to the Model section.
-The number 14 refers to the depth of the bronchial tree, where each layer represents the bifurcation of a branch.
-Thus, each level results in the doubling of the total number of pathways.
-14 branches refers to $2^{14} = 16,384$ endpoints.
-The figure simplifies this to 4 layers and thus $2^4 = 16$ endpoints.
-The figure has been edited to read `14 bifurcations' rather than `14 branches' and the caption was changed to read `vascular bifurcations' for clarity.
+We estimate chemokine production rates, $r$, by adapting the delay differential equation model of influenza infection described in (Mitchell et al., 2011) Eq.
+1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
+\new{Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.
+Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.}  The results of these fits are shown in Table 1 \new{(viral secretion rates are from previous fits in Mitchell et al., 2011)}.
+IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
 \\
 
-\item \textbf{Acronyms in text and tables should be defined (for example, ODE, FOI, NHBE, ATC, ?).}
+... Eq. 1 ... \\
 
-Acronyms in the text have been either defined or removed.  Acronyms in the table are now defined in the caption. \\
+$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
+Table S2 summarizes population and parameter values and descriptions.
+\removed{Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
+The results of these fits are shown in Table 1.
+IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).}
+\end{displayquote}
+\end{response}
 
-\item \textbf{Model definition:}
-\begin{enumerate}
-   \item \textbf{More details concerning the implementation of the ABM need to be given, at least in the supplement: a short description of the spatial arrangement (size of the simulation space, cell arrangement, number of epithelial cells, as well as how T cell movement vs.
-particle diffusion has been implemented.}
+\item \textbf{I am not completely clear on many of the parameter choices in your model.
+Yes, you cite several studies for the parameter choice but I would like to see a more thorough review of which parameters you chose and why.
+The issue is that your model is not ODE model used in many of the studies you cite, so it is not straightforward to go from one model to another assuming that parameters must be the same (or similar).
+Specifically, how do you obtain 1257 cell/hour emergence from LNs?
+T cells decay with average lifespan of 3 days?
+IgM-mediated viral clearance increase of 10 fold?
+Constant T cell production after day 5?}
 
-   \item \textbf{IgM addition is mentioned in section 2.1 of the supplement but not indicated in the main text.}
+\begin{response}
+We agree with the reviewer that there is no easy answer for which parameters to include or exclude from the model.
+We added a new paragraph to the beginning of the Model Parameters subsection of the Modeling and Methods section to better explain our reasoning:
 
-   \item \textbf{What happens in this ABM when a free virus particle infects a cell: is it absorbed and does it disappear from the pool of free viruses?}
+\begin{displayquote}
+\new{We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection.
+Thus, we included parameters that affected virus, T cells, and chemokines.
+With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.
+The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.
+Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.
+We added IgM because many studies have argued for the importance of antibody mediated virus clearance.
+However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.}
+\end{displayquote}
 
-\end{enumerate}
+The reviewer is correct in observing that agent-based modeling is different from ODE models, we think that using parameters from the ODE models is a good first estimate.
+We included parameters that ODE models have determined (cited in Table 2) and in cases where appropriate (such as the viral infection rate), we adapted values from an ODE model to mirror the expected behavior in a spatial environment.
 
-Two new sections were added to the supplement (S1.2, S1.3).
-One describes details of the general CyCells model environment and the other describes details of the model specific to this paper.
-These sections address all of the points raised here by the reviewer except IgM.
-We added a sentence referencing the IgM implementation to the main paper: \\
+The specific parameters identified by Reviewer 1, each one is listed with a citation or justification in Table 2.
+For the case of the T cell production rate, there is also a subsection in the paper that describes our methodology (Models for Parameter Estimation).
+For the case of the T cell decay rate, three days was an error as the correct value was four days.
+This has been fixed in the manuscript.
+These values are first referenced in the Model Implementation subsection of the Materials and Methods section, before the justification for the values are referenced.
+To address this inconsistency, we moved the Model Parameters subsection to appear \textit{before} the Model Implementation subsection, and we added more direct references to Table 2 and the Models for Parameter Estimation subsection for clarity.
 
-\textit{IgM is modeled by increasing the viral decay rate by a factor of three after the third day.} \\
+The main finding of our study through sensitivity analysis shows that few parameters actually affect influenza clearance.
+Our findings confirm that despite the difficulty selecting parameters to populate the model, selection of most parameters is not likely to be a crucial factor in determining our results.​
 
-\item \textbf{Materials, p6: the sentence "treatment of the monolayer with protease" is not understandable without reading reference [19] and requires here more explanation.}
+\end{response}
 
-The description of viral measurement was expanded into two sentences and an explanation of why protease was used was added: \\
+\item \textbf{The number of simulations performed seems small (50). Will statistics be better if you run more reasonable number of simulations as recommended for bootstrap analyses (e.g., 500-1000).}
 
-\textit{Basal media was collected from previously undisturbed triplicate or quadruplicate wells at 0, 6, 10, 12, 16, 20, 24, 30, 36, 42, 48, and 72 hours after infection, and stored at -80C until assay.
-Subsequently, apical fluid for virus secretion was collected before and after treatment of the monolayer with protease (Pronase, Sigma) to optimize the collection of infectious virus (Mitchell 2011).} \\
+\begin{response}
+The spatial nature of the model made it difficult to perform a full 1,000 sample bootstrapping run.
+Further, bootstrapping works by shuffling residuals of a model fit to generate new data sets for the purpose of generating confidence intervals over fit parameters.
+In this case, we were simply interested in sampling the variance of the stochastic model itself.
+We chose 50 runs of each model parameter set as that number was at the larger end of what was computationally tractable.
 
-\item \textbf{Chemokine production, p6:}
+To satisfy the reviewer's request, we performed an additional 50 samples of each modeled strain of influenza, doubling our sample size.
+Avian influenza's maximum standard deviation changed from 37.2 to 35 (5.9\%), seasonal influenza changed from 242 to 212 (12.4\%), and pandemic influenza changed from 230 to 231 (0.4\%).
+The text and Figure 4 have been updated to include the new model runs.
 
-\begin{enumerate}
-	\item \textbf{The definition of all the variables and parameters of the ODE model in the supplement (eqs S1) should be added in order to be more self-contained.}
+We think the small change between the old and new values demonstrates that 100 runs is sufficient, but we can perform more samples if the reviewer deems it necessary.
+\end{response}
 
-The central differential equation was moved to the beginning of the model section of the main paper and a new table was added with descriptions of every population and parameter. \\
+\item \textbf{How important is the assumption of diffusion rate of chemonkines being slower than rate of viral spread? What if chemokine spreads faster? Will control be better?}
 
-	\item \textbf{Fig. 3: "8500 pg/mL" is indicated as a measurement threshold for sH1N1.  Why is this not also the case for aH5N1 and pH1N1?}
+\begin{response}
+This is a good question.
+It is important to note that this is a different question than that regarding the direct comparison between the viral diffusion rate and the chemokine diffusion rate as the chemokine's effect is only seen in the gradient of the concentration and the location of the local maxima, while the viral effect is seen in the spread and concentration of the virions themselves.
 
-The upper limit applied only to sH1N1 because there were insufficient sample dilutions made and the sample could not be repeated.  The sentence in the caption has been rewritten: \\
+We added an explanatory text to the Spatial Effects subsection of the Results section to better address this question:
 
-\textit{sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.} \\
+\begin{displayquote}
+Finally, the spatial nature of our model reveals that chemokines can diffuse more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.
+It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.
+Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
+\new{The differences in the infection outcomes between the three strains illustrate the effect of the infection spreading faster than the chemokine maxima can move in the case of pH1N1.
+The velocity of the chemokine maxima positions depends on a combination of many factors, including the chemokine diffusion, decay, delay, and secretion rates (Table 2).}
+\removed{This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.}
+\end{displayquote}
+\end{response}
 
-\end{enumerate}
+\item \textbf{Your references are not formatted properly. Please correct.}
 
-\pagebreak
-
-\item \textbf{T cell sensitivity, p7, top: The sentence "Because multiple T cells ?
-we hypothesize ?
-more T cells are not more efficient above a critical threshold" is unclear.
-Is it not that the probability of apoptosis induction will increase with the T cell number even if the rate of transition from secreting to apoptotic cells does not scale with T cell number?}
-
-It is {\em not} that the probability would increase with additional T cells at the infected epithelial cell in our model.
-The section has been rearranged to improve clarity.
-The detail regarding multiple T cells was moved to other sections.
-We also added two expanded model description sections to the supplement, which we hope clarifies this point.
-\\
-
-\item \textbf{Spatial effects:}
-\begin{enumerate}
-	\item \textbf{p7, bottom: refer also to Fig. S1}
-
-A reference to Figure S1 was added. \\
-
-	\item \textbf{Figure 6, legend: Can the difference in behaviour for pH1N1 and sH1N1 infections not be explained by the difference in IP-10 production rates (even if small)?}
-
-We performed a new model run combining pH1N1 virus and sH1N1 chemokine and found no difference in the infection profile.  A sentence was to the spatial effects results section: \\
-
-\textit{Furthermore, increasing chemokine production fails to control the runaway pH1N1 infection (data not shown) suggesting that the effect of high viral production rates dominate low chemokine concentrations in this scenario.} \\
-
-\end{enumerate}
-
-\item \textbf{Discussion, p10, 2d §: reference should also be given to section 2.2 of the supplement where combinations of chemokines are considered.}
-
-The Chemokine Directed T cell Search section of the discussion was significantly abbreviated for clarity and removal of information not directly relevant to the modeling exercise.
-The paragraph in question was removed entirely.
-\\
-
-\item \textbf{Equations:}
-\begin{enumerate}
-	\item \textbf{Eqs S1: add definition of variables and parameters}
-
-	We  moved the first equation from the supplement to the main paper to  help address the point that the paper is not self-contained.
-We have added a new table with descriptions and values for each population and parameter of this equation.
-\\
-
-	\item \textbf{Eqs S2: add definition of $t_{rc}$ and $v_{tcell}$; reference to Fig.
-1 of the main text should also be given in section 1.2 of the supplement as eqs.
-S2 relate to the bottom part of this figure.}
-
-	These modifications have all been made. \\
-
-	\item \textbf{Section 2.2: does the conclusion that RANTES does not play a significant role for H1N1 infections not simply follows from the fact that IP-10 has a much higher production rate (about 100 times) than RANTES in the H1N1 strains (Table 2, main text)?}
-
-	Indeed, it does follow given the assumption that T cells are equally sensitive to each chemokine.
-The text in S2.2 has been modified to make that assumption clear and to strengthen the conclusion.
-The following two sentences in particular address the point.
-\\
-
-	\textit{T cell sensitivity depends on receptor density (Desmetz 2006) and this was assumed to be constant.  Thus, the chemokines in combination work additively in our model.} \\
-
-	\textit{This suggests that RANTES does not play a significant role in infections that stimulate an IP-10 response due to the higher production rates of IP-10.} \\
-
-\end{enumerate}
-
-\pagebreak
-
-\item \textbf{Minor points:}
-\begin{enumerate}
-	\item \textbf{p7, Fig. 5, top row: T cells are barely visible. Could this be improved?}
-
-	The figure has been enhanced to improve visibility and a note regarding the change has been added to the caption: \\
-
-	\textit{Individual cell images for days 5.5 and 7 were enhanced (green dots are enlarged) manually to improve T cell visibility.
-Thus, individual groups of T cells, which cover only an area of one or two epithelial cells in the model appear larger in the image.
-Original images available upon request.} \\
-
-	\item \textbf{p7, Fig. 6, legend, 5th line:  add "more expressing cells than sH1N1".}
-
-	This has been done. \\
-
-	\item \textbf{p8, line 4: taking the viral production indicated in Table 2, I find 2632 rather than 2643}
-
-	This was calculated before we rounded to significant digits and has been fixed. (Impressive eye for detail!) \\
-
-	\item \textbf{p9, line 12 from bottom: greater ? than}
-
-	This sentence has been removed. \\
-
-	\item \textbf{p9, line 3 from bottom: replace "into" by "in"}
-
-	This has been done. \\
-
-	\item \textbf{Note that some Table and reference numbers have been replaced by question marks in the supplement}
-
-	These have been fixed.
-
-\end{enumerate}
+\begin{response}
+We use the Mendeley reference manager to manage our citations.
+We used the official JTB style formatter for Mendeley, along with the specific bibliography style file supplied by JTB, \texttt{elsarticle-num-names.bst}, to format our references with latex and bibtex.
+We are happy to continue working with the editor to make our reference style compatible with the expectations of the journal but require further guidance on this point.
+\end{response}
 
@@
 
-\begin{enumerate}
+\begin{itemize}
 
-\item \textbf{Very little detail is given about the agent-based model and how chemokines influence cell migration behavior.
-The authors cite several papers (4,17,18) but as far as I can see these papers do not provide convincing evidence that chemokines direct migration of effector T cells to the sites of infection (foci of infected cells).
-In fact, I don't think that there is any evidence of that process.
-How chemokines direct migration of lymphocytes in tissues is not well understood and the cartoon knowledge of textbooks is mainly applicable to the process of extravasations and may not be used by T cells to find infected cells.}
+%\item \textbf{(Introduction) P5 L3: delete one "the"}
+%
+%\begin{response}
+%Done
+%\end{response}
 
-We agree with the reviewer that for the influenza model the evidence for chemokine-directed migration of T cells to infected foci is incomplete.
-To provide a justification for constructing a math model based on chemokine-directed migration, we have completely rewritten the Introduction in order to thoroughly review the available literature on chemokine function with respect to RNA virus infections.
-We state that there is no evidence for T cells climbing chemokine gradients, and that this activity has been shown only for neutrophils.
-However, we feel that the pieces of evidence, taken together, support a reasonable hypothesis for chemokine-directed migration, in addition to extravasation, providing a reasonable justification for the model-building.
+\item \textbf{(Introduction) P5 L4: ``Pathogenicity is a function of the virus"....too vague. Function of what in the virus?}
+
+\begin{response}
+We added a new sentence to clarify our goals.  The question, ``a function of what in the virus?" is exactly the question this paper tries to answer.
+
+\begin{displayquote}
+Pathogenicity is a function of \new{both} the virus and its interaction with the immune response.
+\new{Our model explores how various features of the virus and the host immune system interact to produce observed differences between strains.}
+\end{displayquote}
+
+\end{response}
+
+\item \textbf{P8: ``T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour.
+T cell travel time from the lymph node to a random location on the lung's surface is six seconds.
+If chemokine is not encountered, the circulating T cell returns to a new location in the lung after another six seconds.
+If a circulating cell encounters chemokine, it changes state to chemotaxing and follows the chemotactic gradient to the FOI.
+Circulating T cells decay exponentially with an average lifespan of three days." Add some references for all these data.}
+
+\begin{response}
+This statement is related to Reviewer 1, point 2 (see above for how we addressed it).
+In summary, we added a new section describing our choice of the parameter set.
+We moved the Model Parameters subsection up to appear before the Model Implementation subsection.
+We also added references to Table 2, which contains citations and justifications for each of our parameter values.
+\end{response}
+
+\item \textbf{P8: ``IgM is modeled by...". IgM never defined.}
+
+\begin{response}
+We have added a new sentence to the Model Definition section.
+
+\begin{displayquote}
+\new{After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.}
+\end{displayquote}
+\end{response}
+
+\item \textbf{P11: ``This ODE was fit to data taken from (Miao et al., 2010) using Matlab's nlinfit function in order to obtain a value for $<$sigma$>$.
+The final value was found to be 1,257 per hour." Any confidence intervals over sigma?}
+
+\begin{response}
+Although we don't report confidence intervals, we explain how we obtained the value in the section ``Models for Parameter Estimation".
+$\sigma$ is part of our sensitivity analysis (T cell secretion rate) and shows its value does not have a strong effect on the model results within a biologically plausible range.
+\end{response}
+
+\item \textbf{P12: ``Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.
+4).
+" So 50 runs with the same parameter set?
+Not clear}
+
+\begin{response}
+We have expanded our sample from 50 to 100 runs as per Reviewer 1, point 3.  We have clarified that all 100 runs use the same parameter set.
+\begin{displayquote}
+Because the CyCells model is stochastic, we conducted \new{100}\removed{50} runs of the model \new{with the same parameter set} for each of the three influenza strains, each run initialized with a unique random seed (Fig.
+4).
+\end{displayquote}
+\end{response}
+
+\item \textbf{P12: ``We estimated variance across runs by calculating the standard deviation at each time point." Standard deviation of what? Viral load? Please specify.}
+
+\begin{response}
+We added a sentence to the beginning paragraph of that (Model Results) section:
+\begin{displayquote}
+Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig.
+4).
+\new{We count the number of infected cells in the simulated plaque at every time point.}  For the avian and seasonal strains, all model runs cleared the infection by day 10.
+Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.
 \\
 
-\item \textbf{How you analyze experimental data is unclear. Which equations did you use?}
+We estimated variance across runs by calculating the standard deviation \new{of the number of infected cells} at each time point. ...
+\end{displayquote}
+\end{response}
 
-This was a common theme in our reviews.
-We have made the paper more self-contained by moving the central differential equation to the main paper.
-We added details of the inner workings of the model to the supplement in two new sections (S1.2 and S1.3).
-This allows the paper to stand alone from our previous work in Mitchell 2011.
-\\
+\item \textbf{P13: Not clear how the Sensitivity Analysis has been performed.
+Were the parameters varied simultaneously?
+It appears that they have been varied independently (``Each of the remaining sixteen parameters was varied independently").
+See PMCID: PMC2570191 for a comprehensive review of global sensitivity analysis methods, especially in the context of an ABM/stochastic model.}
 
-\pagebreak
+\begin{response}
 
-\item \textbf{Please check parameter values that you provided in Table 1. It appears to several of those are made up and are not really measured in the referenced work.}
+We thank the reviewer for this suggestion and reference.
+We performed a new sensitivity analysis using Latin hypercube sampling (LHS) and partial rank correlation coefficient (PRCC) analysis following the suggested citation to supplement our original analysis.
+The new sensitivity analysis was performed only for the sH1N1 data due to computational constraints.
+333 sample points involving 17 parameters (those of Table 3 plus the viral secretion rate) were generated using LHS and each point was evaluated three times, for a total of 999 model runs.
+One sample point caused undefined model behavior and was removed, leaving 996 samples.
 
-Table 2 was split into a referenced parameter section and an estimated parameter section to make the distinction clear.
-We reviewed all of our references and annotated two parameters in Table 2: Apoptosis Time and T Cell Speed as follows:
+The results of the PRCC analysis confirm and quantitate the original sensitivity analysis.
+As we found previously, all sensitive parameters remain sensitive under PRCC ($P < 0.01$ with a high absolute Spearman's $\rho$).
+Infected Cell expression time and T Cell Expected Lifespan at the FOI were confirmed to be significant with lower values of $\rho$.
+These results have been added as a new paragraph to the Sensitivity Analysis subsection of the Results section in the main paper (text), Table 3 in the main paper (below), and expanded upon in the Supplement with text (S2.4) and three new figures (Figs.
+S6-S8).
 
-\begin{itemize}
+\begin{displayquote}
+\new{Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).
+The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).
+This indicates our less computationally intensive OFAT analysis is accurate.}
+\end{displayquote}
 
-\item Apoptosis time has been rewritten as an estimated parameter and a note has been made that this value is specific to low T cell densities.
+%\newpage
 
-\item T cell speed has been rewritten as an estimated parameter: \textit{The values were selected to be an order of magnitude faster than those calculated for inside the lymph node (Miller 2003).} \\
+\begin{displayquote}
+
+\setcounter{table}{2}
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{| c | l | c c c |}
+  \hline                        
+  Category & Parameter & Avian H5N1 & Seasonal H1N1 & Pandemic H1N1 \\
+  \hline
+  \multirow{3}{*}{Chemokine} & Chemokine Decay Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable \\
+  & Chemokine Diffusion Rate & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable& \cellcolor{green!50}stable \\
+  & Chemokine Secretion Rate & \cellcolor{blue!30}bounded stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
+  \hline
+  \multirow{6}{*}{T Cell} & Circulation Time & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Kill Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Speed & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Age in Blood& \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable& \cellcolor{blue!30}bounded stable \\
+  & T Cell Age at FOI& \cellcolor{green!50}stable & \cellcolor{blue!30}\textbf{bounded stable$^\dagger$} & \cellcolor{blue!30}bounded stable \\
+  & T Cell Production Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  \hline
+  \multirow{3}{*}{Delay} & Apoptosis Time & \cellcolor{green!50}stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
+  & Expression Time & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}\textbf{peak change$^\dagger$} & \cellcolor{yellow!50}peak change \\
+  & Incubation Time &  \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change \\
+  \hline 
+  \multirow{4}{*}{Virus} & Viral Response to IgM & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^\dagger$} & \cellcolor{red!40}sensitive \\
+  & Infectivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
+  & Viral Decay Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
+  & Viral Diffusion Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
+  \hline  
+\end{tabular}
+\caption{\textbf{Sensitivity Results:} The above parameters were varied over predetermined ranges in isolation, resulting in new model runs for every new value tested (Figs.
+S3-S5).
+The results of the sensitivity analysis were then qualitatively evaluated for each individual parameter.
+A model run's behavior was determined by examining the height of the peak of the infection at day 5 post-infection and the number of infected cells at day 10 post-infection.
+Each combination of influenza strain and free parameter was classified as belonging to one of four categories.
+Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{bounded stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.
+\new{PRCC analysis was also performed for the sH1N1 strain over these parameters (Figs.
+S6-S8).
+Bold text in the seasonal column denotes significant Spearman rank correlation ($P < 0.01$) over the time period where the parameter was active.
+$\dagger$ indicates a maximum \textit{absolute} Spearman's $\rho$ of less than 0.5.
+$*$ indicates a maximum \textit{absolute} Spearman's $\rho$ of over 0.5.}}
+\label{tab:sensitivity}
+\end{center}
+\end{table}
+
+\end{displayquote}
+\end{response}
 
@@
 
-\end{enumerate}
-
-\section*{Reviewer 3}
-
-\textbf{A minor comment is that I think it would be interesting to comment in the discussion on the fact that most pathogenic isolates kill the host (in animal models, for sure) presumably before an large CTL infiltration could occur, but that's really not the focus of the manuscript.} \\
-
-While it does not address early mortality in animal hosts, we added a note in the discussion addressing the fact that most humans did not die from pH1N1 infection: \\
-
-in vivo \textit{pH1N1 infection did not usually exhibit a resurgence due to features of the immune response not modeled, specifically antibody and local T cell proliferation}
-
 \end{document}

### 2016-01-28 paper/Levin_JTB_Clean.tex
In process of drafting Response 3

--- 2015-09-07 paper/chemokine_main_revised.tex
@@
 % amssymb package, useful for mathematical symbols
-\usepackage{amssymb}
+%\usepackage{amssymb}
 
@@
 \usepackage[usenames,dvipsnames,table]{xcolor}
-%\usepackage{ulem}
+\usepackage{ulem}
+
+%\usepackage[normalem]{ulem}
 
@@
 \newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
+\newcommand{\new}[1]{{\color{dkgreen}#1}}
 \newcommand{\drew}[1]{{\color{dkgreen}#1}}
@@
 Here, we develop an agent based spatial model to focus on T cell migration from lymph nodes through the vascular system to sites of infection.
-We use our model to investigate how different strains of influenza modulate this process.
+We use our model to investigate whether different strains of influenza modulate this process.
 
@@
 In this model comparing three influenza viruses, plaque expansion is governed primarily by the replication rate of the virus strain, and the efficiency of the T cell search-and-kill is limited by the density of infected epithelial cells in each plaque.
-Thus for each virus there is a different threshold of T cell search time after which recruited T cells are unable to control further expansion.
+Thus for each virus there is a different threshold of T cell search time above which recruited T cells are unable to control further expansion.
 Future models could use this relationship to more accurately predict control of the infection.
@@
 Influenza has worldwide epidemic potential with 5 million infections annually and half a million deaths \citep{Who2009}.
-Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the the mechanisms that control this variability are not well understood.
-Pathogenicity is a function of the virus and its interaction with the host immune response.
+Emerging strains of influenza, such as 2003 H5N1 avian and 2009 H1N1 pandemic, are more pathogenic than seasonal strains, yet the mechanisms that control this variability are not well understood.
+Pathogenicity is a function of both the virus and its interaction with the immune response.
+Our model explores how various features of the virus and the host immune system interact to produce observed differences between strains.
 
@@
 Several earlier theoretical models have been proposed to study specific aspects of influenza infections.
-Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but ignore localized spatial effects.
+Ordinary differential equation (ODE) models that are fit to empirical data \citep{Handel2008, Lee2009, Miao2010a, Saenz2010, Murillo2013, Crauste2015, Price2015} have elucidated viral population level dynamics but are unable to perceive localized spatial effects.
 Spatial models, on the other hand, have been developed to examine interactions between dendritic cells and T cells in the lymph node \citep{Beauchemin2005, Beltman2007, Zheng2008, Mirsky2011, Celli2012, Vroomans2012, Textor2014}, but have not been extended to consider local conditions in the lung.
@@
 Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI.
+After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.
 After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.
 These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.
+
+\subsection*{Model Parameters}
+
+We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection.
+Thus, we included parameters that affected virus, T cells, and chemokines.
+With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.
+The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.
+Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.
+We added IgM because many studies have argued for the importance of antibody mediated virus clearance.
+However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.
+
+Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}.
+These values were taken from the literature when available.
+Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.
+All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).
+
+Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
+We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
+T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
+T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.
 
@@
 \texttt{Healthy} cells remain unchanged unless infected by virus.
-Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay.
-\texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die.
+Once infected, the cell transitions from \texttt{incubating} to \texttt{expressing} after a 10 hour incubation delay (Table \ref{tab:parameters}).
+\texttt{Expressing} cells secrete virus and chemokine for a fixed 16.7 hours and then die (Table \ref{tab:parameters}).
 \texttt{Expressing} cells initiate apoptosis sooner if they are contacted by activated T cells.
-\texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour.
+\texttt{Apoptoic} cells continue to secrete virus and chemokine and then die after one hour (Table \ref{tab:parameters}).
 \texttt{Dead} cells take up space and do not regenerate over the course of an infection.
@@
 T cells are described by two states: \texttt{circulating} and \texttt{chemotaxing}.
-T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour.
+T cells emerge from the lymph node at five days post infection (p.i.) at a rate of 1,257 cells per hour (Models for Parameter Estimation, Table \ref{tab:parameters}).
 T cell travel time from the lymph node to a random location on the lung's surface is six seconds.
@@
 If a circulating cell encounters chemokine, it changes state to \texttt{chemotaxing} and follows the chemotactic gradient to the FOI.
-Circulating T cells decay exponentially with an average lifespan of three days.
+Circulating T cells decay exponentially with an average lifespan of four days.
 Chemotaxing T cells follow the gradient through the two-dimensional lung endothelium, inducing apoptosis when they encounter expressing epithelial cells.
 Chemotaxing T cells decay exponentially with an average lifespan of two hours.
+Justifications for these parameter values are listed in Table \ref{tab:parameters}.
 
@@
 Both particle types decay exponentially.
-IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.
-
-\subsection*{Model Parameters}
-
-Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}.
-These values were taken from the literature when available.
-Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.
-All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).
-
-Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
-We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
-T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
-T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.
+IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day (Table \ref{tab:parameters}).
 
 \subsection*{Models for Parameter Estimation}
+
+To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.
+\ref{fig:data}, Table S1).
+The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus \citep{Mitchell2011}.
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
 
@@
 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
-Initial population sizes and parameter values are taken from the previous study.
+Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.
+Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.
+The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).
+IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
+
 {\footnotesize
@@
 }
-
-$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
-Table S2 summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
-The results of these fits are discussed in the main paper and shown in Table 1.
-IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
+$\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.  Table S2 summarizes population and parameter values and descriptions.
 
@@
 
-%To provide estimates of chemokine concentrations and secretion rates present in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.~\ref{fig:data}).
-The dynamic viral loads at these intervals in these cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus have been reported previously \citep{Mitchell2011}.
-IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
-
-%To estimate per-cell chemokine production rates, we extended the delay differential equation (DDE) model of Ref.
-\citep{Mitchell2011} to include chemokine production from infected cells (Eq.~S1).
-Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
+% Model fits (Table \ref{tab:strains}) were computed for three strains using Matlab's \texttt{nlinfit} function (Levenberg-Marquardt algorithm).
 The resulting chemokine production values were used in the CyCells ABM by combining IP-10 and RANTES into one aggregate rate but also reported separately (Fig.~S2).
@@
 
-Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
-For the avian and seasonal strains, all model runs cleared the virus by day 10.
+Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig.
+4).
+We count the number of infected cells in the simulated plaque at every time point.
+For the avian and seasonal strains, all model runs cleared the infection by day 10.
 Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.
@@
 We estimated variance across runs by calculating the standard deviation at each time point.
-For aH5N1, the maximum standard deviation was 37, which occurred at day 5.05 where the mean value was 259.
-For sH1N1, the maximum standard deviation was 242, which occurred at day 4.93 where the mean value was 2,607.
-For pH1N1, the maximum standard deviation was 230, which occurred at day 5.07 where the mean value was 7,018 (Fig.~\ref{fig:variance}).
+For aH5N1, the maximum standard deviation was 35, which occurred at day 4.8 where the mean value was 257.
+For sH1N1, the maximum standard deviation was 212, which occurred at day 5.1 where the mean value was 2,528.
+For pH1N1, the maximum standard deviation was 231, which occurred at day 5.1 where the mean value was 6,988 (Fig.~\ref{fig:variance}).
 
@@
 Plaque growth in the model is illustrated in Figure \ref{fig:cycells} and Videos S1-S3.
-Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly, according to the Stokes-Einstein equation.
+Because the virus particles are an order of magnitude larger than the chemokine molecules, they diffuse more slowly according to the Stokes-Einstein equation.
 However, chemokine molecules decay more quickly than the virus (Table~\ref{tab:parameters}).
@@
 Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
-This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.
+The differences in the infection outcomes between the three strains illustrate the effect of the infection spreading faster than the chemokine maxima can move in the case of pH1N1.
+The velocity of the chemokine maxima positions depends on a combination of many factors, including the chemokine diffusion, decay, delay, and secretion rates (Table \ref{tab:parameters}).
 
@@
 
-We conducted a sensitivity analysis of sixteen model parameters to identify those that have an effect on determining model outcomes (S2.3).
+We performed two sensitivity analyses of sixteen model parameters to identify those that have an effect on determining model outcomes (S2.3 - S2.4).
 We studied all but five of the parameters listed in Tables~\ref{tab:strains} and \ref{tab:parameters}, excluding T cell radius, T cell sensitivity to chemokine, T cell onset time, IgM onset time, and the viral production rate.
@@
 
-Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.
+Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain using a one-factor-at-a-time (OFAT) approach, and the model was rerun once for each parameter setting (S2.3).
 Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.
-Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters alter the outcome of the simulation for any variation within a biologically plausible range.
+Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable across a biologically plausible range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters substantively change the outcome of the simulation within biologically plausible ranges.
 
 Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.
-In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.
+In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter, varying among the three panels.
 The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.
-Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.
-None of the T cell parameters nor the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
+Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be considered a viral parameter.
+None of the T cell parameters nor the chemokine parameters has a strong effect on the model's prediction of control or lack of control of viral replication.
 This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.
 
+Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).
+The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).
+Consistent results between the two approaches suggests the less computationally intensive OFAT analysis is also accurate for aH5N1 and pH1N1.
+
 \subsection*{Window of Control}
 
-Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).
+Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the course of the infection (Table~\ref{tab:sensitivity}).
 Here we examine the role of T cells by assessing multiple factors that control the overall T cell response.
@@
 We name this threshold the `window of control'.
-This window can be defined as the combination of the the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).
+This window can be defined as the combination of the time of T cell arrival at a virus-secreting cell ($T_{arr}$), the time it takes for the T cell to induce apoptosis ($T_{kill}$) and the time for the infected cell to apoptose ($T_{apop}$).
 We examine how much virus a virus-secreting cell can produce inside this window.
@@
 (Table \ref{tab:parameters}).
-This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (see S2.4 for a more complete explanation of how $E$ was determined).
+This suggests that 50\% of free virus contacts healthy cells and 54\% of that virus succeeds in infecting a healthy cell, implying that $E \approx 0.27$ (see S2.5 for a more complete explanation of how $E$ was determined).
 Solving for $T_{arr}$ where $R_0 < 1$ gives:
@@
 Our model did not incorporate  the potential contributions from other cytokines that did not show different levels in the physical experiment.
-This includes chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \citep{Matsukura1996, Arndt2002} as well as chemokines secreted by immigrant macrophages \citep{Julkunen2000} and amplification of epithelial cell secretion by CD8 T cells \citep{Zhao2000}.
+This includes chemokines such as CXCL8/IL-8 detected in bronchial cell cultures \citep{Matsukura1996, Arndt2002} as well as chemokines secreted by immigrant macrophages \citep{Julkunen2000}, neutrophils \citep{lim2015neutrophil}, and amplification of epithelial cell secretion by CD8 T cells \citep{Zhao2000}.
 The sensitivity analysis classifies the chemokine secretion rate as stable, suggesting the total amount of chemokine is not a significant factor in determining viral clearance in our model.
@@
 While T cell chemotaxis may not be proven, our previous work has shown severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
-Our sensitivity analysis shows that slowing T cell velocity (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
+Our sensitivity analysis shows that slowing T cell speed (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
 These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.
@@
 $^*$ All estimated parameter values are examined in the sensitivity analysis (Table~\ref{tab:sensitivity}, S2.3).
-$^{\dagger}$ denotes parameters determined to be sensitive by the sensitivity analysis.
+$^{\dagger}$ denotes parameters determined to be sensitive by the one-factor-at-a-time sensitivity analysis.
 Values were taken from experimental literature if possible and from earlier modeling papers if not.
@@
   & T Cell Kill Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
-  & T Cell Velocity & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
-  & Circulating Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable& \cellcolor{blue!30}bounded stable \\
-  & Found Decay Rate & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Speed & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
+  & T Cell Age in Blood & \cellcolor{green!50}stable & \cellcolor{blue!30}bounded stable& \cellcolor{blue!30}bounded stable \\
+  & T Cell Age at FOI & \cellcolor{green!50}stable & \cellcolor{blue!30}\textbf{bounded stable$^\dagger$} & \cellcolor{blue!30}bounded stable \\
   & T Cell Production Rate & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable & \cellcolor{blue!30}bounded stable \\
@@
   \multirow{3}{*}{Delay} & Apoptosis Time & \cellcolor{green!50}stable & \cellcolor{green!50}stable & \cellcolor{green!50}stable \\
-  & Expression Time & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change \\
+  & Expression Time & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}\textbf{peak change$^\dagger$} & \cellcolor{yellow!50}peak change \\
   & Incubation Time &  \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change & \cellcolor{yellow!50}peak change \\
   \hline 
-  \multirow{4}{*}{Virus} & Viral Response to IgM & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
-  & Infectivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
-  & Viral Decay Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
-  & Viral Diffusion Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive & \cellcolor{red!40}sensitive \\
+  \multirow{4}{*}{Virus} & Viral Response to IgM & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^\dagger$} & \cellcolor{red!40}sensitive \\
+  & Infectivity & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
+  & Viral Decay Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
+  & Viral Diffusion Rate & \cellcolor{red!40}sensitive & \cellcolor{red!40}\textbf{sensitive$^*$} & \cellcolor{red!40}sensitive \\
   \hline  
@@
 Each combination of influenza strain and free parameter was classified as belonging to one of four categories.
-Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{bounded stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.}
+Parameters were classified as \textbf{stable} if all runs follow the same behavior, \textbf{bounded stable} if intermediate parameter adjustments did not affect the model's behavior, even if the more extreme adjustments did, \textbf{peak change} if the peak of the infection differs but the result at day 10 is the same, and \textbf{sensitive} if any level of change in the parameter affects the resulting model behavior.
+PRCC analysis was also performed for the sH1N1 strain over these parameters (Figs.
+S6-S8).
+Bold text in the seasonal column denotes significant Spearman rank correlation ($p < 0.01$) over the time period where the parameter was active.
+$\dagger$ indicates a maximum \textit{absolute} Spearman's $\rho$ of less than 0.5.
+$*$ indicates a maximum \textit{absolute} Spearman's $\rho$ of over 0.5.}
 \label{tab:sensitivity}

### 2016-01-28 paper/chemokine_main_revised.tex
In process of drafting Response 3

@@
-\documentclass[preprint,10pt,authoryear]{elsarticle}
+\documentclass[preprint,10pt,numbers]{elsarticle}
 
@@
 %\usepackage{cite}
-\usepackage{natbib}
+\usepackage[]{natbib}
 
@@
 \doublespacing
+
+\usepackage{xr}
+\externaldocument{chemokine_supplement_revised}
 
@@
 
+%\bibliographystyle{elsarticle-num}
 \bibliographystyle{elsarticle-num-names}
+%\bibliographystyle{plos2009}
 
@@
 \newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
+\newcommand{\new}[1]{{\color{dkgreen}#1}}
+%\newcommand{\removed}[1]{}
+%\newcommand{\new}[1]{#1}
 \newcommand{\drew}[1]{{\color{dkgreen}#1}}
@@
 The model is parameterized with values taken from the literature when available.
-Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.
+Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and our previously published ODE model \citep{Mitchell2011}.
+
 To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.
@@
 
-The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.
+The analyses of the spatial model reveal several interesting and underappreciated phenomena.
 The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
@@
 Chemokine diffuses from secreting cells, creating a spreading region of stimulation around the FOI.
+After four simulated days, Immunoglobulin M (IgM) is introduced into the model by increasing the viral decay rate.
 After five simulated days, representing lymph node stimulation and T cell proliferation, activated T cells exit the lymph node at a constant rate and travel through the vasculature to the tissue as described above.
 These cell and molecular interactions and state transitions are depicted in Figure~\ref{fig:systemchart}.
+
+\subsection*{Model Parameters}
+
+We include only parameters that directly address the role of T cells and T cell migration because our study focuses on the role of T cells and the chemokine effects of T cells in influenza virus infection.
+Thus, we included parameters that affected virus, T cells, and chemokines.
+With regards to the influenza, we chose parameters involving viral replication, infectivity in epithelial cells, and virus decay and diffusion rates based on our earlier study.
+The chemokine decay rate, diffusion rate, and secretion rate parameters are relevant to dynamic chemokine gradients required for T cell chemotaxis.
+Because we wanted to use our model to test the role of T cells and T cell migration in influenza clearance, we included multiple T cell parameters, such as T cell production rate, T cell death rate, T cell kill time, and T cell migration rate.
+We added IgM because many studies have argued for the importance of antibody mediated virus clearance.
+However, because our study is not addressing the role of antibody in clearance of influenza, we did not test the full range of B cell responses directly by including them in our parameters.
+
+Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}.
+These values were taken from the literature when available.
+Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.
+All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).
+
+Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
+We extended our previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
+T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
+T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.
 
@@
 Both particle types decay exponentially.
-IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day.
-
-\subsection*{Model Parameters}
-
-Model parameters are listed in Tables \ref{tab:strains} and \ref{tab:parameters}.
-These values were taken from the literature when available.
-Other parameters were determined by matching empirical data to ODE models, and some parameters were estimated using biologically plausible ranges.
-All estimated parameters were studied in a sensitivity analysis to test their impact on model behavior (Table~\ref{tab:sensitivity}).
-
-Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
-We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
-T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
-T cell production is assumed to be constant after day 5 \citep{MartIn-Fontecha2003}.
+IgM is modeled by increasing the viral decay rate by a factor of ten after the fourth day (Table \ref{tab:parameters}).
 
 \subsection*{Models for Parameter Estimation}
+
+To provide estimates of chemokine concentrations and secretion rates in lung tissue, chemokine levels were measured at 4-6h intervals during the first 48 h of infection in wells containing approximately one million human bronchial epithelial cells (Fig.
+\ref{fig:data}, Table \ref{tab:data}).
+The dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus \citep{Mitchell2011}.
+IP-10 concentration increases were observed by 8h post-infection (p.i.), and RANTES by 16h p.i..
 
@@
 1 by adding one new equation ($\dot{C}=r I_{1 \tau_3}-dC$) to model chemokine production.
-Initial population sizes and parameter values are taken from the previous study.
+Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table~\ref{tab:data} using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant \new{(Fig.
+S1, Table~\ref{tab:rmse})}.
+Next, 1,000 bootstrapping runs using resampled residuals were performed on each modeled strain to generate confidence intervals over $r$.
+The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).
+IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
+
 {\footnotesize
@@
 }
-
 $\tau$ subscript variables denote delay terms, signifying the value is the population quantity in existence at time $t - \tau$.
-Table S2 summarizes population and parameter values and descriptions.
-Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1.
-The results of these fits are discussed in the main paper and shown in Table 1.
-IP-10 and RANTES secretion rates are aggregated in the spatial model (S2.2).
+Table \ref{tab:dde} summarizes population and parameter values and descriptions.
 
@@
 The area of the infected region is equal to the area of an infected cell (of radius $r_{cell}$) multiplied by the number of infected cells.
-See Fig.
-1 for more details about the model.
+See Fig.\ref{fig:systemchart} for more details about the model.
 The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
-$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
+$a$ and $b$ were fit to 5 experimental runs of a spatial model implemented in CyCells ($R^2 = 0.997$, $P < 0.01$).
 
@@
 The initial number of infected cells is assumed to be zero.
-Parameter values are listed in Table S3.
+Parameter values are listed in Table \ref{tab:supplement}.
 This ODE was fit to data taken from \citep{Miao2010a} using Matlab's \texttt{nlinfit} function in order to obtain a value for $\sigma$.
@@
 Quantitative chemokine levels were performed in 30 $\mu$L aliquots for a panel of chemokines (IL-8, MCP-1, MIP-1$\alpha$, MIP-1$\beta$, RANTES, IP-10, eotaxin) and cytokines (interferon-gamma, IL-1$\alpha$, IL-1$\beta$, IL-1RA, IL-2, IL-3, IL-4, IL-6, IL-10, IL-12p40, IL-15, IL-17, TNF$\alpha$) (Luminex Assay®, Luminex Corp.) and reported as ng/mL basal media sampled from a total volume of 4 mL.
-Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table S1).
+Only IP-10, RANTES, and TNF$\alpha$ showed increases in production (Table \ref{tab:data}).
 TNF activity is not incorporated in the model.
@@
 
-Because the CyCells model is stochastic, we conducted 50 runs of the model for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
-For the avian and seasonal strains, all model runs cleared the virus by day 10.
+Because the CyCells model is stochastic, we conducted 100 runs of the model with the same parameter set for each of the three influenza strains, each run initialized with a unique random seed (Fig.~\ref{fig:variance}).
+We count the number of infected cells in the simulated plaque at every time point.
+For the avian and seasonal strains, all model runs cleared the infection by day 10.
 Conversely, each simulation of the 2009 pandemic influenza led to uncontrolled infection.
@@
 
-Finally, the spatial nature of our model reveals that chemokines can diffuse more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.
+Finally, the spatial nature of our model reveals that \new{the locations of peak chemokine concentration}\removed{chemokines} can \new{move}\removed{diffuse} more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.
 It takes time for infected cells to begin producing chemokine, while the preexisting areas of high chemokine density are slow to decay.
 Thus, T cells whose movements respond to the spatial layout of the chemokine gradient can become trapped, failing to locate new regions of infected cells in the growing plaque.
-This effect is most pronounced in the pH1N1 strain, where the plaque expands more quickly than the chemokine can diffuse.
+The differences in the infection outcomes between the three strains illustrate the effect of the infection spreading faster than the chemokine maxima can move in the case of pH1N1.
+The velocity of the chemokine maxima positions depends on a combination of many factors, including the chemokine diffusion, decay, delay, and secretion rates (Table \ref{tab:parameters}).
 
 % Taken together, the delayed response and low proportion of virus secreting cells contribute strongly to the failure of T cells clearing the infection.
-Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (Fig.
-S5), suggesting that the effect of high viral production rates dominate increased chemokine production rates.
+Furthermore, increasing chemokine production fails to guide T cells to the actively secreting epithelial cells in the runaway pH1N1 infection (Fig.~\ref{fig:psensitivity}), suggesting that the effect of high viral production rates dominate increased chemokine production rates.
 
@@
 
-Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain, and the model was rerun once for each parameter setting.
+Each of the remaining sixteen parameters was varied independently (over multiple orders of magnitude when appropriate) for each viral strain using a one-factor-at-a-time (OFAT) approach, and the model was rerun once for each parameter setting (S2.3).
 Results were compared for the quantitative size of the peak infection and the number of uncleared infected cells at day 10 p.i.
-Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable throughout a reasonable range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters alter the outcome of the simulation for any variation within a biologically plausible range.
+Based on these runs, we classified each parameter for each strain as follows (Table~\ref{tab:sensitivity}): \textit{stable} parameters are those that do not affect the model behavior over any of the different values tested; \textit{bounded stable} parameters are stable across a biologically plausible range of values but model behavior diverges at some threshold; \textit{peak change} parameters affect the peak size of the infection, but do not change the ultimate number of infected cells at the end of the run; and \textit{sensitive} parameters substantively change the outcome of the simulation within biologically plausible ranges.
 
 Table \ref{tab:sensitivity} shows that four parameters are sensitive: viral response to IgM, infectivity, viral decay rate, and viral diffusion.
-In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter due to the difference between the three panels.
+In addition, Figure \ref{fig:variance} shows that viral production rate is a sensitive parameter, varying among the three panels.
 The four parameters (infectivity, viral decay rate, viral diffusion rate, and viral production rate) are specific viral characteristics.
-Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be included in this group.
-None of the T cell parameters nor the chemokine parameters have a strong effect on the model's prediction of control or lack of control of viral replication.
+Further, viral response to IgM, as coded in the model, directly affects viral decay rate and can be considered a viral parameter.
+None of the T cell parameters nor the chemokine parameters has a strong effect on the model's prediction of control or lack of control of viral replication.
 This result supports the hypothesis that virus dynamics dominate the course of an influenza infection.
 
+Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain as described in \cite{Marino2008} using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).
+The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).
+Consistent results between the two approaches suggests the less computationally intensive OFAT analysis is also accurate for aH5N1 and pH1N1.
+
 \subsection*{Window of Control}
 
-Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the the course of the infection (Table~\ref{tab:sensitivity}).
+Our sensitivity analysis suggests that viral parameters are the most sensitive and have the largest effect on the course of the infection (Table~\ref{tab:sensitivity}).
 Here we examine the role of T cells by assessing multiple factors that control the overall T cell response.
@@
 
-The above calculation suggests a combined transition point for sH1N1 at 163 minutes ($R_0 = 1$ for $T_{arr}+T_{kill}+T_{apop}=163m$), which is consistent with the sensitivity analysis (Fig.~S4: Apoptosis Time: 90m - 120m) and suggests that the modeled $T_{arr}$ is between 33 and 63 minutes for sH1N1.
+The above calculation suggests a combined transition point for sH1N1 at 163 minutes ($R_0 = 1$ for $T_{arr}+T_{kill}+T_{apop}=163m$), which is consistent with the sensitivity analysis (Fig.~\ref{fig:ssensitivity}: Apoptosis Time: 90m - 120m) and suggests that the modeled $T_{arr}$ is between 33 and 63 minutes for sH1N1.
 These data show that the balance between viral production and T cell response is a key factor in clearance of infection: pH1N1 can theoretically be cleared with instantaneous T cell effectiveness (Fig.~\ref{fig:instantkill}), but in actuality viral production exceeds T cell response time and this leads to uncontrolled infection.
@@
 Although leukocytes exhibit directional behavior to chemokines \citep{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
-While T cell chemotaxis may not be proven, our previous work has shown severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
-Our sensitivity analysis shows that slowing T cell velocity (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
+While T cell chemotaxis may not be proven, our previous work suggests severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
+Our sensitivity analysis shows that slowing T cell speed (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
 These results are consistent with the hypothesis that T cells are guided by chemokine in the lung epithelium.
@@
 
-Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.
+Previous ODE models have studied how viral dynamics affect the course of infection \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.
 The use of an ABM can complement spatially homogeneous differential equation models \citep{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.
@@
  \end{center}
-\caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (green square) are shown in pg/mL.
-sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL but these three values (empty red triangles) were not included in the model fitting.
+\caption{{\bf Empirical viral and cytokine titers for three strains of influenza: Avian H5N1, Seasonal sH1N1, and Pandemic pH1N1.}  Viral titer (blue circles) is in PFU/mL, and IP-10 (red triangle) and RANTES (yellow square) are shown in pg/mL.
+sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL and these three values (empty red triangles) were not included in the model fitting.
 An extended differential equation model from \citep{Mitchell2011} was fit to IP-10 and RANTES data (Equation 1) .

### 2016-01-28 paper/chemokine_supplement_revised.tex
In process of drafting Response 3

@@
-\documentclass[preprint,10pt,authoryear]{elsarticle}
+%\documentclass[preprint,10pt,authoryear]{elsarticle}
+\documentclass[preprint,10pt,authoryear]{article}
 
@@
 \usepackage{pdflscape}
+
+\usepackage{xr}
+\externaldocument{chemokine_main_revised}
 
@@
 
-An overview of the model implementation is shown in Fig.
-2.
+An overview of the model implementation is shown in Fig.~\ref{fig:modelchart}.
 The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (288,212 total cells).
-The sheet measures 5$mm$ per side, which is large enough to contain the infection (Fig.
-5).
+The sheet measures 5$mm$ per side, which is large enough to contain the infection (Fig.~\ref{fig:cycells}).
 The molecular grid (Text S1 1.3) is initialized with a resolution of 20$\mu m$.
@@
 
-\subsection{Modeling Decisions}
+\subsection{Modeling decisions}
 
@@
 
-The model simulates a chemokine gradient surrounding the infected focus (Fig.
-5), based on the calculated per-infected cell secretion rate (Table 3) and known chemical parameters for a 10 kDa protein (Table 2).
+The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table \ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table \ref{tab:parameters}).
 T cell sensitivity depends on receptor density \citep{Desmetz2006} and this was assumed to be constant.
@@
 % we hypothesize that it renders IP-10 ineffectual.
-We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).
+We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table \ref{tab:strains}).
 Because of this behavior IP-10 was not included in the aH5N1 model runs.
@@
 
-\subsection{General sensitivity analysis}
+\subsection{One-factor-at-a-time sensitivity analysis}
 
@@
 
-The model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Tables 1 and 2).
-Individual parameters were varied over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
+The model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Tables \ref{tab:strains} and \ref{tab:parameters}).
+One-factor-at-a-time (OFAT) analysis was performed by varying Individual parameters over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
 Each parameter was varied over all three influenza strains, creating three sets of sensitivity plots (Figures~\ref{fig:asensitivity}-\ref{fig:psensitivity}).
 
-We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table 3).
+We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table \ref{tab:sensitivity}).
 Parameters are presented along with their range using the following format:  \\
@@
 Specifically, the plausible parameter values in this group are within a range that does not significantly affect the asymptotic behavior of the model.
-We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{bounded stable} parameters.
+We further split this group into two categories in Table \ref{tab:sensitivity}, \textit{stable} parameters and \textit{bounded stable} parameters.
 \textit{Bounded stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.
@@
 Due to the nature of the ABM, inclusion of these parameters is mechanistically necessary.
-Thus, we include and parameterize them using the methods described in the main paper and Table 2.
+Thus, we include and parameterize them using the methods described in the main paper and Table \ref{tab:parameters}.
 
@@
 
-\textbf{Circulating T Cell Expected Age:} [ \textit{1h |  4d | 400d} ] - The expected age determines the decay rate of T cells in blood.
+\textbf{T Cell Age in Blood:} [ \textit{1h |  4d | 400d} ] - The expected age determines the decay rate of T cells in blood.
 
@@
 
-\textbf{Chemotaxing T Cell Expected Age:} [ \textit{12m | 2h | 20h} ] - The expected age determines the decay rate of T cells in the epithelium.
+\textbf{T Cell Age at FOI:} [ \textit{12m | 2h | 20h} ] - The expected age determines the decay rate of T cells at the focus of infection (FOI).
 
@@
 
-An explicit sensitivity analysis was not directly performed.  The range of the parameters are taken from the respective secretion rates of the aH5N1, sH1N1, and pH1N1 influenzas (Table 1).
+An explicit sensitivity analysis was not directly performed.
+The range of the parameters are taken from the respective secretion rates of the aH5N1, sH1N1, and pH1N1 influenzas (Table \ref{tab:strains}).
 
@@
 \\
+
+\subsection{Partial rank correlation coefficient sensitivity analysis}
+
+We performed a partial rank correlation coefficient (PRCC) sensitivity analysis to complement our one factor at a time (OFAT) parameter analysis.
+Due to computational limitations, we computed PRCC values only for the sH1N1 strain, confirming that the results of the PRCC are consistent with the conclusions of the original sensitivity analysis and that viral parameters dominate the model output.
+
+\subsubsection{PRCC Design and Implementation}
+
+PRCC analysis was performed as described in \cite{Marino2008}.
+Samples were generated using Latin hypercube sampling (LHS) to generate 333 sample points for 17 parameters (those from Table \ref{tab:sensitivity}, plus the sH1N1 viral secretion rate).
+Each parameter was assigned a confidence distribution (Table \ref{tab:lhs}) and sampled using its inverse cumulative distribution function to generate points with density proportional to the underlying probability distribution function.
+Each sample point was evaluated three times using unique random seeds for a total of 999 samples.
+One sample point resulted in undefined model behavior and was removed, leaving a total of 996 samples.
+
+Output is defined as the number of infected cells remaining for each time point.
+Each model run generates 144,001 time points (10 day model with 6 second time steps).
+PRCC analysis was performed over every 100th time point, resulting in a time series of 1441 Spearman rank correlation coefficients ($\rho$) and significance values ($p$) for each parameter (Figs.
+\ref{fig:prcc} - \ref{fig:vsec}).
+
+\subsubsection{PRCC Results}
+
+The  analysis confirmed the conclusion of original sensitivity analysis that viral parameters dominate model output.
+Of the 17 parameters tested with PRCC analysis, 7 show significance of $p < 0.01$ over the time period where the parameter was relevant to model dynamics (Figs.
+S6 - S7).
+All five of the viral parameters (viral response to IgM, viral infectivity, viral decay rate, viral diffusion rate, and viral secretion rate) are significant.
+Infected cell expression time and the T cell age at the focus of infection are also significant at $p < 0.01$ (but have $\rho$ values near 0).
+Viral infectivity, viral decay rate, viral diffusion rate, and viral secretion rate are the only parameters that have maximum absolute vales of $\rho$ above 0.5.
+Other parameters showed significant correlations (with low values of $\rho$), but only over time periods where they were not active.
+
+Viral infectivity, viral diffusion rate, and viral secretion rate, values of $\rho$ cross or approach zero over time.
+This merely reflects that the modeled target cells have died (Fig.~\ref{fig:exhaustive}) and so the decline in $\rho$ is irrelevant.
+
+%This effect is due to the spatial constraints built into our agent based model.
+The  model environment was designed to be large enough to contain an infection defined by the default model parameter values.
+Extreme values of the viral parameters cause rapid infection growth to the point that the modeled system is exhausted (Fig.
+\ref{fig:exhaustive}).
+In these cases the number of infected cells tends towards zero at later time points as every modeled target cell has died before the model is complete.
+Thus, the calculated values of $\rho$ for these given parameters should be trusted only at earlier time points.
 
@@
 \begin{center}
- \includegraphics[width=\textwidth]{Figure_S-1}
+ \includegraphics[width=\textwidth]{FigureS1}
+ \end{center}
+\caption{{\bf Preliminary Model 1 fits to data.}  Preliminary Model 1 (Eq.
+1) was fit to experimental data (Table~\ref{tab:data}) using a genetic algorithm as described in the main paper.
+sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL and these three values (empty red triangles) were not included in the model fitting.
+RMSE values of the fits can be seen in Table~\ref{tab:rmse}.
+IP-10 data and model fits shown in red (triangles), RANTES data and model fits shown in yellow (squares).}
+ \label{fig:fits}
+\end{figure}
+
+\begin{figure}[!ht]
+\begin{center}
+ \includegraphics[width=\textwidth]{Figure_S2}
  \end{center}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S2}
+	\includegraphics[width=\textwidth]{Figure_S3}
 	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S3}
+	\includegraphics[width=9in]{Figure_S4}
 	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S4}
+	\includegraphics[width=9in]{Figure_S5}
 	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S5}
+	\includegraphics[width=9in]{Figure_S6}
 	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.
@@
 
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=9in]{Figure_S7}
+	\caption{\textbf{PRCC sensitivity analysis.} PRCC analysis generated Spearman correlation coefficients ($\rho$) and significance values ($p$) over 1441 time points.
+Plots show values of $\rho$ vs.
+time for each parameter tested.
+Dark blue indicates significance levels at $p < 0.01$.
+Gold indicates significance of $p < 0.05$.
+Light gray indicates $p \geq 0.05$.
+Six parameters show significant effect over the time period where they were relevant to model dynamics: infected cell expression time, viral response to IgM, viral infectivity, T cell expected lifetime at the FOI, viral decay rate, and the viral diffusion rate.}
+	\label{fig:prcc}
+\end{center}
+\end{figure}
+
 \end{landscape}
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{Figure_S8}
+	\caption{\textbf{PRCC sensitivity analysis for the viral secretion rate.} Similar to Figure S6, PRCC analysis generated Spearman correlation coefficients ($\rho$) and significance values ($p$) over 1441 time points.
+Plot shows the values of $\rho$ vs.
+time for the viral secretion rate.
+Dark blue indicates significance levels at $p < 0.01$.
+Gold indicates significance of $p < 0.05$.
+Light gray indicates $p \geq 0.05$.
+Viral secretion shows significant effect on model output.
+The decline of $\rho$ over time reflects the early death of every modeled target cell (Figure~\ref{fig:exhaustive})).}
+	\label{fig:vsec}
+\end{center}
+\end{figure}
+
+\begin{figure}[ht!]
+\begin{center}
+	\includegraphics[width=\textwidth]{Figure_S9}
+	\caption{\textbf{Exhaustive Growth:} Certain parameter sets generated by Latin hypercube sampling caused rapid uncontrolled infection growth (Samples 1-3).
+This resulted in the early death of every modeled target cell (saturation at day 1) and the inevitable decline of the infection size to zero.
+A larger, unbounded model environment would show unconstrained growth.
+This effect explains the decline of the $\rho$ value of several viral parameters to or past zero over time in Figures \ref{fig:prcc} and \ref{fig:vsec}.}
+	\label{fig:exhaustive}
+\end{center}
+\end{figure}
 
@@
 %\begin{center}
-%	\includegraphics[width=\textwidth]{Figure_S6}
+%	\includegraphics[width=\textwidth]{Figure_SX}
 %	\caption{\textbf{pH1N1 sensitivity analysis.} Model results and fits of the viral effectiveness rate, $E$.}
@@
 All parameters and populations are taken from \citep{Mitchell2011} except for $C$, $d$, and $\tau_3$.
-The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.
-3).
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
 Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
@@
 \begin{center}
-\begin{tabular}{ | c | c | c | }
+\begin{tabular}{| c | c | c |}
+  \hline
+  Strain & Chemokine & RMSE \\
+  \hline
+  \multirow{2}{*}{aH5N1} & RANTES & 166 \\
+  & IP-10 & 7,950 \\
+  \hline
+  \multirow{2}{*}{sH1N1} & RANTES & 90 \\
+  & IP-10 & 2,732 \\
+  \hline
+  \multirow{2}{*}{pH1N1} & RANTES & 2,671  \\
+  & IP-10 & 64,342 \\
+  \hline
+\end{tabular}
+\caption{\textbf{Root mean squared error of Preliminary Model 1 fits.}  Preliminary Model 1 (Eq.
+1) was fit to experimental data (Table~\ref{tab:data}) using a genetic algorithm as described in the main paper.
+The resulting model fits can be seen in Figure~\ref{fig:fits}.}
+\label{tab:rmse}
+\end{center}
+\end{table}
+
+\begin{table}[!ht]
+\begin{center}
+\begin{tabular}{ | c | c | c |}
   \hline                        
@@
 \end{tabular}
+\begin{tabular}{| c | c |}
+  \hline
+  I & C \\
+  \hline
+  4 & 329 \\
+  58 & 997 \\
+  816 & 2550 \\
+  3,264 & 4,410 \\
+  10,050 & 8,069 \\
+  \hline
+\end{tabular}
 \caption{\textbf{Parameters for Preliminary Model 2 (Equation 2).}  Values are chosen to match the equations borrowed from both \citep{Mitchell2011} and \citep{Miao2010a}.
-Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM.}
+Values \textit{a} and \textit{b} were fit to multiple runs of a simplified CyCells ABM using the linear model $C = a\sqrt{I}+b$ with a resulting $R^2 = 0.997$, $P < 0.01$.}
 \label{tab:supplement}
 \end{center}
+\end{table}
+
+\begin{table}[!ht]
+\centering
+\begin{tabular}{| c | l | c | l |}
+  \hline                        
+  Category & Parameter & Units & Distribution \\
+  \hline
+  \multirow{3}{*}{Chemokine} & Chemokine Decay Rate & \textit{Hz} &  LN(-3.4, 1.0) \\
+  & Chemokine Diffusion Rate & \textit{$\mu$m$^2$/s} & LN(-0.5, 1.0) \\
+  & Chemokine Secretion Rate & $(pg/s\cdot cell)$ & LN(-3.7, 0.1) \\
+  \hline
+  \multirow{6}{*}{T Cell} & Circulation Time & \textit{seconds} & LN(-.75, 0.5)\\
+  & T Cell Kill Rate & \textit{min} & LN(1.0, 0.5) \\
+  & T Cell Speed & \textit{$\mu$m/min} & LN(-.75, 1), \\
+  & T Cell Age in Blood & \textit{days} & LN(0.6, 1.0) \\
+  & T Cell Age at FOI& \textit{min} & LN(2.1, 0.5) \\
+  & T Cell Production Rate & \textit{cells/h} & LN(3.1, 0.15) \\
+  \hline
+  \multirow{3}{*}{Delay} & Apoptosis Time & \textit{hours} & U(0.0, 2.0) \\
+  & Expression Time & \textit{min} & LN(3.0, 0.15) \\
+  & Incubation Time & \textit{hours} & LN(-1.0, 0.15) \\
+  \hline 
+  \multirow{4}{*}{Virus} & Viral Response to IgM & | & LN(1.0, 0.5) \\
+  & Infectivity & \textit{min/virion} & LN(2.1, 0.5) \\
+  & Viral Decay Rate &  \textit{day$^{-1}$} & LN(0, 1.0) \\
+  & Viral Diffusion Rate & \textit{$\mu$m$^2$/s} & LN(-1.5, 1.0) \\
+  & Viral Secretion & $(PFU/s\cdot cell)$ & LN(-3.4, 0.3) \\
+  \hline  
+\end{tabular}
+\caption{\textbf{Latin hypercube sampling distributions.}  LHS was used to generate sample points containing each of the 16 parameters from Table \ref{tab:sensitivity}, plus the sH1N1 viral secretion rate.
+Parameters were sampled over biologically plausible ranges listed here for use in PRCC analysis.
+LN signifies a log-normal distribution.
+Values were sampled from a regular normal distribution with the listed parameters and then transformed to linear space from $log_{10}$ space.
+U signifies a uniform distribution.}
+\label{tab:lhs}
 \end{table}

### 2016-01-28 paper/cover_letter.tex
In process of drafting Response 3

--- 2014-01-27 paper/cover_letter.tex
@@
 
-\begin{flushleft}Dear Editors, \end{flushleft}
+\pagenumbering{gobble}
 
-We submit this manuscript entitled "Spatially explicit model of the efficiency of the lymphocyte diaspora in influenza-infected lung reveals thresholds on chemokine directed migration" for consideration of publication in PLOS ONE.
+\begin{flushleft}Dear Dr. Denise Kirschner \\
+Co-Editor-in-Chief \\
+Journal of Theoretical Biology, \end{flushleft}
+
+%We submit this manuscript entitled "A spatial model of the efficiency of T cell search in the influenza-infected lung" for consideration of publication in Royal Society Interface.
 Our paper implements an agent-based model of bronchial influenza infection and the subsequent T cell response using new empirical values of chemokine secretion.
 The paper shows that chemokine levels have a limited yet important role in directing T cell migration and that the spatial effects of chemokine diffusion can actually hinder T cell search in rapidly spreading viruses.
+Our sensitivity analysis reveals the individual contributions of the model parameters on the course of the infection.
+Finally, we calculate a maximum threshold for T cell search time, beyond which leads to uncontrolled infection.
 We feel this work will be of interest to investigators modeling the immune response for the ultimate purpose of improving vaccine design.
+
+We are pleased to submit the revised manuscript, ``A spatial model of the efficiency of T cell search in the influenza-infected lung" for JTB's consideration.
+This is a resubmission of manuscript number JTB-D-15-00501, received by us on August 12, 2015 with an invitation for resubmission within 60 days.
 \\
 
-This work has not been published elsewhere, and is complementary to a recent publication from our group (H Mitchell, D Levin, et al. J Virol 2011;85(2):1125-1135). \\
-
-In review we have noticed that our bibliography does not use annotated journal names, which seem to be the standard for PLOS ONE.
-We have {\em not} changed our bibliography as of yet as it was generated automatically using the Mendeley Desktop suite.
-If this is an issue, we would be happy to return and reconstruct the references manually.
+The reviewer comments were extremely helpful, and we have addressed each of their suggestions in the attached Response to Reviewers.
+To summarize the more important changes, we performed the recommended PRCC sensitivity analysis for one viral strain.
+Because it confirmed the results of our original sensitivity analysis, we conducted this computationally intensive analysis for only this strain.
+Beyond the PRCC runs, we doubled the number of model runs using our standard parameters.
+This resulted in a very small decrease in the variance of the results and did not affect the mean.
+This gives us additional confidence in the validity of the reported results.
+Next, we reorganized the Methods and Model section to make the rationale for our parameter choices more clear.
+We note that the PRCC sensitivity analysis confirms that many of these parameters do not have a strong effect on the model output.
+Finally, we corrected the typographical errors pointed out by the referees.
 \\
 
-% This work is a resubmission of a rejected PLOS CB manuscript (PCOMPBIOL-D-12-01516R1).
+We feel that the paper is much improved with these changes, and we appreciate your patience during the revision process.
 
-All authors have read and approved the revised manuscript.
+\begin{flushleft}Sincerely, \end{flushleft}
 
-\begin{flushleft}Respectfully yours,\end{flushleft}
-Drew Levin \\
-Stephanie Forrest \\
-Soumya Banerjee \\
-Candice Clay \\
-Melanie Moses \\
-Frederick Koster
+\noindent Drew Levin \\
+University of New Mexico Department of Computer Science \\
+MSC01 1130 \\
+1 University of New Mexico \\
+Albuquerque, NM 87131 \\
+drew@cs.unm.edu
 

### 2016-01-28 paper/cover_letter2.tex
In process of drafting Response 3

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Professional Formal Letter
% LaTeX Template
% Version 1.0 (28/12/13)
%
% This template has been downloaded from:
% http://www.LaTeXTemplates.com
%
% Original author:
% Brian Moses (http://www.ms.uky.edu/~math/Resources/Templates/LaTeX/)
% with extensive modifications by Vel (vel@latextemplates.com)
%
% License:
% CC BY-NC-SA 3.0 (http://creativecommons.org/licenses/by-nc-sa/3.0/)
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%----------------------------------------------------------------------------------------
%	PACKAGES AND OTHER DOCUMENT CONFIGURATIONS
%----------------------------------------------------------------------------------------

\documentclass[11pt,a4paper]{letter} % Specify the font size (10pt, 11pt and 12pt) and paper size (letterpaper, a4paper, etc)

\usepackage{graphicx} % Required for including pictures
\usepackage{microtype} % Improves typography
%\usepackage{gfsdidot} % Use the GFS Didot font: http://www.tug.dk/FontCatalogue/gfsdidot/
%\usepackage[T1]{fontenc} % Required for accented characters

% Create a new command for the horizontal rule in the document which allows thickness specification
\makeatletter
\def\vhrulefill#1{\leavevmode\leaders\hrule\@height#1\hfill \kern\z@}
\makeatother

%----------------------------------------------------------------------------------------
%	DOCUMENT MARGINS
%----------------------------------------------------------------------------------------

\textwidth 6.75in
\textheight 10.5in
\oddsidemargin -.25in
\evensidemargin -.25in
\topmargin -1.5in
\longindentation 0.50\textwidth
\parindent 0.4in

%----------------------------------------------------------------------------------------
%	SENDER INFORMATION
%----------------------------------------------------------------------------------------

\def\Who{Drew Levin} % Your name
\def\What{, PhD Candidate} % Your title
\def\Where{Dept. of Computer Science} % Your department/institution
\def\Address{MSC01 1130 \\ 1 Univ. of New Mexico} % Your address
\def\CityZip{Albuquerque, NM 87131} % Your city, zip code, country, etc
\def\Email{E-mail: drew@cs.unm.edu} % Your email address

%----------------------------------------------------------------------------------------
%	HEADER AND FROM ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\address{
\includegraphics[width=1in]{logo.jpg} % Include the logo of your institution
\hspace{5.1in} % Position of the institution logo, increase to move left, decrease to move right
\vskip -1.07in~\\ % Position of the text in relation to the institution logo, increase to move down, decrease to move up
\Large\hspace{1.5in}UNIVERSITY \hfill ~\\[0.05in] % First line of institution name, adjust hspace if your logo is wide
\hspace{1.5in}OF NEW MEXICO \hfill \normalsize % Second line of institution name, adjust hspace if your logo is wide
\makebox[0ex][r]{\bf \Who \What }\hspace{0.08in} % Print your name and title with a little whitespace to the right
~\\[-0.11in] % Reduce the whitespace above the horizontal rule
\hspace{1.5in}\vhrulefill{1pt} \\ % Horizontal rule, adjust hspace if your logo is wide and \vhrulefill for the thickness of the rule
\hspace{\fill}\parbox[t]{2.85in}{ % Create a box for your details underneath the horizontal rule on the right
\footnotesize % Use a smaller font size for the details
\Who \\ \em % Your name, all text after this will be italicized
\Where\\ % Your department
\Address\\ % Your address
\CityZip\\ % Your city and zip code
\Email\\ % Your email address
}
\hspace{-1.4in} % Horizontal position of this block, increase to move left, decrease to move right
\vspace{-1in} % Move the letter content up for a more compact look
}

%----------------------------------------------------------------------------------------
%	TO ADDRESS STRUCTURE
%----------------------------------------------------------------------------------------

\def\opening#1{\thispagestyle{empty}
{\centering\fromaddress \vspace{0.6in} \\ % Print the header and from address here, add whitespace to move date down
\hspace*{\longindentation}\today\hspace*{\fill}\par} % Print today's date, remove \today to not display it
{\raggedright \toname \\ \toaddress \par} % Print the to name and address
\vspace{0.4in} % White space after the to address
\noindent #1 % Print the opening line
% Uncomment the 4 lines below to print a footnote with custom text
%\def\thefootnote{}
%\def\footnoterule{\hrule}
%\footnotetext{\hspace*{\fill}{\footnotesize\em Footnote text}}
%\def\thefootnote{\arabic{footnote}}
}

%----------------------------------------------------------------------------------------
%	SIGNATURE STRUCTURE
%----------------------------------------------------------------------------------------

\signature{\Who \\ \Where \\ University of New Mexico} % The signature is a combination of your name and title

\long\def\closing#1{
\vspace{0.1in} % Some whitespace after the letter content and before the signature
\noindent % Stop paragraph indentation
\hspace*{\longindentation} % Move the signature right
\parbox{\indentedwidth}{\raggedright
#1 % Print the signature text
\vskip 0.65in % Whitespace between the signature text and your name
\fromsig}} % Print your name and title

%----------------------------------------------------------------------------------------

\begin{document}

%----------------------------------------------------------------------------------------
%	TO ADDRESS
%----------------------------------------------------------------------------------------

\begin{letter}
{Dr. Denise Kirschner \\
Co-Editor-in-Chief \\
Journal of Theoretical Biology
}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

\opening{Dear Dr. Kirschner,}

We are pleased to resubmit the revised manuscript, ``A spatial model of the
efficiency of T cell search in the influenza-infected lung" for JTB's 
consideration. This is a resubmission of manuscript number JTB-D-15-00501R1,
received by us on October 26, 2015 with an invitation for resubmission within 
two months.

We are happy to learn that the major concerns of Reviewer 2 and the previous 
concerns of Reviewer 1 have been addressed. Reviewer 1 now asks for alternative
models (major point 1) to achieve higher confidence in our estimation of specific
parameters. We have addressed this concern in our current response and we 
note that two separate sensitivity analyses confirmed that the parameters in 
question (T cell production rate and the chemokine secretion rate) do not have 
a strong effect on the model results. Reviewer 1 has also raised several new 
interesting questions and suggested further extensions to our model, including
new parameters describing the innate immune response (major point 4) and 
environmental factors such as vascular structure (major point 3).  We feel that these 
requests point to questions that are outside the scope of our current study, 
which focuses on comparing the effectiveness of T cells at controlling infection
in three specific influenza strains.

Our revised manuscript discusses how the model takes these additional factors
into account, but we have not explicitly added new parameters to our already 
complicated model. We anticipate several follow-on projects to explore these 
questions using new data sets and alternative models.

With this set of revisions, we hope that you find the paper acceptable 
for publication in the Journal of Theoretical Biology.

\closing{Sincerely,}

%----------------------------------------------------------------------------------------

\end{letter}
\end{document}

### 2016-01-28 paper/final/source/Levin_JTB.tex
In process of drafting Response 3

--- 2016-01-28 paper/Levin_JTB_Clean.tex
@@
 % amssymb package, useful for mathematical symbols
-%\usepackage{amssymb}
+\usepackage{amssymb}
 
@@
 
-\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
-\newcommand{\new}[1]{{\color{dkgreen}#1}}
+%\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
+%\newcommand{\new}[1]{{\color{dkgreen}#1}}
+\newcommand{\removed}[1]{}
+\newcommand{\new}[1]{#1}
 \newcommand{\drew}[1]{{\color{dkgreen}#1}}
@@
 The model is parameterized with values taken from the literature when available.
-Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and an earlier ODE model \citep{Mitchell2011}.
+Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and \removed{an earlier}\new{our previously published} ODE model \citep{Mitchell2011}.
+
 To obtain the physical data, wells of human epithelial cells were infected with one of the three strains and then measured for virus and cytokine concentrations over a 48-hour period.
@@
 
-The spatial model and the supporting experiments reveal several interesting and underappreciated phenomena.
+The \new{analyses of the} spatial model \removed{and the supporting experiments} reveal several interesting and underappreciated phenomena.
 The model identified physical constraints on T cells' ability to clear the rapidly replicating pH1N1 infection.
@@
 Because we are interested in the interplay between the virus and the induced immune response, we infected human epithelial cells with the three different strains of influenza \textit{in vitro} and measured the resulting cytokine and chemokine responses.
-We extended a previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
+We extended \removed{a}\new{our} previously published differential equation model \citep{Mitchell2011} to obtain values for per-cell chemokine production rates.
 T cell production rates were derived from measured replication rates \textit{in vitro} \citep{Miao2010a} using another differential equation model.
@@
 Strain-specific values for $r$ were found by fitting the equations to the experimental data in Table S1 using a genetic algorithm to minimize the log squared error between the model and the data while holding the rest of the parameter values constant.
-Next, 1,000 bootstrapping runs were performed on each modeled strain to generate confidence intervals over $r$.
+Next, 1,000 bootstrapping runs \new{using resampled residuals} were performed on each modeled strain to generate confidence intervals over $r$.
 The results of these fits are shown in Table 1 (viral secretion rates are from previous fits in \cite{Mitchell2011}).
@@
 The area of the region expressing chemokine was found to be related non-linearly to the number of infected cells ($\pi r'^{2} = a \sqrt{I} + b$) where $a$ and $b$ are constants that depend on the viral strain.
-$a$ and $b$ were fit using 12 experimental runs of a spatial model implemented in CyCells.
+$a$ and $b$ were fit \removed{using 12}\new{to 5} experimental runs of a spatial model implemented in CyCells \new{($R^2 = 0.997$, $P < 0.01$)}.
 
@@
 
-Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).
+Partial rank correlation coefficient (PRCC) analysis was performed on the sH1N1 strain \new{as described in \cite{Marino2008}} using the same 16 parameters plus the viral secretion rate to determine the strength and significance of the parameters' partial correlation with the model output (S2.4).
 The  five viral parameters are confirmed by the PRCC analysis to be significantly correlated with model output (Table \ref{tab:sensitivity}).
@@
 Although leukocytes exhibit directional behavior to chemokines \citep{LiJeon2002, McDonald2010}, CD8 T cells have not yet been shown to climb chemokine gradients.
-While T cell chemotaxis may not be proven, our previous work has shown severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
+While T cell chemotaxis may not be proven, our previous work \removed{has shown}\new{suggests} severe limitations to viral clearance in the absence of this ability \citep{Banerjee2011}.
 Our sensitivity analysis shows that slowing T cell speed (and consequentially, the ability of T cells to climb gradients) reduces clearance of sH1N1 and pH1N1 strains.
@@
 
-Previous ODE models have shown that viral dynamics are critical in determining the course of an infection \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.
+\removed{Previous ODE models have shown that viral dynamics are critical in determining the course of an infection}\new{Previous ODE models have studied how viral dynamics affect the course of infection} \citep{Handel2008, Lee2009, Miao2010a, Mitchell2011, Saenz2010, Crauste2015, Price2015}.
 The use of an ABM can complement spatially homogeneous differential equation models \citep{Beltman2007, Textor2014, Vroomans2012, Zheng2008}.

### 2016-01-28 paper/final/source/Levin_JTB_Supplement.tex
In process of drafting Response 3

--- 2016-01-28 paper/chemokine_supplement_revised.tex
@@
 \usepackage{pdflscape}
-
-\usepackage{xr}
-\externaldocument{chemokine_main_revised}
 
@@
 
-An overview of the model implementation is shown in Fig.~\ref{fig:modelchart}.
+An overview of the model implementation is shown in Fig.
+2.
 The model is initialized with one virus-secreting cell placed in the middle of a 2-dimensional sheet of hexagonally-tiled epithelial cells (288,212 total cells).
-The sheet measures 5$mm$ per side, which is large enough to contain the infection (Fig.~\ref{fig:cycells}).
+The sheet measures 5$mm$ per side, which is large enough to contain the infection (Fig.
+5).
 The molecular grid (Text S1 1.3) is initialized with a resolution of 20$\mu m$.
@@
 
-The model simulates a chemokine gradient surrounding the infected focus (Fig.~\ref{fig:cycells}), based on the calculated per-infected cell secretion rate (Table \ref{tab:strains}) and known chemical parameters for a 10 kDa protein (Table \ref{tab:parameters}).
+The model simulates a chemokine gradient surrounding the infected focus (Fig.
+5), based on the calculated per-infected cell secretion rate (Table 3) and known chemical parameters for a 10 kDa protein (Table 2).
 T cell sensitivity depends on receptor density \citep{Desmetz2006} and this was assumed to be constant.
@@
 % we hypothesize that it renders IP-10 ineffectual.
-We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table \ref{tab:strains}).
+We hypothesize that this leads to the elevated RANTES secretion rates measured in aH5N1 compared to the other two strains (Table 3).
 Because of this behavior IP-10 was not included in the aH5N1 model runs.
@@
 
-The model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Tables \ref{tab:strains} and \ref{tab:parameters}).
+The model parameters were chosen from literature when available and estimated within plausible ranges otherwise (Tables 1 and 2).
 One-factor-at-a-time (OFAT) analysis was performed by varying Individual parameters over ranges of plausible (and sometimes even implausible) values while the rest of the parameter set was held constant.
@@
 
-We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table \ref{tab:sensitivity}).
+We then categorized the model parameters into one of three qualitative groups: parameters that do not affect the model's qualitative behavior (stable parameters), parameters that affect the model's quantitative peak infection size but do not affect final clearance, (peak change parameters), and those parameters that affect the final clearance of the infection (sensitive parameters) (Table 3).
 Parameters are presented along with their range using the following format:  \\
@@
 Specifically, the plausible parameter values in this group are within a range that does not significantly affect the asymptotic behavior of the model.
-We further split this group into two categories in Table \ref{tab:sensitivity}, \textit{stable} parameters and \textit{bounded stable} parameters.
+We further split this group into two categories in Table 3, \textit{stable} parameters and \textit{bounded stable} parameters.
 \textit{Bounded stable} parameters are only stable within a specific range of values, while \textit{stable} parameters seem to be stable over all the values that were tested.
@@
 Due to the nature of the ABM, inclusion of these parameters is mechanistically necessary.
-Thus, we include and parameterize them using the methods described in the main paper and Table \ref{tab:parameters}.
+Thus, we include and parameterize them using the methods described in the main paper and Table 2.
 
@@
 
-An explicit sensitivity analysis was not directly performed.
-The range of the parameters are taken from the respective secretion rates of the aH5N1, sH1N1, and pH1N1 influenzas (Table \ref{tab:strains}).
+An explicit sensitivity analysis was not directly performed.  The range of the parameters are taken from the respective secretion rates of the aH5N1, sH1N1, and pH1N1 influenzas (Table 1).
 
@@
 PRCC analysis was performed as described in \cite{Marino2008}.
-Samples were generated using Latin hypercube sampling (LHS) to generate 333 sample points for 17 parameters (those from Table \ref{tab:sensitivity}, plus the sH1N1 viral secretion rate).
+Samples were generated using Latin hypercube sampling (LHS) to generate 333 sample points for 17 parameters (those from Table 3, plus the sH1N1 viral secretion rate).
 Each parameter was assigned a confidence distribution (Table \ref{tab:lhs}) and sampled using its inverse cumulative distribution function to generate points with density proportional to the underlying probability distribution function.
@@
 PRCC analysis was performed over every 100th time point, resulting in a time series of 1441 Spearman rank correlation coefficients ($\rho$) and significance values ($p$) for each parameter (Figs.
-\ref{fig:prcc} - \ref{fig:vsec}).
+S6 - S7).
 
@@
 Viral infectivity, viral diffusion rate, and viral secretion rate, values of $\rho$ cross or approach zero over time.
-This merely reflects that the modeled target cells have died (Fig.~\ref{fig:exhaustive}) and so the decline in $\rho$ is irrelevant.
+This merely reflects that the modeled target cells have died (Fig.
+S8) and so the decline in $\rho$ is irrelevant.
 
@@
 Extreme values of the viral parameters cause rapid infection growth to the point that the modeled system is exhausted (Fig.
-\ref{fig:exhaustive}).
+S8).
 In these cases the number of infected cells tends towards zero at later time points as every modeled target cell has died before the model is complete.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{FigureS1}
- \end{center}
-\caption{{\bf Preliminary Model 1 fits to data.}  Preliminary Model 1 (Eq.
-1) was fit to experimental data (Table~\ref{tab:data}) using a genetic algorithm as described in the main paper.
-sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL and these three values (empty red triangles) were not included in the model fitting.
-RMSE values of the fits can be seen in Table~\ref{tab:rmse}.
-IP-10 data and model fits shown in red (triangles), RANTES data and model fits shown in yellow (squares).}
- \label{fig:fits}
-\end{figure}
-
-\begin{figure}[!ht]
-\begin{center}
- \includegraphics[width=\textwidth]{Figure_S2}
+ \includegraphics[width=\textwidth]{Figure_S-1}
  \end{center}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S3}
+	\includegraphics[width=\textwidth]{Figure_S2}
 	\caption{\textbf{Effects of different chemokine combinations.}  A) aH5N1 does not stimulate an IP-10 response.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S4}
+	\includegraphics[width=9in]{Figure_S3}
 	\caption{\textbf{aH5N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the aH5N1 model simulation.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S5}
+	\includegraphics[width=9in]{Figure_S4}
 	\caption{\textbf{sH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the sH1N1 model simulation.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S6}
+	\includegraphics[width=9in]{Figure_S5}
 	\caption{\textbf{pH1N1 sensitivity analysis.} The sensitivity analysis results for 16 parameters applied to the pH1N1 model simulation.
@@
 \begin{center}
-	\includegraphics[width=9in]{Figure_S7}
+	\includegraphics[width=9in]{Figure_S6}
 	\caption{\textbf{PRCC sensitivity analysis.} PRCC analysis generated Spearman correlation coefficients ($\rho$) and significance values ($p$) over 1441 time points.
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S8}
+	\includegraphics[width=\textwidth]{Figure_S7}
 	\caption{\textbf{PRCC sensitivity analysis for the viral secretion rate.} Similar to Figure S6, PRCC analysis generated Spearman correlation coefficients ($\rho$) and significance values ($p$) over 1441 time points.
@@
 Viral secretion shows significant effect on model output.
-The decline of $\rho$ over time reflects the early death of every modeled target cell (Figure~\ref{fig:exhaustive})).}
+The decline of $\rho$ over time reflects the early death of every modeled target cell (Figure S8).}
 	\label{fig:vsec}
@@
 \begin{center}
-	\includegraphics[width=\textwidth]{Figure_S9}
+	\includegraphics[width=\textwidth]{Figure_S8}
 	\caption{\textbf{Exhaustive Growth:} Certain parameter sets generated by Latin hypercube sampling caused rapid uncontrolled infection growth (Samples 1-3).
@@
 A larger, unbounded model environment would show unconstrained growth.
-This effect explains the decline of the $\rho$ value of several viral parameters to or past zero over time in Figures \ref{fig:prcc} and \ref{fig:vsec}.}
+This effect explains the decline of the $\rho$ value of several viral parameters to or past zero over time in Figures S6 and S7.}
 	\label{fig:exhaustive}
@@
 All parameters and populations are taken from \citep{Mitchell2011} except for $C$, $d$, and $\tau_3$.
-The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.~\ref{fig:data}).
+The value chosen for $d$ corresponds to a 30 minute half-life and values for $\tau_3$ were set from the observed chemokine data (Fig.
+3).
 Interferon ($F$) is an abstracted unitless quantity and thus $e$ is a unitless multiplier.}
 \label{tab:dde}
-\end{center}
-\end{table}
-
-\begin{table}[!ht]
-\begin{center}
-\begin{tabular}{| c | c | c |}
-  \hline
-  Strain & Chemokine & RMSE \\
-  \hline
-  \multirow{2}{*}{aH5N1} & RANTES & 166 \\
-  & IP-10 & 7,950 \\
-  \hline
-  \multirow{2}{*}{sH1N1} & RANTES & 90 \\
-  & IP-10 & 2,732 \\
-  \hline
-  \multirow{2}{*}{pH1N1} & RANTES & 2,671  \\
-  & IP-10 & 64,342 \\
-  \hline
-\end{tabular}
-\caption{\textbf{Root mean squared error of Preliminary Model 1 fits.}  Preliminary Model 1 (Eq.
-1) was fit to experimental data (Table~\ref{tab:data}) using a genetic algorithm as described in the main paper.
-The resulting model fits can be seen in Figure~\ref{fig:fits}.}
-\label{tab:rmse}
 \end{center}
@@
 \end{tabular}
-\caption{\textbf{Latin hypercube sampling distributions.}  LHS was used to generate sample points containing each of the 16 parameters from Table \ref{tab:sensitivity}, plus the sH1N1 viral secretion rate.
+\caption{\textbf{Latin hypercube sampling distributions.}  LHS was used to generate sample points containing each of the 16 parameters from Table 3, plus the sH1N1 viral secretion rate.
 Parameters were sampled over biologically plausible ranges listed here for use in PRCC analysis.

### 2016-01-28 paper/response2.tex
In process of drafting Response 3

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color} 
\usepackage[usenames,dvipsnames,table]{xcolor}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

\usepackage{multirow}

\usepackage{color}
\usepackage{csquotes}
\usepackage{ulem}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

\definecolor{dkred}{rgb}{0.75,0,0}
\definecolor{dkgreen}{rgb}{0,0.5,0}
\definecolor{dkblue}{rgb}{0,0,0.75}
\definecolor{dkpurple}{rgb}{.375,0,.375}
\definecolor{gray}{rgb}{0.5,0.5,0.5}

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
%\newcommand{\removed}[1]{\textcolor{dkred}{#1}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}

\newenvironment{response}{\fontfamily{cms}\selectfont\small}{\par}

\renewcommand{\rmdefault}{cmr}
\renewcommand{\sfdefault}{lmss}

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung (Response to Reviewers)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin, 
Stephanie Forrest, 
Soumya Banerjee,
Candice Clay,
Judy Cannon, 
Melanie Moses, 
Frederick Koster
\end{flushleft}
\vspace{0.5cm}

%We thank the reviewers for their constructive and insightful comments.  We have done our best to address each point raised and feel the incorporation of the reviewers' comments has greatly improved the paper.  We include detailed responses to each comment below.
We thank the reviewers for their efforts in providing insightful critiques. Below we have provided detailed responses to each individual point raised by the reviewers. As the response to Reviewer 2 is brief, we address Reviewer 2's critique first. We hope that the reviewers and editor will find these detailed responses sufficient to address all their current critiques and find our manuscript suitable for publication in the Journal of Theoretical Biology.

\section*{Reviewer 2}

\textbf{One last point that can be addressed is that they performed a new detailed uncertainty and Sensitivity analysis, but no references are listed for the method that they employed. It would be helpful to readers to know what approach they took to allow for reproducing their results or for following their method for future work. This must be rectified.} \\

\begin{response}
We are glad that Reviewer 2 is satisfied with the previous revisions, and we have addressed the one remaining suggestion by adding in the recommended reference (Marino et. al., 2008) in the Sensitivity Analysis section of the main paper.
\end{response}

\section*{Reviewer 1}

\begin{enumerate}

\item \textbf{You provided more detail of how you estimate the rate of chemokine production and how T cell production from LNs are calculated. However, from your description I cannot evaluate whether your numbers can be trust. Specifically, for model (Eq. 1) I have no idea whether your model predicts accurately chemokine production. No fits are shown for the data. How do you know that the parameters from Mitchell et al 2011 are compatible with your experiments. Although we hope that different experiments give the same parameter measurements, this is often not true. See paper by De Boer's group (Althaus et al JI 2008). Furthermore, your model fits for Eqn. 2 to estimate sigma (page 8). Again, I have no idea whether your model is good for the data as you claim. To me it appears that you think that estimating model parameters from data is a trivial exercise. It is not and one must try different alternative models to see how they describe the data. See book by Hilborn Ecological detective.}

\begin{response}

We are glad that the reviewer is happy with the additional detail provided with our previous revision. In response to this critique, we emphasize that the Mitchell 2011 reference is a paper from our group (including multiple authors of the current manuscript), and the data modeled in the earlier paper were used here to calibrate our new model in the current manuscript. We clarify the appropriateness of using parameters from the Mitchell 2011 paper as follows:

\begin{displayquote}
``The model is parameterized with values taken from the literature when available. Because chemokine secretion rates are central to our model and appropriate data values are not available, we estimate the parameters using data from physical experiments and \removed{an earlier} \new{our previously published} ODE model (Mitchell et. al., 2011)." (p. 3)
\end{displayquote}

\begin{displayquote}
``We extended \removed{a} \new{our}  previously published differential equation model (Mitchell et al., 2011) to obtain values for per-cell chemokine production rates. T cell production rates were derived from measured replication rates in vitro (Miao et al., 2010) using another differential equation model. T cell production is assumed to be constant after day 5 (Martin-Fontecha et al., 2003).'' (p. 5) 
\end{displayquote}

In addition, there is clarifying text on p. 6:

\begin{displayquote}
``the dynamic viral loads at these intervals have been reported previously by us for cultures infected with seasonal H1N1 virus, pandemic H1N1 virus, and avian H5N1 virus (Mitchell et al., 2011).''
\end{displayquote}

However, as the reviewer correctly points out, no single model is likely to fit all experimental data pefectly.   For the question we are asking, the goal is less to look at the exact fits than to study the range of qualitative behaviors of T cell / chemokine interactions in the context of infection by three strains of influenza.  We have a deep appreciation for the difficulty of estimating parameter values from models, which is why we conducted a thorough sensitivity analysis, including the PRCC extension helpfully suggested by the reviewer earlier.

Our sensitivity analysis, shown in Table 3 and Figures S3-S8, demonstrates that the two parameters that the reviewer is concerned about (chemokine production rate and T cell production rate) do not clear pandemic flu.  We tested chemokine production rates over a wide range of values: $\pm$ 2 orders of magnitude from strain-specific baseline, and  T cell production ranging from  $125/h$ to $3,750/h$, see Table 2.  Thus, we acknowledge the reviewer’s comment that estimating model parameters is not trivial, but our model results suggest that while the T cell response is sufficient to clear seasonal flu strains, chemokine production and T cell production rates are not likely to impact clearance of pandemic flu.  We note that to our knowledge this is the first such analysis, and we hope that subsequent models and experiments will shed further light on this important question.
\end{response}

\item \textbf{You ignored my request to investigate if virus control can be good by T cells (irrespectively of the type of virus) if chemokines diffuse faster than virus particles. Please do the simulations, and rephrase your discussion depending on the results you find.}

\begin{response}
We apologize that our previous response to this was not clear. In fact, we explicitly model chemokines diffusing faster than virus particles (see Table 2). Thus, all of our conclusions are based on modeling chemokine diffusing faster than virus.
\end{response}

\item \textbf{One additional issue I thought of is the ability of T cells to recruit more T cells to the site of infection by also secreting cytokines or changing vasculature structure to allow more T cell entry. Will this improve viral control?}

\begin{response}
The reviewer correctly points out that T cells may recruit more T cells to the site by changing the inflammatory environment. While no model can incorporate all potential aspects of T cell effects on the environment, we do address the reviewer's question by varying the number of recruited T cells over a wide range (see Sensitivity Analysis: Table 3 and Figures S3-S5). If T cells secrete more cytokines or change the vasculature in order to recruit more T cells, increasing the number of T cells would take this potential effect into account without explicitly modeling cytokines or vasculature.  As stated in the paper, we find that while T cells are sufficient to control seasonal H1N1,  increasing T cell secretion from 1,257 per hour to  3,750 per hour does not affect T cell control of pandemic flu (Figure S5).
\end{response}

\item \textbf{I wonder if your model does not represent flu dynamics appropriately. For all the data on flu dynamics in mice that I have seen, flu titers reach peak very early in infection, about 2-3 days p.i, while T cells start arriving to the lung 5-6 days post-infection (e.g., work by Doherty/Thomas). It seems to be that difference in pathogenesis of different viruses would come not from the inability of T cells to control virus spread but how much of the lung is being infected within 1-5 days post infection. Less virulent viruses infect fewer epithelian cells, are all cleared and this does not result in much pathology. Virulent viruses are able to infect much wider areas of the lung and T cells by clearing virus in the whole lung cause death - via immunopathology. Innate immunity is controlling early virus spread. Please investigate.}

\begin{response}
We concur with the reviewer's observation that innate immunity is important to early viral control. Even though we do not explicitly model each possible innate mechanism specifically, our model represents viral spread using parameters which likely reflect innate control mechanisms (virus secretion, infectivity, diffusion rate, and decay rate, see Tables 1 and 2). We appreciate the reviewer's perspective, but our goal in this paper is to study the effect of T cells and not to fully represent the interaction between influenza strains and the entirety of the immune response (Introduction 3rd paragraph shown below).

\begin{displayquote}
``Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, cytokines, and replicating influenza virus. ... Therefore, instead of developing a comprehensive immune system model, we present a spatially explicit agent-based model (ABM) to
describe T cell interactions with chemotactic signals and a dynamically growing plaque." (p. 1)
\end{displayquote}

We feel that careful investigation of the role of early innate control, with its many complex mechanisms, in different influenza strains is best left to future studies.

\end{response}

\end{enumerate}

\subsection*{Minor Issues}

\begin{itemize}

\item \textbf{Your reference list must be either numbered or be alphabetical. When you cite a reference, I cannot easily find it in the list of your references because currently references are not alphabetical (or do not have any numbering). I am surprised that you did not understand the issue yourself.}

\begin{response}
We apologize for the inaccuracies in the reference list and have corrected this as requested.
\end{response}

\item \textbf{page 3 - ``spatial model and the supporting experiments" - which experiments? Chemokine production? This is only very remotely related to your model.}

\begin{response}
The text has been changed to: ``\new{The analyses of} the spatial model reveal..."
\end{response}

\item \textbf{How bootsraps were performed (page 7) is not stated. Resampled data, residuals, parametric or non-parametric bootstrap. Please address}

\begin{response}
The text has been updated to read: ``1,000 bootstrapping runs \new{using resampled residuals} were performed..."
\end{response}

\item \textbf{page 8. Estimation of parameters a and b is not. This has to be extended, fits are shown and argued that these are appropriate fits (by doing all the needed statistics for nonlinear regressions!)}

\begin{response}
$a$ and $b$ were fit using a linear regression with a resulting $R^2 = 0.997$ and $P < 0.01$.  These values have been added to the main text and supplement, and the data points used for fitting have also been added to the supplement (Table S3).
\end{response}

\item \textbf{page 15 - reference to Banerjee et al. 2011 as showing "severe limitation to viral clearance ... ``is not appropriate as this is theoretical work. Only experiments may show this, model may only suggest.}

\begin{response}
The phrase ``has shown" has been changed to ``suggests".
\end{response}

\item \textbf{page 15. ``virus dynamics are crucial in determining the course of infection". }

\begin{response}
We changed the text to say \new{previous ODE models have studied how viral dynamics affect the course of infection}.
\end{response}

\end{itemize}

\end{document}

### 2016-01-28 paper/response3.tex
In process of drafting Response 3

\documentclass[10pt]{article}

% amsmath package, useful for mathematical formulas
\usepackage{amsmath}
% amssymb package, useful for mathematical symbols
\usepackage{amssymb}

% graphicx package, useful for including eps and pdf graphics
% include graphics with the command \includegraphics
\usepackage{graphicx}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

\usepackage{color} 
\usepackage[usenames,dvipsnames,table]{xcolor}

% Use doublespacing - comment out for single spacing
\usepackage{setspace} 
%\doublespacing

% Text layout
\topmargin 0.0cm
\oddsidemargin 0.5cm
\evensidemargin 0.5cm
\textwidth 16cm 
\textheight 21cm

% Bold the 'Figure #' in the caption and separate it with a period
% Captions will be left justified
\usepackage[labelfont=bf,labelsep=period,justification=raggedright]{caption}

% Use the PLoS provided bibtex style
%\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

\pagestyle{myheadings}
%% ** EDIT HERE **

\usepackage{multirow}

\usepackage{color}
\usepackage{csquotes}
\usepackage{ulem}

%% ** EDIT HERE **
%% PLEASE INCLUDE ALL MACROS BELOW

\definecolor{dkred}{rgb}{0.75,0,0}
\definecolor{dkgreen}{rgb}{0,0.5,0}
\definecolor{dkblue}{rgb}{0,0,0.75}
\definecolor{dkpurple}{rgb}{.375,0,.375}
\definecolor{gray}{rgb}{0.5,0.5,0.5}

\newcommand{\removed}[1]{{\color{dkred}\sout{#1}}}
%\newcommand{\removed}[1]{\textcolor{dkred}{#1}}
\newcommand{\new}[1]{{\color{dkgreen}#1}}

%\newenvironment{response}{\fontfamily{cms}\selectfont\small}{\par}
\newenvironment{response}{\fontfamily{cmr}}{\par}

\renewcommand{\rmdefault}{cmr}
\renewcommand{\sfdefault}{lmss}

% figure files reside in the figures/ directory
\graphicspath{
{figures/}
}

%% END MACROS SECTION

\begin{document}

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf{A spatial model of the efficiency of T cell search in the influenza-infected lung (Response to Reviewers)}
}
% Insert Author names, affiliations and corresponding author email.
\\
Drew Levin, 
Stephanie Forrest, 
Soumya Banerjee,
Candice Clay,
Judy Cannon, 
Melanie Moses, 
Frederick Koster
\end{flushleft}
\vspace{0.5cm}

%We thank the reviewers for their constructive and insightful comments.  We have done our best to address each point raised and feel the incorporation of the reviewers' comments has greatly improved the paper.  We include detailed responses to each comment below.

%We thank the reviewers for their efforts in providing insightful critiques. Below we have provided detailed responses to each individual point raised by the reviewers. As the response to Reviewer 2 is brief, we address Reviewer 2's critique first. We hope that the reviewers and editor will find these detailed responses sufficient to address all their current critiques and find our manuscript suitable for publication in the Journal of Theoretical Biology.

We are grateful for the opportunity to respond to Reviewer 1's comments. We regret that Reviewer 1 did not ultimately agree with our rebuttal, but we believe that we have directly addressed all of his/her concerns in the previous rebuttal. We are providing additional evidence and discussion as requested by the last round of revisions. While we are grateful for the time given by Reviewer 1 to review our manuscript, we respectfully disagree with his/her contention that we did not understand or address his/her concerns. We provide additional rebuttals to Reviewer 1's comments below. We believe that we have adequately addressed all comments from both Reviewer 1 and 2.

\section*{Reviewer 1}

Reviewer 1: The authors misinterpreted what my comments meant, and this is third time.

\begin{enumerate}

\item \textbf{If you estimate chemokine production rate, I want to know how. You show no 
model fits, so I cannot trust your numbers/analysis. The fact that chemokine 
secretion rate does not impact clearance of pandemic virus raises a question -
why do you need to estimate secretion rate then?}

\begin{response}

There is no known literature with chemokine secretion rate estimations.  In this paper, we present new empirical data of \textit{in vitro} chemokine levels and extend a previously published ODE model to estimate chemokine production rates.  We are as surprised as Reviewer 1 that these production rates do not have a strong effect on our agent-based model results as determined by our sensitivity analyses.  Nevertheless, because our agent-based model examines chemokine interactions, we require a non-zero value for the production rate.  We believe that fitting the extended ODE model to our new empirical data is the best approach available to determine biologically plausible values.  Our paper contains a subsection, \textit{Models for Parameter Estimation}, that details the exact process used to estimate the chemokine production rates (page 6).  We fit the system of differential equations to empirical data using a genetic algorithm to minimize the log squared error between the model and the data (the genetic algorithm performed favorably when compared to standard Levenberg-Marquardt nonlinear regression techniques).

We have added a figure displaying the model fits to the supplement (Fig. S1, also below) and Table S3 showing the root mean squared error of the fits for completeness.

%\begin{displayquote}
%``quote from paper"
%\end{displayquote}
%
%more response stuff

\end{response}

\item \textbf{Mitchell et al. work is on in vitro culture while you model in vivo lung 
infection. These are two completely different scenarios, parameter can and are 
likely to be different.}

\begin{response}
All biological data have a foundation using \textit{in vitro} research. We agree with Reviewer 1 that while this raises the potential caveat that some of the values that we use to parameterize our models using \textit{in vitro} generated data may not reflect \textit{in vivo} values, we, along with most other researchers in the field, use \textit{in vitro} generated values as a starting point. We explicitly state this in the manuscript. This is precisely the reason that we undertook an extensive sensitivity analysis as we acknowledge that the values we use to parameterize the model may vary from initial starting values. We do not claim that the values we estimate are absolutely reflective of \textit{in vivo} values. Instead, we use best fit estimates that are currently available as a starting point, and use sensitivity analysis to confirm the role of individual parameters on the model.​
\end{response}

\item \textbf{You state that chemokine diffuses SLOWER that infection spread. What if it diffuses faster?}

\begin{response}

Chemokine does have a higher diffusion rate than virus in our model (Table 2).  It is important to note that our discussion of the infection outpacing the chemokine is different than the direct comparison between the viral diffusion rate and the chemokine diffusion rate.  Because T cells climb the chemokine gradient, a chemokine's effect is seen in the location of its local maxima, while the viral effect is seen in the spread and concentration of the virions themselves.  We have edited the text to be more clear of this distinction (\textit{Results: Spatial Effects}, page 10).

\begin{displayquote}
Finally, the spatial nature of our model reveals that \new{the locations of peak chemokine concentration}\removed{chemokines} can \new{move}\removed{diffuse} more slowly than the rate infected cells and virus expand, thus misdirecting the T cells.
\end{displayquote}

To complement our spatial analysis we performed multiple sensitivity analyses on our model's free parameters.  In doing so, we examined scenarios where the chemokine's diffusion rate was increased by up to two orders of magnitude (Table 2).  Each sensitivity analysis showed that the chemokine diffusion rate was an insensitive parameter in our model (Table 3 and Supplement)

\end{response}

\item \textbf{Secretion of T cells is irrelevant parameter, recruitment to the focus of 
infection is the important one.}

\begin{response}

We agree with the reviewer that the T cell secretion rate is an insensitive parameter in our model as determined by our multiple sensitivity analyses.  Because our agent-based model investigates the recruitment of T cells to the focus of infection, we required a non-zero secretion rate to introduce these T cells into the model.  Similar to the chemokine secretion rate as discussed in the first point and in our paper (\textit{Models for Parameter Estimation}, page 7), we fit an ODE model to empirical data to best estimate a biologically plausible value for the T cell secretion rate for use in our agent-based model.

We also agree that the recruitment of T cells to the focus of the infection is important.  We discuss this in the \textit{Spatial Effects} subsection (page 9-10) and in the discussion (page 15) as follows:

\begin{displayquote}
In our spatial model, CD8 T cells climb a chemokine gradient to find infected epithelial cells and
cluster at local maxima of chemokine concentration. Because T cells are clustered, they cannot cover the
expanding plaque effectively, where infected cells on the periphery become more highly dispersed as the
plaque grows. Thus, T cells in the model become redundant at a relatively low threshold, beyond which
additional T cells do not improve clearance rates.
\end{displayquote}

\end{response}

\item \textbf{You don't understand my point about virus control and relevance of your work
for interpretation of how CD8 T cells control flu replication.}

\begin{response}

The term `viral control' has been used in three different contexts.  We have explicitly addressed each point in previous responses.  In summary:

\begin{enumerate}
\item \textit{You ignored my request to investigate if virus control can be good by T cells (irrespectively of the type of virus) if chemokines diffuse faster than virus particles. Please do the simulations, and rephrase your discussion depending on the results you find.}
\begin{displayquote}
We have performed these simulations numerous times in the form of our sensitivity analyses.  Further, you seem to be misunderstanding the distinction between the differing roles of viral diffusion and chemokine diffusion.  Please refer to Response 3 of this document for more information.
\end{displayquote}

\item \textit{One additional issue I thought of is the ability of T cells to recruit more T cells to the site of infection by also secreting cytokines or changing vasculature structure to allow more T cell entry. Will this improve viral control?}

\begin{displayquote}
Here we summarize our previous response (emphasis added). While no model can incorporate all potential aspects of T cell effects on the environment, we do address the reviewer’s question by varying the number of recruited T cells over a wide range (see Sensitivity Analysis: Table 3 and Figures S4-S7). If T cells secrete more cytokines or change the vasculature in order to recruit more T cells, increasing the number of T cells would take this potential effect into account without explicitly modeling cytokines or vasculature. As stated in the paper, we find that while T cells are sufficient to control seasonal H1N1, increasing T cell secretion from 1,257 per hour to 3,750 per hour \emph{does not affect} T cell control of pandemic flu (Figure S6).
\end{displayquote}

\item \textit{It seems to be that difference in pathogenesis of different viruses would come not from the inability of T cells to control virus spread but how much of the lung is being infected within 1-5 days post infection. Less virulent viruses infect fewer epithelian cells, are all cleared and this does not result in much pathology. Virulent viruses are able to infect much wider areas of the lung and T cells by clearing virus in the whole lung cause death - via immunopathology. Innate immunity is controlling early virus spread. Please investigate.}

\begin{displayquote}
In summary, our paper makes no attempt to model the immune response in its entirety.  Our model does represent viral
spread using parameters which likely reflect innate control mechanisms (virus secretion, infectivity, diffusion
rate, and decay rate, see Tables 1 and 2). We appreciate the reviewer’s perspective, but our goal in this
paper is to study the effect of T cells and not to fully represent the interaction between influenza strains
and the entirety of the immune response (Introduction 3rd paragraph shown below). \\

\textit{``Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, cy-
tokines, and replicating influenza virus. ... Therefore, instead of developing a comprehensive
immune system model, we present a spatially explicit agent-based model (ABM) to describe T
cell interactions with chemotactic signals and a dynamically growing plaque." (page 1)} \\

We feel that careful investigation of the role of early innate control, with its many complex mechanisms, in
different influenza strains is best left to future studies.

\end{displayquote}

\end{enumerate}

\end{response}

\item \textbf{Please read the paper by Oreskes et al. Science 1994 about how mathematical 
modeling should be used to understand natural sciences.}

\begin{response}
We have reviewed Oreskes et. al.  The paper correctly points out that scientists often use terms incorrectly when describing their model's veracity which leads to incorrect claims and conclusions.  Specifically, Oreskes et. al. examines four terms: `verification', `validation', `calibration', and `confirmation'.
\begin{itemize}
\item \textbf{Verification:} Oreskes et. al. states that it is not possible to verify a numerical model by comparing it to an analytical model.  We use the term `verification' one time (page 15, concluding paragraph) in the context of obtaining empirical data only (emphasis added).
\begin{displayquote}
``Empirical \textbf{verification} of the model's sensitive parameters (viral response to IgM, infectivity, viral decay rate, viral diffusion rate, and viral production rate) will be valuable to future studies."
\end{displayquote}
\item \textbf{Validation:} We do not use the term in our paper.
\item \textbf{Calibration:} Oreskes et. al. warn of the dangers of using the calibration of free parameters to emprical data as a basis for claiming model validity.  We use the term `calibration' only once (Abstract) and make no erroneous claims (emphasis added).
\begin{displayquote}
``We \textbf{calibrate} the model using viral and chemokine secretion rates we measure \textit{in vitro} together with values taken from literature."
\end{displayquote}
\item \textbf{Confirmation:} Oreskes et. al. states that observations consistent to a theory can `confirm' a theory, but warn that confirmation cannot lead to verification.  We use the term `confirm' twice and make no claims of verification or validity (emphasis added).

Page 11:
\begin{displayquote}
``The  five viral parameters are \textbf{confirmed} by the PRCC analysis to be significantly correlated with model output (Table 3)."
\end{displayquote}

Page 12:
\begin{displayquote}
``We \textbf{confirm} this effect in the model by first setting both relevant parameters (apoptosis time and T cell kill time) to zero, and as expected all three strains were cleared (Fig. 7)."
\end{displayquote}

\end{itemize}

At no point have we made claims inconsistent with Oreskes et. al.  We agree that models can inform research and use our model to propose a hypothesis of how additional T cells are limited in their effectiveness due to the spatial effects of the chemokine gradient and T cell chemotaxis.
\end{response}

\end{enumerate}

\setcounter{figure}{0}
\renewcommand{\thefigure}{S\arabic{figure}}

\begin{figure}[b]
\begin{center}
\includegraphics[width=\columnwidth]{FigureS1.eps}
\caption{{\bf Preliminary Model 1 fits to data.}  Preliminary Model 1 (Eq. 1) was fit to experimental data (Table S1) using a genetic algorithm as described in the main paper.  sH1N1 IP-10 secretion exceeded measurement accuracy above 8500 pg/mL and these three values (empty red triangles) were not included in the model fitting.  RMSE values of the fits can be seen in Table S3.  IP-10 data and model fits shown in red (triangles), RANTES data and model fits shown in yellow (squares).}
\label{fig:fits}
\end{center}
\end{figure}

\end{document}

### 2016-01-28 log_rmse.m

@@ -1,0 +1,29 @@
+% rmse(data_time, data_points, fn_time, fn_points)
+%   Calculates the root mean squared error between a set of data points and 
+% a series of points representing a function output.
+%
+% Inputs:
+%   data_time   - A 1D array containing the time points of the data
+%   data_points - A 1D array of the data values
+%   fit_time    - A 1D array of the time points for the function output
+%   fit_points  - A 1D array of the function values
+% Output:
+%   error = the root mean squared error of the data vs. the function
+
+function error = log_rmse(data_time, data_points, fn_time, fn_points)
+   
+    % Intepolate the fit to match the time points of the data
+    fn_values = interp1(fn_time, fn_points, data_time);
+    
+    sq_err = 0;
+    count = 0;
+    for i=1:length(data_time)
+       if data_points(i) > 0 && fn_values(i) > 0
+          sq_err = sq_err + (log10(data_points(i)) - log10(fn_values(i)))^2; 
+          count = count+1;
+       end
+    end
+
+    % Square the error values and sum them up
+    error = sqrt(sq_err/count);
+ 

### 2016-01-28 plot_data.m

@@ -69,8 +69,6 @@ subplot(1, 3, 1)
 hold on
 hline3 = plot(ait, ai);
 hline4 = errorbar(avt, av, ave*2, 'ob');
-%hline4 = plot(aft/60, aif);
-%hline5 = plot(aft/60, arf);
 hold off
 
 set(hline1, 'LineWidth', 2)
@@ -98,6 +96,9 @@ set(hline3, 'Color', colors(1,:))
 set(hline3, 'MarkerEdgeColor', colors(5,:))
 set(hline3, 'MarkerFaceColor', colors(1,:))
 
+%set(hline4, 'LineWidth', 2)
+%set(hline4, 'LineStyle', '--')
+%set(hline4, 'Color', colors(1,:))
 set(hline4, 'LineWidth', 2)
 set(hline4, 'LineStyle', 'none')
 set(hline4, 'Marker', 'o')
@@ -106,9 +107,9 @@ set(hline4, 'Color', colors(4,:))
 set(hline4, 'MarkerEdgeColor', colors(8,:))
 set(hline4, 'MarkerFaceColor', colors(4,:))
 
-%set(hline5, 'LineWidth', 2)
-%set(hline5, 'LineStyle', '--')
-%set(hline5, 'Color', colors(2,:))
+% set(hline5, 'LineWidth', 2)
+% set(hline5, 'LineStyle', '--')
+% set(hline5, 'Color', colors(2,:))
 
 axes(haxes(1))
 set(haxes(1), 'FontSize', font_size)
@@ -146,8 +147,6 @@ hold on
 hline4 = plot(siit, sii);
 hline3 = plot(sit, si);
 hline5 = errorbar(svt, sv, sve*2, 'ob');
-%hline5 = plot(sft/60, sif);
-%hline6 = plot(sft/60, srf);
 hold off
 
 set(hline1, 'LineWidth', 2)
@@ -181,9 +180,9 @@ set(hline4, 'Marker', '^')
 set(hline4, 'MarkerSize', 10)
 set(hline4, 'Color', colors(1,:))
 
-%set(hline5, 'LineWidth', 2)
-%set(hline5, 'LineStyle', '--')
-%set(hline5, 'Color', colors(1,:))
+% set(hline5, 'LineWidth', 2)
+% set(hline5, 'LineStyle', '--')
+% set(hline5, 'Color', colors(1,:))
 set(hline5, 'LineWidth', 2)
 set(hline5, 'LineStyle', 'none')
[…]

### 2016-01-28 plot_fits.m

@@ -1,0 +1,236 @@
+% Creates Data Plot
+
+width = 20;
+height = 8;
+
+font_size = 18;
+title_font_size = 24;
+legend_font_size = 14;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
+
+% --- Avian Data ---
+avt = [480, 600, 960, 1200, 1440, 1680, 1920,  2280, 2520]/60;
+av  = [ 100, 150, 825, 2563, 988, 1650, 2838, 17375, 5775];
+ave = [   0,  29, 229,  348, 166,  390,  112,  3062, 1261];
+
+ait = [120, 480, 600, 960, 1200, 1440, 1680,  1920,  2280,  2880]/60;
+ai  = [ 24,  87, 296, 770, 3048, 3901, 8261, 12935, 26970, 24120];
+
+art = [120, 480, 600, 960, 1200, 1440, 1680, 1920, 2280, 2880]/60;
+ar  = [ 45,  50,  57,  87,  134,  151,  150,  245,  780,  695];
+
+% --- Seasonal Data ---
+svt = [600,  720,  960,  1200,  1440,  1800,  2280,  2880]/60;
+sv  = [ 20, 5846, 6900, 38500, 39738, 20238, 15138, 20453];
+sve = [  0, 1676, 1578, 15695, 17271,  2992,  2886,  3837];
+
+sit = [120,  480,  720,  960, 1200, 1440]/60;
+si  = [391, 1723, 3462, 4618, 6807, 8164] - 300;
+
+siit  = [1440, 1800, 2280, 2880]/60;
+%sii   = [8164, 10200, 12913, 16306] - 300;
+sii   = [8164, 8500, 8500, 8500] - 300;
+
+srt = [120, 480, 720, 960, 1200, 1440, 1800, 2280, 2880]/60;
+sr  = [  1,   1,   1,   2,    6,   16,   68,  239,  655];
+
+% --- Swine Data ---24
+wvt = [120,   720,   960,   1200,   1440,   1800,   2880]/60;
+wv  = [ 20, 64620, 84053, 254587, 665559, 814586, 654773];
[…]

### 2016-01-28 rmse.m

@@ -1,0 +1,18 @@
+% rmse(data_time, data_points, fn_time, fn_points)
+%   Calculates the root mean squared error between a set of data points and 
+% a series of points representing a function output.
+%
+% Inputs:
+%   data_time   - A 1D array containing the time points of the data
+%   data_points - A 1D array of the data values
+%   fit_time    - A 1D array of the time points for the function output
+%   fit_points  - A 1D array of the function values
+% Output:
+%   error = the root mean squared error of the data vs. the function
+
+function error = rmse(data_time, data_points, fn_time, fn_points)
+   
+    sq_err = squared_error(data_time, data_points, fn_time, fn_points);
+
+    error = sqrt(sq_err/length(data_time));
+ 

### 2016-01-28 variance.m

@@ -17,26 +17,16 @@ set(fig, 'PaperPosition', [0 0 width height]);
 
 subplot(1, 3, 1)
 
-avar = [infect(vara00) infect(vara01) infect(vara02) infect(vara03) infect(vara04) ...
-        infect(vara05) infect(vara06) infect(vara07) infect(vara08) infect(vara09) ...
-        infect(vara10) infect(vara11) infect(vara12) infect(vara13) infect(vara14) ...
-        infect(vara15) infect(vara16) infect(vara17) infect(vara18) infect(vara19) ...
-        infect(vara20) infect(vara21) infect(vara22) infect(vara23) infect(vara24) ...
-        infect(vara25) infect(vara26) infect(vara27) infect(vara28) infect(vara29) ...
-        infect(vara30) infect(vara31) infect(vara32) infect(vara33) infect(vara34) ...
-        infect(vara35) infect(vara36) infect(vara37) infect(vara38) infect(vara39) ...        
-        infect(vara40) infect(vara41) infect(vara42) infect(vara43) infect(vara44) ...
-        infect(vara45) infect(vara46) infect(vara47) infect(vara48) infect(vara49)];
+aavg = mean(vara);
+asorted = sort(vara);
 
-aavg = mean(avar, 2);
-asorted = sort(avar, 2);
-
-plot(time, avar, 'color', [.75 .75 .75]);
+plot(time(1:100:end), vara(:,1:100:end), 'color', [.75 .75 .75]);
 hold on
-plot(time, aavg, '-k', 'LineWidth', 2)
-plot(time, asorted(:,2), 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2);
-plot(time, asorted(:,49), 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2)
+plot(time(1:100:end), aavg(:,1:100:end), '-k', 'LineWidth', 2)
+plot(time(1:100:end), asorted(3,1:100:end), 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2);
+plot(time(1:100:end), asorted(98,1:100:end), 'Color', [.75 0 0], 'LineStyle', '--', 'LineWidth', 2)
 hold off
+
 
 set(gca, 'FontSize', font_size)
 set(gca, 'YColor', 'k')
@@ -49,7 +39,7 @@ ylabel('Infected Cell Count')
[…]

### 2016-02-15 paper/cover_letter3.tex
added prcc sensitivity and edits

--- 2016-01-28 paper/cover_letter2.tex
@@
 
-We are pleased to resubmit the revised manuscript, ``A spatial model of the
+We are pleased to submit the revised manuscript, ``A spatial model of the
 efficiency of T cell search in the influenza-infected lung" for JTB's 
-consideration. This is a resubmission of manuscript number JTB-D-15-00501R1,
-received by us on October 26, 2015 with an invitation for resubmission within 
-two months.
+consideration. This is a resubmission of manuscript number JTB-D-15-00501R2,
+received by us on December 16, 2015 with an invitation for resubmission within 
+sixty days.
 
-We are happy to learn that the major concerns of Reviewer 2 and the previous 
-concerns of Reviewer 1 have been addressed. Reviewer 1 now asks for alternative
-models (major point 1) to achieve higher confidence in our estimation of specific
-parameters. We have addressed this concern in our current response and we 
-note that two separate sensitivity analyses confirmed that the parameters in 
-question (T cell production rate and the chemokine secretion rate) do not have 
-a strong effect on the model results. Reviewer 1 has also raised several new 
-interesting questions and suggested further extensions to our model, including
-new parameters describing the innate immune response (major point 4) and 
-environmental factors such as vascular structure (major point 3).  We feel that these 
-requests point to questions that are outside the scope of our current study, 
-which focuses on comparing the effectiveness of T cells at controlling infection
-in three specific influenza strains.
+We are grateful for the opportunity to (re)respond to Reviewer 1's
+comments. We regret that Reviewer 1 was not satisfied by our
+earlier responses, but we believe that we have addressed all of
+his/her concerns.  In our response, we have provided additional evidence and discussion
+specifically to address Reviewer 1's comments.  We appreciate the time
+spent by Reviewer 1 on the manuscript, but we respectfully disagree
+with his/her opinion that we did not understand or address his/her
+concerns.   At this point, we believe that we 
+have adequately addressed all comments from both Reviewer 1 and 2.
 
-Our revised manuscript discusses how the model takes these additional factors
-into account, but we have not explicitly added new parameters to our already 
-complicated model. We anticipate several follow-on projects to explore these 
-questions using new data sets and alternative models.
+%In summary, we agree with the statements and implications in Points 1, 2, 4 and 6.
+%We have answered the questions in Points 3 and 5. We include detailed responses 
+%to these points, but we respectfully submit that nowhere in the review is there a 
+%reference to an incorrect or incomplete analysis.
 

### 2016-02-15 paper/response3.tex
added prcc sensitivity and edits

@@
 
-We are grateful for the opportunity to respond to Reviewer 1's comments.
-We regret that Reviewer 1 did not ultimately agree with our rebuttal, but we believe that we have directly addressed all of his/her concerns in the previous rebuttal.
-We are providing additional evidence and discussion as requested by the last round of revisions.
-While we are grateful for the time given by Reviewer 1 to review our manuscript, we respectfully disagree with his/her contention that we did not understand or address his/her concerns.
-We provide additional rebuttals to Reviewer 1's comments below.
-We believe that we have adequately addressed all comments from both Reviewer 1 and 2.
+We are grateful for the opportunity to (re)respond to Reviewer 1's
+comments. We regret that Reviewer 1 was not satisfied by our
+earlier responses, but we believe that we directly addressed all of
+his/her concerns.  Here we provide additional evidence and discussion
+to respond to Reviewer 1's comments.  We appreciate the time
+spent by Reviewer 1 on the manuscript, but we respectfully disagree
+with his/her opinion that we did not understand or address his/her
+concerns.  At this point, we believe that we have adequately addressed
+all comments from both Reviewer 1 and 2.
+
+In summary, we agree with the statements and implications in Points 1,
+2, 4 and 6. We have answered the questions in Points 3 and 5. Below we
+include detailed responses to these points, but we respectfully submit
+that nowhere in the review is there a reference to an incorrect
+or incomplete analysis.
 
@@
 
-Reviewer 1: The authors misinterpreted what my comments meant, and this is third time.
+%\textbf{Reviewer 1: The authors misinterpreted what my comments meant, and this is third time.}
 
@@
 
-There is no known literature with chemokine secretion rate estimations.
-In this paper, we present new empirical data of \textit{in vitro} chemokine levels and extend a previously published ODE model to estimate chemokine production rates.
-We are as surprised as Reviewer 1 that these production rates do not have a strong effect on our agent-based model results as determined by our sensitivity analyses.
-Nevertheless, because our agent-based model examines chemokine interactions, we require a non-zero value for the production rate.
-We believe that fitting the extended ODE model to our new empirical data is the best approach available to determine biologically plausible values.
-Our paper contains a subsection, \textit{Models for Parameter Estimation}, that details the exact process used to estimate the chemokine production rates (page 6).
-We fit the system of differential equations to empirical data using a genetic algorithm to minimize the log squared error between the model and the data (the genetic algorithm performed favorably when compared to standard Levenberg-Marquardt nonlinear regression techniques).
+This comment seems contradictory: On the one hand it asks for more
+details about how we estimate chemokine production rate, and on the
+other hand suggests that chemokine production rate doesn't need to
+be estimated because it does not impact clearance.  Our analysis
+agrees with his statement that it does not impact clearance. For
+thoroughness, we explain how we estimated chemokine production rate
+below.
 
-We have added a figure displaying the model fits to the supplement (Fig. S1, also below) and Table S3 showing the root mean squared error of the fits for completeness.
+There is no known literature with chemokine secretion rate
+estimations.  In this paper, we present new empirical data of
+\textit{in vitro} chemokine levels and extend a previously published
+ODE model to estimate chemokine production rates.  We are as surprised
+as Reviewer 1 that these production rates do not have a strong effect
+on the agent-based modeling results (see sensitivity analyses).
+Nevertheless, because our agent-based model examines chemokine
+interactions, we require \textit{some} non-zero value for the production
+rate.  We believe that fitting the extended ODE model to our new
+empirical data is the best approach available to determine
+biologically plausible values.  Our paper contains a subsection,
+\textit{Models for Parameter Estimation}, that details the exact
+process used to estimate the chemokine production rates (page 6).  We
+fit the system of differential equations to empirical data using a
+genetic algorithm to minimize the log squared error between the model
+and the data (the genetic algorithm gave better fits than the more standard Levenberg-Marquardt nonlinear regression techniques).
+
+We have now added a figure displaying the model fits to the supplement (Fig. S1, also below) and Table S3 showing the root mean squared error of the fits for completeness.
 
@@
 \begin{response}
-All biological data have a foundation using \textit{in vitro} research.
-We agree with Reviewer 1 that while this raises the potential caveat that some of the values that we use to parameterize our models using \textit{in vitro} generated data may not reflect \textit{in vivo} values, we, along with most other researchers in the field, use \textit{in vitro} generated values as a starting point.
-We explicitly state this in the manuscript.
-This is precisely the reason that we undertook an extensive sensitivity analysis as we acknowledge that the values we use to parameterize the model may vary from initial starting values.
-We do not claim that the values we estimate are absolutely reflective of \textit{in vivo} values.
-Instead, we use best fit estimates that are currently available as a starting point, and use sensitivity analysis to confirm the role of individual parameters on the model.​
+
+  We agree that in vitro and in vivo parameters may be different, but
+  this is a caveat that applies throughout biological and biomedical
+  sciences.  We, along with most other researchers in the field, use
+  \textit{in vitro} generated values as a starting point when
+  \textit{in vivo} values are not available, and we state when we do this explicitly
+  in the manuscript.  In addition, the paper contains an extensive
+  sensitivity analysis to address the concern that the values used to
+  parameterize the model might vary from the initial starting
+  values. We do not claim that the values we estimate correspond
+  exactly to  \textit{in vivo} values. Instead, we use the best fit
+  estimates that are currently available as a starting point, and
+  conduct the sensitivity analysis to confirm the role of individual parameters on the model.
 \end{response}
@@
 
-Chemokine does have a higher diffusion rate than virus in our model (Table 2).
-It is important to note that our discussion of the infection outpacing the chemokine is different than the direct comparison between the viral diffusion rate and the chemokine diffusion rate.
-Because T cells climb the chemokine gradient, a chemokine's effect is seen in the location of its local maxima, while the viral effect is seen in the spread and concentration of the virions themselves.
-We have edited the text to be more clear of this distinction (\textit{Results: Spatial Effects}, page 10).
+This question is answered directly in the paper---if diffusion of
+chemokine is faster than infection spread, it does not impact our
+results (Table 3, Figs. S4-S7).  In the paper, we argue that direct comparison of
+viral and chemokine diffusion rates is not informative. 
+Because T cells climb the chemokine gradient, a chemokine's effect is
+determined by the location of its local maxima, while the viral effect
+is determined by the spread and concentration of the virions
+themselves.  We have edited the text to make this distinction more clear (\textit{Results: Spatial Effects}, page 10).
 
@@
 
-To complement our spatial analysis we performed multiple sensitivity analyses on our model's free parameters.
-In doing so, we examined scenarios where the chemokine's diffusion rate was increased by up to two orders of magnitude (Table 2).
-Each sensitivity analysis showed that the chemokine diffusion rate was an insensitive parameter in our model (Table 3 and Supplement)
+To complement the spatial analysis, we also performed multiple
+sensitivity analyses of the model's free parameters.
+This included scenarios where the chemokine's diffusion rate was increased by up to two orders of magnitude (Table 2).
+Each sensitivity analysis showed that the chemokine diffusion rate was an insensitive parameter in the model (Table 3 and Supplement).
 
@@
 \begin{response}
-
-We agree with the reviewer that the T cell secretion rate is an insensitive parameter in our model as determined by our multiple sensitivity analyses.
-Because our agent-based model investigates the recruitment of T cells to the focus of infection, we required a non-zero secretion rate to introduce these T cells into the model.
-Similar to the chemokine secretion rate as discussed in the first point and in our paper (\textit{Models for Parameter Estimation}, page 7), we fit an ODE model to empirical data to best estimate a biologically plausible value for the T cell secretion rate for use in our agent-based model.
-
-We also agree that the recruitment of T cells to the focus of the infection is important.
-We discuss this in the \textit{Spatial Effects} subsection (page 9-10) and in the discussion (page 15) as follows:
+This comment restates as fact one of the conclusions of our analysis,
+so we have no disagreement with this point.
+%We agree with the reviewer that the T cell secretion rate is an
+%insensitive parameter in our model as determined by our multiple
+%sensitivity analyses.  
+However, because our agent-based model investigates the recruitment of
+T cells to the focus of infection, we required a non-zero secretion
+rate to introduce T cells into the model.  To select a biologically
+plausible value, we fit an ODE model to empirical data in our paper
+(\textit{Models for Parameter Estimation}, page 7) and similarly to
+the chemokine secretion rate discussed in Point 1.  We also agree that
+the recruitment of T cells to the focus of the infection is important,
+a point that is discussed in the \textit{Spatial Effects} subsection (page 9-10)
+and in the discussion (page 15) as follows:
 
@@
 
-The term `viral control' has been used in three different contexts.  We have explicitly addressed each point in previous responses.  In summary:
+The reviewer asks if changing certain parameters in our model would
+change the outcome for virus control. Our model does not produce the
+viral control that the reviewer hypothesizes it could produce, but
+we believe that reporting results that do not conform to his/her
+expectations should not be confused with ignoring the request to perform
+the analysis.
+
+The term `viral control' has been used by Reviewer 1 in three different contexts.  We have explicitly addressed each point in previous responses.  In summary:
 
@@
 \begin{displayquote}
-We have performed these simulations numerous times in the form of our sensitivity analyses.
-Further, you seem to be misunderstanding the distinction between the differing roles of viral diffusion and chemokine diffusion.
-Please refer to Response 3 of this document for more information.
+We performed these simulations numerous times as part of the
+sensitivity analyses, and Response 3 of this document reviews the
+different roles of viral diffusion and chemokine diffusion.  
 \end{displayquote}
@@
 \begin{displayquote}
-In summary, our paper makes no attempt to model the immune response in its entirety.  Our model does represent viral
-spread using parameters which likely reflect innate control mechanisms (virus secretion, infectivity, diffusion
+In summary, our model does not attempt to include the entire immune response. It represents viral
+spread using parameters that likely reflect innate control mechanisms (virus secretion, infectivity, diffusion
 rate, and decay rate, see Tables 1 and 2). We appreciate the reviewer’s perspective, but our goal in this
@@
 
-\textit{``Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, cy-
-tokines, and replicating influenza virus. ... Therefore, instead of developing a comprehensive
+\textit{``Specifically, we focus on the interactions between activated antigen-specific CD8 T cells, 
+cytokines, and replicating influenza virus. ... Therefore, instead of developing a comprehensive
 immune system model, we present a spatially explicit agent-based model (ABM) to describe T
@@
 different influenza strains is best left to future studies.
-
 \end{displayquote}
@@
 \begin{response}
-We have reviewed Oreskes et.
-al.
-The paper correctly points out that scientists often use terms incorrectly when describing their model's veracity which leads to incorrect claims and conclusions.
+We have reviewed Oreskes et. al.  and outline how our paper is
+consistent with it.
+This excellent paper observes that scientists often use terms incorrectly when describing their model's veracity which leads to incorrect claims and conclusions.
 Specifically, Oreskes et.
@@
 \begin{itemize}
-\item \textbf{Verification:} Oreskes et.
-al.
-states that it is not possible to verify a numerical model by comparing it to an analytical model.
-We use the term `verification' one time (page 15, concluding paragraph) in the context of obtaining empirical data only (emphasis added).
+\item \textbf{Verification:} Oreskes et. al. states that it is not
+  possible to verify a numerical model by comparing it to an
+  analytical model.  We use the term `verification' once (page 15, concluding paragraph) in the context of obtaining empirical data (emphasis added).
 \begin{displayquote}
@@
 \end{displayquote}
-\item \textbf{Validation:} We do not use the term in our paper.
-\item \textbf{Calibration:} Oreskes et.
-al.
-warn of the dangers of using the calibration of free parameters to emprical data as a basis for claiming model validity.
-We use the term `calibration' only once (Abstract) and make no erroneous claims (emphasis added).
+\item \textbf{Validation:} This term does not appear in our paper.
+\item \textbf{Calibration:} Oreskes et. al. warn of the dangers of
+  using the calibration of free parameters to emprical data as a basis
+  for claiming model validity.  We use the term `calibration' only
+  once (Abstract) and make no claims that calibration implies model validity (emphasis added).
 \begin{displayquote}
@@
 
-At no point have we made claims inconsistent with Oreskes et.
-al.
-We agree that models can inform research and use our model to propose a hypothesis of how additional T cells are limited in their effectiveness due to the spatial effects of the chemokine gradient and T cell chemotaxis.
+We agree that models can inform and guide empirical research and use our model to propose
+a hypothesis that additional T cells are limited in their
+effectiveness due to the spatial effects of the chemokine gradient and
+T cell chemotaxis.  We cannot find a place where our paper's
+conclusions are inconsistent with the contents of Oreskes et al.
 \end{response}

### 2016-02-15 sensitivity/PRCC.m

@@ -1,0 +1,94 @@
+%% It calculates PRCCs and their significances
+%% (uncorrected p-value, Bonferroni correction and 
+%% Benjamini and Hochberg False Discovery Rate correction)
+%% LHSmatrix: LHS matrix (N x k) %%
+%% Y: output matrix (time x N) %%
+%% s: time points to test (row vector) %%
+%% PRCC_var: vector of strings {'p1','p2',...,'pk'}
+%%            to label all the parameters varied in the LHS
+%% by Simeone Marino, May 29 2007 %%
+%% N.B.: this version uses ONE output at a time
+
+function [prcc sign sign_label]=PRCC(LHSmatrix,Y,s,PRCC_var,alpha);
+
+%LHSmatrix=LHS; % Define LHS matrix
+Y=Y(s,:)';% Define the output. Comment out if the Y is already 
+          % a subset of all the time points and it already comprises
+          % ONLY the s rows of interest
+[~, k] = size(LHSmatrix); % Define the size of LHS matrix
+[~, out] = size(Y);
+
+prcc = zeros(k, out);
+prcc_sign = zeros(k, out);
+
+for i=1:k  % Loop for the whole submatrices
+
+    LHStemp = LHSmatrix;
+    LHStemp(:,i) = [];
+%    Z = LHStemp;
+    
+    [rho, p] = partialcorr([LHSmatrix(:,i), Y], LHStemp,'type','Spearman');
+    
+    for j=1:out      
+        prcc(i,j) = rho(1, j+1);
+        prcc_sign(i,j) = p(1, j+1);
+    end
+end
+
+%[length(s) k out];
+prcc = prcc';
+uncorrected_sign = prcc_sign';
+sign = uncorrected_sign;
+
+%% Multiple tests correction: Bonferroni and FDR
+%tests=length(s)*k; % # of tests performed
+%correction_factor=tests;
+%Bonf_sign=uncorrected_sign*tests;
+%sign_new=[];
+%for i=1:length(s)
+%    sign_new=[sign_new;(1:k)',ones(k,1)*s(i),sign(i,:)'];
+%end
+%sign_new=sortrows(sign_new,3);
+%for j=2:k
+%    sign_new(j,3)=sign_new(j,3)*(tests/(tests-j+1));
+%end
+%sign_new=sortrows(sign_new,[2 1]); % FDR correction
[…]

### 2016-02-15 sensitivity/PRCC_PLOT.m

@@ -1,0 +1,42 @@
+%% Plot the residual of the partial regression of X (input - LHS matrix) and Y (output)
+%% at column s (time points saved). PCC Coefficients are calculated on these
+%% var: labels of the parameters varied in the X (as legend)
+%% The Title of the plot is the Pearson correlation coefficient of the
+%% transformed data, that is  the PRCC calculated on the original data.
+%% The p-value is also showed in the title
+%% by Simeone Marino, June 5 2007 %%
+
+function [r, rx] = PRCC_PLOT(X,Y,s,PRCC_var,y_var)
+
+Y = Y(s,:);
+[a, k] = size(X); % Define the size of LHS matrix
+Xranked = rankingN(X);
+Yranked = ranking1(Y);
+
+Z = ones(a, k, k);
+r = zeros(a, k);
+rx = zeros(a, k);
+
+for i=1:k  % Loop for the whole submatrices, Zi 
+    LHStemp = Xranked;
+    LHStemp(:,i) = ones(1,a);
+    Z(:,:,i) = LHStemp;
+end
+for i=1:k
+    [~, ~, r(:,i)]  = regress(Yranked, Z(:,:,i));
+    [~, ~, rx(:,i)] = regress(Xranked(:,i), Z(:,:,i));   
+end
+
+display = 0;
+if display
+    for i=1:k        
+        [R, P] = corr(r(:,i), rx(:,i));
+        a = ['[PRCC , p-value] = ' '[' num2str(R) ' , ' num2str(P) '].'];% ' Time point=' num2str(s-1)];   
+        figure
+        plot(r(:,i), rx(:,i), '.')
+        title(a)
+        legend(PRCC_var{i})
+        xlabel(PRCC_var{i})
+        ylabel(y_var)
+    end
+end

### 2016-02-15 sensitivity/gen_deffile.py

@@ -1,0 +1,179 @@
+# gen_deffile.py
+# Drew Levin
+# August 30, 2015
+#
+# Generates a .def file for model definition in CyCells.  Takes in a list of
+# parameters and a file name.  The file creation is a side effect.
+
+# parameters is a dictionary of parameter values with keys as defined in
+# hypercube.py
+
+import os
+
+def gen_file(params, filename, is_large, is_avian):
+
+  # Get OS compatible new line string
+#  eof = os.linsep
+  eof = '\n'
+
+  # Percent of lung represented
+  p_in = 0.005625 if is_large else 0.0025
+  p_out = 1.0 - p_in
+
+
+  # Chemokine onset time is 8h for IP10, 16h for RANTES
+  chem_onset = 56700 if is_avian else 28800
+
+  # Write the output to the string
+  output = ''
+
+  output += '#DefFormat 8' + eof
+  output += eof
+  output += 'cell_names { healthy infected expressing dead apoptosis t_found t_search t_circ oracle }' + eof
+  output += eof
+  output += 'molecule_type virus {' + eof
+  output += '  diffusion_rate ' + str(params['v_dif']) + eof
+  output += '  decay_until ' + str(params['v_decay']/86400.0) + ' 345600' + eof
+  output += '  decay_rate ' + str(params['v_decay']*params['igm']/86400.0) + eof
+  output += '}' + eof
+  output += eof
+  output += 'molecule_type chemokine {' + eof
+  output += '  diffusion_rate ' + str(params['c_dif'])  + eof
+  output += '  decay_rate ' + str(params['c_decay']) + eof
+  output += '}' + eof
+  output += eof
+  output += 'cell_type healthy {' + eof
+  output += '  radius 5' + eof
+  output += eof
+  output += '  attribute age fixed 0 fixed 0' + eof
+  output += '  attribute apop_age fixed 0 fixed 0' + eof
+  output += '  attribute virusC fixed 0 fixed 0' + eof
+  output += '  attribute infection_age gaussian ' + str(params['t_inc']*3600) + ' ' + str(params['t_inc']*360) + ' gaussian ' + str(params['t_inc']*3600) + ' ' + str(params['t_inc']*360) + eof
+  output += '  attribute sense_tcell fixed 0 fixed 0' + eof
+  output += eof
+  output += '  sense virusC copy_conc virus' + eof
+  output += eof
[…]

### 2016-02-15 sensitivity/init_sensitivity.m

@@ -1,0 +1,90 @@
+
+% Import the LHS
+temp = importdata('/nfs/adaptive/drew/chemokine/sensitivity/seasonal_hcube.txt');
+hcube = temp.data;
+rephcube = repmat(hcube, 3, 1);
+labels = temp.colheaders;
+
+% Initialize the matrices
+n = 333;
+n3 = n*3;
+t = 1441;
+s = (1:t)';
+Y = zeros(t, n);
+Y3 = zeros(t, n3);
+N = zeros(1, n);
+N3 = zeros(1, n3);
+
+shuffle = [5 6 11 3 1 2 14 9 17 15 13 7 8 4 10 12 16];
+
+% Iterate through each file of the 999 possible.  Update data if it exists
+for r = 1:3
+    disp(r)
+    for i=1:n
+        name = ['seasonal_', num2str(r), '_', num2str(i, '%0.3u'), '.hist'];
+        disp(name);
+        if exist(['/nfs/adaptive/drew/chemokine/sensitivity/hist/', name], 'file')
+            N(i) = N(i) + 1;
+            N3(n*(r-1)+i) = N3(n*(r-1)+i) + 1;
+            temp = importdata(['/nfs/adaptive/drew/chemokine/sensitivity/hist/', name]);
+            temp = infect(temp.data);
+            Y(:, i) = Y(:, i) + temp(1:100:end);
+            Y3(:, n*(r-1)+i) = Y3(:, n*(r-1)+i) + temp(1:100:end);
+        end
+    end
+end
+
+for i = 1:t
+    Y(i,:) = Y(i,:) ./ N;
+end
+
+% Clear empty rows
+m = n;
+for i=0:n-1
+    if N(n-i) == 0
+        N(n-i) = [];
+        Y(:,n-i) = [];
+        hcube(n-i,:) = [];
+        m = m-1;
+    end
+end
+
+m3 = n3;
+for i=0:n3-1
+    if N3(n3-i) == 0
+        N3(n3-i) = [];
+        Y3(:,n3-i) = [];
+        rephcube(n3-i,:) = [];
+        m3 = m3-1;
+    end
+end
+
+n = m;
+n3 = m3;
+display([n, n3])
+
+% Rank transform the hypercube and model output
+rankH = zeros(n, 17);
+rankY = zeros(t, n);
+
+rankH3 = zeros(n3, 17);
[…]

### 2016-02-15 sensitivity/plot_extreme.m

@@ -1,0 +1,31 @@
+width = 8;
+height = 6;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
+
+font_size = 14;
+title_font_size = 16;
+legend_font_size = 11;
+
+temp = importdata('/nfs/adaptive/drew/chemokine/sensitivity/hist/seasonal_1_023.hist');
+extreme_1 = infect(temp.data);
+temp = importdata('/nfs/adaptive/drew/chemokine/sensitivity/hist/seasonal_2_023.hist');
+extreme_2 = infect(temp.data);
+temp = importdata('/nfs/adaptive/drew/chemokine/sensitivity/hist/seasonal_3_023.hist');
+extreme_3 = infect(temp.data);
+
+plot(time, [extreme_1, extreme_2, extreme_3, infect(vars00)]);
+lh = legend('Sample 1', 'Sample 2', 'Sample 3', 'Baseline');
+title('Exhaustive Growth Example', 'FontSize', title_font_size);
+xlabel('Time (days)')
+ylabel('Infected Cells')
+
+set(gca, 'FontSize', font_size)
+set(gca, 'FontWeight', 'normal')
+set(lh, 'FontSize', legend_font_size)

### 2016-02-15 sensitivity/plot_sensitivity.m

@@ -1,0 +1,62 @@
+width = 20;
+height = 13.7;
+
+fig = figure();
+
+set(fig, 'PaperUnits', 'inches');
+set(fig, 'PaperSize', [width height]);
+set(fig, 'Units', 'inches');
+set(fig, 'Position', [0 0 width height]);
+set(fig, 'PaperPosition', [0 0 width height]);
+
+[n, ~] = size(prcc3);
+x = 1:n;
+
+colors = [196, 196, 196; ...
+          221, 156, 25; ...
+          31,  66,  148; ...
+          0,   0,   0] / 256;
+
+
+for i = 1:16
+
+    % generate color vector
+    
+    c = colors(1,:)' * ones(1,n);
+    
+    % P < 0.05
+    mask = find(sign3(:,i) < 0.05);
+    c(:, mask) = colors(2, :)' * ones(1, length(mask));
+    
+    % P < 0.01
+    mask = find(sign3(:,i) < 0.01);
+    c(:, mask) = colors(3, :)' * ones(1, length(mask));
+    
+    % P < 0.001
+%    mask = find(sign3(:,i) < 0.001);
+%    c(:, mask) = colors(4, :)' * ones(1, length(mask));
+    
+    subplot(4, 4, i);
+
+    plot(x*6/864, zeros(1,n), 'k--');
+    hold on
+    sh = scatter(x*6/864, prcc3(:,i), 16, c', 'filled');
+    hold off
+    
+    title(labels_long(i), 'FontSize', title_font_size);
+    
+    if i > 12
+        xlabel('Time (days)');
+    end
+    
+    if mod(i, 4) == 1
+        ylabel('Spearman Correlation');
+    end
+    
+    ylim([-1 1]);
+    xlim([0, 10]);
+
+    set(gca, 'FontSize', font_size)
+    set(gca, 'FontWeight', 'normal')
+end
+

### 2016-02-15 sensitivity/run_cycells.py

@@ -1,0 +1,198 @@
+# gen_deffile.py
+# Drew Levin
+# September 2 2015
+#
+# Automates the evaluation of many CyCells models on a machine.
+# Requires in the same directory:
+#   1) The CyCells executable
+#   2) gen_deffile.py
+#   3) hypercube text files
+#   4) immune.init
+#
+# Takes 5 command line arguments in this order:
+# 1: strain letter (a, s, p)
+# 2: replicate # (1, 2, 3)
+# 3: number of cores to use (model evals to run at once)
+# 4: start row of hypercube file (1 indexed)
+# 5: end row of hypercube file (1 indexed)
+
+# Python Dependencies
+import csv
+import subprocess as sp
+import threading
+import sys
+import os
+import socket
+import time
+import logging as log
+
+# My Dependencies
+import gen_deffile as gen
+
+# Generate filename for status log file
+log_filename = 'status_' + socket.gethostname() + '_' + str(os.getpid()) + '.txt'
+
+# Configure the logger
+log.basicConfig(
+  level = log.INFO,
+  filename = log_filename,
+  format = '[%(asctime)s] (%(threadName)-9s) %(message)s',
+  datefmt = '%m/%d/%Y %H:%M:%S'
+)
+log.getLogger().addHandler(log.StreamHandler())
+
+# Global thread container and lock
+threads = []
+lock = threading.Lock()
+
+# Global command line argments
+strain    = ''
+rep       = ''
+cores     = 0
+row_start = 0
+row_end   = 0
+
+# Global value for the next available row
+current_row = 0
+
+# Global list of parameters values
+parameters = []
+
+# Long form names for file naming purposes
+strain_long = {'a': 'avian', 's': 'seasonal', 'p': 'pandemic'}
+
+# Thread-safe function to assign a thread a specific row of the hypercube
+def getNextRow():
+  global current_row, strain, rep
+  with(lock):
+    while True:
+      next_row = current_row
+      current_row += 1
+
+      # Skip runs that have already completed and recorded the hist file
+      filename = '_'.join([strain_long[strain], rep, str(next_row).zfill(3)])
+      if not os.path.isfile('hist/'+filename+'.hist'):
+        return next_row
+
+# Thread functionality, runs CyCells instances until satisfied
+def run(i):
+  global strain, rep, cores, row_start, row_end, parameters
+  while True:
+    # Obtain current set of parameters
+    row = getNextRow()
[…]

### 2016-02-15 sensitivity/still_running.py

@@ -1,0 +1,39 @@
+# still_running.py
+# Drew Levin
+# September 4, 2015
+#
+# Scans the supplied status.txt file to return a list of parameter sets still
+# being evaluated by run_cycells.py
+
+import sys
+import re
+
+running = {}
+
+pattern = re.compile('_._')
+
+if len(sys.argv) < 2:
+  print('Too few arguments')
+  print('  python still_running.py <file>')
+  sys.exit(1)
+
+# Read in the status file as a list of strings (one per row)
+with open(sys.argv[1]) as f:
+  lines = f.read().splitlines()
+
+# Go through the file, line by line
+for row in sorted(lines):
+  if "Starting Row" in row:
+    start = row.find('Row ')+4
+    n = int(row[start:])
+    running[n] = (n, row[1:24])
+  elif "failed to complete" in row or "completed successfully" in row:
+    start = pattern.search(row).start()
+    n = int(row[start+3:start+6])
+    if n in running:
+      del running[n]
+
+for (n, t) in running.values():
+  print(t + ' ' + str(n))
+
+

### 2016-02-15 sensitivity/testing/run_cycells.py

@@ -1,0 +1,193 @@
+# gen_deffile.py
+# Drew Levin
+# September 2 2015
+#
+# Automates the evaluation of many CyCells models on a machine.
+# Requires in the same directory:
+#   1) The CyCells executable
+#   2) gen_deffile.py
+#   3) hypercube text files
+#   4) immune.init
+#
+# Takes 5 command line arguments in this order:
+# 1: strain letter (a, s, p)
+# 2: replicate # (1, 2, 3)
+# 3: number of cores to use (model evals to run at once)
+# 4: start row of hypercube file (1 indexed)
+# 5: end row of hypercube file (1 indexed)
+
+# Python Dependencies
+import csv
+import subprocess as sp
+import threading
+import sys
+import os
+import socket
+import time
+import logging as log
+
+# My Dependencies
+import gen_deffile as gen
+
+# Generate filename for status log file
+log_filename = 'status_' + socket.gethostname() + '.txt'
+
+# Configure the logger
+log.basicConfig(
+  level = log.INFO,
+  filename = log_filename,
+  format = '[%(asctime)s] (%(threadName)-10s) %(message)s',
+  datefmt = '%m/%d/%Y %I:%M:%S %p'
+)
+log.getLogger().addHandler(log.StreamHandler())
+
+# Global thread container and lock
+threads = []
+lock = threading.Lock()
+
+# Global command line argments
+strain    = ''
+rep       = ''
+cores     = 0
+row_start = 0
+row_end   = 0
+
+# Global value for the next available row
+current_row = 0
+
+# Global list of parameters values
+parameters = []
+
+# Long form names for file naming purposes
+strain_long = {'a': 'avian', 's': 'seasonal', 'p': 'pandemic'}
+
+# Thread-safe function to assign a thread a specific row of the hypercube
+def getNextRow():
+  global current_row
+  with(lock):
+    next_row = current_row
+    current_row += 1
+    return next_row
+
+# Thread functionality, runs CyCells instances until satisfied
+def run(i):
+  global strain, rep, cores, row_start, row_end, parameters
+  while True:
+    # Obtain current set of parameters
+    row = getNextRow()
+    if row > row_end:
+      break
+    log.info('Starting Row ' + str(row))
+
+    params = dict(zip(parameters[0], map(float, parameters[row])))
+
+    # Generate the CyCells definition file
+    filename = '_'.join([strain_long[strain], rep, str(row).zfill(3)])
[…]
